---
title: "Extreme rainfall reveals human amplification of land-to-sea nutrient transport across tropical coastlines"
subtitle: "Kona Low March 2026 — Clean Reproducible Analysis"
author:
  - name: Nyssa J. Silbiger
    affiliation: University of Hawaiʻi at Mānoa
  - name: Andrea Kealoha
    affiliation: University of Hawaiʻi at Mānoa
  - name: Sara Kahanamoku
    affiliation: University of Hawaiʻi at Mānoa
date: today
date-format: "MMMM D, YYYY"

format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    embed-resources: true
    theme: flatly
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 6
    lang: en
    css: |
      body { font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto,
             "Helvetica Neue", Arial, sans-serif; }
      .callout { margin-top: 1rem; }

execute:
  warning: false
  message: false
  cache: false   # set to true for expensive chunks individually with #| cache: true
---

```{r}
#| label: global-seed
#| include: false

# Global random seed — set once, used everywhere.
# All stochastic operations (PAM, RF, Bayesian samplers, block CV splits)
# must reference this object rather than hard-coding numbers.
SEED <- 2847
set.seed(SEED)
```

# Setup {#sec-setup}

```{r}
#| label: packages

# ── Core tidyverse ────────────────────────────────────────────────────────────
library(tidyverse)
library(lubridate)
library(scales)
library(here)

# ── Tables & reporting ────────────────────────────────────────────────────────
library(knitr)
library(kableExtra)
library(gt)

# ── Visualisation ─────────────────────────────────────────────────────────────
library(patchwork)
library(cowplot)
library(ggrepel)

# ── Spatial ───────────────────────────────────────────────────────────────────
library(sf)
library(terra)
library(gstat)
library(spdep)
library(exactextractr)

# ── Statistics & modelling ────────────────────────────────────────────────────
library(cluster)     # PAM clustering
library(ranger)      # random forest
library(broom)       # tidy model output
library(infer)       # tidy inference

# Note: brms, sdmTMB, glmmTMB, mgcv, blockCV loaded in the sections
# where they are first used to keep startup time reasonable.
```

```{r}
#| label: output-directories

# Create output directory tree. Safe to re-run — dirs already present are skipped.
dirs <- here(c(
  "output/main_figures",
  "output/supplement",
  "output/tables",
  "output/model_objects",
  "output/diagnostics"
))
walk(dirs, \(d) dir.create(d, showWarnings = FALSE, recursive = TRUE))
```

```{r}
#| label: global-constants

# ── Island colour palette ─────────────────────────────────────────────────────
island_pal <- c(
  "O'ahu"    = "#1565C0",
  "Maui"     = "#2E7D32",
  "Moloka'i" = "#E65100",
  "Lāna'i"   = "#6A1B9A"
)

# Normalise okina/smart-quote variants so island-name joins are consistent
fix_island <- function(x) {
  x |>
    stringr::str_replace_all("[ʻʼ'']", "'") |>
    stringr::str_replace_all("ā", "a") |>   # for join matching only
    stringr::str_trim()
}

# Canonical island factor levels (display form)
island_levels <- c("O'ahu", "Maui", "Moloka'i", "Lāna'i")

# ── Nutrient response variables ───────────────────────────────────────────────
response_vars  <- c("NO3NO2", "PO4", "NH3", "SiO2")
response_labels <- c(
  NO3NO2 = "NO\u2083+NO\u2082 (\u03bcmol L\u207b\u00b9)",
  PO4    = "PO\u2084 (\u03bcmol L\u207b\u00b9)",
  NH3    = "NH\u2084 (\u03bcmol L\u207b\u00b9)",
  SiO2   = "SiO\u2082 (\u03bcmol L\u207b\u00b9)"
)

# ── Reef health thresholds (µmol/L; Reef Check / Hawaii benchmarks) ───────────
reef_thresholds <- c(NO3NO2 = 1.0, PO4 = 0.1, SiO2 = 5.0, NH3 = 1.0)

# ── PAM water-type cluster palette ───────────────────────────────────────────
cluster_pal <- c(
  "Marine background"  = "#1565C0",
  "Nearshore enriched" = "#2E7D32",
  "Runoff-dominated"   = "#E65100"
)
cluster_levels <- c("Marine background", "Nearshore enriched", "Runoff-dominated")

# ── Storm & sampling time windows ─────────────────────────────────────────────
storm_start  <- as.Date("2026-03-10")
storm_peak   <- as.Date("2026-03-14")
storm_end    <- as.Date("2026-03-25")
samp_start   <- as.Date("2026-03-28")
samp_end     <- as.Date("2026-04-08")

# Shared annotate() rectangles for time-series plots
ts_windows <- list(
  annotate("rect", xmin = storm_start, xmax = storm_end,
           ymin = -Inf, ymax = Inf, fill = "#BBDEFB", alpha = 0.70),
  annotate("rect", xmin = samp_start, xmax = samp_end,
           ymin = -Inf, ymax = Inf, fill = "#C8E6C9", alpha = 0.70)
)
```

```{r}
#| label: theme-kona-paper

# Custom ggplot2 theme for all publication figures.
# Requirements: white bg, minimal grid, ~9pt text, no unnecessary legends,
# panel letters added via labs(tag = "A"), accessible palette,
# interpretable in grayscale.

theme_kona_paper <- function(base_size = 9, ...) {
  theme_minimal(base_size = base_size, ...) +
    theme(
      # Typography
      text              = element_text(colour = "gray10"),
      plot.title        = element_text(size = rel(1.15), face = "bold",
                                       margin = margin(b = 4)),
      plot.subtitle     = element_text(size = rel(0.95), colour = "gray40",
                                       margin = margin(b = 6)),
      plot.caption      = element_text(size = rel(0.82), colour = "gray50",
                                       hjust = 0, margin = margin(t = 6)),
      plot.tag          = element_text(face = "bold", size = rel(1.1)),
      # Axes
      axis.title        = element_text(size = rel(0.95)),
      axis.title.x      = element_text(margin = margin(t = 5)),
      axis.title.y      = element_text(margin = margin(r = 5)),
      axis.text         = element_text(size = rel(0.88), colour = "gray20"),
      axis.ticks        = element_line(colour = "gray70", linewidth = 0.3),
      # Panel
      panel.background  = element_rect(fill = "white", colour = NA),
      plot.background   = element_rect(fill = "white", colour = NA),
      panel.grid.major  = element_line(colour = "gray92", linewidth = 0.35),
      panel.grid.minor  = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.border      = element_blank(),
      # Facets
      strip.background  = element_rect(fill = "gray96", colour = NA),
      strip.text        = element_text(face = "bold", size = rel(0.92)),
      # Legends
      legend.position   = "right",
      legend.title      = element_text(size = rel(0.9), face = "bold"),
      legend.text       = element_text(size = rel(0.85)),
      legend.key.size   = unit(0.9, "lines"),
      legend.background = element_rect(fill = NA, colour = NA),
      # Margins
      plot.margin       = margin(8, 10, 6, 8)
    )
}

# Make it the default for this document
theme_set(theme_kona_paper())

# Helper: save a figure as both PDF (vector) and PNG (600 dpi)
save_fig <- function(plot, name, width, height,
                     dir = here("output", "main_figures")) {
  ggsave(file.path(dir, paste0(name, ".pdf")), plot,
         width = width, height = height, device = cairo_pdf)
  ggsave(file.path(dir, paste0(name, ".png")), plot,
         width = width, height = height, dpi = 600)
  invisible(plot)
}

save_supp <- function(plot, name, width, height) {
  save_fig(plot, name, width, height,
           dir = here("output", "supplement"))
}
```

---

# Data Inventory and QA/QC {#sec-data-inventory}

> **Step 1.** Audit the sample data before doing any modeling.
> Target: explicit sample-flow table, site_id creation, coordinate QA.

```{r}
#| label: load-raw-chemistry

chem_raw <- read_csv(here("data", "clean_chemdata.csv"), show_col_types = FALSE)
```

```{r}
#| label: load-spatial-support-files

ahupuaa   <- read_csv(here("data", "ahupuaa_cache.csv"),             show_col_types = FALSE)
cess_dist <- read_csv(here("data", "sample_cesspool_distances.csv"), show_col_types = FALSE)
cess_buf  <- read_csv(here("data", "cesspool_buffer_counts.csv"),    show_col_types = FALSE)
cess_ahu  <- read_csv(here("data", "cesspool_by_ahupuaa.csv"),       show_col_types = FALSE)
ccap      <- read_csv(here("data", "ccap_ahupuaa_summary.csv"),      show_col_types = FALSE)
slope     <- read_csv(here("data", "ahupuaa_slope.csv"),             show_col_types = FALSE)
ksat      <- read_csv(here("data", "ahupuaa_ksat.csv"),              show_col_types = FALSE)
discharge <- read_csv(here("data", "usgs_discharge_march2026.csv"),  show_col_types = FALSE)
fetch     <- read_csv(here("data", "wave_fetch.csv"),                show_col_types = FALSE)
```

```{r}
#| label: resolve-duplicate-ahupuaa
# Several ahupuaa-level tables contain duplicate rows from multi-polygon
# boundary intersections during raster extraction. Resolved by averaging
# within the composite key ahupuaa + mokupuni (island).
#
# CRITICAL: ahupuaa names are NOT unique across islands. Seven names recur
# on multiple islands (Hālawa, Kahana, Kalihi, Kawela, Waiheʻe, Wailua, Waimea).
# All joins MUST use ahupuaa + mokupuni as a composite key.
# cesspool_by_ahupuaa uses column name 'island' (same encoding as mokupuni);
# rename to mokupuni before joining.

slope_clean <- slope |>
  group_by(ahupuaa, mokupuni) |>
  summarise(mean_slope_deg = mean(mean_slope_deg, na.rm = TRUE), .groups = "drop")

ksat_clean <- ksat |>
  group_by(ahupuaa, mokupuni) |>
  summarise(mean_ksat_um_s = mean(mean_ksat_um_s, na.rm = TRUE), .groups = "drop")

ccap_clean <- ccap |>
  group_by(ahupuaa, mokupuni) |>
  summarise(
    pct_impervious  = mean(pct_impervious,  na.rm = TRUE),
    pct_agriculture = mean(pct_agriculture, na.rm = TRUE),
    .groups = "drop"
  )

cess_ahu_clean <- cess_ahu |>
  rename(mokupuni = island) |>   # harmonise column name; encoding matches mokupuni
  group_by(ahupuaa, mokupuni) |>
  summarise(
    cesspool_density_per_km2 = mean(cesspool_density_per_km2, na.rm = TRUE),
    n_cesspools              = mean(n_cesspools,              na.rm = TRUE),
    .groups = "drop"
  )

discharge_clean <- discharge |>
  group_by(ahupuaa, mokupuni) |>
  summarise(
    total_cfs_march            = mean(total_cfs_march,            na.rm = TRUE),
    specific_discharge_cfs_km2 = mean(specific_discharge_cfs_km2, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) if_else(is.nan(x), NA_real_, x)))
```

```{r}
#| label: filter-coords-assign-site-id
# One sample has completely missing metadata (NA sample_id, island, coordinates).
# Drop it; all downstream analyses require coordinates.
has_coords <- chem_raw |>
  filter(!is.na(latitude), !is.na(longitude))

# Assign a stable site_id: each unique lat/lon pair is one site.
# This is an exact-coordinate match — coordinates are already stored to ~5 dp.
site_lookup <- has_coords |>
  distinct(latitude, longitude) |>
  arrange(latitude, longitude) |>
  mutate(site_id = sprintf("SITE_%04d", row_number()))

chem_audit <- has_coords |>
  left_join(site_lookup, by = c("latitude", "longitude"))
```

*Table S1 (sample attrition) and Table S2 (missingness matrix) are generated after all
covariates are joined — see the end of [§ Master analytical dataset](#sec-master-dataset).*

```{r}
#| label: coordinate-duplicate-audit
# Classify the 33 multi-sample sites
site_type_tbl <- chem_audit |>
  group_by(site_id) |>
  filter(n() > 1) |>
  summarise(
    n_samples  = n(),
    n_dates    = n_distinct(as.Date(collected_hst)),
    cv_NO3     = sd(NO3NO2, na.rm = TRUE) / mean(NO3NO2, na.rm = TRUE),
    cv_Si      = sd(SiO2,   na.rm = TRUE) / mean(SiO2,   na.rm = TRUE),
    site_type  = case_when(
      n_dates > 1                                                       ~ "same_site_repeat_visit",
      (is.na(cv_NO3) | cv_NO3 < 0.25) & (is.na(cv_Si) | cv_Si < 0.25) ~ "probable_technical_replicate",
      TRUE                                                              ~ "same_coord_independent_samples"
    ),
    .groups = "drop"
  )

site_type_tbl |>
  count(site_type, name = "n_sites") |>
  kable(caption = "Multi-sample sites by classification (33 sites share coordinates with ≥1 other sample).",
        col.names = c("Site type", "n sites")) |>
  kable_styling(bootstrap_options = "condensed", full_width = FALSE)
```

::: {.callout-note}
**Coordinate duplicate decisions.** Of 708 unique lat/lon positions, 33 have ≥2 samples.
Nine sites have nearly identical chemistry on the same date (probable technical replicates),
18 sites have substantially different chemistry on the same date (independent nearby
observations — retain all; shared coordinates may reflect GPS rounding),
and 6 sites have samples from different dates (genuine repeat visits).
All 746 samples are retained for nutrient-specific analyses; hierarchical random effects
(site nested in ahupuaʻa nested in island) will account for pseudoreplication in
landscape-scale models.
:::

```{r}
#| label: build-master-dataset
# Build one-row-per-sample master dataset with all covariates joined.
# This is the base for all downstream analyses.

chem_master <- chem_audit |>
  left_join(ahupuaa |> select(sample_id, ahupuaa, moku, mokupuni),
            by = "sample_id") |>
  left_join(cess_dist,                                          by = "sample_id") |>
  left_join(cess_buf,                                           by = "sample_id") |>
  left_join(fetch |> select(sample_id, fetch_mean_km, fetch_min_km),
            by = "sample_id") |>
  # All ahupuaa-level joins use composite key ahupuaa + mokupuni to prevent
  # cross-island name collisions (7 ahupuaa names recur on multiple islands)
  left_join(slope_clean,    by = c("ahupuaa", "mokupuni")) |>
  left_join(ksat_clean,     by = c("ahupuaa", "mokupuni")) |>
  left_join(ccap_clean,     by = c("ahupuaa", "mokupuni")) |>
  left_join(cess_ahu_clean  |> select(ahupuaa, mokupuni, cesspool_density_per_km2, n_cesspools),
            by = c("ahupuaa", "mokupuni")) |>
  left_join(discharge_clean |> select(ahupuaa, mokupuni, total_cfs_march, specific_discharge_cfs_km2),
            by = c("ahupuaa", "mokupuni"))

# site_lookup saved here; chem_master saved after BDL substitution and cesspool correction
write_csv(site_lookup, here("data", "site_lookup.csv"))
```



::: {.callout-warning}
**Key data gaps to note before modeling.**

- **USGS stream discharge** (61.5% missing): **this is gauge-coverage missingness, not
  streamlessness.** Of the 470 ungauged samples, 91% (n = 426) are within 5 km of an
  OSM stream mouth, and median distance to the nearest stream mouth is ≤ 0.5 km on
  Maui, Molokaʻi, and Lānaʻi. USGS gauges exist only for major perennial streams; the
  many smaller coastal drainages that were sampled are simply ungauged. Treating missing
  discharge as zero would be scientifically incorrect. Use `dist_to_stream_mouth_m`
  (computed from OSM stream mouths; now in `chem_master`) as the primary within-watershed
  connectivity variable; use spatially resolved CHIRPS storm rainfall as the storm-forcing
  variable. Reserve USGS discharge for sensitivity analysis and supplementary context.
- **NH₄** (18.5% missing): missingness is distributed across islands; NH₄-specific models
  will use a smaller sample. Do not impute; report n separately.
- **Cesspool metrics — island-specific data gaps.** HCPT layer 32 does not cover Molokaʻi
  (103 samples); all cesspool variables (`dist_to_nearest_cesspool_m`,
  `cesspool_density_per_km2`, `n_cess_500m`, `n_cess_1km`) are set to `NA` for Molokaʻi.
  For Lānaʻi (20 samples), Lānaʻi does not appear in `cesspool_by_ahupuaʻa.csv` —
  `cesspool_density_per_km2` is `NA` for all 20 samples. Buffer counts (`n_cess_500m`,
  `n_cess_1km`) are correctly 0 (no cesspools in the HCPT layer within those buffers
  on-island). `dist_to_nearest_cesspool_m` is `NA` because sample-level distances
  (16–34 km) referenced cesspools on neighboring islands. An earlier observation of
  "9 cesspools in Kamoku (Lānaʻi)" was a cross-island join artefact: that record
  belongs to Kamoku ahupuaʻa on Hawaiʻi Island; the composite-key fix eliminated this.
  For inferential models, exclude Molokaʻi from cesspool analyses; for Lānaʻi, only
  buffer counts (= 0) are usable — no density or distance metric is available.
:::

```{r}
#| label: bdl-substitution
# All reported zeros are below-detection-limit (BDL), not true zeros.
# Substitution rule: replace with DL/2 where DL = minimum observed non-zero value.
# BDL indicator columns (bdl_*) retain the original flag for sensitivity analyses.

dl_half <- c(
  NO3NO2 = min(chem_master$NO3NO2[chem_master$NO3NO2 > 0], na.rm = TRUE) / 2,
  PO4    = min(chem_master$PO4[chem_master$PO4 > 0],       na.rm = TRUE) / 2,
  NH3    = min(chem_master$NH3[chem_master$NH3 > 0],       na.rm = TRUE) / 2,
  SiO2   = min(chem_master$SiO2[chem_master$SiO2 > 0],     na.rm = TRUE) / 2
)

chem_master <- chem_master |>
  mutate(
    bdl_NO3NO2 = !is.na(NO3NO2) & NO3NO2 == 0,
    bdl_PO4    = !is.na(PO4)    & PO4    == 0,
    bdl_NH3    = !is.na(NH3)    & NH3    == 0,
    bdl_SiO2   = !is.na(SiO2)  & SiO2   == 0,
    NO3NO2 = if_else(bdl_NO3NO2, dl_half["NO3NO2"], NO3NO2),
    PO4    = if_else(bdl_PO4,    dl_half["PO4"],    PO4),
    NH3    = if_else(bdl_NH3,    dl_half["NH3"],     NH3),
    SiO2   = if_else(bdl_SiO2,  dl_half["SiO2"],    SiO2)
  )

# Summary table
bdl_tbl <- tibble(
  Nutrient    = c("NO\u2083+NO\u2082", "PO\u2084", "NH\u2084", "SiO\u2082"),
  n_bdl       = c(sum(chem_master$bdl_NO3NO2), sum(chem_master$bdl_PO4),
                  sum(chem_master$bdl_NH3),    sum(chem_master$bdl_SiO2)),
  n_available = c(sum(!is.na(chem_master$NO3NO2)), sum(!is.na(chem_master$PO4)),
                  sum(!is.na(chem_master$NH3)),     sum(!is.na(chem_master$SiO2))),
  pct_bdl     = round(100 * n_bdl / n_available, 1),
  `DL/2 substituted (umol/L)` = round(dl_half, 4)
)

bdl_tbl |>
  kable(caption = "BDL substitution summary. All zeros replaced with DL/2 (half the minimum observed non-zero value). BDL flags retained in `bdl_*` columns for sensitivity analyses.",
        align = c("l", "r", "r", "r", "r")) |>
  kable_styling(bootstrap_options = "condensed", full_width = FALSE)
```

::: {.callout-warning}
**NH₄ has 30.4% BDL observations.** DL/2 substitution is defensible for modest BDL fractions,
but at 30% the substituted values can meaningfully inflate apparent precision.
The NH₄ model should be treated as a candidate for censored regression (e.g., `survival::Surv`
left-censored, or a Bayesian truncated-normal likelihood in `brms`). A sensitivity analysis
comparing DL/2 results to a censored model is required for any inferential claim about NH₄.
:::

```{r}
#| label: cesspool-island-corrections
# Molokaʻi: HCPT layer 32 has no coverage — all cesspool metrics set to NA.
# Lānaʻi: no on-island cesspools in the layer for most ahupuaʻa; buffer counts
# and density are correctly 0 where present, but dist_to_nearest_cesspool_m
# values (16–34 km) reference off-island sources and are set to NA.
# Molokaʻi: layer does not cover island → all cesspool metrics NA.
# Lānaʻi: island absent from cesspool_by_ahupuaa → density NA.
#   dist_to_nearest set to NA (sample-level values pointed off-island, 16–34 km).
#   Buffer counts correctly 0 (no on-island cesspools in HCPT layer).
#   The earlier "Kamoku 9-cesspool" flag was a cross-island join artefact
#   (Hawaiʻi Island's Kamoku); resolved by the composite ahupuaa+mokupuni key.
chem_master <- chem_master |>
  mutate(
    dist_to_nearest_cesspool_m = if_else(island == "Moloka'i",             NA_real_, dist_to_nearest_cesspool_m),
    n_cess_500m                = if_else(island == "Moloka'i",             NA_real_, n_cess_500m),
    n_cess_1km                 = if_else(island == "Moloka'i",             NA_real_, n_cess_1km),
    cesspool_density_per_km2   = if_else(island == "Moloka'i",             NA_real_, cesspool_density_per_km2),
    n_cesspools                = if_else(island == "Moloka'i",             NA_real_, n_cesspools),
    dist_to_nearest_cesspool_m = if_else(grepl("n.?a.?i", island, ignore.case = TRUE),
                                          NA_real_, dist_to_nearest_cesspool_m)
  )
```

```{r}
#| label: dist-to-stream-mouth
# Compute distance from each sample to the nearest OSM stream mouth.
# Source: osm_stream_mouths.csv (OpenStreetMap, filtered to 400 m coastal buffer).
# Missing discharge data reflects gauge-coverage gaps, not absent streams:
# 91% of ungauged samples are within 5 km of a mapped stream mouth.
# dist_to_stream_mouth_m is therefore the primary connectivity variable.

stream_mouths <- read_csv(here("data", "osm_stream_mouths.csv"), show_col_types = FALSE)

samples_sf <- chem_master |>
  select(sample_id, latitude, longitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)   # UTM zone 4N

streams_sf <- stream_mouths |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

nearest_idx <- st_nearest_feature(samples_sf, streams_sf)
chem_master <- chem_master |>
  mutate(dist_to_stream_mouth_m = as.numeric(
    st_distance(samples_sf, streams_sf[nearest_idx, ], by_element = TRUE)
  ))
```

```{r}
#| label: offshore-flags
# Flag samples that fall outside the stream-connectivity framework.
# offshore_reference: samples in the "Open Ocean" ahupuaa — deliberately
#   collected far offshore as marine background (n = 11; 15–114 km from
#   any stream mouth). Exclude from connectivity and landscape models;
#   retain as marine background context.
# far_from_stream: dist_to_stream_mouth_m > 10 km (n = 35 total).
#   Includes the 11 offshore reference samples plus 24 leeward O'ahu
#   coastal samples (Honouliuli, Nānākuli, Waiʻanae, Waipiʻo ahupuaʻa)
#   in genuinely stream-sparse dry terrain. Leeward samples retain
#   valid CHIRPS land-pixel data and are kept in landscape models with
#   the caveat that stream-proximity effects are poorly constrained there.

chem_master <- chem_master |>
  mutate(
    offshore_reference = ahupuaa == "Open Ocean",
    far_from_stream    = dist_to_stream_mouth_m > 10000
  )

# CHIRPS -9999 fill values indicate ocean/off-land pixels — same set as offshore samples.
# These receive NA for all CHIRPS metrics.
```

```{r}
#| label: chirps-storm-rainfall
# Spatially resolved CHIRPS daily rainfall, March 10–31 2026 (storm onset to end).
# Extracted at 0.05° (~5 km) resolution for all unique sample pixel locations.
# Source: CHC server (data.chc.ucsb.edu); cached to data/chirps_march2026_daily_raw.rds.
# April 1–8 data (34 samples) not currently available from CHC; rainfall during
# this post-storm window was near-zero (0–2.4 mm/day in test pixels), so the
# March 10–31 cumulative captures effectively all storm rainfall.
# CHIRPS fill value -9999 (ocean pixels) set to NA.

library(chirps)

chirps_march_all <- readRDS(here("data", "chirps_march2026_daily_raw.rds"))

unique_locs <- chem_master |>
  mutate(chirps_lon = round(longitude, 2), chirps_lat = round(latitude, 2)) |>
  distinct(chirps_lon, chirps_lat) |>
  arrange(chirps_lat, chirps_lon) |>
  mutate(id = row_number())

chirps_pixel <- as_tibble(chirps_march_all) |>
  left_join(unique_locs, by = "id") |>
  mutate(chirps = if_else(chirps == -9999, NA_real_, chirps)) |>
  group_by(chirps_lon, chirps_lat) |>
  summarise(
    chirps_storm_cumulative_mm = sum(chirps, na.rm = TRUE),
    chirps_peak_day_mm         = max(chirps, na.rm = TRUE),
    chirps_storm_days_nonzero  = sum(chirps > 0, na.rm = TRUE),
    chirps_ocean_pixel         = any(is.na(chirps)),
    .groups = "drop"
  ) |>
  mutate(
    chirps_storm_cumulative_mm = if_else(chirps_ocean_pixel, NA_real_, chirps_storm_cumulative_mm),
    chirps_peak_day_mm         = if_else(chirps_ocean_pixel | is.infinite(chirps_peak_day_mm),
                                          NA_real_, chirps_peak_day_mm)
  )

chem_master <- chem_master |>
  mutate(chirps_lon = round(longitude, 2), chirps_lat = round(latitude, 2)) |>
  left_join(
    chirps_pixel |> select(chirps_lon, chirps_lat, chirps_storm_cumulative_mm,
                           chirps_peak_day_mm, chirps_storm_days_nonzero, chirps_ocean_pixel),
    by = c("chirps_lon", "chirps_lat")
  )
```

```{r}
#| label: save-master-dataset
write_csv(chem_master, here("data", "chem_master.csv"))
```

---

# Sampling Coverage {#sec-sampling-coverage}

> **Step 6 (Figure 1).** Four-island map, event timeline, summary statistics.
> Show the scale of the natural experiment — do not show nutrient concentrations yet.

*← Analysis code goes here when we work through Step 6 together.*

---

# Event Magnitude and Long-Term Context {#sec-event-magnitude}

## Nutrient enrichment — storm vs background {#sec-enrichment}

> **Step 7 (Figure 2).** Quantify the magnitude of nutrient enrichment.
> Four-panel: NO₃+NO₂, PO₄, NH₄, Si — storm-influenced vs marine-background.

*← Analysis code goes here when we work through Step 7 together.*

## Long-term Hui baseline comparison {#sec-hui-baseline}

> **Step 8.** Load Hui o Ka Wai Ola data, compute station-specific historical baselines,
> calculate fold-changes relative to pre-2026 medians.

*← Analysis code goes here when we work through Step 8 together.*

## Was 2026 statistically extreme? {#sec-extreme-test}

> **Step 9 (Figure 3).** Test whether March 2026 anomalies exceed the range expected
> from the decade-long rainfall–nutrient relationship.

*← Analysis code goes here when we work through Step 9 together.*

---

# Spatial Nutrient Patterns {#sec-spatial-patterns}

## Spatial autocorrelation structure {#sec-morans}

> **Step 10.** Moran's I, spatial correlograms, variograms — per nutrient per island.

*← Analysis code goes here when we work through Step 10 together.*

## Interpolated maps {#sec-kriging}

> **Step 10 (Figure 4).** Spatial interpolation for visualization only.
> Mask to coastal band; show observed points on top.

*← Analysis code goes here when we work through Step 10 together.*

---

# Terrestrial Connectivity {#sec-connectivity}

## Master analytical dataset {#sec-master-dataset}

> **Steps 2–5.** Load and join all spatial covariates; build `analysis_master`;
> missingness matrix; predictor correlations; define four predictor groups.

```{r}
#| label: step2-load-remaining-covariates
# Load the five covariate sources not yet joined in §2 (Data Inventory):
#   wave height, population density, soil moisture, livestock, NPDES distance.
# All ahupuaa-level joins use composite key ahupuaa + mokupuni.
# pop_density uses 'island' with ʻokina encoding = mokupuni.
# soilmoist and livestock use straight-apostrophe encoding = island;
#   apostrophe variants normalised before joining.

waveheight <- read_csv(here("data", "sample_waveheight.csv"),      show_col_types = FALSE)
pop_dens   <- read_csv(here("data", "pop_density_ahupuaa.csv"),    show_col_types = FALSE)
epa        <- read_csv(here("data", "epa_facilities_hi.csv"),      show_col_types = FALSE)
soilmoist  <- read_csv(here("data", "nasa_power_soilmoist.csv"),   show_col_types = FALSE)
livestock  <- read_csv(here("data", "usda_nass_livestock_hi.csv"), show_col_types = FALSE)

# Normalise apostrophe variants for island-level joins
normalize_island <- function(x) str_replace_all(x, "['\u2018\u2019\u02bc]", "'")

# 1. Wave height — sample_id join
chem_master <- chem_master |>
  left_join(waveheight |> select(sample_id, hs_at_collection_m, hs_mean_m, hs_max_m),
            by = "sample_id")

# 2. Population density — composite key (pop_dens$island uses ʻokina = mokupuni)
pop_clean <- pop_dens |>
  rename(mokupuni = island) |>
  group_by(ahupuaa, mokupuni) |>
  summarise(pop_density_km2 = mean(pop_density_km2, na.rm = TRUE), .groups = "drop")

chem_master <- chem_master |>
  left_join(pop_clean |> select(ahupuaa, mokupuni, pop_density_km2),
            by = c("ahupuaa", "mokupuni"))

# 3. Soil moisture & 4. Livestock — island-level; normalise apostrophe first
chem_master <- chem_master |>
  mutate(island_key = normalize_island(island)) |>
  left_join(soilmoist |> mutate(island = normalize_island(island)) |>
              select(island, mean_gwetroot),
            by = c("island_key" = "island")) |>
  left_join(livestock |> mutate(island = normalize_island(island)),
            by = c("island_key" = "island")) |>
  select(-island_key)

# 5. NPDES distance — spatial computation
npdes_pts <- epa |>
  filter(!is.na(lat), !is.na(lon), between(lat, 18, 23), between(lon, -161, -154)) |>
  distinct(lat, lon) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(32604)

samples_sf <- chem_master |>
  select(sample_id, latitude, longitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

nearest_npdes <- st_nearest_feature(samples_sf, npdes_pts)
chem_master <- chem_master |>
  mutate(
    dist_to_nearest_npdes_m = as.numeric(
      st_distance(samples_sf, npdes_pts[nearest_npdes, ], by_element = TRUE)
    ),
    log_dist_npdes = log(dist_to_nearest_npdes_m)
  )
```

::: {.callout-note}
**Step 2 missingness summary.**

| Variable | n missing | % | Note |
|---|--:|--:|---|
| `total_cfs_march` | 470 | 63.0 | Gauge coverage gap — not absent streams |
| `cesspool_density_per_km2` | 165 | 22.1 | Molokaʻi (103) + Lānaʻi (20) + ungauged ahupuaʻa |
| `dist_to_nearest_cesspool_m` | 123 | 16.5 | Molokaʻi (103) + Lānaʻi (20) |
| `n_cess_500m / n_cess_1km` | 103 | 13.8 | Molokaʻi only (layer 32 gap) |
| `chirps_storm_cumulative_mm` | 32 | 4.3 | Ocean-pixel locations (offshore samples) |
| `mean_ksat_um_s` | 19 | 2.5 | Ahupuaʻa with no SSURGO coverage |
| `mean_slope_deg / pct_impervious / pct_agriculture / pop_density_km2` | 11 | 1.5 | Open Ocean ahupuaʻa |
| All other predictors | 0 | 0 | — |

Apostrophe encoding mismatch between `soilmoist`/`livestock` (ASCII `'`, byte 39) and
`chem_master$island` (Unicode left-quote U+2018, byte 8216) — resolved by the
`normalize_island()` helper before joining.
:::

```{r}
#| label: sample-flow-table

# Helper: n samples, n unique sites, n islands for any filtered tibble
flow_row <- function(label, df) {
  tibble(
    Stage       = label,
    `n samples` = nrow(df),
    `n sites`   = df |> distinct(latitude, longitude) |> nrow(),
    `n islands` = df |> distinct(island) |> nrow()
  )
}

full_ids <- chem_master |>
  filter(
    !is.na(ahupuaa),
    !is.na(pct_impervious),
    !is.na(chirps_storm_cumulative_mm),
    !is.na(hs_mean_m),
    !is.na(dist_to_nearest_npdes_m),
    !is.na(pop_density_km2)
  ) |>
  pull(sample_id)

bind_rows(
  tibble(Stage       = "Raw chemistry dataset",
         `n samples` = nrow(chem_raw),
         `n sites`   = chem_raw |> filter(!is.na(latitude)) |>
                         distinct(latitude, longitude) |> nrow(),
         `n islands` = chem_raw |> filter(!is.na(island)) |>
                         distinct(island) |> nrow()),
  flow_row("Coordinates available",             chem_master),
  flow_row("NO\u2083+NO\u2082 available",       chem_master |> filter(!is.na(NO3NO2))),
  flow_row("PO\u2084 available",                chem_master |> filter(!is.na(PO4))),
  flow_row("NH\u2084 available",                chem_master |> filter(!is.na(NH3))),
  flow_row("Si available",                      chem_master |> filter(!is.na(SiO2))),
  flow_row("Ahupua\u02BBa assigned",            chem_master |> filter(!is.na(ahupuaa))),
  flow_row("USGS stream discharge",             chem_master |> filter(!is.na(total_cfs_march))),
  flow_row("Land-cover data (C-CAP)",           chem_master |> filter(!is.na(pct_impervious))),
  flow_row("Cesspool data (sample-level dist)", chem_master |> filter(!is.na(dist_to_nearest_cesspool_m))),
  flow_row("CHIRPS storm rainfall",             chem_master |> filter(!is.na(chirps_storm_cumulative_mm))),
  flow_row("Wave height (WW3 PacIOOS)",         chem_master |> filter(!is.na(hs_mean_m))),
  flow_row("Population density",                chem_master |> filter(!is.na(pop_density_km2))),
  flow_row("NPDES outfall distance",            chem_master |> filter(!is.na(dist_to_nearest_npdes_m))),
  flow_row("Full landscape model dataset",      chem_master |> filter(sample_id %in% full_ids))
) |>
  kable(caption = paste0("Table S1. Sample availability through data stages (n = ",
                         nrow(chem_master), " samples with coordinates)."),
        align = c("l", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r}
#| label: missingness-matrix

chem_master |>
  summarise(across(
    c(NO3NO2, PO4, NH3, SiO2,
      total_cfs_march, dist_to_stream_mouth_m, chirps_storm_cumulative_mm,
      mean_slope_deg, mean_ksat_um_s, mean_gwetroot,
      fetch_mean_km, hs_mean_m, hs_at_collection_m,
      cesspool_density_per_km2, dist_to_nearest_cesspool_m,
      n_cess_500m, n_cess_1km,
      pct_impervious, pct_agriculture,
      pop_density_km2, dist_to_nearest_npdes_m, cattle_per_km2),
    ~ sum(is.na(.x)),
    .names = "{.col}"
  )) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "n_missing") |>
  mutate(
    pct_missing = round(100 * n_missing / nrow(chem_master), 1),
    Group = case_when(
      Variable %in% c("NO3NO2", "PO4", "NH3", "SiO2")
        ~ "A — Response",
      Variable %in% c("total_cfs_march", "dist_to_stream_mouth_m",
                      "chirps_storm_cumulative_mm", "mean_slope_deg",
                      "mean_ksat_um_s", "mean_gwetroot",
                      "fetch_mean_km", "hs_mean_m", "hs_at_collection_m")
        ~ "B — Hydrologic / physical",
      Variable %in% c("cesspool_density_per_km2", "dist_to_nearest_cesspool_m",
                      "n_cess_500m", "n_cess_1km",
                      "pct_impervious", "pct_agriculture",
                      "pop_density_km2", "dist_to_nearest_npdes_m", "cattle_per_km2")
        ~ "C — Anthropogenic"
    )
  ) |>
  arrange(Group, desc(n_missing)) |>
  select(Group, Variable, n_missing, pct_missing) |>
  kable(caption = paste0("Table S2. Missingness across all predictor variables (n = ",
                         nrow(chem_master), " samples with coordinates)."),
        col.names = c("Group", "Variable", "n missing", "% missing"),
        align = c("l", "l", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Si connectivity model {#sec-si-model}

> **Step 11.** Model log(Si) as a function of physical/hydrologic predictors.
> First mechanistic model; establishes terrestrial-water delivery baseline.

*← Analysis code goes here when we work through Step 11 together.*

---

# Human Amplification {#sec-human-amplification}

## NO₃+NO₂ model — the central analysis {#sec-nox-model}

> **Step 12.** log(NO₃+NO₂) ~ log(Si) + cesspool + agriculture + impervious + ...
> Key question: do anthropogenic variables explain variation *beyond* hydrologic delivery?

*← Analysis code goes here when we work through Step 12 together.*

## PO₄ model {#sec-po4-model}

> **Step 13.** Parallel PO₄ model. Same predictor groups as NO₃+NO₂.

*← Analysis code goes here when we work through Step 13 together.*

## NH₄ model — the contrast {#sec-nh4-model}

> **Step 14.** NH₄ model. Expect weaker explanatory power — treat that as a result.

*← Analysis code goes here when we work through Step 14 together.*

---

# Nutrient-Specific Models — Comparison {#sec-model-comparison}

## Standardized effect sizes (Figure 5) {#sec-effect-sizes}

> **Step 15.** Coefficient plot — one row per predictor, four columns (nutrients).
> Group rows: hydrologic forcing / watershed transport / anthropogenic / coastal dilution.

*← Analysis code goes here when we work through Step 15 together.*

## Explicit human-amplification test {#sec-amplification-test}

> **Step 17.** Compare Model A (physical only) vs B (+anthropogenic) vs C (+interactions).

*← Analysis code goes here when we work through Step 17 together.*

## Priority interactions {#sec-interactions}

> **Step 21.** StreamDistance × Slope; Rainfall × CesspoolDensity;
> Rainfall × Agriculture; Rainfall × Ksat.

*← Analysis code goes here when we work through Step 21 together.*

---

# Variance Partitioning {#sec-variance-partitioning}

> **Step 16 (Figure 6).** Unique and shared explanatory power for each driver group
> across all four nutrients.

*← Analysis code goes here when we work through Step 16 together.*

---

# Out-of-Sample Hotspot Prediction {#sec-hotspot-prediction}

## Spatial cross-validation {#sec-spatial-cv}

> **Step 18.** Blocked spatial CV and leave-one-island-out CV.
> Report R², RMSE, MAE, calibration slope.

*← Analysis code goes here when we work through Step 18 together.*

## Random forest — applied prediction {#sec-random-forest}

> **Step 19.** Continuous-response RF for all four nutrients.
> Use for hotspot identification, not inference.

*← Analysis code goes here when we work through Step 19 together.*

## Vulnerability maps {#sec-vulnerability-maps}

> **Step 20.** Cross-validated predicted nutrient concentrations mapped to coastline.
> Mask extrapolation beyond training predictor space.

*← Analysis code goes here when we work through Step 20 together.*

---

# Sensitivity Analyses {#sec-sensitivity}

> **Step 23.** Repeat key results under 11 analytical alternatives.
> Produce supplementary stability table.

*← Analysis code goes here when we work through Step 23 together.*

---

# Publication Figures {#sec-pub-figures}

> Final, camera-ready assembly of all six main figures.
> Saved to `output/main_figures/` as PDF + PNG (600 dpi).

*← Assembled when all upstream analysis sections are complete.*

---

# Supplementary Analyses {#sec-supplementary}

## Nutrient stoichiometry {#sec-stoichiometry}

> **Step 22.** Si:N and N:P ratios as supplementary descriptive context only.

*← Analysis code goes here when we work through Step 22 together.*

## Pseudoreplication diagnostics {#sec-pseudoreplication}

> **Step 24.** Document hierarchical structure; demonstrate random-effects approach
> compared to naïve sample-level analysis.

*← Analysis code goes here when we work through Step 24 together.*

---

# Model Diagnostics {#sec-diagnostics}

> **Step 27.** Automated diagnostic figures for every final model.
> Residual vs fitted, QQ, spatial residual map, residual Moran's I,
> leverage/influence, observed vs predicted, CV performance.

*← Generated automatically once models in §6–8 are finalised.*

---

# Reproducibility Information {#sec-reproducibility}

> **Step 34.** Session info, checklist, and "What the analyses actually support."

```{r}
#| label: session-info
#| eval: false   # set to true when producing the final document

sessioninfo::session_info() |>
  capture.output() |>
  writeLines(here("output", "session_info.txt"))
```

```{r}
#| label: checklist
#| eval: false   # evaluate interactively as steps are completed

checklist <- tribble(
  ~item,                                          ~done,
  "Sample attrition audited",                     TRUE,
  "Response-specific sample sizes reported",       TRUE,
  "Repeated sites identified",                     TRUE,
  "Predictor collinearity evaluated",              FALSE,
  "Spatial rainfall calculated",                   FALSE,
  "Long-term Hui comparison completed",            FALSE,
  "Four nutrient spatial structures evaluated",    FALSE,
  "Si connectivity model completed",              FALSE,
  "NO3+NO2 human-amplification model completed",  FALSE,
  "PO4 model completed",                          FALSE,
  "NH4 model completed",                          FALSE,
  "Model residual spatial autocorrelation checked",FALSE,
  "Hierarchical structure accounted for",          FALSE,
  "Spatial cross-validation completed",            FALSE,
  "Leave-one-island-out prediction completed",     FALSE,
  "Variance partitioning completed",               FALSE,
  "Key interactions tested",                       FALSE,
  "Sensitivity analyses completed",                FALSE,
  "Main figures exported",                         FALSE,
  "Supplementary figures exported",                FALSE,
  "Tables exported",                               FALSE,
  "All model objects saved",                       FALSE,
  "session_info() saved",                          FALSE,
  "All claims updated to match actual results",    FALSE
)

checklist |>
  mutate(status = if_else(done, "\u2713", "\u25cb")) |>
  select(status, item) |>
  kable(col.names = c("", "Checklist item"), align = "cl") |>
  kable_styling(bootstrap_options = c("condensed"), full_width = FALSE,
                font_size = 12) |>
  row_spec(which(checklist$done), color = "gray50")
```

---

*Document rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`*
*Global seed: `r SEED`*
