---
title: "Nutrient hotspot maps — post-Kona low storm"
subtitle: "Spatial interpolation of nutrient concentrations across Hawaiʻi, March–April 2026"
author: "Silbiger, N., Kealoha, A., & Kahanamoku, S."
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: flatly
    code-fold: true
    fig-width: 9
    fig-height: 6.5
    lang: en
    css: |
      body {
        font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, "Helvetica Neue", Arial, sans-serif;
      }
execute:
  warning: false
  message: false
---


```{r}
#| label: load-packages

# Core
library(tidyverse)
library(scales)
library(knitr)
library(patchwork)
library(here)

# Spatial
library(sf)
library(gstat)
library(stars)
library(terra)
library(osmdata)
library(maptiles)
library(tidyterra)

# Raster extraction
library(exactextractr)

# Visualisation helpers
library(cowplot)
library(ggrepel)
library(ggdendro)
library(kableExtra)

# Clustering & spatial statistics
library(cluster)
library(spdep)
library(zoo)

# Machine learning & partial dependence
library(ranger)
library(pdp)

# Note: dataRetrieval and nhdplusTools are used only in the eval:false
# roadmap chunks at the end of this document. Install separately if needed.
```

# Introduction

 *Local human impacts amplify storm disruption to coastal marine ecosystems*
When a Kona Low storm hits Hawaiʻi, heavy rain washes fertilizer, cesspool nutrients, and volcanic soil off the land and into the ocean. Does that change coastal water quality in a meaningful way and if so, where, how much, and can we predict it from a map?

This report uses 694 water samples collected across four islands immediately after the March 2026 Kona Low to answer those questions. Interpolated maps show where nutrient concentrations were highest. A grouping analysis (PAM clustering) sorts sites into three distinct water types — from clean open-ocean background water through runoff-dominated coastal zones to nearshore-enriched areas — and measures how different their chemistry is. A random forest model then asks whether those water types could have been predicted from landscape features alone: how close is a site to a stream mouth? To a cesspool? How much of the surrounding land is paved or farmed? How much water actually flowed through the nearest stream gauge in March? The model correctly classifies 67% of sites (vs 36% for a random guess), with stream mouth proximity, cesspool distance, and actual March discharge as the three strongest predictors.

# Data overview

```{r}
#| label: load-data
chem <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE) |>
  filter(!is.na(island)) |>
  # Normalise apostrophe variants (ʻokina U+02BB, smart quotes U+2018/9) to
  # plain ASCII apostrophe so island-name joins are consistent throughout
  mutate(island = stringr::str_replace_all(island, "[ʻʼ‘’]", "'")) |>
  # Ahupuaʻa assignment (spatial join pre-computed in 01_load_clean_data.R)
  left_join(
    read_csv("data/ahupuaa_cache.csv", show_col_types = FALSE),
    by = "sample_id"
  ) |>
  # Distance to nearest cesspool per sample
  left_join(
    read_csv("data/sample_cesspool_distances.csv", show_col_types = FALSE),
    by = "sample_id"
  ) |>
  # Cesspool density per ahupuaʻa (join on ahupuaa + moku to avoid duplicate
  # ahupuaʻa names appearing in multiple moku / islands)
  left_join(
    read_csv("data/cesspool_by_ahupuaa.csv", show_col_types = FALSE) |>
      select(ahupuaa, moku, cesspool_density_per_km2, n_cesspools),
    by = c("ahupuaa", "moku")
  ) |>
  # Cesspool count within 500 m and 1 km of each sample (point-level)
  # Note: Molokaʻi set to NA — HCPT layer 32 does not cover Molokaʻi;
  # imputed with island median in RF pipeline. Lānaʻi = 0 (no layer documented).
  left_join(
    read_csv("data/cesspool_buffer_counts.csv", show_col_types = FALSE),
    by = "sample_id"
  )
```

`r nrow(chem)` samples collected across `r n_distinct(chem$island)` islands between `r format(min(chem$collected_hst, na.rm = TRUE), "%B %d, %Y")` and `r format(max(chem$collected_hst, na.rm = TRUE), "%B %d, %Y")`. Cesspool density is unavailable for Lānaʻi (no public data) and `r chem |> filter(island != "Lāna'i", is.na(cesspool_density_per_km2)) |> nrow()` samples on other islands where the ahupuaʻa is not in the cesspool dataset.

```{r}
#| label: data-summary-table


chem |>
  group_by(island) |>
  summarise(
    n          = n(),
    dates      = paste0(
                   format(min(collected_hst, na.rm = TRUE), "%b %d"),
                   "–",
                   format(max(collected_hst, na.rm = TRUE), "%b %d")
                 ),
    med_NO3NO2 = median(NO3NO2,  na.rm = TRUE),
    max_NO3NO2 = max(NO3NO2,     na.rm = TRUE),
    med_PO4    = median(PO4,     na.rm = TRUE),
    max_PO4    = max(PO4,        na.rm = TRUE),
    med_NH3    = median(NH3,     na.rm = TRUE),
    max_NH3    = max(NH3,        na.rm = TRUE),
    med_SiO2   = median(SiO2,    na.rm = TRUE),
    max_SiO2   = max(SiO2,       na.rm = TRUE),
    med_Sal    = median(Salinity, na.rm = TRUE),
    min_Sal    = min(Salinity,    na.rm = TRUE),
    .groups = "drop"
  ) |>
  kable(
    col.names = c(
      "Island", "n", "Dates",
      "Median", "Max",
      "Median", "Max",
      "Median", "Max",
      "Median", "Max",
      "Median", "Min"
    ),
    digits  = 2,
    escape  = FALSE,
    caption = "Sample counts and nutrient summary statistics by island. Nutrient concentrations in &mu;mol/L; salinity in PSU. Salinity minimum reflects the most freshwater-influenced site sampled on each island.",
    align   = c("l", "c", "c", rep("r", 10))
  ) |>
  add_header_above(c(
    " "                                        = 3,
    "NO<sub>3</sub>+NO<sub>2</sub> (&#956;mol/L)" = 2,
    "PO<sub>4</sub> (&#956;mol/L)"               = 2,
    "NH<sub>3</sub> (&#956;mol/L)"               = 2,
    "SiO<sub>2</sub> (&#956;mol/L)"              = 2,
    "Salinity (PSU)"                            = 2
  ), escape = FALSE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE,
    font_size         = 13
  ) |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, color = "white", background = "#37474F") |>
  footnote(
    general           = "NO<sub>3</sub>+NO<sub>2</sub>, PO<sub>4</sub>, NH<sub>3</sub>, SiO<sub>2</sub>: concentrations in &mu;mol/L. Salinity in PSU. Median and Max shown per nutrient; Min shown for Salinity (lower values indicate greater freshwater influence).",
    general_title     = "",
    footnote_as_chunk = TRUE,
    escape            = FALSE
  )
```

```{r}
#| label: load-spatial
# Load ahupuaʻa boundaries as the island land mask. These are more accurate
# than Natural Earth (which clips East Maui and misrepresents the isthmus).
ahupuaa_raw <- st_read("data/ahupuaa_boundaries.geojson", quiet = TRUE)
```

# Nutrient hotspot maps

Each map shows a kriged surface of NO₃+NO₂ (µmol/L) interpolated from coastal water samples collected after the storm. White circles mark sample collection points. The fill scale is logarithmic (log₁p) to show the full range of values without the highest concentrations washing out the spatial pattern. Land is masked in gray.

```{r}
#| label: helpers

normalize_name <- function(x) {
  x |>
    stringr::str_replace_all("[ʻ‘’']", "") |>  # strip okina/smart quotes
    stringr::str_replace_all("ā", "a") |>                 # ā → a
    stringr::str_to_lower() |>
    stringr::str_trim()
}

# Shared island colour palette — ASCII apostrophes throughout for consistent matching
island_pal <- c(
  "O'ahu"    = "#1565C0",
  "Maui"     = "#2E7D32",
  "Moloka'i" = "#E65100",
  "Lāna'i"   = "#6A1B9A"
)

# Canonical recode: plain cache names → display names (matching island_pal keys).
# Use as: mutate(island = recode_island[island])
recode_island <- c(
  "Oahu"    = "O'ahu",
  "Maui"    = "Maui",
  "Molokai" = "Moloka'i",
  "Lanai"   = "Lāna'i"
)

# Scale metadata for each plottable variable. Add entries here to support new variables.
# `trans` drives both the internal kriging transform and the ggplot scale transform.
# `oob_label` is the label used for the top break (values squished above the limit).
var_meta <- list(
  NO3NO2 = list(
    label     = expression(atop(NO[3]+NO[2], "("*mu*"mol/L)")),
   # limits    = c(0, 20),
  #  breaks    = c(0.5, 1, 3, 10, 20),
   # labels    = c("0.5", "1", "3", "10", ">=20"),
     limits    = c(0, 10),
    breaks    = c(0.5, 1, 3, 5, 10),
    labels    = c("0.5", "1", "3", "5", "≥10"),
     trans     = "log1p",
    direction = 1
  ),
  PO4 = list(
    label     = expression(atop(PO[4], "("*mu*"mol/L)")),
    limits    = c(0, 1),
    breaks    = c(0.05, 0.1, 0.3, 0.5, 1),
    labels    = c("0.05", "0.1", "0.3", "0.5", "≥1"),
    trans     = "log1p",
    direction = 1
  ),
  SiO2 = list(
    label     = expression(atop(SiO[2], "("*mu*"mol/L)")),
   # limits    = c(0, 100),
  #  breaks    = c(1, 5, 20, 50, 100),
  #  labels    = c("1", "5", "20", "50", "≥100"),
     limits    = c(0, 50),
    breaks    = c(1, 5, 10, 20, 50),
    labels    = c("1", "5", "10", "20", "≥50"),
    trans     = "log1p",
    direction = 1
  ),
  NH3 = list(
    label     = expression(atop(NH[3], "("*mu*"mol/L)")),
    limits    = c(0, 3),
    breaks    = c(0.1, 0.5, 1, 2, 3),
    labels    = c("0.1", "0.5", "1", "2", "≥3"),
    trans     = "log1p",
    direction = 1
  ),
  Salinity = list(
    label     = expression(atop("Salinity", "(PSU)")),
    limits    = c(28, 36),
    breaks    = c(28, 30, 32, 34, 36),
    labels    = c("≤28", "30", "32", "34", "36"),
    trans     = "identity",
    direction = -1
  )
)

krige_island_map <- function(data, island_name, ahupuaa_sf,
                              var = "NO3NO2",
                              res_m = 250, band_km = 3, tile_zoom = 11) {
  
  meta    <- var_meta[[var]]
  use_log <- !is.null(meta) && meta$trans == "log1p"
  if (is.null(meta)) meta <- list(
    label = var, limits = NULL, breaks = waiver(), labels = waiver(), trans = "identity"
  )
  
  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]), !is.na(latitude), !is.na(longitude))
  
  # Dissolve ahupuaʻa polygons; drop parts < 1 km² to remove errant geometry
  match_idx <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union  <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  land_sf_map <- st_as_sf(st_sf(geometry = land_union))
  
  coast_outer <- st_buffer(land_union, dist = band_km * 1000)
  coast_band  <- st_difference(coast_outer, land_union)
  coast_bbox  <- st_bbox(coast_outer)
  
  # Deduplicate sample locations — duplicate coordinates cause a singular
  # kriging covariance matrix and NA predictions across the whole island
  pts_proj <- island_data |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(raw_val = .data[[var]],
           log_val = if (use_log) log1p(raw_val) else raw_val,
           X = st_coordinates(geometry)[, 1],
           Y = st_coordinates(geometry)[, 2]) |>
    group_by(X, Y) |>
    summarise(log_val = mean(log_val, na.rm = TRUE),
              raw_val = mean(raw_val, na.rm = TRUE), .groups = "drop") |>
    st_as_sf(coords = c("X", "Y"), crs = 32604)
  
  # Display points clipped to coast_bbox; carries variable values for shared color scale
  pts_proj_all <- island_data |>
    filter(!is.na(latitude), !is.na(longitude)) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(val = .data[[var]],
           X   = st_coordinates(geometry)[,1],
           Y   = st_coordinates(geometry)[,2]) |>
    filter(X >= coast_bbox["xmin"], X <= coast_bbox["xmax"],
           Y >= coast_bbox["ymin"], Y <= coast_bbox["ymax"])
  
  # Variogram with enforced minimum range of island_diag / 5
  v_emp       <- variogram(log_val ~ 1, pts_proj)
  init_psill  <- var(pts_proj$log_val, na.rm = TRUE)
  island_diag <- sqrt(diff(st_bbox(pts_proj)[c("xmin","xmax")])^2 +
                      diff(st_bbox(pts_proj)[c("ymin","ymax")])^2)
  v_auto <- tryCatch(
    fit.variogram(v_emp, vgm(psill = init_psill * 0.8, model = "Sph",
                             range = island_diag / 5, nugget = init_psill * 0.2)),
    error = function(e) NULL)
  min_range <- island_diag / 5
  v_fit <- if (!is.null(v_auto) && v_auto$range[2] >= min_range) v_auto else {
    psill  <- if (!is.null(v_auto)) v_auto$psill[2] else init_psill * 0.8
    nugget <- if (!is.null(v_auto)) v_auto$psill[1] else init_psill * 0.2
    vgm(psill = psill, model = "Sph", range = min_range, nugget = nugget)
  }
  
  # Prediction grid: coastal band intersected with a buffer around sample points.
  # Grid cells farther than min_range from any sample are excluded — no data
  # nearby means no interpolation shown (prevents extrapolation into unsurveyed coast).
  sample_coverage <- st_buffer(st_union(pts_proj), dist = min_range)
  grid_pts <- st_make_grid(coast_band, cellsize = res_m, what = "centers") |>
    st_as_sf() |>
    st_filter(coast_band) |>
    st_filter(sample_coverage)
  
  k_out <- suppressWarnings(
    krige(log_val ~ 1, pts_proj, grid_pts, model = v_fit, nmax = 40)
  )
  pred_df <- st_coordinates(k_out) |>
    as.data.frame() |>
    mutate(pred = if (use_log) expm1(k_out$var1.pred) else k_out$var1.pred)
  
  krige_label <- if (use_log) paste0("log1p(", var, ")") else var
  
  ggplot() +
    geom_spatraster_rgb(data = get_tiles(st_transform(coast_outer, 4326),
                                         provider = "Esri.WorldImagery", zoom = tile_zoom)) +
    geom_raster(data = pred_df, aes(x = X, y = Y, fill = pred), alpha = 0.8) +
    geom_sf(data = land_sf_map, fill = NA, color = "white", linewidth = 0.4) +
    geom_sf(data = pts_proj_all, aes(fill = val), size = 1.0, shape = 21,
            color = "white", stroke = 0.3, alpha = 0.5) +
    scale_fill_viridis_c(
      name      = meta$label,
      option    = "plasma",
      direction = meta$direction,
      trans     = meta$trans,
      limits    = meta$limits,
      breaks    = meta$breaks,
      labels    = meta$labels,
      oob       = scales::squish,
      na.value  = "transparent",
      guide     = guide_colorbar(barheight = unit(10, "lines"),
                                 barwidth  = unit(0.6, "lines"),
                                 ticks.linewidth = 0.5)
    ) +
    coord_sf(crs = 32604,
             xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = island_name,
         subtitle = paste0("n = ", nrow(pts_proj),
                           " unique locations | ordinary kriging on ", krige_label)) +
    theme_minimal(base_size = 12) +
    theme(axis.title = element_blank(), axis.text = element_text(size = 7),
          plot.title = element_text(face = "bold", size = 14),
          plot.subtitle = element_text(size = 9, color = "gray40"))
}

raw_points_map <- function(data, island_name, ahupuaa_sf,
                            var = "NO3NO2", tile_zoom = 12) {
  
  meta <- var_meta[[var]]
  if (is.null(meta)) meta <- list(
    label = var, limits = NULL, breaks = waiver(), labels = waiver(), trans = "identity"
  )
  
  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]), !is.na(latitude), !is.na(longitude)) |>
    mutate(val = .data[[var]]) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604)
  
  # Same land polygon logic as krige_island_map
  match_idx <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |>
    st_make_valid() |>
    st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union  <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  land_sf_map <- st_as_sf(st_sf(geometry = land_union))
  
  coast_outer <- st_buffer(land_union, dist = 5000)
  coast_bbox  <- st_bbox(coast_outer)
  
  sat_tiles <- get_tiles(st_transform(coast_outer, 4326),
                         provider = "Esri.WorldImagery", zoom = tile_zoom)
  
  ggplot() +
    geom_spatraster_rgb(data = sat_tiles) +
    geom_sf(data = land_sf_map, fill = NA, color = "white", linewidth = 0.4) +
    geom_sf(data = island_data, aes(fill = val), size = 3.5, shape = 21,
            color = "white", stroke = 0.5, alpha = 0.9) +
    scale_fill_viridis_c(
      name      = meta$label,
      option    = "plasma",
      direction = meta$direction,
      trans     = meta$trans,
      limits    = meta$limits,
      breaks    = meta$breaks,
      labels    = meta$labels,
      oob       = scales::squish,
      guide     = guide_colorbar(barheight = unit(8, "lines"),
                                 barwidth  = unit(0.6, "lines"),
                                 ticks.linewidth = 0.5)
    ) +
    coord_sf(crs = 32604,
             xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = island_name,
         subtitle = paste0("n = ", nrow(island_data),
                           " samples | raw observations (insufficient coverage for kriging)")) +
    theme_minimal(base_size = 12) +
    theme(axis.title = element_blank(), axis.text = element_text(size = 7),
          plot.title = element_text(face = "bold", size = 14),
          plot.subtitle = element_text(size = 9, color = "gray40"))
}

# ── krige_surface_data: returns pred_df + land polygon + coast_bbox ───────────
krige_surface_data <- function(data, island_name, ahupuaa_sf,
                               var = "NO3NO2", res_m = 300, band_km = 3) {
  meta    <- var_meta[[var]]
  use_log <- !is.null(meta) && meta$trans == "log1p"
  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]),
           !is.na(latitude), !is.na(longitude))
  match_idx  <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  coast_outer <- st_buffer(land_union, dist = band_km * 1000)
  coast_band  <- st_difference(coast_outer, land_union)
  coast_bbox  <- st_bbox(coast_outer)
  pts_proj <- island_data |>
    st_as_sf(coords = c("longitude","latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(raw_val = .data[[var]],
           log_val = if (use_log) log1p(raw_val) else raw_val,
           X = st_coordinates(geometry)[,1],
           Y = st_coordinates(geometry)[,2]) |>
    group_by(X, Y) |>
    summarise(log_val = mean(log_val, na.rm = TRUE),
              raw_val = mean(raw_val, na.rm = TRUE), .groups = "drop") |>
    st_as_sf(coords = c("X","Y"), crs = 32604)
  island_diag <- sqrt(diff(st_bbox(pts_proj)[c("xmin","xmax")])^2 +
                      diff(st_bbox(pts_proj)[c("ymin","ymax")])^2)
  min_range <- island_diag / 5
  v_emp  <- variogram(log_val ~ 1, pts_proj)
  v_auto <- tryCatch(
    fit.variogram(v_emp, vgm(var(pts_proj$log_val)*0.8, "Sph",
                              min_range, var(pts_proj$log_val)*0.2)),
    error = function(e) NULL)
  v_fit <- if (!is.null(v_auto) && v_auto$range[2] >= min_range) v_auto else
    vgm(var(pts_proj$log_val)*0.8, "Sph", min_range, var(pts_proj$log_val)*0.2)
  sample_cov <- st_buffer(st_union(pts_proj), dist = min_range)
  grid_pts <- st_make_grid(coast_band, cellsize = res_m, what = "centers") |>
    st_as_sf() |> st_filter(coast_band) |> st_filter(sample_cov)
  k_out <- suppressWarnings(krige(log_val ~ 1, pts_proj, grid_pts, model = v_fit, nmax = 40))
  pred_df <- st_coordinates(k_out) |> as.data.frame() |>
    mutate(pred = if (use_log) expm1(k_out$var1.pred) else k_out$var1.pred)
  list(pred_df    = pred_df,
       land_sf    = st_as_sf(st_sf(geometry = land_union)),
       coast_bbox = coast_bbox)
}

# ── maui_validation_map: kriging surface + optional Hui station overlay ───────
# Pass bg_hui_loc to show Hui triangles (only meaningful for var = "NO3NO2").
maui_validation_map <- function(data, ahupuaa_sf, var = "NO3NO2",
                                 bg_hui_loc = NULL, tile_zoom = 11) {
  meta <- var_meta[[var]]
  if (is.null(meta)) meta <- list(
    label = var, limits = NULL, breaks = waiver(), labels = waiver(),
    trans = "identity", direction = 1
  )

  kd     <- krige_surface_data(data, "Maui", ahupuaa_sf, var = var)
  cs_pts <- data |>
    filter(island == "Maui", !is.na(.data[[var]]), !is.na(latitude), !is.na(longitude)) |>
    mutate(val = .data[[var]]) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604)
  sat <- get_tiles(
    st_as_sfc(kd$coast_bbox) |> st_set_crs(32604) |> st_transform(4326),
    provider = "Esri.WorldImagery", zoom = tile_zoom
  )

  # Build Hui station sf internally when bg_hui_loc is supplied
  # Map chem var names to the corresponding Hui column (all in µmol/L)
  hui_col_map <- c(NO3NO2 = "NNN_umol", PO4 = "Phosphate_umol",
                   SiO2   = "Silicate_umol", NH3 = "NH4_umol")
  hui_col  <- hui_col_map[var]
  show_hui <- !is.null(bg_hui_loc) && !is.na(hui_col) && hui_col %in% names(bg_hui_loc)

  if (show_hui) {
    hui_post <- bg_hui_loc |>
      filter(year(Date) == 2026, month(Date) == 3,
             Date >= as.Date("2026-03-14"), !is.na(.data[[hui_col]])) |>
      group_by(Station, SiteName, region, Lat, Long) |>
      summarise(val_umol = mean(.data[[hui_col]], na.rm = TRUE), .groups = "drop")
    hui_hist_med <- bg_hui_loc |>
      filter(!is.na(.data[[hui_col]])) |>
      group_by(Station) |>
      summarise(med_hist = median(.data[[hui_col]], na.rm = TRUE), .groups = "drop")
    log_lo   <- log1p(1); log_hi <- log1p(20)
    grey_pal <- colorRampPalette(c("gray20", "white"))(1000)
    hui_sf <- bg_hui_loc |>
      distinct(Station, SiteName, region, Lat, Long) |>
      filter(!is.na(Lat), !is.na(Long)) |>
      left_join(hui_post |> select(Station, val_umol), by = "Station") |>
      left_join(hui_hist_med, by = "Station") |>
      mutate(
        fold_change = val_umol / med_hist,
        fc_safe     = pmin(pmax(replace_na(fold_change, 1), 1), 20),
        tri_fill    = grey_pal[round((log1p(fc_safe) - log_lo) /
                                       (log_hi - log_lo) * 999) + 1],
        tri_fill    = case_when(
          is.na(fold_change) & !is.na(med_hist) ~ "gray50",
          is.na(med_hist)                        ~ "gray70",
          TRUE                                   ~ tri_fill
        ),
        fold_change = replace_na(fold_change, 1)
      ) |>
      st_as_sf(coords = c("Long","Lat"), crs = 4326) |>
      st_transform(32604)
  }

  p <- ggplot() +
    geom_spatraster_rgb(data = sat) +
    geom_raster(data = kd$pred_df, aes(x = X, y = Y, fill = pred), alpha = 0.8) +
    geom_sf(data = kd$land_sf, fill = NA, color = "white", linewidth = 0.35) +
    geom_sf(data = cs_pts, aes(fill = val),
            shape = 21, size = 1.0, color = "white", stroke = 0.3, alpha = 0.5)

  # fill scale used only for kriging raster + CS circles (plasma)
  p <- p +
    scale_fill_viridis_c(
      name      = meta$label,
      option    = "plasma",
      direction = meta$direction,
      trans     = meta$trans,
      limits    = meta$limits,
      breaks    = meta$breaks,
      labels    = meta$labels,
      oob       = scales::squish,
      na.value  = "transparent",
      guide     = guide_colorbar(barheight = unit(10, "lines"),
                                 barwidth  = unit(0.6, "lines"),
                                 ticks.linewidth = 0.5)
    )

  if (show_hui) {
    hui_breaks <- c(1, 3, 5, 10, 20)
    hui_labels <- c("1", "3", "5", "10", "≥20")
    hui_name   <- "Fold-change\n(Mar 2026 /\nall-time median)"
    p <- p +
      # Shape 24 (fill + white outline); I() bypasses the plasma fill scale
      geom_sf(data = hui_sf,
              aes(size = fold_change, fill = I(tri_fill)),
              color = "white", shape = 24, stroke = 0.8, alpha = 0.85) +
      ggrepel::geom_label_repel(
        data = hui_sf |> filter(fold_change > 15),
        aes(label    = paste0(SiteName, "\n", round(fold_change, 0), "×"),
            geometry = geometry),
        stat = "sf_coordinates", size = 2.5, fontface = "bold", color = "#B71C1C",
        fill = alpha("white", 0.85), label.size = 0.2,
        nudge_y = 4500, min.segment.length = 0
      ) +
      scale_size_continuous(
        name = hui_name, range = c(1.0, 8), limits = c(1, 20),
        breaks = hui_breaks, labels = hui_labels,
        guide = guide_legend(
          theme = theme(legend.key = element_rect(fill = "gray30", color = NA)),
          override.aes = list(shape = 24, color = "white", stroke = 0.8, alpha = 0.85,
                              fill = c("#333333", "#6F6F6F", "#929292", "#C6C6C6", "#FFFFFF"))
        )
      )
  }

  hui_note <- if (show_hui) {
    paste0("    △ Hui o Ka Wai Ola stations (size = Mar 2026 fold-change above all-time median)\n",
           "△ small = no March 2026 sample    △ gray = insufficient data    ",
           "Labels: stations with fold-change > 15×")
  } else ""

  p +
    coord_sf(crs = 32604,
             xlim = c(kd$coast_bbox["xmin"], kd$coast_bbox["xmax"]),
             ylim = c(kd$coast_bbox["ymin"], kd$coast_bbox["ymax"]),
             expand = FALSE) +
    labs(
  #    subtitle = paste0("● community science samples (this study)", hui_note),
      caption  = if (show_hui) "Hui data: Falinski et al. / Hui o Ka Wai Ola." else NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.title    = element_blank(),
      axis.text     = element_text(size = 7),
      plot.subtitle = element_text(size = 8.5, color = "gray40"),
      plot.caption  = element_text(size = 8, color = "gray50", hjust = 0),
      legend.position = "right"
    )
}
```

## O'ahu

Oʻahu's highest nutrient concentrations cluster along the northwestern (Waiʻanae) coast and around Pearl Harbor, consistent with those areas having both dense cesspool infrastructure and streams that drain heavily developed land. The salinity map shows freshwater dilution in the same locations, confirming this is land-sourced water.

```{r}
#| label: map-oahu
#| fig-height: 14
#| fig-cap: "Spatially interpolated nutrient and salinity maps for Oʻahu, March 2026. Color shows concentration — yellow-green = highest, dark purple = lowest (all log scale). Circles are individual sample points colored on the same scale. Low-salinity (freshwater-diluted) hotspots on the northwest coast line up with the highest nitrogen and silica readings."
oahuNN<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var = "NO3NO2" )

oahuPO<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="PO4" )

oahuSi<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="SiO2" )

oahuNH3<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="NH3" )

oahuSal<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="Salinity" )

(oahuNN|oahuPO|oahuNH3)/(oahuSi|oahuSal|plot_spacer())
ggsave(here("output","Oahu_spatial.png"), height = 12, width = 18)

```

## Maui

Maui shows the most dramatic hotspot signal of any island, concentrated on the north and west coasts. The Mālā Ramp area and Honoköwai in West Maui stand out clearly — these sites also showed the largest departures from their long-term monitoring baselines (see Hui comparison below). North Maui's windward streams drain some of the island's most intensively farmed land, explaining the particularly high silica (SiO₂) signature there.

```{r}
#| label: map-maui
#| fig-height: 14
#| fig-cap: "Spatially interpolated nutrient and salinity maps for Maui, March 2026. West Maui's leeward coast and the windward north coast show the highest concentrations. The match between low-salinity patches and high-nutrient zones confirms that dilution with freshwater runoff is driving the signal, not local biological processes."

MauiNN<-krige_island_map(chem, "Maui", ahupuaa_raw,var = "NO3NO2" )

MauiPO<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="PO4" )

MauiSi<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="SiO2" )

MauiNH3<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="NH3" )

MauiSal<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="Salinity" )

(MauiNN|MauiPO|MauiNH3)/(MauiSi|MauiSal|plot_spacer())
ggsave(here("output","Maui_spatial.png"), height = 12, width = 18)
```

## Moloka'i

Molokaʻi shows elevated nitrogen and silica across most of its southern coast, with the signal strongest near Kaunakakai Harbor. The pattern is more spatially uniform than Maui, consistent with Molokaʻi's lower topographic relief producing more diffuse runoff rather than concentrated stream plumes.

```{r}
#| label: map-molokai
#| fig-height: 14
#| fig-cap: "Spatially interpolated nutrient and salinity maps for Molokaʻi, March 2026. Elevated nutrients are distributed broadly along the south coast rather than in discrete plumes, reflecting the island's lower elevation and more evenly distributed runoff pathways."

molokaiNN<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var = "NO3NO2", tile_zoom = 12 )

molokaiPO<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="PO4" , tile_zoom = 12)

molokaiSi<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="SiO2", tile_zoom = 12 )

molokaiNH3<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="NH3", tile_zoom = 12 )

molokaiSal<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="Salinity", tile_zoom = 12 )

(molokaiNN|molokaiPO|molokaiNH3)/(molokaiSi|molokaiSal|plot_spacer())
ggsave(here("output","molokai_spatial.png"), height = 12, width = 18)


```

## Lāna'i

Only raw sample locations are shown for Lānaʻi — with just 20 samples, there aren't enough data points to produce a reliable interpolated surface. The samples that do exist show moderate nutrient concentrations with no extreme hotspots, though sampling coverage is too sparse to draw firm conclusions.

```{r}
#| label: map-lanai
#| fig-height: 6
lanaiNN<-raw_points_map(chem, "Lāna'i", ahupuaa_raw, var = "NO3NO2", tile_zoom = 12)

lanaiPO<-raw_points_map(chem, "Lāna'i", ahupuaa_raw, var = "PO4", tile_zoom = 12)
lanaiSi<-raw_points_map(chem, "Lāna'i", ahupuaa_raw, var = "SiO2", tile_zoom = 12)
lanaiNH3<-raw_points_map(chem, "Lāna'i", ahupuaa_raw, var = "NH3", tile_zoom = 12)
lanaiSal<-raw_points_map(chem, "Lāna'i", ahupuaa_raw, var = "Salinity", tile_zoom = 12)

(lanaiNN|lanaiPO|lanaiNH3)/(lanaiSi|lanaiSal|plot_spacer())
ggsave(here("output","lanai_spatial.png"), height = 12, width = 18)
```


```{r}
#| label: map-all
#| fig-height: 6

#oahu / maui /molokai + plot_layout(guides = "collect")
#ggsave(here("output","maps_all.png"), height = 10, width = 10)
 
```


# Independent validation: Hui o Ka Wai Ola stations on the Maui kriging surface

The kriging surface is derived entirely from community science samples. The Hui o Ka Wai Ola programme provides an independent check: do the long-term fixed stations agree with where the community science kriging shows hotspots? Triangles show Hui station post-storm with size relative to fold-increase from baseline values.

```{r}
#| label: maui-validation-map
#| fig-height: 14
#| fig-cap: " Background: the interpolated nitrogen (NO₃+NO₂) map built from community science samples. Triangles: Hui o Ka Wai Ola monitoring stations. The Hui stations with the most extreme readings — Mālā Ramp (43×), Camp Olowalu (32×), and others on West Maui — sit precisely where the community science map shows the darkest hotspot colors. The agreement between two completely independent datasets confirms the hotspot pattern is real."


bg_hui_loc <- bind_rows(
  readxl::read_excel("data/background_nutrients/hui-south-maui-thru-2026-1st-quarter.0.xlsx",
    sheet = "data") |>
    mutate(Turbidity = as.numeric(Turbidity), region = "South Maui"),
  readxl::read_excel(
    "data/background_nutrients/hui-west-maui-thru-2026-1st-quarter.0.xlsx",
    sheet = "data") |>
    mutate(Turbidity = as.numeric(Turbidity), region = "West Maui")
) |>
  mutate(Date           = as.Date(Date),
         NNN_umol       = as.numeric(NNN)       / 14.007,   # µg/L N  → µmol/L
         Phosphate_umol = as.numeric(Phosphate) / 30.974,   # µg/L P  → µmol/L
         Silicate_umol  = as.numeric(Silicate)  / 28.086,   # µg/L Si → µmol/L
         NH4_umol       = as.numeric(NH4)        / 14.007)  # µg/L N  → µmol/L

HuiNN<-maui_validation_map(chem, ahupuaa_raw, var = "NO3NO2", bg_hui_loc = bg_hui_loc)

HuiSi<-maui_validation_map(chem, ahupuaa_raw, var = "SiO2", bg_hui_loc = bg_hui_loc)

HuiNH3<-maui_validation_map(chem, ahupuaa_raw, var = "NH3", bg_hui_loc = bg_hui_loc)

HuiPO<-maui_validation_map(chem, ahupuaa_raw, var = "PO4", bg_hui_loc = bg_hui_loc)

no_sub  <- theme(plot.subtitle = element_blank(), plot.caption = element_blank())
no_size <- guides(size = "none")

size_legend <- get_legend(
  HuiNN + guides(fill = "none") + theme(legend.position = "right")
)

plot_grid(
  plot_grid(
    HuiNN  + no_size + no_sub,
    HuiPO  + no_size + no_sub,
    HuiNH3 + no_size + no_sub,
    HuiSi  + no_size + no_sub,
    ncol = 2
  ),
  size_legend,
  ncol = 2,
  rel_widths = c(1, 0.13)
) |>
  ggdraw() +
  theme(plot.background = element_rect(fill = "white", color = NA))

ggsave(here("output","hui_spatial.png"), height = 12, width = 12)
```

The Hui stations on West Maui's leeward coast show amber-to-orange colors, meaning their post-storm readings were 8–43× higher than their own ten-year averages. The calmer eastern coast shows lower readings in both datasets. The fact that community science volunteers and trained monitoring staff — sampling at different times with different equipment — independently found the same hotspot pattern in the same places gives us high confidence the signal is real, not a measurement artifact.

---

# Storm cascade: disruption of coastal biogeochemistry

The storm set off a chain of events visible in satellite data, stream gauges, and water chemistry. Rain (panel A) filled streams (panel B), which flushed nutrients into the coast, and leaving elevated chemical concentrations still measurable when volunteers sampled two weeks later (panel C). Peak rainfall hit March 13–14, reaching 42–100 mm/day averaged across each island. Streams responded immediately, with peak flows of 231–1,040 cubic feet per second depending on watershed size. When community scientists sampled in late March and early April, nitrogen (NO₃+NO₂) and silica (SiO₂) were the most clearly elevated nutrients, consistent with runoff from volcanic soils.

```{r}
#| label: storm-cascade-figure
#| fig-height: 8
#| fig-width: 10
#| fig-cap: "Rain → streams → algae → chemistry: the full storm cascade in four panels. (A) Average daily rainfall per island from atmospheric model data (NCEP). (B) Stream discharge at three USGS gauges — Makaha Stream (Oʻahu), Halawa Stream (Molokaʻi), Hanawi Stream (Maui) — on a log scale because flows span three orders of magnitude. (C) Post-storm nutrient concentrations by island from community science sampling (Mar 28–Apr 8). Blue shading = the Kona Low storm window (Mar 10–25); green shading = sampling window."



# ── shared constants ────────────────────────────────────────────────
storm_xmin <- as.Date("2026-03-10"); storm_xmax <- as.Date("2026-03-25")
samp_xmin  <- as.Date("2026-03-28"); samp_xmax  <- as.Date("2026-04-08")
plot_xlim  <- as.Date(c("2026-01-15", "2026-06-05"))

windows <- list(
  annotate("rect", xmin = storm_xmin, xmax = storm_xmax,
           ymin = -Inf, ymax = Inf, fill = "#BBDEFB", alpha = 0.75),
  annotate("rect", xmin = samp_xmin,  xmax = samp_xmax,
           ymin = -Inf, ymax = Inf, fill = "#C8E6C9", alpha = 0.75)
)

ts_base <- theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title.x       = element_blank(),
    axis.text.x        = element_blank(),
    axis.ticks.x       = element_blank(),
    legend.position    = "none",
    plot.margin        = margin(3, 6, 1, 6),
    axis.title.y       = element_text(size = 8.5)
  )

# Panel A: Rainfall
rain_df <- read_csv("hawaii_ncep_janjun_cache.csv", show_col_types = FALSE) |>
  filter(island %in% c("Oahu", "Maui", "Molokai"),
         date >= plot_xlim[1], date <= plot_xlim[2]) |>
  mutate(island = recode_island[island]) |>
  group_by(island, date) |>
  summarise(precip_mm = mean(precip_mmday, na.rm = TRUE), .groups = "drop")

pA <- ggplot(rain_df, aes(x = date, y = precip_mm, fill = island, color = island)) +
  windows +
  geom_area(position = "identity", alpha = 0.30, linewidth = 0.35) +
  annotate("text", x = as.Date("2026-02-17"), y = 172,
           label = "Feb event", size = 2.6, color = "gray35", hjust = 0.5, vjust = 0) +
  annotate("text", x = as.Date("2026-03-17"), y = 108,
           label = "Kona Low", size = 2.6, color = "#1A237E", hjust = 0.5, vjust = 0) +
  scale_fill_manual(values  = island_pal) +
  scale_color_manual(values = island_pal) +
  scale_x_date(limits = plot_xlim, expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(y = "Rainfall\n(mm/day)", tag = "A") +
  ts_base

# Panel B: Stream discharge (USGS NWIS — cached in data/usgs_discharge_cache.rds)
gauge_meta <- tibble(
  site_no = c("16211600", "16400000", "16508000"),
  island  = c("O'ahu",    "Moloka'i", "Maui")
)

dis_df <- read_csv("data/usgs_discharge_cache.csv", show_col_types = FALSE,
                   col_types = cols(site_no = col_character())) |>
  dataRetrieval::renameNWISColumns() |>
  filter(site_no %in% gauge_meta$site_no,
         Date >= plot_xlim[1], Date <= plot_xlim[2], Flow > 0.5) |>
  left_join(gauge_meta, by = "site_no")

pB <- ggplot(dis_df, aes(x = Date, y = Flow, color = island, group = site_no)) +
  windows +
  geom_line(linewidth = 0.65, alpha = 0.9, na.rm = TRUE) +
  scale_color_manual(values = island_pal) +
  scale_x_date(limits = plot_xlim, date_labels = "%b %Y", expand = c(0, 0)) +
  scale_y_log10(breaks = c(1, 10, 100, 1000),
                labels = scales::label_number(big.mark = ",")) +
  coord_cartesian(ylim = c(0.8, 3000)) +
  labs(y = "Discharge\n(cfs, log)", tag = "B") +
  ts_base +
  theme(
    axis.text.x  = element_text(size = 8),
    axis.ticks.x = element_line(linewidth = 0.3)
  )

# Panel C: Coastal chlorophyll (VIIRS NOAA-20 — cached in hawaii_chlorophyll_cache.rds)
chla_df <- readRDS("hawaii_chlorophyll_cache.rds") |>    # satellite array — kept as RDS
  filter(island %in% c("Oahu", "Maui", "Molokai"),
         date >= plot_xlim[1], date <= plot_xlim[2]) |>
  mutate(island = recode_island[island]) |>
  group_by(island, date) |>
  summarise(mean_chla = mean(chlor_a, na.rm = TRUE),
            n_valid   = sum(!is.na(chlor_a)), .groups = "drop") |>
  filter(n_valid >= 8, mean_chla > 0)

pC <- ggplot(chla_df, aes(x = date, y = mean_chla, color = island)) +
  windows +
  geom_point(size = 0.8, alpha = 0.55, na.rm = TRUE) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE,
              linewidth = 0.9, na.rm = TRUE) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_x_date(limits = plot_xlim, date_labels = "%b %Y", expand = c(0, 0)) +
  scale_y_log10(breaks = c(0.1, 0.3, 1, 3, 10),
                labels = scales::label_number(accuracy = 0.1)) +
  labs(y = expression("Chlorophyll-a\n(mg/m"^3*", log)"), tag = "C") +
  ts_base +
  theme(
    axis.text.x      = element_text(size = 8),
    axis.ticks.x     = element_line(linewidth = 0.3),
    legend.position  = "bottom",
    legend.text      = element_text(size = 9),
    legend.key.width = unit(1.2, "lines"),
    plot.margin      = margin(1, 6, 2, 6)
  )

# Panel D: Post-storm nutrient distributions
nut_long <- chem |>
  filter(island %in% c("O'ahu", "Maui", "Moloka'i")) |>
  mutate(
    island = factor(island, levels = c("Maui", "Moloka'i", "O'ahu"))
  ) |>
  select(island, NO3NO2, NH3, PO4, SiO2) |>
  pivot_longer(c(NO3NO2, NH3, PO4, SiO2), names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>
  mutate(nutrient = factor(
    recode(nutrient, "NO3NO2" = "NO[3]+NO[2]", "NH3" = "NH[4]",
           "PO4" = "PO[4]", "SiO2" = "SiO[2]"),
    levels = c("NO[3]+NO[2]", "NH[4]", "PO[4]", "SiO[2]")
  ))

pD <- ggplot(nut_long, aes(x = island, y = conc, fill = island, color = island)) +
  geom_violin(alpha = 0.28, linewidth = 0.3, trim = TRUE) +
  geom_boxplot(width = 0.17, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = island_pal) +
  scale_color_manual(values = island_pal) +
  scale_y_log10(breaks = c(0.01, 0.1, 1, 10, 100),
                labels = scales::label_number(drop0trailing = TRUE)) +
  facet_wrap(~nutrient, ncol = 1, scales = "free_y",
             labeller = label_parsed) +
  labs(tag = "C", y = expression(paste(mu, "mol L"^{-1})), x = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x        = element_text(angle = 30, hjust = 1, vjust = 1, size = rel(0.85)),
    strip.text         = element_text(size = 9, face = "bold"),
    legend.position    = "none",
    plot.margin        = margin(4, 8, 4, 2),
    axis.title.y       = element_text(size = 8.5)
  )

# Assemble
((pA / pB) | pD) +
  plot_layout(widths = c(3, 1.5)) +
  plot_annotation(
    title   = "Kona Low storm cascade: terrestrial runoff disrupts coastal biogeochemistry",
    caption = paste0(
      "Blue shading = Kona Low storm window (Mar 10–25).  ",
      "Green shading = community science sampling (Mar 28–Apr 8).\n",
      "",
      "Stream gauges: Makaha Str (Oʻahu), Halawa Str (Molokaʻi), Hanawi Str (Maui)."
    ),
    theme = theme(
      plot.title   = element_text(size = 11, face = "bold", margin = margin(b = 4)),
      plot.caption = element_text(size = 7.5, color = "gray40",
                                  hjust = 0, margin = margin(t = 6)),
      plot.margin  = margin(8, 8, 6, 8)
    )
  )

ggsave(here("output","nutrient_gradient.png"))
```

# Spatial structure of the biogeochemical signal

If the storm's chemical impact was driven by watershed runoff, then nearby coastal sites should have more similar nutrient levels than sites far apart — high readings near stream mouths or cesspools, lower readings away from them. This would produce a non-random, geographically coherent pattern. We test this with Moran's I, a standard measure of spatial clustering where a positive, statistically significant value means nearby sites are more alike than you'd expect by chance. The test was run separately for each nutrient and island using the eight nearest sampling neighbours. Lāna'i had too few samples and is excluded.

```{r}
#| label: morans-i
#| fig-height: 5
#| fig-cap: "Do high-nutrient sites cluster together spatially? Moran's I measures this: values above zero mean nearby sites are more similar than random chance would produce. Filled circles = significant (p < 0.05); open circles = not significant. Error bars = ±1.96 standard errors. SiO₂ and NO₃+NO₂ show clear geographic clustering on both Maui and Molokaʻi — their hotspots follow the shape of watersheds and stream mouths. NH₃ shows no significant clustering in Maui or Molokaʻi, meaning it is scattered randomly across the coast rather than concentrated near any particular landscape feature."

nutrients  <- c("SiO2", "NO3NO2", "PO4", "NH3")
islands    <- c("O'ahu", "Maui", "Moloka'i")

chem_sf_mi <- chem |>
  filter(!is.na(latitude), !is.na(longitude), !is.na(island)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

moran_results <- purrr::map_dfr(islands, function(isl) {
  purrr::map_dfr(nutrients, function(nut) {
    pts <- chem_sf_mi |> filter(island == isl, !is.na(.data[[nut]]))
    if (nrow(pts) < 15) return(NULL)
    nb <- knn2nb(knearneigh(st_coordinates(pts), k = 8), sym = TRUE)
    lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
    mt <- moran.test(pts[[nut]], lw, zero.policy = TRUE, alternative = "greater")
    tibble(
      island   = isl,
      nutrient = nut,
      n        = nrow(pts),
      moran_I  = unname(mt$estimate["Moran I statistic"]),
      expected = unname(mt$estimate["Expectation"]),
      variance = unname(mt$estimate["Variance"]),
      p_value  = mt$p.value
    )
  })
})

# Plot
nutrient_labels_mi <- c(
  "SiO2"   = "SiO[2]",
  "NO3NO2" = "NO[3]+NO[2]",
  "PO4"    = "PO[4]",
  "NH3"    = "NH[3]"
)

moran_results |>
  mutate(
    island   = factor(island,   levels = c("O'ahu", "Maui", "Moloka'i")),
    nutrient = factor(nutrient, levels = nutrients),
    sig_label = if_else(p_value < 0.05, "p < 0.05", "n.s."),
    sig_label = factor(sig_label, levels = c("p < 0.05", "n.s."))
  ) |>
  ggplot(aes(x = moran_I, y = nutrient, color = island, shape = sig_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  geom_errorbar(
    aes(xmin = moran_I - 1.96 * sqrt(variance),
        xmax = moran_I + 1.96 * sqrt(variance)),
    width = 0, linewidth = 0.4, alpha = 0.35,
    orientation = "y",
    position = position_dodge(width = 0.55)
  ) +
  geom_point(size = 3.2, position = position_dodge(width = 0.55)) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_shape_manual(
    values = c("p < 0.05" = 16, "n.s." = 1),
    name   = "Significance"
  ) +
  scale_x_continuous(limits = c(-0.05, 0.5), breaks = seq(0, 0.5, 0.1)) +
  scale_y_discrete(labels = \(x) parse(text = nutrient_labels_mi[x])) +
  labs(
    x        = "Moran's I  (k = 8 nearest neighbours)",
    y        = NULL,
    subtitle = "Positive values indicate spatial clustering; error bars = ±1.96 SE"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, color = "gray90"),
    legend.position    = "right",
    plot.subtitle      = element_text(size = 9, color = "gray40")
  )
```

```{r}
#| label: morans-i-table
moran_results |>
  mutate(
    sig = case_when(
      p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",   TRUE ~ "n.s."
    )
  ) |>
  select(Island = island, Nutrient = nutrient, n,
         `Moran's I` = moran_I, `p-value` = p_value, ` ` = sig) |>
  kable(digits = 3,
        caption = "Moran's I spatial autocorrelation statistics per island and nutrient. Spatial weights: k = 8 nearest neighbours (UTM zone 4N). One-sided test against the alternative of positive spatial autocorrelation.")
```

Silica (SiO₂) shows the tightest geographic clustering on Maui (Moran's I = 0.41) and Molokaʻi (I = 0.40). Silica is a reliable marker of volcanic rock weathering — it enters the ocean almost exclusively through stream runoff and SGD, so its spatial pattern closely traces where streams met the coast during the storm. Nitrogen (NO₃+NO₂) clusters almost as strongly. Ammonium (NH₃), by contrast, shows no significant spatial pattern on Maui or Molokaʻi — it appears scattered randomly across the coast. This contrast is telling: the storm stamped a watershed-shaped footprint on nitrogen and silica, but NH₃ — which builds up through biological decomposition in sediments — was  less affected by the event.

# Land–sea connectivity: salinity–nutrient mixing

Salinity is a natural tracer of freshwater: the lower it is, the more land-sourced water is mixed in. If runoff from the storm is carrying the nutrients, we'd expect to see a clear pattern — the more freshwater at a site (lower salinity), the more nutrients. The plots below test this for all four measured nutrients. Sites with Sal ≤ 34 PSU are coloured orange (storm-influenced), while marine-background sites (Sal > 34 PSU) are blue. Zeros are excluded as below-detection readings.

```{r}
#| label: salinity-nutrient-mixing
#| fig-height: 7
#| fig-cap: "As salinity drops (more freshwater mixing in), nutrients rise — but only for the nutrients carried by runoff. Orange points and trend line: storm-influenced sites (Sal ≤ 34 PSU). Blue points and horizontal line: marine-background sites (Sal > 34 PSU) shown at their median concentration. NO₃+NO₂ and SiO₂ show steep, consistent increases as sites become more freshwater-influenced — confirming that land-sourced water is delivering these nutrients. PO₄ shows a weaker version of the same pattern. NH₃ has a flat, scattered relationship, meaning it is not being delivered by runoff in the same way — it comes from other sources (biological breakdown in sediments) that are unrelated to how much freshwater is present."

sal_zone_pal <- c("Storm-influenced\n(≤ 34 PSU)" = "#E65100",
                  "Marine background\n(> 34 PSU)"   = "#1565C0")

mixing_long <- chem |>
  filter(!is.na(Salinity)) |>
  select(island, Salinity, NO3NO2, PO4, SiO2, NH3) |>
  pivot_longer(c(NO3NO2, PO4, SiO2, NH3),
               names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>               # drop zeros (below detection)
  mutate(
    nutrient = factor(nutrient,
      levels = c("NO3NO2", "PO4", "SiO2", "NH3"),
      labels = c("NO[3]+NO[2]~(mu*mol/L)",
                 "PO[4]~(mu*mol/L)",
                 "SiO[2]~(mu*mol/L)",
                 "NH[3]~(mu*mol/L)")),
    sal_zone = factor(
      if_else(Salinity > 34,
              "Marine background\n(> 34 PSU)",
              "Storm-influenced\n(≤ 34 PSU)"),
      levels = c("Storm-influenced\n(≤ 34 PSU)",
                 "Marine background\n(> 34 PSU)"))
  )

# Median nutrient concentration for marine background sites (horizontal reference)
marine_medians <- mixing_long |>
  filter(sal_zone == "Marine background\n(> 34 PSU)") |>
  group_by(nutrient) |>
  summarise(med = median(conc, na.rm = TRUE), .groups = "drop")

ggplot(mixing_long, aes(x = Salinity, y = conc, color = sal_zone)) +
  geom_vline(xintercept = 34, linetype = "dotted",
             color = "gray40", linewidth = 0.6) +
  geom_point(size = 1.2, alpha = 0.45) +
  geom_smooth(data = \(d) filter(d, sal_zone == "Storm-influenced\n(≤ 34 PSU)"),
              method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15,
              color = "#E65100") +
  geom_hline(data = marine_medians, aes(yintercept = med),
             color = "#1565C0", linetype = "solid", linewidth = 0.8,
             inherit.aes = FALSE) +
  scale_color_manual(values = sal_zone_pal, name = NULL) +
  scale_y_log10(labels = label_number(drop0trailing = TRUE)) +
  scale_x_continuous(limits = c(NA, 36.5)) +
  facet_wrap(~nutrient, scales = "free_y", ncol = 2, labeller = label_parsed) +
  labs(x = "Salinity (PSU)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

**NO₃+NO₂ and SiO₂** behave as expected if they are coming from land runoff: concentrations fall steeply as salinity rises, and level off once water becomes fully marine. This is the textbook pattern for a land-sourced contaminant diluting into the ocean. **PO₄** follows the same pattern but also remains elevated in fully marine water — suggesting a chronic background phosphorus source (likely cesspools leaching through groundwater) beyond the storm signal alone. **NH₃** shows little relationship with salinity in either zone.

# Methods

## Sampling

Water samples were collected by community volunteers and scientists across O'ahu, Maui, Moloka'i, and Lāna'i between `r format(min(chem$collected_hst, na.rm = TRUE), "%B %d")` and `r format(max(chem$collected_hst, na.rm = TRUE), "%B %d, %Y")`, following the March 2026 Kona Low storm. NO₃+NO₂ was analyzed colorimetrically.

## Spatial interpolation

Ordinary kriging was used to interpolate NO₃+NO₂ concentrations between sample locations for islands with sufficient coverage (O'ahu, Maui, Moloka'i). Kriging was performed in R using the `gstat` package. Prior to kriging, concentrations were log₁p-transformed (log(x + 1)) to stabilize variance and approximate normality — a standard approach for right-skewed geochemical data.

Where multiple samples were collected at the same GPS location, values were averaged prior to kriging — duplicate coordinates produce a singular covariance matrix that causes predictions to fail. An empirical variogram was computed for each island and fit with a spherical model. Because the auto-fit variogram range (~5–6 km) is driven by local sample clustering and leaves sections of the coastline with no near-neighbor data, a minimum range of 1/5 of the island's spatial diagonal is enforced to ensure gap-free predictions across the full coast. The prediction grid was constructed as a 3 km coastal band derived by buffering the island polygon outward and subtracting the land area, then further clipped to the union of sample-point buffers (radius = variogram range). This prevents kriging from extrapolating into unsurveyed coastal stretches. The 40 nearest neighbors were used for each prediction. Back-transformed predictions (expm1) are displayed on a log₁p color scale. Sample points are colored on the same scale as the kriging surface, so extreme observed values (e.g. 265 µmol/L on O'ahu) are visible even where the smooth interpolation surface cannot reach those peaks.

Island land boundaries are derived from the Hawaiʻi ahupuaʻa dataset, which provides more accurate coastlines than Natural Earth for all four islands (Natural Earth clips East Maui at the sample extent boundary). Small errant polygon fragments (< 1 km²) in the ahupuaʻa data are removed before computing the coastal band.

---

# Salinity-based enrichment above marine background

Within the same sampling event, sites with salinity above 34 PSU were minimally affected by freshwater runoff — they represent what the ocean chemistry looked like without the storm signal. Comparing these `r sum(chem$Salinity > 34, na.rm = TRUE)` "background" sites to the `r sum(chem$Salinity < 32, na.rm = TRUE)` most storm-influenced sites (Sal < 32 PSU) gives a direct, within-study measure of how much the storm changed things.

```{r}
#| label: salinity-baseline-enrichment
#| fig-height: 6.5
#| fig-cap: "How much higher were nutrient levels in storm-influenced water compared to clean marine water sampled at the same time? (A) The fold-enrichment for each nutrient — how many times higher the typical storm-influenced concentration was relative to the marine background. Point size reflects the magnitude; error bars are 95% confidence intervals from 2,000 bootstrap resamples. SiO₂ was 6× higher, consistent with volcanic rock dissolving in freshwater and flowing into the coast. NO₃+NO₂ was 3× higher and PO₄ was about 2.3× higher, reflecting runoff from farms, cesspools, and soils. NH₃ was 2× higher but more variable, consistent with it coming partly from runoff and partly from biological processes. (B) The full distribution of concentrations in each zone. Dashed lines show the background zone means."

# ── Define salinity zones ────────────────────────────────────────────────────
sal_zones <- chem |>
  filter(!is.na(Salinity)) |>
  mutate(zone = case_when(
    Salinity > 34.0 ~ "Background\n(Sal > 34 PSU)",
    Salinity < 32.0 ~ "Storm-influenced\n(Sal < 32 PSU)",
    TRUE            ~ NA_character_
  )) |>
  filter(!is.na(zone))

bg   <- sal_zones |> filter(str_detect(zone, "Background"))
strm <- sal_zones |> filter(str_detect(zone, "Storm"))

nutrients_vec    <- c("NO3NO2", "PO4", "SiO2", "NH3")
nut_labels_ascii <- c("NO3+NO2", "PO4", "SiO2", "NH3")
zone_pal <- c(
  "Background\n(Sal > 34 PSU)"       = "#1565C0",
  "Storm-influenced\n(Sal < 32 PSU)" = "#E65100"
)

# ── Bootstrap fold-changes ───────────────────────────────────────────────────
set.seed(4492)
ci_df <- map_dfr(seq_along(nutrients_vec), function(i) {
  nut  <- nutrients_vec[i]
  bg_v <- bg[[nut]];   bg_v <- bg_v[!is.na(bg_v)]
  st_v <- strm[[nut]]; st_v <- st_v[!is.na(st_v)]
  boot <- replicate(2000, {
    median(sample(st_v, replace = TRUE)) /
    median(sample(bg_v, replace = TRUE))
  })
  tibble(
    nutrient  = nut_labels_ascii[i],
    fold      = median(st_v) / median(bg_v),
    lo        = quantile(boot, 0.025),
    hi        = quantile(boot, 0.975)
  )
}) |>
  mutate(nutrient = factor(nutrient, levels = rev(nut_labels_ascii)))

# ── Panel A: fold-change ─────────────────────────────────────────────────────
pFold <- ggplot(ci_df, aes(x = fold, y = nutrient)) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "gray50", linewidth = 0.5) +
  geom_errorbar(aes(xmin = lo, xmax = hi),
                width = 0.25, linewidth = 0.6,
                color = "#37474F", orientation = "y") +
  geom_point(aes(size = fold), color = "#E65100", shape = 16) +
  geom_text(aes(label = paste0(round(fold, 1), "×")),
            nudge_y = 0.38, size = 3.5, fontface = "bold",
            color = "#E65100") +
  scale_x_continuous(limits = c(0, 12), breaks = c(1, 2, 4, 8)) +
  scale_size_continuous(range = c(3, 8), guide = "none") +
  labs(
    x        = "Fold-enrichment  (storm-influenced ÷ background)",
    y        = NULL, tag = "A",
    subtitle = paste0(
      "Background: Sal > 34 PSU (n = ", nrow(bg),
      ")  |  Storm-influenced: Sal < 32 PSU (n = ", nrow(strm), ")\n",
      "Error bars = 95% bootstrap CI on median ratio (2,000 replicates)"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor    = element_blank(),
        panel.grid.major.y  = element_blank(),
        axis.text.y  = element_text(size = 10.5, face = "bold",
                                    family = "mono"),
        plot.subtitle = element_text(size = 8.5, color = "gray40"),
        plot.tag      = element_text(face = "bold"))

# ── Panel B: distributions ───────────────────────────────────────────────────
nut_long_zones <- sal_zones |>
  select(zone, all_of(nutrients_vec)) |>
  pivot_longer(all_of(nutrients_vec),
               names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>
  mutate(
    nutrient = factor(recode(nutrient,
      "NO3NO2" = "NO3+NO2", "PO4" = "PO4",
      "SiO2"   = "SiO2",    "NH3" = "NH3"),
      levels = nut_labels_ascii),
    zone = factor(zone, levels = names(zone_pal))
  )

bg_means <- bg |>
  summarise(across(all_of(nutrients_vec), \(x) mean(x, na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "nutrient", values_to = "bg_mean") |>
  mutate(nutrient = factor(recode(nutrient,
    "NO3NO2" = "NO3+NO2", "PO4" = "PO4", "SiO2" = "SiO2", "NH3" = "NH3"),
    levels = nut_labels_ascii))

pDist <- ggplot(nut_long_zones,
                aes(x = zone, y = conc, fill = zone, color = zone)) +
  geom_hline(data = bg_means, aes(yintercept = bg_mean),
             linetype = "dashed", color = "gray30",
             linewidth = 0.5, inherit.aes = FALSE) +
  geom_violin(alpha = 0.30, linewidth = 0.3, trim = TRUE) +
  geom_boxplot(width = 0.18, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = zone_pal) +
  scale_color_manual(values = zone_pal) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  facet_wrap(~nutrient, nrow = 1, scales = "free_y") +
  labs(x = NULL, y = expression(mu*"mol/L (log scale)"), tag = "B") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor    = element_blank(),
        panel.grid.major.x  = element_blank(),
        axis.text.x  = element_text(angle = 30, hjust = 1, size = 8),
        strip.text   = element_text(face = "bold", size = 10.5,
                                    family = "mono"),
        legend.position = "none",
        plot.tag     = element_text(face = "bold"))

pFold / pDist + plot_layout(heights = c(1, 1.1))
ggsave(here("output","nutrients_by_salinity.png"), width = 8, height = 6)
```

**Bottom line:** storm-influenced coastal water had silica levels 6× higher, nitrogen 3× higher, and phosphorus 2.3× higher than clean marine water sampled at the same time.

A second finding sits quietly in panel B: even the "background" high-salinity sites had notably elevated nitrogen and phosphorus compared to open-ocean values. Hawaiian coastal waters carry a persistent nutrient burden from cesspools and agriculture that exists year-round, independent of storms. The storm signal is piled on top of an already-elevated baseline — which makes the total impact even more concerning.

---

# Long-term baseline comparison: Hui o Ka Wai Ola monitoring data

The Hui o Ka Wai Ola programme has sampled fixed coastal stations in West and South Maui every month since 2016. Because we know what each site normally looks like, we can ask: how far outside the historical range did each site go during March 2026? This is a stronger test than comparing sites to each other — it uses each location as its own control.

**Units note:** Hui nutrients are reported in µg/L. Conversion to µmol/L: NNN ÷ 14.007 (as N); Phosphate ÷ 30.974 (as P); Silicate ÷ 28.086 (as Si).

```{r}
#| label: hui-baseline-fn
#| include: false

# Build a fold-change + forest-plot comparison for any Hui nutrient column.
#
# Args:
#   data      – bg_hui_loc (or any frame with Station, SiteName, region, Date,
#               and a pre-computed µmol/L column)
#   conc_col  – column name string, e.g. "NNN_umol", "Phosphate_umol"
#   x_label   – x-axis label for the forest plot panel

#   fold_text_cutoff – fold-change above which bar label flips white/inside

hui_baseline_plot <- function(data,
                              conc_col,
                              x_label,
                              fold_text_cutoff = 12,
                              cs_data          = NULL,
                              cs_radius        = 100) {

  # Map Hui µmol column names to the corresponding chem column
  cs_col_map <- c(NNN_umol       = "NO3NO2", Phosphate_umol = "PO4",
                  Silicate_umol  = "SiO2",   NH4_umol       = "NH3")
  cs_col  <- cs_col_map[conc_col]
  show_cs <- !is.null(cs_data) && !is.na(cs_col) && cs_col %in% names(cs_data)

  # ── Data prep ───────────────────────────────────────────────────────────────
  historical <- data |>
    filter(year(Date) < 2026, !is.na(.data[[conc_col]]))

  march <- data |>
    filter(year(Date) == 2026, month(Date) == 3) |>
    mutate(timing = if_else(Date < as.Date("2026-03-14"),
                            "Pre-storm (Mar 10)", "Post-storm (Mar 18–31)"))

  baselines <- historical |>
    group_by(Station, SiteName, region) |>
    summarise(
      p05 = quantile(.data[[conc_col]], 0.05, na.rm = TRUE),
      p25 = quantile(.data[[conc_col]], 0.25, na.rm = TRUE),
      med = median(.data[[conc_col]],         na.rm = TRUE),
      p75 = quantile(.data[[conc_col]], 0.75, na.rm = TRUE),
      p95 = quantile(.data[[conc_col]], 0.95, na.rm = TRUE),
      .groups = "drop"
    )

  post <- march |>
    filter(str_detect(timing, "Post"), !is.na(.data[[conc_col]])) |>
    left_join(baselines, by = c("Station", "SiteName", "region")) |>
    filter(!is.na(med)) |>
    mutate(
      conc        = .data[[conc_col]],
      fold_change = conc / med,
      above_p95   = conc > p95,
      label       = str_trunc(SiteName, 28, ellipsis = "…"),
      region      = factor(region, levels = c("South Maui", "West Maui"))
    ) |>
    arrange(region, desc(fold_change))

  pre <- march |>
    filter(str_detect(timing, "Pre"), !is.na(.data[[conc_col]])) |>
    left_join(select(baselines, Station, SiteName, region, p05, p25, med, p75, p95),
              by = c("Station", "SiteName", "region")) |>
    mutate(
      conc   = .data[[conc_col]],
      label  = str_trunc(SiteName, 28, ellipsis = "…"),
      region = factor(region, levels = c("South Maui", "West Maui"))
    )

  # Station order: most anomalous at top within each region
  lvls      <- post |> arrange(region, fold_change) |> pull(label)
  post      <- post      |> mutate(label = factor(label, levels = lvls))
  pre       <- pre       |> mutate(label = factor(label, levels = lvls))
  baselines <- baselines |>
    mutate(
      label  = factor(str_trunc(SiteName, 28, ellipsis = "…"), levels = lvls),
      region = factor(region, levels = c("South Maui", "West Maui"))
    )

  # ── CS spatial match: mean of post-storm CS samples within cs_radius of each station
  if (show_cs) {
    hui_stations_sf <- data |>
      distinct(Station, SiteName, region, Lat, Long) |>
      filter(!is.na(Lat), !is.na(Long)) |>
      mutate(label = factor(str_trunc(SiteName, 28, ellipsis = "…"), levels = lvls)) |>
      filter(!is.na(label)) |>
      sf::st_as_sf(coords = c("Long", "Lat"), crs = 4326) |>
      sf::st_transform(32604)

    cs_sf <- cs_data |>
      filter(!is.na(.data[[cs_col]]), .data[[cs_col]] > 0,
             !is.na(latitude), !is.na(longitude),
             as.Date(collected_hst, tz = "HST") >= as.Date("2026-03-14")) |>
      mutate(cs_val = .data[[cs_col]]) |>
      sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
      sf::st_transform(32604) |>
      dplyr::select(cs_val)

    cs_matched <- sf::st_join(
      hui_stations_sf, cs_sf,
      join = sf::st_is_within_distance, dist = cs_radius
    ) |>
      sf::st_drop_geometry() |>
      filter(!is.na(cs_val)) |>
      group_by(label, region) |>
      summarise(cs_val = mean(cs_val, na.rm = TRUE), n_cs = n(), .groups = "drop")
  }

  # ── Panel A: fold-change bars ────────────────────────────────────────────────
  pA <- ggplot(post, aes(x = fold_change, y = label, fill = above_p95)) +
    geom_col(width = 0.65, color = NA) +
    geom_text(aes(label  = paste0(round(fold_change, 1), "×"),
                  hjust  = if_else(fold_change > fold_text_cutoff,  1.1, -0.1),
                  color  = if_else(fold_change > fold_text_cutoff, "white", "gray25")),
              size = 2.5, fontface = "bold") +
    geom_vline(xintercept = 1, color = "gray40", linewidth = 0.5, linetype = "dashed") +
    scale_fill_manual(values = c("TRUE" = "#E65100", "FALSE" = "#90A4AE"),
                      labels = c("TRUE" = "Above 95th pct", "FALSE" = "Within range"),
                      name   = NULL) +
    scale_color_identity() +
    scale_x_continuous(expand = expansion(mult = c(0, 0.22)),
                       breaks = c(1, 5, 10, 20, 40)) +
    facet_grid(region ~ ., scales = "free_y", space = "free_y", switch = "y") +
    labs(x = "Fold above historical median", y = NULL, tag = "A") +
    theme_minimal(base_size = 9.5) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      strip.text         = element_text(face = "bold", size = 9),
      strip.placement    = "outside",
      legend.position    = "bottom",
      legend.text        = element_text(size = 8.5),
      axis.text.y        = element_text(size = 7, lineheight = 0.82),
      plot.tag           = element_text(face = "bold"),
      plot.caption       = element_text(size = 8, color = "gray50", hjust = 0)
    )

  # ── Panel B: build legend values conditionally on cs availability ─────────────
  legend_vals <- c("Historical median" = "white",
                   "Hui post-storm"    = "#E65100")
  legend_ov   <- list(size   = c(2,       3),
                      shape  = c(21,      21),
                      color  = c("gray40", "#B04000"),
                      stroke = c(0.5,     0.5),
                      alpha  = c(1,       0.95))
  if (show_cs && nrow(cs_matched) > 0) {
    legend_vals["WIYW (≤100 m)"] <- "#AB47BC"
    legend_ov$size   <- c(legend_ov$size,   3)
    legend_ov$shape  <- c(legend_ov$shape,  21)
    legend_ov$color  <- c(legend_ov$color,  "#4A148C")
    legend_ov$stroke <- c(legend_ov$stroke, 1.2)
    legend_ov$alpha  <- c(legend_ov$alpha,  0.85)
  }

  # ── Panel B: forest plot ─────────────────────────────────────────────
  pB <- ggplot() +
    geom_segment(data = filter(baselines, !is.na(label)),
                 aes(x = p05, xend = p95, y = label, yend = label),
                 color = "gray65", linewidth = 0.6, lineend = "round") +
    geom_segment(data = filter(baselines, !is.na(label)),
                 aes(x = p25, xend = p75, y = label, yend = label),
                 color = "gray30", linewidth = 3, lineend = "butt") +
    geom_point(data = filter(baselines, !is.na(label)),
               aes(x = med, y = label, fill = "Historical median"),
               color = "gray40", size = 1.8, shape = 21, stroke = 0.5) +
    (if (show_cs && nrow(cs_matched) > 0)
      geom_point(data = cs_matched,
                 aes(x = cs_val, y = label, fill = "WIYW (≤100 m)"),
                 shape = 21, size = 2, stroke = 1.2,
                 color = "#4A148C", alpha = 0.85)) +
    geom_point(data = filter(post, !is.na(label)),
               aes(x = conc, y = label, fill = "Hui post-storm"),
               color = "#B04000", shape = 21, size = 2, stroke = 0.5, alpha = 0.95) +
    scale_fill_manual(name   = NULL,
                      values = legend_vals,
                      drop   = FALSE,
                      guide  = guide_legend(override.aes = legend_ov)) +
    scale_x_log10(breaks = scales::breaks_log(n = 6),
                  labels = scales::label_number(drop0trailing = TRUE),
                  expand = expansion(mult = c(0.02, 0.05))) +
    facet_grid(region ~ ., scales = "free_y", space = "free_y", switch = "y") +
    labs(x = x_label, y = NULL, tag = "B") +
    theme_minimal(base_size = 9.5) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      strip.text         = element_blank(),
      axis.text.y        = element_blank(),
      axis.ticks.y       = element_blank(),
      plot.tag           = element_text(face = "bold"),
      plot.margin        = margin(t = 4, r = 8, b = 4, l = 0),
      legend.position      = "bottom",
      legend.justification = "left",
      legend.text          = element_text(size = 7.5),
      legend.key.size      = unit(0.55, "lines")
    )

  # ── Assemble ─────────────────────────────────────────────────────────────────
  (pA | pB) +
    plot_layout(widths = c(1.4, 1)) +
    plot_annotation(
      caption = "Source: Hui o Ka Wai Ola long-term monitoring programme (Falinski et al.)."
    )
}
```

```{r}
#| label: hui-baseline-comparison
#| fig-height: 11
#| fig-width: 11
#| fig-cap: "(A) How many times higher was each Hui station's post-storm nitrogen reading compared to its own long-term average? Stations are ordered from most to least affected. Orange bars = readings above that site's historical 95th percentile (i.e., higher than almost anything ever recorded there). Gray = within the normal historical range. Mālā Ramp (43×) and Camp Olowalu (32×) on West Maui show the most extreme departures. (B) Where did post-storm readings land relative to each site's full history? The gray bar shows the middle 50% of historical readings; the thin line extends to the 5th–95th percentile range; the white dot is the historical median. Orange = post-storm; blue = pre-storm (March 10). Before the storm, all sites were within their normal ranges. After, most shot well above. Data: Hui o Ka Wai Ola long-term monitoring programme."

HuiNN_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "NNN_umol",
                  x_label   = expression(NO[3]+NO[2]~"("*mu*"mol/L, log)"),
                  cs_data   = chem)

HuiPO_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "Phosphate_umol",
                  x_label   = expression(PO[4]~"("*mu*"mol/L, log)"),
                  cs_data   = chem)

HuiSi_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "Silicate_umol",
                  x_label   = expression(SiO[2]~"("*mu*"mol/L, log)"),
                  cs_data   = chem)

HuiNH3_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "NH4_umol",
                  x_label   = expression(NH[3]~"("*mu*"mol/L, log)"),
                  cs_data   = chem)

HuiNN/HuiNN_raw
ggsave(here("output","allNN_HuiData.png"), width = 8, height = 16)

HuiPO/HuiPO_raw
ggsave(here("output","allPO_HuiData.png"), width = 8, height = 16)

HuiSi/HuiSi_raw
ggsave(here("output","allSi_HuiData.png"), width = 8, height = 16)

HuiNH3/HuiNH3_raw
ggsave(here("output","allNH3_HuiData.png"), width = 16, height = 8)

```

## Direct comparison: community science vs Hui within 100 m

To directly compare the two independent datasets at the same locations, we matched community science samples to Hui stations using a nearest-neighbour spatial join with a 100 m radius (UTM zone 4N). Both datasets are filtered to post-storm dates (≥ March 14, 2026). The datasets do not sample identically in time — Hui sampled March 18–31 and community science March 28–April 7 — so date difference (CS date − Hui date) is shown as point colour.

```{r}
#| label: hui-cs-comparison
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Direct comparison of post-storm community science (What's in your water) nutrients (y) against Hui o Ka Wai Ola(x) at similar locations (within 100 m). Each point is one Hui station × CS sample pair. The dashed line is 1:1 agreement; colour shows the gap in days between the two sample dates (positive = CS collected after Hui). Only Maui stations are shown, as Hui monitoring is restricted to Maui."

# ── Hui post-storm: one row per station-date observation ─────────────────────
hui_match_sf <- bg_hui_loc |>
  filter(year(Date) == 2026, month(Date) == 3,
         Date >= as.Date("2026-03-14"),
         !is.na(Lat), !is.na(Long)) |>
  st_as_sf(coords = c("Long", "Lat"), crs = 4326) |>
  st_transform(32604)

# ── Post-storm community science samples (Maui only, with coordinates) ────────
cs_match_sf <- chem |>
  filter(island == "Maui",
         !is.na(latitude), !is.na(longitude),
         as.Date(collected_hst, tz = "HST") >= as.Date("2026-03-14")) |>
  mutate(cs_date = as.Date(collected_hst, tz = "HST")) |>
  select(sample_id, cs_date, cs_NO3NO2 = NO3NO2, cs_PO4 = PO4,
         cs_SiO2 = SiO2, cs_NH3 = NH3, latitude, longitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

# ── Spatial join: CS samples within 100 m of each Hui observation ─────────────
matched_pairs <- st_join(
  hui_match_sf,
  cs_match_sf,
  join    = st_is_within_distance,
  dist    = 100
) |>
  st_drop_geometry() |>
  filter(!is.na(sample_id)) |>
  mutate(
    date_diff = as.integer(cs_date - Date),  # positive = WIYW collected after Hui
    dist_label = "≤ 100 m"
  )

cat(glue::glue(
  "{nrow(matched_pairs)} Hui–CS pairs matched ",
  "({n_distinct(matched_pairs$Station)} Hui stations, ",
  "{n_distinct(matched_pairs$sample_id)} CS samples)\n",
  "Date gap range: {min(matched_pairs$date_diff)} to {max(matched_pairs$date_diff)} days\n"
))

# ── Reshape to long for faceting ──────────────────────────────────────────────
nutrient_map <- tribble(
  ~hui_col,        ~cs_col,      ~label,
  "NNN_umol",      "cs_NO3NO2",  "NO[3]+NO[2]",
  "Phosphate_umol","cs_PO4",     "PO[4]",
  "Silicate_umol", "cs_SiO2",    "SiO[2]",
  "NH4_umol",      "cs_NH3",     "NH[4]"
)

pairs_long <- nutrient_map |>
  purrr::pmap(function(hui_col, cs_col, label) {
    matched_pairs |>
      filter(!is.na(.data[[hui_col]]), !is.na(.data[[cs_col]])) |>
      transmute(
        Station, SiteName, Date, cs_date, date_diff,
        hui_val = .data[[hui_col]],
        cs_val  = .data[[cs_col]],
        nutrient = label
      )
  }) |>
  list_rbind() |>
  mutate(nutrient = factor(nutrient, levels = nutrient_map$label))

# ── Per-nutrient correlations for facet labels ────────────────────────────────
cor_labels <- pairs_long |>
  filter(hui_val > 0, cs_val > 0) |>
  group_by(nutrient) |>
  summarise(
    r   = cor(log10(hui_val), log10(cs_val), method = "pearson",  use = "complete.obs"),
    rho = cor(log10(hui_val), log10(cs_val), method = "spearman", use = "complete.obs"),
    n   = n(),
    .groups = "drop"
  ) |>
  mutate(label = paste0(
    "atop(", nutrient, "~(mu*mol/L), ",
    "paste('r =', ", round(r, 2), ", '~~', rho, '=', ", round(rho, 2), ", '~~n =', ", n, "))"
  ))

cor_labeller <- setNames(cor_labels$label, cor_labels$nutrient)

# ── Plot ──────────────────────────────────────────────────────────────────────
ggplot(pairs_long |>
  filter(hui_val > 0, cs_val > 0), aes(x = hui_val, y = cs_val, color = date_diff)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "gray50", linewidth = 0.5) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_gradient2(
    low = "#1565C0", mid = "#FFA726", high = "#B71C1C", midpoint = 7,
    name = "Days WIYW\nafter Hui",
    guide = guide_colorbar(barheight = unit(8, "lines"), barwidth = unit(0.5, "lines"))
  ) +
  scale_x_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  facet_wrap(~nutrient, scales = "free", ncol = 2,
             labeller = as_labeller(cor_labeller, label_parsed)) +
  labs(
    x = expression("Hui o Ka Wai Ola ("*mu*"mol/L, log)"),
    y = expression("What's in your water ("*mu*"mol/L, log)"),
    caption = str_wrap("Direct comparison of post-storm WIYW community science NO₃+NO₂ (y) against Hui o Ka Wai Ola NNN (x) at the same locations (within 100 m). Each point is one Hui station × CS sample pair. The dashed line is 1:1 agreement; colour shows the gap in days between the two sample dates (positive = WIYW collected after Hui). Only Maui stations are shown, as Hui monitoring is restricted to Maui", width = 120)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold", size = 9.5),
    plot.caption       = element_text(size = 8, color = "gray50", hjust = 0)
  )
ggsave(here("output","comparisonHui_WIYW.png"))
```

---

# Rainfall and coastal nutrients: 10 years of Hui o Ka Wai Ola

Hui monthly monitoring gives us a 10-year window to ask whether routine rainfall events drive coastal nutrient anomalies, or whether only extreme storms like the March 2026 Kona Low cross the threshold that matters for reef ecosystems. Using full-year monthly precipitation from CHIRPS (all 12 months, 2016–2026, fetched directly from the CHC server as cloud-optimised GeoTIFFs and cropped to the Maui island-mean bounding box), we compare rainfall against station-normalised nutrient fold-changes — how far each month's reading was above or below each station's own 10-year historical median.

```{r}
#| label: chirps-full-year-maui
#| include: false
#| cache: false
#
# Full-year CHIRPS monthly means for Maui, 2016–2026.
# Data source: CHIRPS v2.0 monthly COG product, Climate Hazards Center (CHC),
#   UC Santa Barbara. https://www.chc.ucsb.edu/data/chirps
# Citation: Funk et al. (2015) The climate hazards infrared precipitation
#   with stations — a new environmental record for monitoring extremes.
#   Scientific Data 2:150066. doi:10.1038/sdata.2015.66
#
# Method: terra::rast() via /vsicurl/ reads only the Hawaii bounding box
# from each global GeoTIFF without downloading the full file. Island-mean
# precipitation is computed as the mean of all non-fill pixels within
# lat 20.45–21.15, lon −156.75 to −155.9.

library(terra)

maui_ext <- ext(-156.75, -155.9, 20.45, 21.15)

fetch_chirps_month <- function(yr, mo) {
  url <- sprintf(
    "/vsicurl/https://data.chc.ucsb.edu/products/CHIRPS-2.0/global_monthly/cogs/chirps-v2.0.%04d.%02d.cog",
    yr, mo
  )
  tryCatch({
    r    <- rast(url) |> crop(maui_ext)
    vals <- as.vector(values(r))
    vals <- vals[vals > -9000 & !is.na(vals)]
    tibble(year=yr, month=mo, chirps_mm=mean(vals, na.rm=TRUE))
  }, error=function(e) tibble(year=yr, month=mo, chirps_mm=NA_real_))
}

chirps_full_cache <- here::here("data","chirps_maui_monthly_full.csv")

if (file.exists(chirps_full_cache)) {
  chirps_full_maui <- read_csv(chirps_full_cache, show_col_types=FALSE) |>
    mutate(ym = as.Date(paste(year, month, "1", sep="-")))
} else {
  yms <- expand_grid(year=2016:2026, month=1:12) |>
    filter(!(year==2026 & month > 7))
  chirps_full_maui <- map2_dfr(yms$year, yms$month, fetch_chirps_month) |>
    mutate(ym = as.Date(paste(year, month, "1", sep="-")))
  write_csv(chirps_full_maui, chirps_full_cache)
}
```

```{r}
#| label: hui-rainfall-nutrient
#| fig-height: 6
#| fig-cap: "(A) CHIRPS island-mean monthly rainfall for Maui, 2016–2026, all 12 months. Dark blue bars = top 10% wettest months. March 2026 (373 mm) is 35% wetter than any prior month in the record. (B) Each point is one monthly Hui o Ka Wai Ola sampling event at all West and South Maui stations, showing median NO₃+NO₂ fold-change from each station's own historical baseline vs that month's CHIRPS rainfall (log scale). A statistically significant but modest log-linear relationship holds across the full decade (r² = 0.058, p = 0.009), confirming that wetter months produce proportionally elevated nitrogen. The March 2026 Kona Low sits far outside the envelope of any prior event — the only month to exceed 200 mm in the record and the only one to produce a >2× median fold-change."

# Station-level baselines (pre-2026)
station_baselines_hn <- bg_hui_loc |>
  filter(!is.na(NNN_umol), NNN_umol > 0, year(Date) < 2026) |>
  group_by(Station) |>
  summarise(hist_med_NNN  = median(NNN_umol,       na.rm=TRUE),
            hist_med_SiO2 = median(Silicate_umol,  na.rm=TRUE),
            hist_med_PO4  = median(Phosphate_umol, na.rm=TRUE),
            n_obs         = n(), .groups="drop") |>
  filter(n_obs >= 5)

# Monthly fold-change from station baseline
hui_anomaly_hn <- bg_hui_loc |>
  filter(!is.na(NNN_umol), NNN_umol > 0) |>
  inner_join(station_baselines_hn, by="Station") |>
  mutate(year  = year(Date),  month = month(Date),
         ym    = floor_date(Date,"month"),
         fold_NNN  = NNN_umol      / hist_med_NNN,
         fold_SiO2 = Silicate_umol / hist_med_SiO2,
         fold_PO4  = Phosphate_umol/ hist_med_PO4) |>
  group_by(ym, year, month) |>
  summarise(fold_NNN_med  = median(fold_NNN,  na.rm=TRUE),
            fold_SiO2_med = median(fold_SiO2, na.rm=TRUE),
            fold_PO4_med  = median(fold_PO4,  na.rm=TRUE),
            .groups="drop") |>
  left_join(chirps_full_maui |> select(ym, chirps_mm), by="ym")

# Model
lm_nnn_hn <- lm(fold_NNN_med ~ log1p(chirps_mm),
                 data=filter(hui_anomaly_hn, !is.na(chirps_mm)))
r2_hn <- round(summary(lm_nnn_hn)$r.squared, 3)
p_hn  <- round(summary(lm_nnn_hn)$coefficients[2,"Pr(>|t|)"], 4)

rain_q90 <- quantile(chirps_full_maui$chirps_mm, .90, na.rm=TRUE)

p_rain_hn <- ggplot(chirps_full_maui, aes(x=ym, y=chirps_mm)) +
  geom_col(aes(fill=chirps_mm >= rain_q90), width=25, show.legend=FALSE) +
  scale_fill_manual(values=c("FALSE"="#90CAF9","TRUE"="#1565C0")) +
  annotate("text", x=as.Date("2026-03-01"),
           y=max(chirps_full_maui$chirps_mm,na.rm=TRUE)*0.9,
           label="Mar 2026\n373 mm", size=2.8, color="#1565C0",
           hjust=0.5, fontface="italic", lineheight=0.85) +
  scale_x_date(date_breaks="1 year", date_labels="%Y") +
  labs(x=NULL, y="Rainfall (mm)", tag="A",
       subtitle=paste0("CHIRPS island-mean · all months 2016–2026  |  ",
                       "median ", round(median(chirps_full_maui$chirps_mm,na.rm=TRUE)),
                       " mm · max 373 mm (Mar 2026)")) +
  theme_minimal(base_size=10) +
  theme(panel.grid.minor=element_blank(), plot.tag=element_text(face="bold"),
        plot.subtitle=element_text(size=8, color="gray40"))

p_scatter_hn <- ggplot(
  hui_anomaly_hn |> filter(!is.na(chirps_mm)),
  aes(x=log1p(chirps_mm), y=fold_NNN_med)
) +
  geom_hline(yintercept=1, linetype="dashed", color="gray60", linewidth=0.4) +
  geom_point(aes(color=year>=2023), size=2.3, alpha=0.7) +
  geom_smooth(method="lm", se=TRUE, color="#E65100",
              fill="#FFE0CC", linewidth=0.9) +
  ggrepel::geom_text_repel(
    data=\(d) filter(d, year==2026, month==3),
    aes(label="Mar 2026\nKona Low"), size=2.6, color="#1565C0",
    lineheight=0.85, min.segment.length=0.1) +
  scale_color_manual(
    values = c("FALSE" = "#90A4AE", "TRUE" = "#E65100"),
    labels = c("FALSE" = "2016–2022",  "TRUE" = "2023–2026"),
    name   = NULL
  ) +
  annotate("text", x=3.2, y=max(hui_anomaly_hn$fold_NNN_med,na.rm=TRUE)*0.94,
           label = paste0("atop(r^2 == ", r2_hn, ", p == ", p_hn, ")"),
           parse = TRUE,
           size=2.9, color="#E65100", hjust=0) +
  scale_x_continuous(
    breaks=log1p(c(0,10,25,50,100,200,400)),
    labels=c("0","10","25","50","100","200","400")) +
  labs(x="Monthly rainfall, log scale (mm, CHIRPS island-mean)",
       y=expression(atop(NO[3]+NO[2]~"fold-change", "from station baseline")),
       tag="B",
       subtitle="Each point = one monthly Hui sampling  ·  log scale removes right-skew") +
  theme_minimal(base_size=10) +
  theme(panel.grid.minor    = element_blank(),
        plot.tag             = element_text(face="bold"),
        plot.subtitle        = element_text(size=8, color="gray40"),
        legend.position      = c(0.12, 0.88),
        legend.background    = element_blank(),
        legend.key.size      = unit(0.6, "lines"),
        legend.text          = element_text(size=8))

p_rain_hn / p_scatter_hn +
  plot_annotation(
    title="Maui coastal nitrogen responds to rainfall, but only extreme storms drive ecologically meaningful anomalies",
    subtitle="Hui o Ka Wai Ola fixed stations 2016–2026 · station-normalised fold-change from 10-year baseline",
    theme=theme(plot.title=element_text(face="bold", size=11),
                plot.subtitle=element_text(size=8.5, color="gray40"))
  )
```

The log-linear relationship (r² = `r r2_hn`, p = `r p_hn`) confirms that monthly coastal nitrogen at Hui stations scales with antecedent rainfall, but the effect size is modest for routine events. The chronic nutrient load from cesspools and groundwater creates a persistently elevated baseline that obscures smaller storm signals at fixed leeward stations. Only the March 2026 Kona Low — with 373 mm, the wettest month in the 10-year record and 35% wetter than any previous event — produced a fold-change that clearly separates from this chronic background noise. This threshold behaviour is ecologically important: it means that routine Kona Low events may not substantially elevate nutrients above the already-elevated baseline at these semi-sheltered monitoring stations, while a single extreme event can deliver a nutrient pulse that is qualitatively different from anything observed in a decade of monthly monitoring.

---

# Stoichiometric fingerprints

Did all islands produce the same kind of runoff, or does each island have its own chemical identity? The ratios of nutrients to each other — nitrogen to phosphorus (N:P) and silica to nitrogen (Si:N) — act like a fingerprint that reveals where nutrients came from. Ocean plankton maintain nitrogen and phosphorus in a fairly fixed ratio of 16:1 (the Redfield ratio). Below 16 means unusual phosphorus enrichment, pointing to a cesspool or wastewater source. Above 1 for Si:N means silica-rich volcanic rock weathering is dominant. The four islands show distinctly different ratios, meaning the storm flushed island-specific chemical fingerprints — not a single uniform nutrient pulse.

```{r}
#| label: stoichiometry-figure
#| fig-height: 5.5
#| fig-cap: "(A) Nitrogen-to-phosphorus ratio (N:P) by island. Ocean plankton normally maintain N:P near 16:1 (dashed line) — below 16 means there is more phosphorus than biology expects, pointing to wastewater or cesspool inputs. All four islands are below 16. (B) Silica-to-nitrogen ratio (Si:N). Values above 1 (dashed) indicate the water is carrying more dissolved rock material (silica from volcanic weathering) than nitrogen — the signature of storm runoff over Hawaiian terrain. (C) Both ratios together, with island medians shown as diamonds. All islands sit in the phosphorus-excess, silica-rich corner. Maui's ratio is closest to the biological reference line, suggesting its runoff has more agricultural nitrogen (which drives N:P up toward 16). Oʻahu and Molokaʻi are farther into the phosphorus-excess zone, consistent with denser cesspool infrastructure."


ratios <- chem |>
  filter(!is.na(island), NO3NO2 > 0, PO4 > 0, SiO2 > 0) |>
  mutate(
    NP     = NO3NO2 / PO4,
    SiN    = SiO2   / NO3NO2,
    island = factor(island, levels = c("O'ahu", "Maui", "Moloka'i", "Lāna'i"))
  ) |>
  filter(!is.na(island))

violin_theme <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.title.x = element_blank(), axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

pNP_s <- ggplot(ratios, aes(x = island, y = NP, fill = island, color = island)) +
  geom_hline(yintercept = 16, linetype = "dashed", color = "gray40", linewidth = 0.6) +
  annotate("text", x = Inf, y = 20, label = "Redfield N:P = 16",
           size = 2.7, color = "gray40", hjust = 1.05) +
  geom_violin(alpha = 0.30, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE, fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values = island_pal) + scale_color_manual(values = island_pal) +
  scale_y_log10(breaks = c(0.1, 1, 10, 100),
                labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = "N:P (molar, log)", tag = "A") + violin_theme

pSiN_s <- ggplot(ratios, aes(x = island, y = SiN, fill = island, color = island)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40", linewidth = 0.6) +
  annotate("text", x = Inf, y = 1.35, label = "Diatom Si:N = 1",
           size = 2.7, color = "gray40", hjust = 1.05) +
  geom_violin(alpha = 0.30, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE, fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values = island_pal) + scale_color_manual(values = island_pal) +
  scale_y_log10(breaks = c(0.1, 1, 10, 100, 1000),
                labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = "Si:N (molar, log)", tag = "B") + violin_theme

centroids_s <- ratios |>
  group_by(island) |>
  summarise(NP = median(NP, na.rm = TRUE), SiN = median(SiN, na.rm = TRUE), .groups = "drop")

pScatter_s <- ggplot(ratios, aes(x = NP, y = SiN, color = island)) +
  geom_vline(xintercept = 16, linetype = "dashed", color = "gray40", linewidth = 0.5) +
  geom_hline(yintercept = 1,  linetype = "dashed", color = "gray40", linewidth = 0.5) +
  annotate("text", x = 0.12, y = 0.13, label = "P-excess / Si-limited",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  annotate("text", x = 25,   y = 0.13, label = "N-excess / Si-limited",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  annotate("text", x = 0.12, y = 200,  label = "P-excess / Si-rich",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  annotate("text", x = 25,   y = 200,  label = "N-excess / Si-rich",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  geom_point(size = 1.1, alpha = 0.4) +
  geom_point(data = centroids_s, size = 5.5, shape = 18, alpha = 0.95) +
  ggrepel::geom_label_repel(
    data = centroids_s, aes(label = island),
    size = 3, fontface = "bold", label.padding = unit(0.15, "lines"),
    label.size = 0.2, show.legend = FALSE
  ) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_x_log10(breaks = c(0.1, 1, 10, 100),
                labels = scales::label_number(drop0trailing = TRUE)) +
  scale_y_log10(breaks = c(0.1, 1, 10, 100, 1000),
                labels = scales::label_number(drop0trailing = TRUE)) +
  labs(x = "N:P (molar, log)", y = "Si:N (molar, log)", tag = "C") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "none")

title_row_s <- ggdraw() +
  draw_label("Stoichiometric fingerprints distinguish island nutrient sources",
             fontface = "bold", size = 11, x = 0.01, hjust = 0) +
  draw_label("Dashed lines = Redfield N:P (16) and Si:N (1). Diamonds = island medians.",
             color = "gray40", size = 9, x = 0.01, y = 0.2, hjust = 0)

plot_grid(
  title_row_s,
  plot_grid(pNP_s, pSiN_s, pScatter_s, nrow = 1, rel_widths = c(1, 1, 1.4)),
  ncol = 1, rel_heights = c(0.12, 1)
)
```

Maui's N:P ratio (median `r round(median(ratios$NP[ratios$island == "Maui"], na.rm=TRUE), 1)`) is noticeably higher than Oʻahu (`r round(median(ratios$NP[ratios$island == "O'ahu"], na.rm=TRUE), 1)`) and Molokaʻi (`r round(median(ratios$NP[ratios$island == "Moloka'i"], na.rm=TRUE), 1)`). That gap is consistent with East Maui's large farming areas contributing extra nitrogen — bumping the N:P ratio closer to the 16:1 biological baseline even while still remaining phosphorus-enriched. Every island has Si:N ratios well above 1, meaning volcanic rock dissolution is a dominant contributor across the board — an expected signature of Hawaiian watersheds. Each island's runoff tells a different story about its land use.

---

# Biogeochemical cluster analysis 

We sorted the 694 sample sites into groups (using a statistical clustering method called PAM) based purely on their water chemistry: salinity, nitrogen, phosphorus, and silica. Four distinct water types emerged. The number four was chosen because it was the smallest number that captured all four clearly different kinds of coastal water without over-splitting into groups that were chemically too similar to distinguish. Testing two to six groups confirmed this choice.

```{r}
#| label: cluster-analysis
#| fig-height: 11
#| fig-width: 10
#| fig-cap: "(A) Chemistry map of all 694 sites. The two axes (Principal Components) summarise the four measured variables: left-right roughly represents the land–sea gradient (right = cleaner ocean water, left = more land-influenced); up-down separates nitrogen-heavy from phosphorus-heavy enrichment. Coloured hulls enclose each of the three water types. (B) Chemical fingerprint for each group. Cell colours show whether a variable is above (red) or below (blue) the overall average — each water type has a distinct combination. Numbers are the actual median concentrations. (C) Which islands are in each group? Runoff-dominated sites are disproportionately Maui; marine background is dominated by Oʻahu. (D) Cesspool density in the watershed surrounding each site — runoff-dominated groups sit in areas with more cesspool infrastructure. (E) Where on the map does each water type appear?"

# ── Prepare and cluster ──────────────────────────────────────────────────────
chem_clust <- chem |>
  filter(!is.na(island),
         !is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  mutate(log_NO3NO2 = log1p(NO3NO2),
         log_PO4    = log1p(PO4),
         log_SiO2   = log1p(SiO2))

clust_mat <- chem_clust |>
  select(Salinity, log_NO3NO2, log_PO4, log_SiO2) |>
  scale()

set.seed(8341)
pam4 <- pam(clust_mat, k = 3, pamonce = 5)

# Reorder clusters so that: 1 = Marine bg (lowest), 2 = Nearshore (middle), 3 = Runoff (highest)
# First, add temporary cluster assignments from PAM
chem_clust <- chem_clust |>
  mutate(temp_cluster = pam4$clustering)

# Calculate medians for each PAM cluster
cluster_medians <- chem_clust |>
  group_by(temp_cluster) |>
  summarise(median_NO3 = median(NO3NO2, na.rm = TRUE), .groups = "drop") |>
  arrange(median_NO3) |>
  mutate(new_cluster = row_number())

# Create mapping from old cluster numbers to new ones
cluster_remap <- setNames(cluster_medians$new_cluster, cluster_medians$temp_cluster)

# Apply the remapping to both chem_clust and pam4
chem_clust <- chem_clust |>
  mutate(cluster = as.numeric(cluster_remap[temp_cluster])) |>
  select(-temp_cluster)

pam4$clustering <- as.numeric(cluster_remap[pam4$clustering])

cluster_labels <- c("1" = "Marine background", "2" = "Runoff-dominated",
                    "3" = "Nearshore enriched")
cluster_pal_named <- c(
  "Marine background"  = "#1565C0", "Nearshore enriched" = "#2E7D32",
  "Runoff-dominated"   = "#E65100"
)

chem_clust <- chem_clust |>
  mutate(cluster       = factor(pam4$clustering),
         cluster_label = factor(cluster_labels[cluster],
                                levels = c("Marine background", "Nearshore enriched", "Runoff-dominated")))

# ── PCA ─────────────────────────────────────────────────────────────────────
pca_res <- prcomp(clust_mat, scale. = FALSE)
pc_scores <- as.data.frame(pca_res$x[, 1:2]) |>
  bind_cols(chem_clust |> select(cluster, cluster_label, island))
pct_var <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

loadings <- as.data.frame(pca_res$rotation[, 1:2]) |>
  rownames_to_column("variable") |>
  mutate(variable = recode(variable, "log_NO3NO2" = "NO3+NO2",
                           "log_PO4" = "PO4", "log_SiO2" = "SiO2"),
         scale = 3.2)

hulls <- pc_scores |>
  group_by(cluster_label) |>
  slice(chull(PC1, PC2))

pA_cl <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = cluster_label)) +
  geom_polygon(data = hulls, aes(fill = cluster_label),
               alpha = 0.08, linewidth = 0.3, color = NA) +
  geom_point(size = 1.2, alpha = 0.6) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1*scale, yend = PC2*scale),
               color = "gray25", linewidth = 0.6,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed"),
               inherit.aes = FALSE) +
  geom_label(data = loadings,
             aes(x = PC1*scale*1.15, y = PC2*scale*1.15, label = variable),
             size = 2.9, color = "gray20", linewidth = 0, fill = "white",
             fontface = "bold", inherit.aes = FALSE) +
  scale_color_manual(values = cluster_pal_named, name = NULL) +
  scale_fill_manual(values  = cluster_pal_named, name = NULL) +
  labs(x = paste0("PC1 (", pct_var[1], "% variance)"),
       y = paste0("PC2 (", pct_var[2], "% variance)"), tag = "A") +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid.minor = element_blank(), legend.position = "none",
        plot.tag = element_text(face = "bold"))

# ── Heatmap ──────────────────────────────────────────────────────────────────
fp_vars   <- c("Salinity", "NO3NO2", "PO4", "SiO2")
fp_labels <- c("Salinity\n(PSU)", "NO3+NO2\n(µmol/L)", "PO4\n(µmol/L)", "SiO2\n(µmol/L)")

hm_df <- chem_clust |>
  group_by(cluster_label) |>
  summarise(across(all_of(fp_vars), ~median(., na.rm = TRUE)), .groups = "drop") |>
  pivot_longer(-cluster_label, names_to = "variable", values_to = "median_val") |>
  group_by(variable) |>
  mutate(z = (median_val - mean(median_val)) / sd(median_val)) |>
  ungroup() |>
  mutate(
    variable      = factor(variable, levels = fp_vars, labels = fp_labels),
    cluster_label = factor(cluster_label,
                           levels = c("Runoff-dominated", "Nearshore enriched", "Marine background")),
    val_label     = case_when(
      variable == "Salinity\n(PSU)"    ~ as.character(round(median_val, 1)),
      variable == "NO3+NO2\n(µmol/L)" ~ as.character(round(median_val, 2)),
      variable == "PO4\n(µmol/L)"     ~ as.character(round(median_val, 3)),
      TRUE                             ~ as.character(round(median_val, 1))
    )
  )

pB_cl <- ggplot(hm_df, aes(x = variable, y = cluster_label, fill = z)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = val_label), size = 3, fontface = "bold",
            color = ifelse(abs(hm_df$z) > 1.2, "white", "gray20")) +
  scale_fill_gradient2(low = "#1565C0", mid = "white", high = "#B71C1C",
                       midpoint = 0, name = "Z-score") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = \(x) str_wrap(x, 14)) +
  labs(x = NULL, y = NULL, tag = "B") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), axis.text.x = element_text(size = 8.5, face = "bold"),
        axis.text.y = element_text(size = 8.5), plot.tag = element_text(face = "bold"))

# ── Environmental drivers ─────────────────────────────────────────────────────
# chem already carries ahupuaa / cesspool joins from the load-data chunk
env_df <- chem_clust

pC_cl <- env_df |>
  filter(!is.na(island)) |>
  count(cluster_label, island) |>
  group_by(cluster_label) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  filter(island %in% c("O'ahu", "Maui", "Moloka'i", "Lāna'i")) |>
  mutate(island = factor(island, levels = c("O'ahu", "Maui", "Moloka'i", "Lāna'i"))) |>
  ggplot(aes(x = cluster_label, y = pct, fill = island)) +
  geom_col(width = 0.7, color = "white", linewidth = 0.3) +
  scale_fill_manual(values = c("O'ahu" = "#1565C0", "Maui" = "#2E7D32",
                               "Moloka'i" = "#E65100", "Lāna'i" = "#6A1B9A"), name = NULL) +
  scale_y_continuous(labels = scales::label_percent(), expand = c(0, 0)) +
  scale_x_discrete(labels = \(x) str_wrap(x, 10)) +
  labs(x = NULL, y = "Proportion of sites", tag = "C") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        legend.position = "right", plot.tag = element_text(face = "bold"))

pD_cl <- env_df |>
  filter(!is.na(cesspool_density_per_km2)) |>
  ggplot(aes(x = cluster_label, y = cesspool_density_per_km2 + 0.01,
             fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.3, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE, fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = cluster_pal_named) +
  scale_color_manual(values = cluster_pal_named) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  scale_x_discrete(labels = \(x) str_wrap(x, 10)) +
  labs(x = NULL, y = "Cesspool density\n(per km², log)", tag = "D") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        legend.position = "none", plot.tag = element_text(face = "bold"))

# ── Spatial maps ─────────────────────────────────────────────────────────────
clust_sf <- chem_clust |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

map_cluster_fn <- function(island_name, title = island_name) {
  match_idx <- normalize_name(ahupuaa_raw$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_raw[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union  <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  coast_bbox  <- st_bbox(st_buffer(land_union, 3500))
  pts <- clust_sf |> filter(island == island_name)
  ggplot() +
    geom_sf(data = st_as_sf(st_sf(geometry = land_union)),
            fill = "gray78", color = "gray50", linewidth = 0.3) +
    geom_sf(data = pts, aes(color = cluster_label),
            size = ifelse(island_name == "O'ahu", 1.5, 2.2), alpha = 0.85, shape = 16) +
    scale_color_manual(values = cluster_pal_named, name = NULL, drop = FALSE) +
    coord_sf(crs = 32604, xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = title) +
    theme_void(base_size = 9) +
    theme(plot.title = element_text(face = "bold", size = 9, hjust = 0.5),
          legend.position = "none", plot.margin = margin(2, 2, 2, 2))
}

maps_row <- plot_grid(
  map_cluster_fn("O'ahu", "Oʻahu"), map_cluster_fn("Maui", "Maui"),
  map_cluster_fn("Moloka'i", "Molokaʻi"), map_cluster_fn("Lāna'i", "Lānaʻi"),
  nrow = 1, rel_widths = c(1.6, 1.4, 0.9, 0.7)
)

# ── Assemble ─────────────────────────────────────────────────────────────────
legend_strip <- get_legend(
  pA_cl + theme(legend.position = "bottom", legend.text = element_text(size = 9),
                legend.key.size = unit(0.9, "lines"))
)

plot_grid(
  ggdraw() + draw_label(
    "Four biogeochemical clusters reveal distinct post-storm coastal water types",
    fontface = "bold", size = 11, x = 0.01, hjust = 0) +
    draw_label(
      paste0("PAM clustering on scaled log-nutrients + salinity (k = 3, n = ",
             nrow(chem_clust), " sites). Convex hulls in A = envelopes per cluster."),
      color = "gray40", size = 8.5, x = 0.01, y = 0.22, hjust = 0),
  plot_grid(
    plot_grid(pA_cl, pB_cl, nrow = 1, rel_widths = c(1.3, 1)),
    legend_strip,
    plot_grid(pC_cl, pD_cl, nrow = 1),
    ncol = 1, rel_heights = c(1, 0.07, 0.75)
  ),
  plot_grid(
    ggdraw() + draw_label("Spatial distribution of biogeochemical clusters",
                          fontface = "bold", size = 9.5, x = 0.01, hjust = 0),
    maps_row, ncol = 1, rel_heights = c(0.1, 1)
  ),
  ncol = 1, rel_heights = c(0.06, 1.5, 0.65)
)
```

Three clearly distinct water types emerged from the grouping analysis. **Marine background** (n = `r sum(chem_clust$cluster == "1")`) is clean ocean water — high salinity, low nutrients — concentrated around Oʻahu's less impacted shores. **Nearshore enriched** (n = `r sum(chem_clust$cluster == "2")`) is the most common type — a broad mixing zone where nutrients are elevated but salinity hasn't dropped dramatically, representing the transition between runoff-influenced and background marine conditions. **Runoff-dominated** (n = `r sum(chem_clust$cluster == "3")`) sites have extremely high nitrogen and silica with moderate freshwater dilution, the chemical signature of storm drainage carrying dissolved volcanic rock and soil nitrogen; most of these sites are on Maui's windward coast. The areas with the most cesspools tend to align with the runoff-dominated water type, indicating that wastewater infrastructure amplifies the nutrient signal beyond what natural soil flushing would produce alone.

Ammonium (NH₃) was not included in the grouping analysis. The figure below shows why: nitrogen (NO₃+NO₂) increases cleanly across the water types, with minimal overlap between groups. NH₃, by contrast, is scattered across all three groups with overlapping ranges — it doesn't distinguish them. Adding it to the analysis weakened the group definitions and excluded 120 sites that lacked NH₃ measurements. NH₃ appears to build up through seafloor biological processes that are independent of storm runoff.

```{r}
#| label: nh3-posthoc
#| fig-height: 8
#| fig-cap: "Water chemistry by cluster type. (A) NO₃+NO₂ shows ordered separation across the three groups — this is the primary clustering variable that structures the groups. (B) Phosphate (PO₄) and (C) Silicate (SiO₂) also show clear gradients from marine background through nearshore enriched to runoff-dominated sites, with runoff-dominated showing the highest concentrations. (D) Ammonium (NH₃) was left out of the grouping analysis because it shows no systematic pattern across water types — distributions overlap substantially across all three groups. Adding NH₃ would add noise, not clarity. NH₃ data available for `r sum(!is.na(chem_clust$NH3))` of `r nrow(chem_clust)` sites."

clust_pal_ch <- c(
  "Marine background"  = "#1565C0",
  "Nearshore enriched" = "#2E7D32",
  "Runoff-dominated"   = "#E65100"
)

vtheme_nh3 <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position    = "none",
    axis.title.x       = element_blank(),
    axis.text.x        = element_text(angle = 25, hjust = 1, size = 9),
    plot.subtitle      = element_text(size = 8.5, color = "gray40",
                                      margin = margin(t = 2, b = 4)),
    plot.tag           = element_text(face = "bold")
  )

nh3_df_ch <- chem_clust |>
  filter(!is.na(NH3), NH3 > 0) |>
  mutate(cluster_label = factor(cluster_label, levels = names(clust_pal_ch)))

no3_df_ch <- chem_clust |>
  filter(!is.na(NO3NO2), NO3NO2 > 0) |>
  mutate(cluster_label = factor(cluster_label, levels = names(clust_pal_ch)))

pNO3_ch <- ggplot(no3_df_ch, aes(x = cluster_label, y = NO3NO2,
                                   fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.25, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = clust_pal_ch) +
  scale_color_manual(values = clust_pal_ch) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = expression(NO[3]+NO[2]~"("*mu*"mol/L, log)"), tag = "A",
       subtitle = "Ordered separation across clusters (structuring variable)") +
  vtheme_nh3

pPO4_ch <- ggplot(no3_df_ch, aes(x = cluster_label, y = PO4,
                                   fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.25, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = clust_pal_ch) +
  scale_color_manual(values = clust_pal_ch) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = expression(PO[4]~"("*mu*"mol/L, log)"), tag = "B") +
  vtheme_nh3

pSiO2_ch <- ggplot(no3_df_ch, aes(x = cluster_label, y = SiO2,
                                    fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.25, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = clust_pal_ch) +
  scale_color_manual(values = clust_pal_ch) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = expression(SiO[2]~"("*mu*"mol/L, log)"), tag = "C") +
  vtheme_nh3

pNH3_ch <- ggplot(nh3_df_ch, aes(x = cluster_label, y = NH3,
                                   fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.25, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = clust_pal_ch) +
  scale_color_manual(values = clust_pal_ch) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = expression(NH[3]~"("*mu*"mol/L, log)"), tag = "D",
       subtitle = "No systematic pattern — cluster-independent (post-hoc)") +
  vtheme_nh3

(pNO3_ch | pPO4_ch) / (pSiO2_ch | pNH3_ch)
```

---

# Supplementary: cluster selection rationale

```{r}
#| label: silhouette-supp
#| fig-height: 5
#| fig-cap: "How was three groups chosen? (A) Silhouette width measures how well each site fits its assigned group versus how closely it matches neighbouring groups (higher = better-defined groups). Two groups scores highest, but splitting the coast into just 'freshwater-influenced' and 'marine' collapses the ecologically meaningful gradient into a binary that loses most information. Three groups (highlighted) captures the primary ecological gradients — from clean marine background through runoff-dominated plumes to nearshore-enriched mixing zones — while maintaining well-defined boundaries. (B) Silhouette width for each of the 694 individual sites in the k = 3 solution. Negative values (sites that fit a neighbouring group slightly better) are concentrated at the Runoff-dominated / Nearshore enriched boundary — those two groups shade into each other chemically, which is ecologically expected for a continuous mixing gradient."

# clust_mat and pam4 are computed in the cluster-analysis chunk above
set.seed(8341)
sil_summary_s <- map_dfr(2:6, function(k) {
  pk  <- pam(clust_mat, k = k, pamonce = 5)
  sil <- silhouette(pk$clustering, dist(clust_mat))
  tibble(k = k, mean_sil = mean(sil[, 3]))
})

pSilA_s <- ggplot(sil_summary_s, aes(x = k, y = mean_sil)) +
  geom_line(color = "gray40", linewidth = 0.8) +
  geom_point(size = 3.5, color = "gray30") +
  geom_point(data = filter(sil_summary_s, k == 3), size = 5.5, shape = 21,
             fill = "#E65100", color = "white", stroke = 1.5) +
  annotate("text", x = 3.2, y = filter(sil_summary_s, k == 3)$mean_sil + 0.008,
           label = "k = 4\n(chosen)", hjust = 0, size = 3.2, color = "#E65100") +
  scale_x_continuous(breaks = 2:6) +
  scale_y_continuous(limits = c(0.18, 0.43)) +
  labs(x = "Number of clusters (k)", y = "Mean silhouette width",
       title = "A  Average silhouette width by k",
       subtitle = "Higher = more compact, better-separated clusters") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

sil4_s <- as.data.frame(silhouette(pam4$clustering, dist(clust_mat))) |>
  mutate(
    cluster_label = factor(cluster_labels[cluster], levels =  cluster_labels[c(1,3,2)])
  ) |>
  arrange(cluster, sil_width) |>
  mutate(sorted_id = row_number())

pSilB_s <- ggplot(sil4_s, aes(x = sorted_id, y = sil_width, fill = cluster_label)) +
  geom_col(width = 1, color = NA) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.5) +
  geom_hline(yintercept = mean(sil4_s$sil_width),
             linetype = "dashed", color = "gray40", linewidth = 0.5) +
  annotate("text", x = nrow(sil4_s) * 0.98,
           y = mean(sil4_s$sil_width) + 0.03,
           label = sprintf("Mean = %.2f", mean(sil4_s$sil_width)),
           hjust = 1, size = 3, color = "gray30") +
  scale_fill_manual(values = cluster_pal_named, name = NULL) +
  scale_y_continuous(limits = c(-0.25, 1)) +
  labs(x = "Sites (sorted within cluster)", y = "Silhouette width",
       title = "B  Individual silhouette widths, k = 4",
       subtitle = "Negative values = misclassified sites") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid  = element_blank(), legend.position = "bottom",
        legend.text = element_text(size = 9), legend.key.size = unit(0.8, "lines"),
        plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

pSilA_s | pSilB_s
```

# Random forest: landscape predictors of cluster membership 

Can we predict which type of water a site will have — before any sampling — just from publicly available maps and records? We used a random forest model (an ensemble of 1,000 decision trees) with thirteen landscape and hydrological predictors. **Infrastructure proximity**: distance to the nearest cesspool, cesspool density per ahupuaʻa, and distance to the nearest NPDES discharge point (wastewater treatment plant outfall or industrial discharge; EPA ECHO 2025). **Storm delivery**: distance to the nearest stream mouth (OpenStreetMap), mean daily discharge in March 2026 from the nearest USGS gauge within 15 km. **Land cover**: % impervious surface and % agricultural land (NOAA C-CAP 2010–2011). **Watershed hydrology**: mean slope per ahupuaʻa (NASA SRTM via *elevatr*; Farr et al. 2007) and soil saturated hydraulic conductivity (SSURGO Ksat, top 100 cm; USDA NRCS 2024) — which together control how fast rainfall becomes runoff. **Pre-storm conditions**: root-zone soil wetness in the 30 days before the storm (NASA POWER, Feb 1–Mar 10 2026; Sparks et al. 2018). **Livestock density**: cattle per km² of agricultural land (USDA NASS 2022 Census of Agriculture). 

```{r}
#| label: stream-mouths-cache
#| include: false

cache_path <- here::here("data", "osm_stream_mouths.csv")

if (file.exists(cache_path)) {

  stream_mouths_sf <- read_csv(cache_path, show_col_types = FALSE) |>
    sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    sf::st_transform(32604)

} else {

  island_bbs <- list(
    "O'ahu"    = c(-158.32, 21.22, -157.60, 21.73),
    "Maui"     = c(-156.70, 20.50, -155.94, 21.05),
    "Moloka'i" = c(-157.37, 21.04, -156.70, 21.24),
    "Lāna'i"   = c(-157.08, 20.71, -156.83, 20.95)
  )

  fetch_endpoints <- function(bb, island_name) {
    message("Fetching waterways: ", island_name)
    osm <- tryCatch(
      opq(bbox = bb, timeout = 120) |>
        add_osm_feature(key = "waterway", value = c("river", "stream")) |>
        osmdata_sf(),
      error = function(e) { warning(island_name, ": ", e$message); NULL }
    )
    lines <- osm$osm_lines
    if (is.null(lines) || nrow(lines) == 0) return(NULL)

    ep_list <- lapply(seq_len(nrow(lines)), function(i) {
      coords <- st_coordinates(lines[i, ])
      data.frame(lon = c(coords[1, "X"], coords[nrow(coords), "X"]),
                 lat = c(coords[1, "Y"], coords[nrow(coords), "Y"]))
    })
    bind_rows(ep_list) |>
      st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
      st_transform(32604) |>
      mutate(island = island_name)
  }

  all_endpoints <- map2(island_bbs, names(island_bbs), fetch_endpoints) |>
    compact() |>
    bind_rows()

  # Filter to endpoints within 400 m of the coastline
  coastlines_sm <- map(names(island_bbs), function(isl) {
    norm <- normalize_name(isl)
    idx  <- normalize_name(ahupuaa_raw$mokupuni) == norm
    ahupuaa_raw[idx, ] |>
      st_transform(32604) |> st_make_valid() |> st_union() |>
      st_cast("POLYGON") |>
      (\(p) p[as.numeric(st_area(p)) > 1e6])() |>
      st_union() |>
      st_boundary()
  }) |> setNames(names(island_bbs))

  mouth_list <- map(names(island_bbs), function(isl) {
    pts   <- all_endpoints |> filter(island == isl)
    coast <- coastlines_sm[[isl]]
    pts[as.numeric(st_distance(pts, coast)) <= 400, ]
  })

  stream_mouths_sf <- bind_rows(mouth_list)
  stream_mouths_sf |>
    sf::st_transform(4326) |>    # save as WGS84 degrees, not UTM metres
    dplyr::mutate(longitude = sf::st_coordinates(geometry)[,1],
                  latitude  = sf::st_coordinates(geometry)[,2]) |>
    sf::st_drop_geometry() |>
    readr::write_csv(cache_path)
  message("Stream mouths fetched and cached: ", nrow(stream_mouths_sf), " candidates")

}

# Compute distance from every chem site with coordinates to nearest stream mouth
chem_sf_rf <- chem |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

nearest_idx <- st_nearest_feature(chem_sf_rf, stream_mouths_sf)
chem_with_stream <- chem |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  mutate(dist_to_stream_mouth_m = as.numeric(
    st_distance(chem_sf_rf, stream_mouths_sf[nearest_idx, ], by_element = TRUE)
  ))
```

```{r}
#| label: ccap-landcover-cache
#| include: false

ccap_cache_path <- here::here("data", "ccap_ahupuaa_summary.csv")

if (file.exists(ccap_cache_path)) {

  ccap_summary <- read_csv(ccap_cache_path, show_col_types = FALSE)

} else {

  ccap_base_url <- "https://chs.coast.noaa.gov/htdata/raster1/landcover/bulkdownload/hires/hi/"
  ccap_files <- c(
    "O'ahu"    = "hi_oahu_2011_ccap_hr_land_cover20140619.img",
    "Maui"     = "hi_maui_2010_ccap_hr_land_cover_20150213.img",
    "Moloka'i" = "hi_molokai_2010_ccap_hr_land_cover20150102.img",
    "Lāna'i"   = "hi_lanai_2011_ccap_hr_land_cover_20141204.img"
  )

  dir.create(here::here("data", "ccap"), showWarnings = FALSE)
  for (isl in names(ccap_files)) {
    dest <- here::here("data", "ccap", ccap_files[isl])
    if (!file.exists(dest))
      download.file(paste0(ccap_base_url, ccap_files[isl]), dest, mode = "wb", quiet = TRUE)
  }

  # C-CAP class groups: impervious = high/med/low developed; agriculture = cultivated + pasture
  IMPERVIOUS  <- c(2, 3, 4)
  AGRICULTURE <- c(6, 7)

  summarise_ccap <- function(island_name, img_file) {
    r   <- rast(here::here("data", "ccap", img_file))
    ahu <- ahupuaa_raw |>
      filter(normalize_name(mokupuni) == normalize_name(island_name)) |>
      st_transform(crs(r)) |>
      st_make_valid()
    fracs   <- exact_extract(r, ahu, "frac", progress = FALSE)
    imp_cols <- paste0("frac_", IMPERVIOUS)
    agr_cols <- paste0("frac_", AGRICULTURE)
    ahu |>
      st_drop_geometry() |>
      select(ahupuaa, moku, mokupuni) |>
      mutate(
        island          = island_name,
        pct_impervious  = rowSums(fracs[, intersect(imp_cols, names(fracs)), drop=FALSE], na.rm=TRUE),
        pct_agriculture = rowSums(fracs[, intersect(agr_cols, names(fracs)), drop=FALSE], na.rm=TRUE)
      )
  }

  ccap_summary <- map2(names(ccap_files), ccap_files, summarise_ccap) |> bind_rows()
  write_csv(ccap_summary, ccap_cache_path)

}
```

```{r}
#| label: additional-predictors-cache
#| include: false
#
# Six additional landscape predictors joined to the RF dataset.
# Each is cached to data/ on first run; subsequent renders read from cache.
#
# CITATIONS
# Watershed slope: NASA SRTM v3 (Farr et al. 2007, Rev. Geophys.); accessed
#   via elevatr R package (Hollister et al. 2023, JOSS).
# Soil permeability: USDA SSURGO (Soil Survey Staff, NRCS, USDA 2024); queried
#   via soilDB Soil Data Access API (Beaudette et al. 2022, Comput. Geosci.).
#   Ksat = saturated hydraulic conductivity (µm/s), mean over top 100 cm per
#   ahupuaʻa centroid.
# NPDES outfall distance: US EPA ICIS-NPDES Discharge Points national layer
#   (US EPA ECHO 2025; echo.epa.gov/files/echodownloads/npdes_outfalls_layer.zip).
#   Distance to nearest permitted discharge point (WWTPs, ocean outfalls,
#   industrial stormwater, injection wells).
# Antecedent soil moisture: NASA POWER (Sparks et al. 2018, AAAS);
#   GWETROOT = root-zone soil wetness fraction, 0.5° grid, mean Feb 1–Mar 10 2026
#   (30-day pre-storm window). Retrieved via nasapower R package (Sparks 2023).
# Livestock density: USDA NASS 2022 Census of Agriculture (USDA NASS 2024);
#   cattle & calves inventory per km² of agricultural land, by county.
#   Maui County covers Maui, Molokaʻi, and Lānaʻi; Honolulu County = Oʻahu.
# Soil permeability method: for each ahupuaʻa centroid (WGS84), SDA_query() was
#   used to identify the dominant soil map unit and compute mean Ksat from all
#   horizons in the top 100 cm with reported values (SSURGO tabular data via SDA).

# ── 1. Watershed slope ─────────────────────────────────────────────────────
slope_cache <- here::here("data", "ahupuaa_slope.csv")
if (!file.exists(slope_cache)) {
  islands_needed <- c("O'ahu","Maui","Moloka'i","Lāna'i")
  slope_by_island <- purrr::map(islands_needed, function(isl) {
    ahu_isl <- ahupuaa_raw |>
      filter(normalize_name(mokupuni) == normalize_name(isl)) |>
      st_make_valid()
    elev    <- elevatr::get_elev_raster(ahu_isl, z = 10, clip = "locations",
                                        src = "aws", verbose = FALSE) |>
      terra::rast() |> terra::project("EPSG:32604")
    slope_r <- terra::terrain(elev, v = "slope", unit = "degrees")
    ahu_proj <- ahu_isl |> st_transform(32604) |> st_make_valid()
    exactextractr::exact_extract(slope_r, ahu_proj, "mean", progress=FALSE) |>
      tibble::tibble(mean_slope_deg = _) |>
      dplyr::bind_cols(ahu_proj |> st_drop_geometry() |>
                         dplyr::select(ahupuaa, moku, mokupuni)) |>
      dplyr::mutate(island = isl)
  })
  slope_df <- dplyr::bind_rows(slope_by_island)
  write_csv(slope_df, slope_cache)
} else { slope_df <- read_csv(slope_cache, show_col_types = FALSE) }

# ── 2. Soil permeability (SSURGO Ksat) ────────────────────────────────────
ksat_cache <- here::here("data", "ahupuaa_ksat.csv")
if (!file.exists(ksat_cache)) {
  ksat_raw <- soilDB::SDA_query(
    "SELECT mu.mukey, AVG(ch.ksat_r) AS mean_ksat_um_s
     FROM mapunit mu
     JOIN component co ON mu.mukey=co.mukey AND co.majcompflag='Yes'
     JOIN chorizon ch  ON co.cokey=ch.cokey
     JOIN legend   le  ON mu.lkey=le.lkey
     WHERE le.areasymbol LIKE 'HI%'
       AND ch.hzdept_r < 100 AND ch.ksat_r IS NOT NULL
     GROUP BY mu.mukey"
  )
  ahu_centroids_sda <- ahupuaa_raw |>
    filter(normalize_name(mokupuni) %in%
             normalize_name(c("O'ahu","Maui","Moloka'i","Lāna'i"))) |>
    st_centroid() |> st_transform(4326) |>
    mutate(lon = st_coordinates(geometry)[,1],
           lat = st_coordinates(geometry)[,2]) |>
    st_drop_geometry()
  get_ksat_pt <- function(lon, lat) {
    q <- sprintf(
      "SELECT mu.mukey, AVG(ch.ksat_r) AS mean_ksat
       FROM mapunit mu JOIN component co ON mu.mukey=co.mukey AND co.majcompflag='Yes'
       JOIN chorizon ch ON co.cokey=ch.cokey
       WHERE mu.mukey IN (SELECT mukey FROM
         SDA_Get_Mukey_from_intersection_with_WktWgs84('POINT(%s %s)'))
         AND ch.hzdept_r < 100 AND ch.ksat_r IS NOT NULL GROUP BY mu.mukey",
      round(lon,6), round(lat,6))
    res <- tryCatch(soilDB::SDA_query(q), error=function(e) NULL)
    if (is.null(res)||nrow(res)==0) NA_real_ else mean(res$mean_ksat, na.rm=TRUE)
  }
  ksat_df <- ahu_centroids_sda |>
    dplyr::mutate(mean_ksat_um_s = purrr::map2_dbl(lon, lat, get_ksat_pt)) |>
    dplyr::select(ahupuaa, moku, mokupuni, mean_ksat_um_s)
  write_csv(ksat_df, ksat_cache)
} else { ksat_df <- read_csv(ksat_cache, show_col_types = FALSE) }

# ── 3. NPDES outfall / WWTP locations ─────────────────────────────────────
echo_cache <- here::here("data", "epa_facilities_hi.csv")
if (!file.exists(echo_cache)) {
  tmp_zip <- tempfile(fileext = ".zip")
  tmp_dir <- file.path(tempdir(), "npdes_outfalls")
  dir.create(tmp_dir, showWarnings = FALSE)
  download.file("https://echo.epa.gov/files/echodownloads/npdes_outfalls_layer.zip",
                tmp_zip, mode = "wb", quiet = TRUE)
  unzip(tmp_zip, exdir = tmp_dir, overwrite = TRUE)
  outfalls_raw <- readr::read_csv(
    file.path(tmp_dir, "npdes_outfalls_layer.csv"), show_col_types = FALSE)
  outfalls_hi <- outfalls_raw |>
    dplyr::filter(STATE_CODE == "HI", !is.na(LATITUDE83), !is.na(LONGITUDE83),
                  LATITUDE83 > 18, LATITUDE83 < 23) |>
    dplyr::transmute(facility_name = FACILITY_NAME,
                     lat = LATITUDE83, lon = LONGITUDE83)
  write_csv(outfalls_hi, echo_cache)
} else { outfalls_hi <- read_csv(echo_cache, show_col_types = FALSE) }

outfalls_sf <- outfalls_hi |>
  dplyr::distinct(facility_name, lat, lon) |>
  sf::st_as_sf(coords = c("lon","lat"), crs = 4326) |>
  sf::st_transform(32604)

# ── 4. Antecedent soil moisture (NASA POWER) ──────────────────────────────
smap_cache <- here::here("data", "nasa_power_soilmoist.csv")
if (!file.exists(smap_cache)) {
  island_pts <- tibble::tribble(
    ~island,    ~lon,      ~lat,
    "O'ahu",   -157.98,   21.47,
    "Maui",    -156.34,   20.80,
    "Moloka'i",-157.02,   21.13,
    "Lāna'i", -156.92, 20.83
  )
  sm_df <- purrr::pmap_dfr(island_pts, function(island, lon, lat) {
    nasapower::get_power(community="ag", lonlat=c(lon,lat),
                         pars="GWETROOT",
                         dates=c("2026-02-01","2026-03-10"),
                         temporal_api="daily") |>
      dplyr::summarise(mean_gwetroot = mean(GWETROOT, na.rm=TRUE)) |>
      dplyr::mutate(island = island)
  })
  write_csv(sm_df, smap_cache)
} else { sm_df <- read_csv(smap_cache, show_col_types = FALSE) }
sm_df <- sm_df |>
  mutate(island = stringr::str_replace_all(island, "[ʻʼ‘’]", "'"))

# ── 5. Livestock density (USDA NASS 2022 Census of Agriculture) ───────────
livestock_cache <- here::here("data", "usda_nass_livestock_hi.csv")
if (!file.exists(livestock_cache)) {
  # Cattle & calves inventory and agricultural land by county
  # Source: USDA NASS 2022 Census of Agriculture, Table 11 (published Feb 2024)
  livestock_by_island <- tibble::tribble(
    ~island,    ~cattle_per_km2,
    "O'ahu",    58.2,   # Honolulu County: 4,467 head / 76.7 km² ag land
    "Maui",     56.1,   # Maui County (Maui+Molokai+Lanai): 35,741 / 636.8 km²
    "Moloka'i", 56.1,
    "Lāna'i", 56.1
  )
  write_csv(livestock_by_island, livestock_cache)
} else { livestock_by_island <- read_csv(livestock_cache, show_col_types = FALSE) }
livestock_by_island <- livestock_by_island |>
  mutate(island = stringr::str_replace_all(island, "[ʻʼ‘’]", "'"))

# ── Join all new predictors to chem_with_discharge ────────────────────────
island_norm <- c("oahu"="O'ahu","maui"="Maui","molokai"="Moloka'i",
                 "lanai"="Lāna'i")

add_island_col <- function(df) {
  df |> dplyr::mutate(island = island_norm[normalize_name(mokupuni)])
}

slope_j <- slope_df |> add_island_col() |> filter(!is.na(island)) |>
  group_by(ahupuaa, island) |>
  summarise(mean_slope_deg = mean(mean_slope_deg, na.rm=TRUE), .groups="drop")

ksat_j <- ksat_df |> add_island_col() |> filter(!is.na(island)) |>
  group_by(ahupuaa, island) |>
  summarise(mean_ksat_um_s = mean(mean_ksat_um_s, na.rm=TRUE), .groups="drop")

sm_j <- sm_df |> dplyr::select(island, mean_gwetroot)
lv_j <- livestock_by_island |> dplyr::select(island, cattle_per_km2)
```

```{r}
#| label: usgs-discharge-cache
#| include: false

discharge_cache <- here::here("data", "usgs_discharge_march2026.csv")
gauge_cache     <- here::here("data", "usgs_gauges_march2026.csv")

if (file.exists(discharge_cache) && file.exists(gauge_cache)) {

  discharge_per_ahu <- read_csv(discharge_cache, show_col_types = FALSE,
                                col_types = cols(site_no = col_character()))
  gauge_sf <- read_csv(gauge_cache, show_col_types = FALSE,
                       col_types = cols(site_no = col_character())) |>
    sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    sf::st_transform(32604)

} else {

  hi_sites <- dataRetrieval::whatNWISsites(
    stateCd = "HI", siteType = "ST",
    parameterCd = "00060", hasDataTypeCd = "dv"
  )

  site_batches <- split(hi_sites$site_no,
                        ceiling(seq_along(hi_sites$site_no) / 100))

  discharge_raw <- purrr::map_dfr(site_batches, function(sites) {
    tryCatch(
      dataRetrieval::readNWISdv(
        siteNumbers = sites, parameterCd = "00060",
        startDate = "2026-03-01", endDate = "2026-03-31"
      ) |> dataRetrieval::renameNWISColumns(),
      error = function(e) NULL
    )
  })

  march_discharge <- discharge_raw |>
    filter(!is.na(Flow), Flow >= 0) |>
    group_by(site_no) |>
    summarise(mean_cfs_march = mean(Flow, na.rm = TRUE),
              n_days         = n(), .groups = "drop") |>
    filter(n_days >= 15)

  gauge_sf <- hi_sites |>
    select(site_no, station_nm, dec_lat_va, dec_long_va) |>
    filter(!is.na(dec_lat_va), !is.na(dec_long_va)) |>
    inner_join(march_discharge, by = "site_no") |>
    st_as_sf(coords = c("dec_long_va", "dec_lat_va"), crs = 4326) |>
    st_transform(32604)

  gauge_ahu <- st_join(
    gauge_sf,
    ahupuaa_raw |> st_transform(32604) |> st_make_valid() |>
      select(ahupuaa, moku, mokupuni),
    join = st_within
  ) |>
    st_drop_geometry() |>
    filter(!is.na(ahupuaa))

  discharge_per_ahu <- gauge_ahu |>
    group_by(ahupuaa, moku, mokupuni) |>
    summarise(total_cfs_march = sum(mean_cfs_march),
              n_gauges        = n(), .groups = "drop")

  write_csv(discharge_per_ahu, discharge_cache)
  gauge_sf |>
    sf::st_transform(4326) |>    # save as WGS84 degrees, not UTM metres
    dplyr::mutate(longitude = sf::st_coordinates(geometry)[,1],
                  latitude  = sf::st_coordinates(geometry)[,2]) |>
    sf::st_drop_geometry() |>
    readr::write_csv(gauge_cache)
  message("Discharge cached: ", nrow(gauge_sf), " gauges, ",
          nrow(discharge_per_ahu), " ahupuaa")
}

# Assign nearest gauge discharge to each chem site (within 15 km)
chem_sf_discharge <- chem_with_stream |>
  filter(!is.na(latitude), !is.na(longitude),
         !is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

nearest_idx  <- st_nearest_feature(chem_sf_discharge, gauge_sf)
nearest_dist <- as.numeric(st_distance(chem_sf_discharge,
                                        gauge_sf[nearest_idx, ],
                                        by_element = TRUE))

chem_with_discharge <- chem_with_stream |>
  filter(!is.na(latitude), !is.na(longitude),
         !is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  mutate(
    mean_cfs_march    = if_else(nearest_dist > 15000, NA_real_,
                                gauge_sf$mean_cfs_march[nearest_idx]),
    log_discharge_cfs = log1p(mean_cfs_march)
  )

# Compute distance from each chem site to nearest NPDES outfall
# (outfalls_sf is created in additional-predictors-cache above)
chem_sf_out    <- chem_with_discharge |>
  sf::st_as_sf(coords = c("longitude","latitude"), crs = 4326) |>
  sf::st_transform(32604)

outfall_nn_idx <- sf::st_nearest_feature(chem_sf_out, outfalls_sf)
outfall_dist_v <- as.numeric(sf::st_distance(chem_sf_out,
                                              outfalls_sf[outfall_nn_idx, ],
                                              by_element = TRUE))

chem_enhanced_base <- chem_with_discharge |>
  mutate(log_dist_outfall = log1p(outfall_dist_v))
```

```{r}
#| label: random-forest-cluster
#| fig-height: 5.5
#| fig-cap: "(A) Which landscape features matter most for predicting water type? Bar length = variable importance (how much accuracy drops when that variable's values are shuffled randomly). March stream discharge (mean daily flow in cubic feet per second from the nearest USGS gauge within 15 km) is the third strongest predictor — behind stream mouth proximity and cesspool proximity, but ahead of land cover. Discharge captures how much water actually flowed during the storm, adding information beyond just knowing where streams are. (B) How often did the model get each water type right? Rows = actual water type; columns = predicted; numbers = % of each row. The model is most accurate for marine background and freshwater plume, and most confused at the storm runoff / nearshore enriched boundary."

# Build cluster assignments; join stream mouth distances from chem_with_stream
chem_rf_base <- chem_enhanced_base |>
  mutate(log_NO3NO2 = log1p(NO3NO2), log_PO4 = log1p(PO4), log_SiO2 = log1p(SiO2))

set.seed(8341)
pam_rf <- pam(
  chem_rf_base |> select(Salinity, log_NO3NO2, log_PO4, log_SiO2) |> scale(),
  k = 4, pamonce = 5
)

rf_base_joined <- chem_rf_base |>
  mutate(cluster       = factor(pam_rf$clustering),
         cluster_label = factor(cluster_labels[cluster], levels = cluster_labels)) |>
  left_join(
    ccap_summary |>
      group_by(ahupuaa, island) |>
      summarise(pct_impervious  = mean(pct_impervious,  na.rm = TRUE),
                pct_agriculture = mean(pct_agriculture, na.rm = TRUE),
                .groups = "drop"),
    by = c("ahupuaa", "island")
  ) |>
  left_join(slope_j, by = c("ahupuaa","island"), relationship="many-to-one") |>
  left_join(ksat_j,  by = c("ahupuaa","island"), relationship="many-to-one") |>
  left_join(sm_j,   by = "island") |>
  left_join(lv_j,   by = "island") |>
  mutate(cesspool_category = case_when(
    cesspool_density_per_km2 < 1.53 ~ "Low",
    cesspool_density_per_km2 < 5.43 ~ "Medium",
    TRUE ~ "High"
  ) |> factor(levels = c("Low", "Medium", "High")))

rf_df_qmd <- rf_base_joined |>
  filter(!is.na(cesspool_density_per_km2),
         !is.na(dist_to_nearest_cesspool_m),
         !is.na(dist_to_stream_mouth_m),
         !is.na(pct_impervious),
         !is.na(log_discharge_cfs)) |>
  mutate(
    across(c(mean_slope_deg, mean_ksat_um_s, log_dist_outfall,
             mean_gwetroot, cattle_per_km2, n_cess_500m, n_cess_1km),
           \(x) if_else(is.na(x), median(x, na.rm = TRUE), x))
  ) |>
  transmute(
    cluster_label,
    log_n_cess_500m       = log1p(n_cess_500m),
    log_n_cess_1km        = log1p(n_cess_1km),
    log_dist_cesspool     = log1p(dist_to_nearest_cesspool_m),
    log_dist_stream_mouth = log1p(dist_to_stream_mouth_m),
    log_discharge_cfs,
    pct_impervious,
    sqrt_ag               = sqrt(pct_agriculture),
    mean_slope_deg,
    log_ksat              = log1p(mean_ksat_um_s),
    log_dist_outfall,
    mean_gwetroot,
    log_cattle_km2        = log1p(cattle_per_km2)
  ) |>
  filter(!is.na(cluster_label), !is.na(log_n_cess_1km),
         !is.na(log_dist_cesspool), !is.na(log_dist_stream_mouth),
         !is.na(pct_impervious), !is.na(sqrt_ag))

# Add explicit interaction terms
rf_df_int <- rf_df_qmd |>
  mutate(
    cess_x_ksat  = log_n_cess_1km * log_ksat,
    imp_x_ag     = pct_impervious * sqrt_ag,
    dist_x_slope = log_dist_stream_mouth * mean_slope_deg
  ) %>%
  mutate(cluster_label = factor(cluster_label, levels = c("Marine background","Nearshore enriched","Runoff-dominated")))

set.seed(5572)
rf_fit_int <- ranger(
  cluster_label ~ .,
  data       = rf_df_int,
  num.trees  = 1000,
  mtry       = 4,
  importance = "permutation",
  seed       = 5572
)

oob_acc_int <- round((1 - rf_fit_int$prediction.error) * 100, 1)
null_acc    <- round(max(table(rf_df_int$cluster_label)) / nrow(rf_df_int) * 100, 1)

label_map <- c(
  "log_n_cess_500m"       = "Cesspools within\n500 m (count, log)",
  "log_n_cess_1km"        = "Cesspools within\n1 km (count, log)",
  "log_dist_cesspool"     = "Dist. to nearest\ncesspool (log m)",
  "log_dist_stream_mouth" = "Dist. to nearest\nstream mouth (log m)",
  "log_discharge_cfs"     = "March discharge\n(log cfs)",
  "pct_impervious"        = "% Impervious surface\n(C-CAP)",
  "sqrt_ag"               = "% Agriculture\n(C-CAP, sqrt)",
  "mean_slope_deg"        = "Watershed slope\n(SRTM, deg)",
  "log_ksat"              = "Soil permeability\n(SSURGO Ksat, log µm/s)",
  "log_dist_outfall"      = "Dist. to nearest\nNPDES outfall (log m)",
  "mean_gwetroot"         = "Pre-storm soil moisture\n(NASA POWER, Feb–Mar10)",
  "log_cattle_km2"        = "Livestock density\n(NASS 2022, log head/km²)",
  "cess_x_ksat"           = "Cesspools (1 km) \u00d7 Ksat\n(interaction)",
  "dist_x_slope"          = "Stream mouth dist. × Slope\n(interaction)"
)

imp_df <- enframe(rf_fit_int$variable.importance, name = "predictor", value = "importance") |>
  mutate(
    predictor  = recode(predictor, !!!label_map),
    importance = importance / max(importance)
  ) |>
  arrange(importance)

pImp_qmd <- ggplot(imp_df, aes(x = importance, y = reorder(predictor, importance))) +
  geom_col(fill = "#37474F", width = 0.65) +
  geom_text(aes(label = round(importance, 2)), hjust = -0.15, size = 3.2) +
  scale_x_continuous(limits = c(0, 1.15), expand = c(0, 0)) +
  labs(x = "Relative importance (permutation, normalised)", y = NULL,
       title = "A  Variable importance",
       subtitle = sprintf("OOB accuracy = %.1f%%  (null = %.1f%%)",
                          oob_acc_int, null_acc)) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

conf_df_qmd <- as.data.frame(rf_fit_int$confusion.matrix) |>
  rename(true_label = true, n = Freq) |>
  group_by(true_label) |>
  mutate(pct = n / sum(n),
         true_label = factor(true_label, levels = rev(cluster_labels)),
         predicted  = factor(predicted,  levels = cluster_labels)) |>
  ungroup()

pConf_qmd <- ggplot(conf_df_qmd, aes(x = predicted, y = true_label, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = ifelse(n > 0, sprintf("%d\n%.0f%%", n, pct * 100), "")),
            size = 2.8, lineheight = 0.9,
            color = ifelse(conf_df_qmd$pct > 0.45, "white", "gray20")) +
  scale_fill_gradient(low = "white", high = "#1565C0",
                      name = "Row\nproportion",
                      labels = scales::label_percent()) +
  scale_x_discrete(labels = \(x) str_wrap(x, 12), position = "top") +
  scale_y_discrete(labels = \(x) str_wrap(x, 12)) +
  labs(x = "Predicted cluster", y = "True cluster",
       title = "B  OOB confusion matrix") +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid   = element_blank(),
        axis.text.x  = element_text(size = 8),
        axis.text.y  = element_text(size = 8),
        plot.title   = element_text(face = "bold", size = 11))

pImp_qmd | pConf_qmd
```

```{r}
#| label: rf-pdp-interaction
#| fig-height: 4
#| fig-width: 10

# Fit probability RF (same predictors as rf_fit_int) for PDP computation
set.seed(5572)
rf_prob <- ranger(
  cluster_label ~ .,
  data        = rf_df_int,
  num.trees   = 1000,
  mtry        = 4,
  probability = TRUE,
  seed        = 5572
)

# Representative Ksat values: median of each tertile
ksat_breaks <- quantile(rf_df_int$log_ksat, c(1/3, 2/3))
ksat_rep <- c(
  "Low"    = median(rf_df_int$log_ksat[rf_df_int$log_ksat <= ksat_breaks[1]]),
  "Medium" = median(rf_df_int$log_ksat[rf_df_int$log_ksat >  ksat_breaks[1] &
                                        rf_df_int$log_ksat <= ksat_breaks[2]]),
  "High"   = median(rf_df_int$log_ksat[rf_df_int$log_ksat >  ksat_breaks[2]])
)

# Cesspools-within-1km grid across observed range
cess_seq <- seq(
  min(rf_df_int$log_n_cess_1km),
  max(rf_df_int$log_n_cess_1km),
  length.out = 60
)

# Marginal PDP: for each grid point, replace log_n_cess_1km + ksat in the full
# data and average predicted class probabilities across all rows
pdp_df <- map_dfr(names(ksat_rep), function(grp) {
  kv <- ksat_rep[[grp]]
  map_dfr(cess_seq, function(cv) {
    nd <- rf_df_int |>
      mutate(
        log_n_cess_1km = cv,
        log_ksat       = kv,
        cess_x_ksat    = cv * kv
      )
    probs <- colMeans(predict(rf_prob, data = nd)$predictions)
    tibble(
      ksat_group     = grp,
      log_n_cess_1km = cv,
      n_cess_1km     = expm1(cv),
      as_tibble_row(probs)
    )
  })
})
ksat_pal <- c(
  "Low Ksat\n(4.6 µm/s)"    = "#1565C0",
  "Medium Ksat\n(7.5 µm/s)" = "#FB8C00",
  "High Ksat\n(28.0 µm/s)"  = "#E53935"
)

pdp_df |>
  pivot_longer(
    cols      = c(`Marine background`, `Nearshore enriched`, `Runoff-dominated`),
    names_to  = "cluster",
    values_to = "probability"
  ) |>
  mutate(
    cluster    = factor(cluster, levels = c("Marine background",
                                            "Nearshore enriched",
                                            "Runoff-dominated")),
    ksat_group = factor(ksat_group, levels = c("Low", "Medium", "High"),
                        labels = c("Low Ksat\n(4.6 µm/s)",
                                   "Medium Ksat\n(7.5 µm/s)",
                                   "High Ksat\n(28.0 µm/s)"))
  ) |>
  ggplot(aes(x = log_n_cess_1km, y = probability, color = ksat_group)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.5, linewidth = 1.1) +
  scale_color_manual(values = ksat_pal, name = "Soil permeability\n(Ksat tertile)") +
  scale_x_continuous(
    name   = "Cesspools within 1 km (count, log scale)",
    breaks = log1p(c(0, 1, 5, 20, 50, 100, 200)),
    labels = c("0", "1", "5", "20", "50", "100", "200")
  ) +
  scale_y_continuous(
    name   = "Predicted probability",
    labels = scales::label_percent(accuracy = 1)
  ) +
  facet_wrap(~cluster, nrow = 1) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    legend.title     = element_text(size = 9),
    legend.text      = element_text(size = 8)
  )
```

The model correctly identified the water type for `r oob_acc_int`% of sites — `r oob_acc_int - null_acc` percentage points better than simply guessing the most common type every time. **Stream mouth proximity is the top predictor; its interaction with watershed slope ranks third** — confirming that proximity to a stream outlet matters most in steep catchments where runoff is concentrated and delivered rapidly to the coast. This stream mouth distance × slope interaction was the strongest of the three interactions included and suggests the two variables are not simply additive: a site close to a stream mouth in a flat watershed faces less nutrient exposure than an equivalently close site in a steep one. Soil permeability (Ksat) ranks second independently; permeable volcanic soils absorb rainfall before it becomes runoff, reducing coastal nutrient loading even under heavy rain. Cesspool proximity ranks fourth. Distance to the nearest NPDES outfall ranks eighth, indicating that point-source discharges contribute a small but statistically detectable independent signal beyond diffuse cesspool leaching. Pre-storm soil moisture and livestock density add the least. Together, fourteen landscape and hydrological predictors — eleven main effects and three pairwise interactions — explain coastal water type at `r oob_acc_int`% OOB accuracy across four islands and 694 sites, with each predictor sourced from a reproducible, openly licensed dataset.

---

# Landscape predictability 

If cesspools are genuinely contributing nutrients to the coast during storm events, watersheds with more cesspools per square kilometre should show higher post-storm nutrient levels. We test this by averaging nutrient concentrations per traditional watershed (ahupuaʻa) and comparing against cesspool density.

```{r}
#| label: cesspool-regression
#| fig-height: 5
#| eval: true
#| echo: false
#| include: false
#| fig-cap: "Watersheds with more cesspools per square kilometre had higher post-storm nitrogen concentrations. Each point is one ahupuaʻa (traditional watershed), with at least 3 samples; point size reflects how many samples it had. Lines show the model predictions per island from a statistical regression. The relationship is real but modest — cesspools explain about 15% of the variation in nitrogen across watersheds, with the rest explained by farmland, stream network, and other factors. The fact that the relationship holds separately within each island rules out island-level confounding."


ahupuaa_df <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE) |>
  filter(!is.na(island)) |>
  left_join(read_csv("data/ahupuaa_cache.csv",       show_col_types = FALSE), by = "sample_id") |>
  left_join(read_csv("data/cesspool_by_ahupuaa.csv", show_col_types = FALSE) |>
              select(ahupuaa, moku, cesspool_density_per_km2, n_cesspools),
            by = c("ahupuaa", "moku")) |>
  filter(!is.na(cesspool_density_per_km2), island != "Lāna'i") |>
  group_by(island, ahupuaa, moku, cesspool_density_per_km2, n_cesspools) |>
  summarise(
    mean_NO3NO2 = mean(NO3NO2, na.rm = TRUE),
    mean_PO4    = mean(PO4,    na.rm = TRUE),
    n_samples   = n(),
    .groups     = "drop"
  ) |>
  filter(n_samples >= 3, mean_NO3NO2 > 0)

fit_cp <- lm(log1p(mean_NO3NO2) ~ log1p(cesspool_density_per_km2) + island,
             data = ahupuaa_df)

pred_cp <- expand.grid(
  cesspool_density_per_km2 = exp(seq(log(0.01), log(150), length.out = 100)) - 1,
  island = c("O'ahu", "Maui", "Moloka'i")
) |>
  mutate(
    mean_NO3NO2 = expm1(predict(fit_cp, newdata = data.frame(
      cesspool_density_per_km2 = cesspool_density_per_km2, island = island)))
  )

r2_cp  <- summary(fit_cp)$r.squared
p_cp   <- pf(summary(fit_cp)$fstatistic[1], summary(fit_cp)$fstatistic[2],
             summary(fit_cp)$fstatistic[3], lower.tail = FALSE)

ggplot(ahupuaa_df,
       aes(x = cesspool_density_per_km2 + 0.01, y = mean_NO3NO2, color = island)) +
  geom_line(data = pred_cp,
            aes(x = cesspool_density_per_km2 + 0.01, y = mean_NO3NO2, color = island),
            linewidth = 0.8, alpha = 0.7) +
  geom_point(aes(size = n_samples), alpha = 0.7) +
  annotate("text", x = Inf, y = Inf,
           label = sprintf("R² = %.2f, p = %.3f", r2_cp, p_cp),
           hjust = 1.05, vjust = 1.5, size = 3, color = "gray30") +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_size_continuous(name = "Samples\nper ahupuaʻa", range = c(1.5, 6),
                        breaks = c(3, 5, 10, 20)) +
  scale_x_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(
    x        = "Cesspool density (per km², log)",
    y        = "Mean NO₃+NO₂ (µmol/L, log)",
    subtitle = "Each point = one ahupuaʻa (n ≥ 3 samples). Lines = island-level model predictions."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "right",
        plot.subtitle = element_text(size = 9, color = "gray40"))
```

The positive relationship between cesspool density and coastal NO₃+NO₂ (R² = `r round(summary(fit_cp)$r.squared, 2)`, p = `r round(p_cp, 3)`) indicates that the storm amplified nutrient delivery in proportion to pre-existing wastewater infrastructure. The modest explained variance reflects the reality that storm runoff integrates multiple nitrogen sources — agricultural fertilizers, natural soil nitrogen, and atmospheric deposition — in addition to cesspools. Phosphate, which is more specifically linked to wastewater, shows a cleaner cesspool signal (R² = 0.10, p = 0.016 without island covariate), consistent with P being a more conservative wastewater tracer than N.

## Cesspool density categories

Grouping samples by cesspool density reveals significant differences in nitrate concentrations. Samples from high-density areas show substantially elevated median and mean nitrate, with greater variability overall.

```{r}
#| label: cesspool-category
#| fig-height: 6
#| fig-width: 6
#| fig-cap: "Water samples from areas with high cesspool density (5.43–89.3 cesspools/km²) had significantly higher nitrate concentrations (mean: 8.56 µmol/L) compared to low density areas (mean: 2.40 µmol/L) or medium density areas (mean: 5.47 µmol/L). Each colored dot represents one water sample; the violin outline shows the full distribution of data within each category; the white circle indicates the mean ± 1 SE (black error bars). ANOVA confirms significant differences across density categories (F = 7.52, p < 0.001), though substantial overlap remains due to multiple nitrogen sources from the storm."

top_cess <- rf_base_joined |>
  filter(!is.na(cesspool_category), NO3NO2 > 0, !is.na(ahupuaa)) |>
  group_by(ahupuaa, island, cesspool_category) |>
  summarise(median_NN = median(NO3NO2, na.rm = TRUE), .groups = "drop") |>
  group_by(cesspool_category) |>
  slice_max(median_NN, n = 3) |>
  ungroup() |>
  mutate(label = paste0(ahupuaa, " – ", island))

p_cess_cat <- rf_base_joined %>%
  drop_na(cesspool_category, NO3NO2) %>%
  filter(NO3NO2 > 0) %>%
  mutate(cesspool_category = factor(cesspool_category, levels = c("Low","Medium","High"))) %>%
  ggplot(aes(x = cesspool_category, y = NO3NO2, color = cesspool_category, fill = cesspool_category)) +
  geom_violin(alpha = 0.15, color = NA) +
  geom_jitter(alpha = 0.25, size = 1.5, width = 0.1) +
  stat_summary(fun = function(y) 10^mean(log10(y[y > 0])), geom = "point", size = 5, 
               color = "black", shape = 21, fill = "white", stroke = 2) +
  stat_summary(
    fun.data = function(y) {
      y <- y[y > 0]
      log_y <- log10(y)
      m <- mean(log_y)
      se <- sd(log_y) / sqrt(length(log_y))
      data.frame(y = 10^m, ymin = 10^(m - se), ymax = 10^(m + se))
    },
    geom = "errorbar", width = 0.15, color = "black", linewidth = 1, alpha = 0.8
  ) +
  scale_color_manual(values = c("#43A047","#FB8C00","#E53935"), guide = "none") +
  scale_fill_manual(values = c("#43A047","#FB8C00","#E53935"), guide = "none") +
  scale_x_discrete(
    labels = c("Low\n(0.03–1.52)", "Medium\n(1.53–5.43)", "High\n(5.43–89.3)")
  ) +
  scale_y_log10() +
  labs(x = expression("Cesspool density (per km"^2*")"),
       y = expression(NO[3]+NO[2]~"("*mu*"mol/L, log)"))+
  annotate("text", x = c(1, 2, 3), y = c(2.40, 5.47, 8.56), 
           label = c("Mean: 2.40\n(0–58.6)", "Mean: 5.47\n(0–79.3)", "Mean: 8.56\n(0–266)"),
           size = 3, color = "gray20", fontface = "bold", vjust = -0.5)+
  annotate("text", x = 2, y = Inf, vjust = 1.2, hjust = 0.5,
           label = "ANOVA: F = 7.52, p < 0.001",
           size = 3.5, color = "gray40", fontface = "italic")+
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "none",
        plot.subtitle = element_text(size = 9, color = "gray40")) 
  # geom_label_repel(
  #   data          = filter(top_cess, ahupuaa != "Malaekahana"),
  #   aes(x = cesspool_category, y = median_NN, label = label),
  #   inherit.aes   = FALSE,
  #   size          = 2.5, label.size = 0.2,
  #   color         = "gray15", fill = alpha("white", 0.85),
  #   direction     = "y", nudge_x = 0.35,
  #   segment.color = "gray60", segment.size = 0.3,
  #   min.segment.length = 0.1, max.overlaps = 20
  # ) +
  # geom_label_repel(
  #   data          = filter(top_cess, ahupuaa == "Malaekahana"),
  #   aes(x = cesspool_category, y = median_NN, label = label),
  #   inherit.aes   = FALSE,
  #   size          = 2.5, label.size = 0.5,
  #   color         = "#7B1FA2", fill = "#F3E5F5",
  #   fontface      = "bold",
  #   direction     = "y", nudge_x = 0.35,
  #   segment.color = "#7B1FA2", segment.size = 0.5,
  #   min.segment.length = 0.1, max.overlaps = 20
  # )

p_cess_cat
ggsave(here("output","cesspool_NN.png"), width = 6, height = 6)
```

## Agricultural land influence

Areas with agricultural land present in their ahupuaʻa show higher nitrate concentrations than areas without agriculture, consistent with fertilizer contributions to coastal nitrogen loading during the storm.

```{r}
#| label: agriculture-category
#| fig-height: 6
#| fig-width: 5
#| fig-cap: "Nitrate concentrations rise with the percentage of agricultural land in the surrounding ahupuaʻa. Sites with no agriculture (mean: 2.98 µmol/L) and medium coverage (>0–<5%; mean: 3.90 µmol/L) are broadly similar, while high agricultural coverage (≥5%; mean: 9.34 µmol/L) shows substantially elevated nitrogen. ANOVA confirms significant differences across groups (F = 7.09, p < 0.001)."

top_ag <- rf_base_joined |>
  filter(!is.na(pct_agriculture), NO3NO2 > 0, !is.na(ahupuaa)) |>
  mutate(ag3 = case_when(
    pct_agriculture == 0   ~ "Absent",
    pct_agriculture < 0.05 ~ "Medium",
    TRUE                   ~ "High"
  ),
  ag3 = factor(ag3, levels = c("Absent", "Medium", "High"))) |>
  group_by(ahupuaa, island, ag3) |>
  summarise(median_NN = median(NO3NO2, na.rm = TRUE), .groups = "drop") |>
  group_by(ag3) |>
  slice_max(median_NN, n = 3) |>
  ungroup() |>
  mutate(label = paste0(ahupuaa, " – ", island))

p_ag <- rf_base_joined |>
  filter(!is.na(pct_agriculture), NO3NO2 > 0) |>
  mutate(ag3 = case_when(
    pct_agriculture == 0   ~ "Absent",
    pct_agriculture < 0.05 ~ "Medium",
    TRUE                   ~ "High"
  ),
  ag3 = factor(ag3, levels = c("Absent", "Medium", "High"))) |>
  ggplot(aes(x = ag3, y = NO3NO2, color = ag3, fill = ag3)) +
  geom_violin(alpha = 0.15, color = NA) +
  geom_jitter(alpha = 0.25, size = 1.5, width = 0.1) +
  stat_summary(fun = "mean", geom = "point", size = 5,
               color = "black", shape = 21, fill = "white", stroke = 2) +
  stat_summary(fun.data = "mean_se", geom = "errorbar",
               width = 0.15, color = "black", linewidth = 1, alpha = 0.8) +
  scale_color_manual(
    values = c("Absent" = "#43A047", "Medium" = "#FB8C00", "High" = "#E53935"),
    guide  = "none"
  ) +
  scale_fill_manual(
    values = c("Absent" = "#43A047", "Medium" = "#FB8C00", "High" = "#E53935"),
    guide  = "none"
  ) +
  scale_x_discrete(labels = c(
    "Absent" = "Absent\n0%",
    "Medium" = "Medium\n>0–<5%",
    "High"   = "High\n≥5%"
  )) +
  scale_y_log10() +
  labs(x = expression("Agricultural land (% of ahupua"*"'a"*")"),
       y = expression(NO[3]+NO[2]~"("*mu*"mol/L, log)")) +
  annotate("text", x = c(1, 2, 3), y = c(2.98, 3.90, 9.34),
           label = c("Mean: 2.98\n(0.06–61.2)", "Mean: 3.90\n(0.06–137)",
                     "Mean: 9.34\n(0.06–266)"),
           size = 3, color = "gray20", fontface = "bold", vjust = -0.5) +
  annotate("text", x = 2, y = Inf, vjust = 1.2, hjust = 0.5,
           label = "ANOVA: F = 7.09, p < 0.001",
           size = 3.5, color = "gray40", fontface = "italic") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "none",
        plot.subtitle = element_text(size = 9, color = "gray40")) +
  geom_label_repel(
    data          = filter(top_ag, ahupuaa != "Malaekahana"),
    aes(x = ag3, y = median_NN, label = label),
    inherit.aes   = FALSE,
    size          = 2.5, label.size = 0.2,
    color         = "gray15", fill = alpha("white", 0.85),
    direction     = "y", nudge_x = 0.35,
    segment.color = "gray60", segment.size = 0.3,
    min.segment.length = 0.1, max.overlaps = 20
  ) +
  geom_label_repel(
    data          = filter(top_ag, ahupuaa == "Malaekahana"),
    aes(x = ag3, y = median_NN, label = label),
    inherit.aes   = FALSE,
    size          = 2.5, label.size = 0.5,
    color         = "#7B1FA2", fill = "#F3E5F5",
    fontface      = "bold",
    direction     = "y", nudge_x = 0.35,
    segment.color = "#7B1FA2", segment.size = 0.5,
    min.segment.length = 0.1, max.overlaps = 20
  )

p_ag / p_cess_cat
ggsave(here("output","cesspool_Ag_NN.png"), width = 8, height = 6)
```

### Agricultural land cover by ahupuaʻa

```{r}
#| label: ag-land-map
#| fig-height: 7
#| fig-width: 7
#| fig-cap: "Agricultural land cover (% of ahupuaʻa area) across the four study islands. Dark blue outlines indicate ahupuaʻa with community science water quality samples collected during the March 2026 Kona Low. Central Maui and the western end of Molokaʻi carry the highest agricultural footprints — the same areas that appear as nitrate hotspots in the kriging maps. Source: NOAA Coastal Change Analysis Program (C-CAP) High-Resolution Land Cover, Hawaiʻi 2010–2011 (NOAA Office for Coastal Management, 2014–2015; coast.noaa.gov/ccapftp)."

island_extents <- list(
  "O\u02BBahu"    = c(xmin = -158.32, xmax = -157.60, ymin = 21.22, ymax = 21.75),
  "Maui"          = c(xmin = -156.72, xmax = -155.96, ymin = 20.54, ymax = 21.05),
  "Moloka\u02BBi" = c(xmin = -157.35, xmax = -156.70, ymin = 21.04, ymax = 21.25),
  "L\u0101na\u02BBi" = c(xmin = -157.07, xmax = -156.81, ymin = 20.73, ymax = 20.93)
)

ccap_ag <- read_csv("data/ccap_ahupuaa_summary.csv", show_col_types = FALSE) |>
  select(ahupuaa, island, pct_agriculture) |>
  # Normalize straight apostrophes to ʻokina to match ahupuaa_raw$mokupuni
  mutate(island = str_replace_all(island, "[\u0027\u2018\u2019\u02BC]", "\u02BB")) |>
  distinct(ahupuaa, island, .keep_all = TRUE)

ag_sf_map <- ahupuaa_raw |>
  filter(mokupuni %in% names(island_extents)) |>
  left_join(ccap_ag, by = c("ahupuaa", "mokupuni" = "island")) |>
  mutate(ag_pct = pct_agriculture * 100,
         island = mokupuni)

cs_ahupuaa <- chem |>
  filter(!is.na(ahupuaa), !is.na(island)) |>
  mutate(island = str_replace_all(island, "[\u0027\u2018\u2019\u02BC]", "\u02BB")) |>
  distinct(ahupuaa, island)

ag_sf_sampled <- ag_sf_map |>
  semi_join(cs_ahupuaa, by = c("ahupuaa", "island"))

plot_island_ag <- function(isl) {
  ext <- island_extents[[isl]]
  ggplot(filter(ag_sf_map, island == isl)) +
    geom_sf(aes(fill = ag_pct), color = "white", linewidth = 0.15) +
    geom_sf(data = filter(ag_sf_sampled, island == isl),
            fill = NA, color = "#1A237E", linewidth = 0.55) +
    scale_fill_gradientn(
      colours  = c("#f7fcf0", "#74c476", "#238b45", "#8B4513"),
      na.value = "gray88",
      name     = "Agricultural\nland (%)",
      limits   = c(0, 70), oob = scales::squish,
      breaks   = c(0, 10, 25, 50),
      labels   = c("0", "10", "25", "\u226550")
    ) +
    coord_sf(xlim = ext[c("xmin","xmax")], ylim = ext[c("ymin","ymax")]) +
    labs(title = isl) +
    theme_void(base_size = 10) +
    theme(plot.title      = element_text(face = "bold", size = 10, hjust = 0.5),
          legend.position = "none")
}

legend_ag <- cowplot::get_legend(
  plot_island_ag("Maui") +
    # Add a dummy point to capture the outline in the legend
    geom_sf(data = filter(ag_sf_sampled, island == "Maui")[1, ],
            aes(color = "Sampled ahupua\u02BBa"),
            fill = NA, linewidth = 0.55, show.legend = TRUE) +
    scale_color_manual(values = c("Sampled ahupua\u02BBa" = "#1A237E"),
                       name = NULL) +
    guides(fill  = guide_colorbar(order = 1,
                                  barheight = unit(8, "lines"),
                                  barwidth  = unit(0.8, "lines"),
                                  title.position = "top"),
           color = guide_legend(order = 2,
                                override.aes = list(linewidth = 1.2,
                                                    fill = NA))) +
    theme(legend.position   = "right",
          legend.key.height = unit(8, "lines"),
          legend.title      = element_text(size = 9),
          legend.text       = element_text(size = 8),
          legend.box        = "vertical")
)

ag_maps <- (plot_island_ag("O\u02BBahu")    | plot_island_ag("Maui")) /
           (plot_island_ag("Moloka\u02BBi") | plot_island_ag("L\u0101na\u02BBi"))

cowplot::plot_grid(ag_maps, legend_ag, ncol = 2, rel_widths = c(1, 0.15))
ggsave(here("output","ag_land.png"), width = 12, height = 8 )
```

### Cesspool density by ahupuaʻa

```{r}
#| label: cesspool-density-map
#| fig-height: 7
#| fig-width: 7
#| fig-cap: "Cesspool density (cesspools per km² of ahupuaʻa area) across the four study islands. Dark blue outlines indicate ahupuaʻa with community science water quality samples collected during the March 2026 Kona Low. Lānaʻi has no cesspool layer in the Hawaiʻi GIS system and is shown in grey throughout. Colour is capped at 50 cesspools/km² (squished). Source: Hawaiʻi Statewide GIS Program — HCPT Cesspools (Layer 32; geodata.hawaii.gov), supplemented with OSDS Class-I systems for Molokaʻi (Layer 23)."

cess_data <- read_csv("data/cesspool_by_ahupuaa.csv", show_col_types = FALSE) |>
  select(ahupuaa, island, cesspool_density_per_km2) |>
  distinct(ahupuaa, island, .keep_all = TRUE)

cess_sf_map <- ahupuaa_raw |>
  filter(mokupuni %in% names(island_extents)) |>
  left_join(cess_data, by = c("ahupuaa", "mokupuni" = "island")) |>
  mutate(island = mokupuni)

cess_sf_sampled <- cess_sf_map |>
  semi_join(cs_ahupuaa, by = c("ahupuaa", "island"))

plot_island_cess <- function(isl) {
  ext <- island_extents[[isl]]
  ggplot(filter(cess_sf_map, island == isl)) +
    geom_sf(aes(fill = cesspool_density_per_km2), color = "white", linewidth = 0.15) +
    geom_sf(data = filter(cess_sf_sampled, island == isl),
            fill = NA, color = "#1A237E", linewidth = 0.55) +
    scale_fill_gradientn(
      colours  = c("#fff5eb", "#fdd0a2", "#fd8d3c", "#d94801", "#7F0000"),
      na.value = "gray88",
      name     = "Cesspools\nper km\u00b2",
      limits   = c(0, 50), oob = scales::squish,
      breaks   = c(0, 10, 25, 50),
      labels   = c("0", "10", "25", "\u226550")
    ) +
    coord_sf(xlim = ext[c("xmin","xmax")], ylim = ext[c("ymin","ymax")]) +
    labs(title = isl) +
    theme_void(base_size = 10) +
    theme(plot.title      = element_text(face = "bold", size = 10, hjust = 0.5),
          legend.position = "none")
}

legend_cess <- cowplot::get_legend(
  plot_island_cess("Maui") +
    geom_sf(data = filter(cess_sf_sampled, island == "Maui")[1, ],
            aes(color = "Sampled ahupua\u02BBa"),
            fill = NA, linewidth = 0.55, show.legend = TRUE) +
    scale_color_manual(values = c("Sampled ahupua\u02BBa" = "#1A237E"), name = NULL) +
    guides(fill  = guide_colorbar(order = 1,
                                  barheight = unit(8, "lines"),
                                  barwidth  = unit(0.8, "lines"),
                                  title.position = "top"),
           color = guide_legend(order = 2,
                                override.aes = list(linewidth = 1.2, fill = NA))) +
    theme(legend.position   = "right",
          legend.key.height = unit(8, "lines"),
          legend.title      = element_text(size = 9),
          legend.text       = element_text(size = 8),
          legend.box        = "vertical")
)

cess_maps <- (plot_island_cess("O\u02BBahu")    | plot_island_cess("Maui")) /
             (plot_island_cess("Moloka\u02BBi") | plot_island_cess("L\u0101na\u02BBi"))

cowplot::plot_grid(cess_maps, legend_cess, ncol = 2, rel_widths = c(1, 0.15))
ggsave(here("output","cesspool_all.png"), width = 12, height = 8 )
```

### Soil permeability (SSURGO Ksat) by ahupuaʻa

```{r}
#| label: ksat-map
#| fig-height: 7
#| fig-width: 7
#| fig-cap: "Saturated hydraulic conductivity (Ksat) by ahupuaʻa — a measure of how quickly water (and dissolved nutrients or cesspool effluent) moves through the soil. Higher values indicate more permeable soils where groundwater recharge is rapid and contaminant filtration is minimal. Dark blue outlines indicate ahupuaʻa with community science water quality samples. Colour is capped at 100 µm/s. Source: USDA-NRCS SSURGO (Soil Survey Geographic Database); Ksat values are area-weighted means of representative horizon values (ksat_r) across all soil map units within each ahupuaʻa."

ksat_data <- read_csv("data/ahupuaa_ksat.csv", show_col_types = FALSE) |>
  select(ahupuaa, mokupuni, mean_ksat_um_s) |>
  distinct(ahupuaa, mokupuni, .keep_all = TRUE)

ksat_sf_map <- ahupuaa_raw |>
  filter(mokupuni %in% names(island_extents)) |>
  left_join(ksat_data, by = c("ahupuaa", "mokupuni")) |>
  mutate(island = mokupuni)

ksat_sf_sampled <- ksat_sf_map |>
  semi_join(cs_ahupuaa, by = c("ahupuaa", "island"))

plot_island_ksat <- function(isl) {
  ext <- island_extents[[isl]]
  ggplot(filter(ksat_sf_map, island == isl)) +
    geom_sf(aes(fill = mean_ksat_um_s), color = "white", linewidth = 0.15) +
    geom_sf(data = filter(ksat_sf_sampled, island == isl),
            fill = NA, color = "#1A237E", linewidth = 0.55) +
    scale_fill_gradientn(
      colours  = c("#F0F4FF", "#90CAF9", "#1565C0", "#0D47A1", "#01002E"),
      na.value = "gray88",
      name     = "Ksat\n(\u03bcm/s)",
      limits   = c(0, 100), oob = scales::squish,
      breaks   = c(0, 25, 50, 100),
      labels   = c("0", "25", "50", "\u2265100")
    ) +
    coord_sf(xlim = ext[c("xmin","xmax")], ylim = ext[c("ymin","ymax")]) +
    labs(title = isl) +
    theme_void(base_size = 10) +
    theme(plot.title      = element_text(face = "bold", size = 10, hjust = 0.5),
          legend.position = "none")
}

legend_ksat <- cowplot::get_legend(
  plot_island_ksat("Maui") +
    geom_sf(data = filter(ksat_sf_sampled, island == "Maui")[1, ],
            aes(color = "Sampled ahupua\u02BBa"),
            fill = NA, linewidth = 0.55, show.legend = TRUE) +
    scale_color_manual(values = c("Sampled ahupua\u02BBa" = "#1A237E"), name = NULL) +
    guides(fill  = guide_colorbar(order = 1,
                                  barheight = unit(8, "lines"),
                                  barwidth  = unit(0.8, "lines"),
                                  title.position = "top"),
           color = guide_legend(order = 2,
                                override.aes = list(linewidth = 1.2, fill = NA))) +
    theme(legend.position   = "right",
          legend.key.height = unit(8, "lines"),
          legend.title      = element_text(size = 9),
          legend.text       = element_text(size = 8),
          legend.box        = "vertical")
)

ksat_maps <- (plot_island_ksat("O\u02BBahu")    | plot_island_ksat("Maui")) /
             (plot_island_ksat("Moloka\u02BBi") | plot_island_ksat("L\u0101na\u02BBi"))

cowplot::plot_grid(ksat_maps, legend_ksat, ncol = 2, rel_widths = c(1, 0.15))
ggsave(here("output","Ksat_all.png"), width = 12, height = 8 )
```

### Composite cesspool conversion priority

```{r}
#| label: cesspool-priority-map
#| fig-height: 7
#| fig-width: 7.5
#| fig-cap: "Composite cesspool conversion priority by ahupuaʻa. Background colour (yellow–dark red) is a 0–100 priority score: the mean of three percentile-ranked variables — cesspool density (HCPT), soil permeability (SSURGO Ksat), and observed median NO₃+NO₂. For ahupuaʻa without water quality samples the score is the mean of cesspool density and Ksat ranks only. Overlaid circles show individual community science water samples coloured by NO₃+NO₂ concentration (log scale, plasma palette); high-nitrate samples sitting in dark-red polygons confirm that the structural risk is being expressed as observed enrichment. Dark blue outlines show sampled ahupuaʻa. Critical-tier ahupuaʻa (score ≥ 90) are concentrated on Maui's south coast and Oʻahu's leeward corridor — areas combining high cesspool density, highly permeable volcanic soils, and measured nitrate enrichment."

priority_base <- cess_sf_map |>
  st_drop_geometry() |>
  distinct(ahupuaa, island, .keep_all = TRUE) |>
  select(ahupuaa, island, cesspool_density_per_km2) |>
  left_join(
    ksat_sf_map |> st_drop_geometry() |>
      distinct(ahupuaa, island, .keep_all = TRUE) |>
      select(ahupuaa, island, mean_ksat_um_s),
    by = c("ahupuaa", "island")
  ) |>
  left_join(
    chem |> filter(!is.na(NO3NO2), !is.na(ahupuaa)) |>
      mutate(island = str_replace_all(island, "[\u0027\u2018\u2019\u02BC]", "\u02BB")) |>
      group_by(ahupuaa, island) |>
      summarise(median_NN = median(NO3NO2, na.rm = TRUE), .groups = "drop"),
    by = c("ahupuaa", "island")
  ) |>
  mutate(
    cess_rank = percent_rank(replace_na(cesspool_density_per_km2, 0)),
    ksat_rank = percent_rank(replace_na(mean_ksat_um_s, 0)),
    nn_rank   = percent_rank(replace_na(median_NN, 0)),
    priority  = case_when(
      !is.na(median_NN) ~ (cess_rank + ksat_rank + nn_rank) / 3,
      TRUE              ~ (cess_rank + ksat_rank) / 2
    ) * 100
  )

priority_sf <- ahupuaa_raw |>
  filter(mokupuni %in% names(island_extents)) |>
  distinct(ahupuaa, mokupuni, .keep_all = TRUE) |>
  left_join(priority_base, by = c("ahupuaa", "mokupuni" = "island")) |>
  mutate(island = mokupuni)

chem_pts <- chem |>
  filter(!is.na(latitude), !is.na(longitude), !is.na(NO3NO2)) |>
  mutate(island = recode(island,
    "O'ahu"    = "O\u02BBahu",
    "Moloka'i" = "Moloka\u02BBi",
    "L\u0101na'i" = "L\u0101na\u02BBi"
  )) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

plot_island_priority <- function(isl) {
  ext  <- island_extents[[isl]]
  poly <- priority_sf |> filter(island == isl)
  pts  <- chem_pts |> filter(island == isl)
  ggplot(poly) +
    geom_sf(aes(fill = priority), color = "white", linewidth = 0.15) +
    geom_sf(data = filter(cess_sf_sampled, island == isl),
            fill = NA, color = "#1A237E", linewidth = 0.45, alpha = 0.6) +
    geom_sf(data = pts, aes(color = log1p(NO3NO2)),
            size = 1.8, shape = 21, fill = NA, stroke = 0.8, alpha = 0.85) +
    scale_fill_gradientn(
      colours  = c("#FFF9C4", "#FFCC02", "#F57C00", "#C62828", "#4A0000"),
      na.value = "gray88",
      name     = "Priority\nscore",
      limits   = c(0, 100), breaks = c(0, 25, 50, 75, 100)
    ) +
    scale_color_viridis_c(
      option = "plasma",
      name   = expression(NO[3]+NO[2]~"(log)"),
      limits = c(0, log1p(10)), oob = scales::squish
    ) +
    coord_sf(xlim = ext[c("xmin","xmax")], ylim = ext[c("ymin","ymax")]) +
    labs(title = isl) +
    theme_void(base_size = 10) +
    theme(plot.title      = element_text(face = "bold", size = 10, hjust = 0.5),
          legend.position = "none")
}

legend_priority <- cowplot::get_legend(
  plot_island_priority("Maui") +
    guides(
      fill  = guide_colorbar(order = 1, barheight = unit(6, "lines"),
                             barwidth = unit(0.8, "lines"), title.position = "top"),
      color = guide_colorbar(order = 2, barheight = unit(6, "lines"),
                             barwidth = unit(0.8, "lines"), title.position = "top")
    ) +
    theme(legend.position = "right", legend.box = "vertical",
          legend.title    = element_text(size = 8.5),
          legend.text     = element_text(size = 7.5))
)

priority_maps <-
  (plot_island_priority("O\u02BBahu")    | plot_island_priority("Maui")) /
  (plot_island_priority("Moloka\u02BBi") | plot_island_priority("L\u0101na\u02BBi"))

cowplot::plot_grid(priority_maps, legend_priority, ncol = 2, rel_widths = c(1, 0.18))
ggsave(here("output","cesspool_priority.png"), width = 12, height = 8 )
```

### Agricultural land × impervious surface interaction

```{r}
#| label: ag-imperv-tile
#| fig-height: 5.5
#| fig-width: 7
#| fig-cap: "Median NO₃+NO₂ concentration across combinations of agricultural land cover (columns) and impervious surface cover (rows). Empty cells reflect combinations that are essentially absent from the data — high agricultural cover and high impervious cover rarely co-occur (ρ = −0.33). The peak concentration (2.91 µmol/L, n = 102) occurs in High ag × Medium impervious ahupuaʻa, likely representing agricultural land with farm access roads and packing infrastructure that concentrate and route runoff coastward. Conversely, Medium ag × High impervious (0.67 µmol/L, n = 152) is among the lowest — dense suburban development with modest agriculture produces less nitrate enrichment, possibly because the agricultural footprint is small (residential gardens) and impervious surfaces intercept and dilute stormwater before it reaches sampling sites."

tile_ag_imperv <- rf_base_joined |>
  filter(!is.na(NO3NO2), NO3NO2 > 0,
         !is.na(pct_agriculture), !is.na(pct_impervious)) |>
  mutate(
    ag_bin = cut(pct_agriculture,
                 breaks = c(-Inf, 0, 0.05, Inf),
                 labels = c("Absent\n0%", "Medium\n>0\u2013<5%", "High\n\u22655%")),
    imperv_bin = cut(pct_impervious,
                     breaks = c(0, 0.05, 0.15, 0.35, Inf),
                     labels = c("Low\n(<5%)", "Medium\n(5\u201315%)",
                                "High\n(15\u201335%)", "Very high\n(>35%)"),
                     include.lowest = TRUE)
  ) |>
  group_by(ag_bin, imperv_bin) |>
  summarise(median_NN = median(NO3NO2, na.rm = TRUE),
            n         = n(), .groups = "drop")

ggplot(tile_ag_imperv, aes(x = ag_bin, y = imperv_bin, fill = median_NN)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f\nn = %d", median_NN, n)),
            size = 3, lineheight = 1.2, color = "gray15") +
  scale_fill_gradientn(
    colours  = c("#FFFDE7", "#FFD54F", "#F57F17", "#B71C1C"),
    na.value = "gray92",
    name     = expression(NO[3]+NO[2]~"median\n("*mu*"mol/L)"),
    limits   = c(0, 3), oob = scales::squish,
    breaks   = c(0, 1, 2, 3),
    labels   = c("0", "1", "2", "\u22653")
  ) +
  scale_y_discrete(limits = rev) +
  labs(
    x        = "Agricultural land cover",
    y        = "Impervious surface cover",
    title    = expression(NO[3]+NO[2]~"by agricultural land \u00d7 impervious surface"),
    subtitle = "Median concentration per cell  \u00b7  n = water samples  \u00b7  empty cells = combination absent in data"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid        = element_blank(),
    axis.text.x       = element_text(size = 9, lineheight = 1.2),
    axis.text.y       = element_text(size = 9, lineheight = 1.2, hjust = 1),
    plot.title        = element_text(size = 11),
    plot.subtitle     = element_text(size = 8.5, color = "gray40"),
    legend.title      = element_text(size = 9),
    legend.key.height = unit(2, "lines")
  )
ggsave(here("output","Ag_imperv.png"), width = 8, height = 5)
```

### Cesspool density × soil permeability interaction

```{r}
#| label: ksat-cess-tile
#| fig-height: 5.5
#| fig-width: 7
#| fig-cap: "Median NO₃+NO₂ concentration across combinations of cesspool density (columns) and soil permeability (rows). Each cell shows the median concentration and number of water samples. Soil permeability acts as an amplifier: the three low-to-high Ksat rows show broadly similar, low concentrations (0.6–1.6 µmol/L) regardless of cesspool density, while the very high permeability row (>50 µm/s) shows dramatically elevated concentrations (5–12 µmol/L) across all cesspool density categories. This suggests that it is the combination of even moderate cesspool presence with highly permeable volcanic soils — which allow rapid unfiltered transport of effluent to coastal groundwater — that most strongly drives nearshore nitrate enrichment."

tile_df <- rf_base_joined |>
  filter(!is.na(NO3NO2), NO3NO2 > 0,
         !is.na(cesspool_density_per_km2),
         !is.na(mean_ksat_um_s)) |>
  mutate(
    cess_bin = cut(cesspool_density_per_km2,
                   breaks = c(0, 1.52, 5.43, 89.3),
                   labels = c("Low\n(0\u20131.5/km\u00b2)",
                              "Medium\n(1.5\u20135.4/km\u00b2)",
                              "High\n(>5.4/km\u00b2)"),
                   include.lowest = TRUE),
    ksat_bin = cut(mean_ksat_um_s,
                   breaks = c(0, 6, 14, 50, Inf),
                   labels = c("Low\n(<6 \u03bcm/s)",
                              "Medium\n(6\u201314 \u03bcm/s)",
                              "High\n(14\u201350 \u03bcm/s)",
                              "Very high\n(>50 \u03bcm/s)"),
                   include.lowest = TRUE)
  ) |>
  filter(!is.na(cess_bin)) |>
  group_by(cess_bin, ksat_bin) |>
  summarise(median_NN = median(NO3NO2, na.rm = TRUE),
            n         = n(), .groups = "drop")

ggplot(tile_df, aes(x = cess_bin, y = ksat_bin, fill = median_NN)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.1f\nn = %d", median_NN, n)),
            size = 3, lineheight = 1.2, color = "gray15") +
  scale_fill_gradientn(
    colours  = c("#FFFDE7", "#FFD54F", "#F57F17", "#B71C1C"),
    name     = expression(NO[3]+NO[2]~"median\n("*mu*"mol/L)"),
    limits   = c(0, 12), oob = scales::squish,
    breaks   = c(0, 3, 6, 9, 12),
    labels   = c("0", "3", "6", "9", "\u226512")
  ) +
  scale_y_discrete(limits = rev) +
  labs(
    x        = "Cesspool density",
    y        = "Soil permeability (Ksat)",
    title    = expression(NO[3]+NO[2]~"concentration by cesspool density \u00d7 soil permeability"),
    subtitle = "Median concentration per cell \u00b7 n = number of water samples"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid    = element_blank(),
    axis.text.x   = element_text(size = 9, lineheight = 1.2),
    axis.text.y   = element_text(size = 9, lineheight = 1.2, hjust = 1),
    plot.title    = element_text(size = 11),
    plot.subtitle = element_text(size = 8.5, color = "gray40"),
    legend.title  = element_text(size = 9),
    legend.key.height = unit(2, "lines")
  )
ggsave(here("output","cesspool_ksat.png"), width = 8, height = 5)
```

### Landscape predictor separation by water type

```{r}
#| label: top5-cluster-violins
#| fig-height: 5
#| fig-width: 16
#| fig-cap: "Distribution of the five highest-importance landscape predictors across the three water type clusters. Horizontal bars show cluster medians. Each predictor separates at least two water types, explaining why the random forest ranks them highly. Runoff-dominated and nearshore-enriched sites sit closer to stream mouths and cesspools, on more permeable soils, and in more agricultural watersheds than marine background sites."

cluster_pal <- c(
  "Marine background"  = "#1565C0",
  "Nearshore enriched" = "#2E7D32",
  "Runoff-dominated"   = "#E65100"
)

violin_clust <- function(var, ylab, ybreaks = waiver(), ylabels = waiver()) {
  ggplot(rf_df_int, aes(x = cluster_label, y = .data[[var]])) +
    geom_violin(aes(fill = cluster_label), alpha = 0.15, color = NA) +
    geom_jitter(aes(color = cluster_label), alpha = 0.25, width = 0.12, size = 0.9) +
    stat_summary(fun = median, geom = "crossbar", width = 0.4, linewidth = 0.6, color = "gray15") +
    scale_fill_manual(values  = cluster_pal, guide = "none") +
    scale_color_manual(values = cluster_pal, guide = "none") +
    scale_x_discrete(labels = c("Marine background" = "Marine\nbkgd.",
                                "Nearshore enriched" = "Nearshore\nenriched",
                                "Runoff-dominated"   = "Runoff-\ndominated")) +
    scale_y_continuous(name = ylab, breaks = ybreaks, labels = ylabels) +
    labs(x = NULL) +
    theme_minimal(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          axis.text.x = element_text(size = 8.5, lineheight = 1.15))
}

p1 <- violin_clust(
  "log_dist_stream_mouth", "Stream mouth distance",
  ybreaks = log1p(c(0, 100, 500, 2000, 5000, 10000)),
  ylabels = c("0", "100 m", "500 m", "2 km", "5 km", "10 km")
)

p2 <- violin_clust(
  "dist_x_slope", "Stream mouth dist. × Slope\n(interaction)"
)

p3 <- violin_clust(
  "log_ksat", "Soil permeability (Ksat)",
  ybreaks = log1p(c(0, 2, 5, 14, 50, 100)),
  ylabels = c("0", "2", "5", "14", "50", "100 µm/s")
)

p4 <- violin_clust(
  "log_dist_cesspool", "Dist. to nearest cesspool",
  ybreaks = log1p(c(0, 100, 500, 2000, 5000)),
  ylabels = c("0", "100 m", "500 m", "2 km", "5 km")
)

p5 <- violin_clust(
  "sqrt_ag", "% Agriculture (C-CAP)",
  ybreaks = sqrt(c(0, 0.05, 0.1, 0.2, 0.4, 0.68)),
  ylabels = c("0%", "5%", "10%", "20%", "40%", "68%")
)

p1 | p2 | p3 | p4 | p5
```

---

*Cite these data:* Silbiger, N., Kealoha, A., & Kahanamoku, S. (2026). *What's In Your Water? — Hawaiʻi Coastal Water Quality Samples, March 2026 Kona Low.* Zenodo. DOI: 10.5281/zenodo.19410137
