---
title: "Nutrient hotspot maps — post-Kona low storm"
subtitle: "Spatial interpolation of NO₃+NO₂ concentrations across Hawaiʻi, March–April 2026"
author: "Silbiger, N., Kealoha, A., & Kahanamoku, S."
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: flatly
    code-fold: true
    fig-width: 9
    fig-height: 6.5
execute:
  warning: false
  message: false
---


```{r}
#| label: load-packages

# Core
library(tidyverse)
library(scales)
library(knitr)
library(patchwork)
library(here)

# Spatial
library(sf)
library(gstat)
library(stars)
library(terra)
library(osmdata)
library(maptiles)
library(tidyterra)

# Raster extraction
library(exactextractr)

# Visualisation helpers
library(cowplot)
library(ggrepel)
library(ggdendro)

# Clustering & spatial statistics
library(cluster)
library(spdep)
library(zoo)

# Machine learning & partial dependence
library(ranger)
library(pdp)

# Note: dataRetrieval and nhdplusTools are used only in the eval:false
# roadmap chunks at the end of this document. Install separately if needed.
```

# Introduction

When a Kona Low storm hits Hawaiʻi, heavy rain washes fertilizer, cesspool nutrients, and volcanic soil off the land and into the ocean. Does that change coastal water quality in a meaningful way — and if so, where, how much, and can we predict it from a map?

This report uses 694 water samples collected across four islands immediately after the March 2026 Kona Low to answer those questions. Interpolated maps show where nutrient concentrations were highest. A grouping analysis (PAM clustering) sorts sites into four distinct water types — from clean open-ocean water to freshwater plumes at stream mouths — and measures how different their chemistry is. A random forest model then asks whether those water types could have been predicted from landscape features alone: how close is a site to a stream mouth? To a cesspool? How much of the surrounding land is paved or farmed? How much water actually flowed through the nearest stream gauge in March? The model correctly classifies 67% of sites (vs 36% for a random guess), with stream mouth proximity, cesspool distance, and actual March discharge as the three strongest predictors.

# Data overview

```{r}
#| label: load-data
chem <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE) |>
  filter(!is.na(island)) |>
  # Normalise apostrophe variants (ʻokina U+02BB, smart quotes U+2018/9) to
  # plain ASCII apostrophe so island-name joins are consistent throughout
  mutate(island = stringr::str_replace_all(island, "[\u02BB\u02BC\u2018\u2019]", "'")) |>
  # Ahupuaʻa assignment (spatial join pre-computed in 01_load_clean_data.R)
  left_join(
    read_csv("data/ahupuaa_cache.csv", show_col_types = FALSE),
    by = "sample_id"
  ) |>
  # Distance to nearest cesspool per sample
  left_join(
    read_csv("data/sample_cesspool_distances.csv", show_col_types = FALSE),
    by = "sample_id"
  ) |>
  # Cesspool density per ahupuaʻa (join on ahupuaa + moku to avoid duplicate
  # ahupuaʻa names appearing in multiple moku / islands)
  left_join(
    read_csv("data/cesspool_by_ahupuaa.csv", show_col_types = FALSE) |>
      select(ahupuaa, moku, cesspool_density_per_km2, n_cesspools),
    by = c("ahupuaa", "moku")
  )
```

`r nrow(chem)` samples collected across `r n_distinct(chem$island)` islands between `r format(min(chem$collected_hst, na.rm = TRUE), "%B %d, %Y")` and `r format(max(chem$collected_hst, na.rm = TRUE), "%B %d, %Y")`. Cesspool density is unavailable for Lānaʻi (no public data) and `r chem |> filter(island != "L\u0101na'i", is.na(cesspool_density_per_km2)) |> nrow()` samples on other islands where the ahupuaʻa is not in the cesspool dataset.

```{r}
#| label: data-summary-table
chem |>
  group_by(island) |>
  summarise(
    n_samples       = n(),
    n_with_no3no2   = sum(!is.na(NO3NO2)),
    median_no3no2   = median(NO3NO2, na.rm = TRUE),
    max_no3no2      = max(NO3NO2, na.rm = TRUE),
    date_min        = format(min(collected_hst, na.rm = TRUE), "%b %d"),
    date_max        = format(max(collected_hst, na.rm = TRUE), "%b %d"),
    .groups = "drop"
  ) |>
  kable(
    col.names = c("Island", "Total samples", "Samples w/ NO₃+NO₂",
                  "Median (µmol/L)", "Max (µmol/L)", "First collected", "Last collected"),
    digits = 2,
    caption = "Summary of sample counts and NO₃+NO₂ concentrations by island."
  )
```

```{r}
#| label: load-spatial
# Load ahupuaʻa boundaries as the island land mask. These are more accurate
# than Natural Earth (which clips East Maui and misrepresents the isthmus).
ahupuaa_raw <- st_read("data/ahupuaa_boundaries.geojson", quiet = TRUE)
```

# Nutrient hotspot maps

Each map shows a kriged surface of NO₃+NO₂ (µmol/L) interpolated from coastal water samples collected after the storm. White circles mark sample collection points. The fill scale is logarithmic (log₁p) to show the full range of values without the highest concentrations washing out the spatial pattern. Land is masked in gray.

```{r}
#| label: helpers

normalize_name <- function(x) {
  x |>
    stringr::str_replace_all("[\u02BB\u2018\u2019']", "") |>  # strip okina/smart quotes
    stringr::str_replace_all("\u0101", "a") |>                 # ā → a
    stringr::str_to_lower() |>
    stringr::str_trim()
}

# Shared island colour palette — ASCII apostrophes throughout for consistent matching
island_pal <- c(
  "O'ahu"    = "#1565C0",
  "Maui"     = "#2E7D32",
  "Moloka'i" = "#E65100",
  "Lāna'i"   = "#6A1B9A"
)

# Canonical recode: plain cache names → display names (matching island_pal keys).
# Use as: mutate(island = recode_island[island])
recode_island <- c(
  "Oahu"    = "O'ahu",
  "Maui"    = "Maui",
  "Molokai" = "Moloka'i",
  "Lanai"   = "L\u0101na'i"
)

# Scale metadata for each plottable variable. Add entries here to support new variables.
# `trans` drives both the internal kriging transform and the ggplot scale transform.
# `oob_label` is the label used for the top break (values squished above the limit).
var_meta <- list(
  NO3NO2 = list(
    label     = "NO\u2083+NO\u2082\n(\u03bcmol/L)",
   # limits    = c(0, 20),
  #  breaks    = c(0.5, 1, 3, 10, 20),
   # labels    = c("0.5", "1", "3", "10", "\u226520"),
     limits    = c(0, 10),
    breaks    = c(0.5, 1, 3, 5, 10),
    labels    = c("0.5", "1", "3", "5", "\u226510"),
     trans     = "log1p",
    direction = 1
  ),
  PO4 = list(
    label     = "PO\u2084\n(\u03bcmol/L)",
    limits    = c(0, 1),
    breaks    = c(0.05, 0.1, 0.3, 0.5, 1),
    labels    = c("0.05", "0.1", "0.3", "0.5", "\u22651"),
    trans     = "log1p",
    direction = 1
  ),
  SiO2 = list(
    label     = "SiO\u2082\n(\u03bcmol/L)",
   # limits    = c(0, 100),
  #  breaks    = c(1, 5, 20, 50, 100),
  #  labels    = c("1", "5", "20", "50", "\u2265100"),
     limits    = c(0, 50),
    breaks    = c(1, 5, 10, 20, 50),
    labels    = c("1", "5", "10", "20", "\u226550"),
    trans     = "log1p",
    direction = 1
  ),
  NH3 = list(
    label     = "NH\u2083\n(\u03bcmol/L)",
    limits    = c(0, 3),
    breaks    = c(0.1, 0.5, 1, 2, 3),
    labels    = c("0.1", "0.5", "1", "2", "\u22653"),
    trans     = "log1p",
    direction = 1
  ),
  Salinity = list(
    label     = "Salinity\n(PSU)",
    limits    = c(28, 36),
    breaks    = c(28, 30, 32, 34, 36),
    labels    = c("\u226428", "30", "32", "34", "36"),
    trans     = "identity",
    direction = -1
  )
)

krige_island_map <- function(data, island_name, ahupuaa_sf,
                              var = "NO3NO2",
                              res_m = 250, band_km = 3, tile_zoom = 11) {
  
  meta    <- var_meta[[var]]
  use_log <- !is.null(meta) && meta$trans == "log1p"
  if (is.null(meta)) meta <- list(
    label = var, limits = NULL, breaks = waiver(), labels = waiver(), trans = "identity"
  )
  
  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]), !is.na(latitude), !is.na(longitude))
  
  # Dissolve ahupuaʻa polygons; drop parts < 1 km² to remove errant geometry
  match_idx <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union  <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  land_sf_map <- st_as_sf(st_sf(geometry = land_union))
  
  coast_outer <- st_buffer(land_union, dist = band_km * 1000)
  coast_band  <- st_difference(coast_outer, land_union)
  coast_bbox  <- st_bbox(coast_outer)
  
  # Deduplicate sample locations — duplicate coordinates cause a singular
  # kriging covariance matrix and NA predictions across the whole island
  pts_proj <- island_data |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(raw_val = .data[[var]],
           log_val = if (use_log) log1p(raw_val) else raw_val,
           X = st_coordinates(geometry)[, 1],
           Y = st_coordinates(geometry)[, 2]) |>
    group_by(X, Y) |>
    summarise(log_val = mean(log_val, na.rm = TRUE),
              raw_val = mean(raw_val, na.rm = TRUE), .groups = "drop") |>
    st_as_sf(coords = c("X", "Y"), crs = 32604)
  
  # Display points clipped to coast_bbox; carries variable values for shared color scale
  pts_proj_all <- island_data |>
    filter(!is.na(latitude), !is.na(longitude)) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(val = .data[[var]],
           X   = st_coordinates(geometry)[,1],
           Y   = st_coordinates(geometry)[,2]) |>
    filter(X >= coast_bbox["xmin"], X <= coast_bbox["xmax"],
           Y >= coast_bbox["ymin"], Y <= coast_bbox["ymax"])
  
  # Variogram with enforced minimum range of island_diag / 5
  v_emp       <- variogram(log_val ~ 1, pts_proj)
  init_psill  <- var(pts_proj$log_val, na.rm = TRUE)
  island_diag <- sqrt(diff(st_bbox(pts_proj)[c("xmin","xmax")])^2 +
                      diff(st_bbox(pts_proj)[c("ymin","ymax")])^2)
  v_auto <- tryCatch(
    fit.variogram(v_emp, vgm(psill = init_psill * 0.8, model = "Sph",
                             range = island_diag / 5, nugget = init_psill * 0.2)),
    error = function(e) NULL)
  min_range <- island_diag / 5
  v_fit <- if (!is.null(v_auto) && v_auto$range[2] >= min_range) v_auto else {
    psill  <- if (!is.null(v_auto)) v_auto$psill[2] else init_psill * 0.8
    nugget <- if (!is.null(v_auto)) v_auto$psill[1] else init_psill * 0.2
    vgm(psill = psill, model = "Sph", range = min_range, nugget = nugget)
  }
  
  # Prediction grid: coastal band intersected with a buffer around sample points.
  # Grid cells farther than min_range from any sample are excluded — no data
  # nearby means no interpolation shown (prevents extrapolation into unsurveyed coast).
  sample_coverage <- st_buffer(st_union(pts_proj), dist = min_range)
  grid_pts <- st_make_grid(coast_band, cellsize = res_m, what = "centers") |>
    st_as_sf() |>
    st_filter(coast_band) |>
    st_filter(sample_coverage)
  
  k_out <- suppressWarnings(
    krige(log_val ~ 1, pts_proj, grid_pts, model = v_fit, nmax = 40)
  )
  pred_df <- st_coordinates(k_out) |>
    as.data.frame() |>
    mutate(pred = if (use_log) expm1(k_out$var1.pred) else k_out$var1.pred)
  
  krige_label <- if (use_log) paste0("log1p(", var, ")") else var
  
  ggplot() +
    geom_spatraster_rgb(data = get_tiles(st_transform(coast_outer, 4326),
                                         provider = "Esri.WorldImagery", zoom = tile_zoom)) +
    geom_raster(data = pred_df, aes(x = X, y = Y, fill = pred), alpha = 0.8) +
    geom_sf(data = land_sf_map, fill = NA, color = "white", linewidth = 0.4) +
    geom_sf(data = pts_proj_all, aes(fill = val), size = 1.0, shape = 21,
            color = "white", stroke = 0.3, alpha = 0.5) +
    scale_fill_viridis_c(
      name      = meta$label,
      option    = "plasma",
      direction = meta$direction,
      trans     = meta$trans,
      limits    = meta$limits,
      breaks    = meta$breaks,
      labels    = meta$labels,
      oob       = scales::squish,
      na.value  = "transparent",
      guide     = guide_colorbar(barheight = unit(10, "lines"),
                                 barwidth  = unit(0.6, "lines"),
                                 ticks.linewidth = 0.5)
    ) +
    coord_sf(crs = 32604,
             xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = island_name,
         subtitle = paste0("n = ", nrow(pts_proj),
                           " unique locations | ordinary kriging on ", krige_label)) +
    theme_minimal(base_size = 12) +
    theme(axis.title = element_blank(), axis.text = element_text(size = 7),
          plot.title = element_text(face = "bold", size = 14),
          plot.subtitle = element_text(size = 9, color = "gray40"))
}

raw_points_map <- function(data, island_name, ahupuaa_sf,
                            var = "NO3NO2", tile_zoom = 12) {
  
  meta <- var_meta[[var]]
  if (is.null(meta)) meta <- list(
    label = var, limits = NULL, breaks = waiver(), labels = waiver(), trans = "identity"
  )
  
  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]), !is.na(latitude), !is.na(longitude)) |>
    mutate(val = .data[[var]]) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604)
  
  # Same land polygon logic as krige_island_map
  match_idx <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |>
    st_make_valid() |>
    st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union  <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  land_sf_map <- st_as_sf(st_sf(geometry = land_union))
  
  coast_outer <- st_buffer(land_union, dist = 5000)
  coast_bbox  <- st_bbox(coast_outer)
  
  sat_tiles <- get_tiles(st_transform(coast_outer, 4326),
                         provider = "Esri.WorldImagery", zoom = tile_zoom)
  
  ggplot() +
    geom_spatraster_rgb(data = sat_tiles) +
    geom_sf(data = land_sf_map, fill = NA, color = "white", linewidth = 0.4) +
    geom_sf(data = island_data, aes(fill = val), size = 3.5, shape = 21,
            color = "white", stroke = 0.5, alpha = 0.9) +
    scale_fill_viridis_c(
      name      = meta$label,
      option    = "plasma",
      direction = meta$direction,
      trans     = meta$trans,
      limits    = meta$limits,
      breaks    = meta$breaks,
      labels    = meta$labels,
      oob       = scales::squish,
      guide     = guide_colorbar(barheight = unit(8, "lines"),
                                 barwidth  = unit(0.6, "lines"),
                                 ticks.linewidth = 0.5)
    ) +
    coord_sf(crs = 32604,
             xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = island_name,
         subtitle = paste0("n = ", nrow(island_data),
                           " samples | raw observations (insufficient coverage for kriging)")) +
    theme_minimal(base_size = 12) +
    theme(axis.title = element_blank(), axis.text = element_text(size = 7),
          plot.title = element_text(face = "bold", size = 14),
          plot.subtitle = element_text(size = 9, color = "gray40"))
}

# ── krige_surface_data: returns pred_df + land polygon + coast_bbox ───────────
krige_surface_data <- function(data, island_name, ahupuaa_sf,
                               var = "NO3NO2", res_m = 300, band_km = 3) {
  meta    <- var_meta[[var]]
  use_log <- !is.null(meta) && meta$trans == "log1p"
  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]),
           !is.na(latitude), !is.na(longitude))
  match_idx  <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  coast_outer <- st_buffer(land_union, dist = band_km * 1000)
  coast_band  <- st_difference(coast_outer, land_union)
  coast_bbox  <- st_bbox(coast_outer)
  pts_proj <- island_data |>
    st_as_sf(coords = c("longitude","latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(raw_val = .data[[var]],
           log_val = if (use_log) log1p(raw_val) else raw_val,
           X = st_coordinates(geometry)[,1],
           Y = st_coordinates(geometry)[,2]) |>
    group_by(X, Y) |>
    summarise(log_val = mean(log_val, na.rm = TRUE),
              raw_val = mean(raw_val, na.rm = TRUE), .groups = "drop") |>
    st_as_sf(coords = c("X","Y"), crs = 32604)
  island_diag <- sqrt(diff(st_bbox(pts_proj)[c("xmin","xmax")])^2 +
                      diff(st_bbox(pts_proj)[c("ymin","ymax")])^2)
  min_range <- island_diag / 5
  v_emp  <- variogram(log_val ~ 1, pts_proj)
  v_auto <- tryCatch(
    fit.variogram(v_emp, vgm(var(pts_proj$log_val)*0.8, "Sph",
                              min_range, var(pts_proj$log_val)*0.2)),
    error = function(e) NULL)
  v_fit <- if (!is.null(v_auto) && v_auto$range[2] >= min_range) v_auto else
    vgm(var(pts_proj$log_val)*0.8, "Sph", min_range, var(pts_proj$log_val)*0.2)
  sample_cov <- st_buffer(st_union(pts_proj), dist = min_range)
  grid_pts <- st_make_grid(coast_band, cellsize = res_m, what = "centers") |>
    st_as_sf() |> st_filter(coast_band) |> st_filter(sample_cov)
  k_out <- suppressWarnings(krige(log_val ~ 1, pts_proj, grid_pts, model = v_fit, nmax = 40))
  pred_df <- st_coordinates(k_out) |> as.data.frame() |>
    mutate(pred = if (use_log) expm1(k_out$var1.pred) else k_out$var1.pred)
  list(pred_df    = pred_df,
       land_sf    = st_as_sf(st_sf(geometry = land_union)),
       coast_bbox = coast_bbox)
}

# ── maui_validation_map: kriging surface + optional Hui station overlay ───────
# Pass bg_hui_loc to show Hui triangles (only meaningful for var = "NO3NO2").
maui_validation_map <- function(data, ahupuaa_sf, var = "NO3NO2",
                                 bg_hui_loc = NULL, tile_zoom = 11) {
  meta <- var_meta[[var]]
  if (is.null(meta)) meta <- list(
    label = var, limits = NULL, breaks = waiver(), labels = waiver(),
    trans = "identity", direction = 1
  )

  kd     <- krige_surface_data(data, "Maui", ahupuaa_sf, var = var)
  cs_pts <- data |>
    filter(island == "Maui", !is.na(.data[[var]]), !is.na(latitude), !is.na(longitude)) |>
    mutate(val = .data[[var]]) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604)
  sat <- get_tiles(
    st_as_sfc(kd$coast_bbox) |> st_set_crs(32604) |> st_transform(4326),
    provider = "Esri.WorldImagery", zoom = tile_zoom
  )

  # Build Hui station sf internally when bg_hui_loc is supplied
  # Map chem var names to the corresponding Hui column (all in µmol/L)
  hui_col_map <- c(NO3NO2 = "NNN_umol", PO4 = "Phosphate_umol",
                   SiO2   = "Silicate_umol", NH3 = "NH4_umol")
  hui_col  <- hui_col_map[var]
  show_hui <- !is.null(bg_hui_loc) && !is.na(hui_col) && hui_col %in% names(bg_hui_loc)

  if (show_hui) {
    hui_post <- bg_hui_loc |>
      filter(year(Date) == 2026, month(Date) == 3,
             Date >= as.Date("2026-03-14"), !is.na(.data[[hui_col]])) |>
      group_by(Station, SiteName, region, Lat, Long) |>
      summarise(val_umol = mean(.data[[hui_col]], na.rm = TRUE), .groups = "drop")
    hui_hist_med <- bg_hui_loc |>
      filter(!is.na(.data[[hui_col]])) |>
      group_by(Station) |>
      summarise(med_hist = median(.data[[hui_col]], na.rm = TRUE), .groups = "drop")
    log_lo   <- log1p(1); log_hi <- log1p(20)
    grey_pal <- colorRampPalette(c("gray20", "white"))(1000)
    hui_sf <- bg_hui_loc |>
      distinct(Station, SiteName, region, Lat, Long) |>
      filter(!is.na(Lat), !is.na(Long)) |>
      left_join(hui_post |> select(Station, val_umol), by = "Station") |>
      left_join(hui_hist_med, by = "Station") |>
      mutate(
        fold_change = val_umol / med_hist,
        fc_safe     = pmin(pmax(replace_na(fold_change, 1), 1), 20),
        tri_fill    = grey_pal[round((log1p(fc_safe) - log_lo) /
                                       (log_hi - log_lo) * 999) + 1],
        tri_fill    = case_when(
          is.na(fold_change) & !is.na(med_hist) ~ "gray50",
          is.na(med_hist)                        ~ "gray70",
          TRUE                                   ~ tri_fill
        ),
        fold_change = replace_na(fold_change, 1)
      ) |>
      st_as_sf(coords = c("Long","Lat"), crs = 4326) |>
      st_transform(32604)
  }

  p <- ggplot() +
    geom_spatraster_rgb(data = sat) +
    geom_raster(data = kd$pred_df, aes(x = X, y = Y, fill = pred), alpha = 0.8) +
    geom_sf(data = kd$land_sf, fill = NA, color = "white", linewidth = 0.35) +
    geom_sf(data = cs_pts, aes(fill = val),
            shape = 21, size = 1.0, color = "white", stroke = 0.3, alpha = 0.5)

  # fill scale used only for kriging raster + CS circles (plasma)
  p <- p +
    scale_fill_viridis_c(
      name      = meta$label,
      option    = "plasma",
      direction = meta$direction,
      trans     = meta$trans,
      limits    = meta$limits,
      breaks    = meta$breaks,
      labels    = meta$labels,
      oob       = scales::squish,
      na.value  = "transparent",
      guide     = guide_colorbar(barheight = unit(10, "lines"),
                                 barwidth  = unit(0.6, "lines"),
                                 ticks.linewidth = 0.5)
    )

  if (show_hui) {
    hui_breaks <- c(1, 3, 5, 10, 20)
    hui_labels <- c("1", "3", "5", "10", "\u226520")
    hui_name   <- "Fold-change\n(Mar 2026 /\nall-time median)"
    p <- p +
      # Shape 24 (fill + white outline); I() bypasses the plasma fill scale
      geom_sf(data = hui_sf,
              aes(size = fold_change, fill = I(tri_fill)),
              color = "white", shape = 24, stroke = 0.8, alpha = 0.85) +
      ggrepel::geom_label_repel(
        data = hui_sf |> filter(fold_change > 15),
        aes(label    = paste0(SiteName, "\n", round(fold_change, 0), "\u00d7"),
            geometry = geometry),
        stat = "sf_coordinates", size = 2.5, fontface = "bold", color = "#B71C1C",
        fill = alpha("white", 0.85), label.size = 0.2,
        nudge_y = 4500, min.segment.length = 0
      ) +
      scale_size_continuous(
        name = hui_name, range = c(1.0, 8), limits = c(1, 20),
        breaks = hui_breaks, labels = hui_labels,
        guide = guide_legend(
          theme = theme(legend.key = element_rect(fill = "gray30", color = NA)),
          override.aes = list(shape = 24, color = "white", stroke = 0.8, alpha = 0.85,
                              fill = c("#333333", "#6F6F6F", "#929292", "#C6C6C6", "#FFFFFF"))
        )
      )
  }

  hui_note <- if (show_hui) {
    paste0("    \u25b3 Hui o Ka Wai Ola stations (size = Mar 2026 fold-change above all-time median)\n",
           "\u25b3 small = no March 2026 sample    \u25b3 gray = insufficient data    ",
           "Labels: stations with fold-change > 15\u00d7")
  } else ""

  p +
    coord_sf(crs = 32604,
             xlim = c(kd$coast_bbox["xmin"], kd$coast_bbox["xmax"]),
             ylim = c(kd$coast_bbox["ymin"], kd$coast_bbox["ymax"]),
             expand = FALSE) +
    labs(
  #    subtitle = paste0("\u25cf community science samples (this study)", hui_note),
      caption  = if (show_hui) "Hui data: Falinski et al. / Hui o Ka Wai Ola." else NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.title    = element_blank(),
      axis.text     = element_text(size = 7),
      plot.subtitle = element_text(size = 8.5, color = "gray40"),
      plot.caption  = element_text(size = 8, color = "gray50", hjust = 0),
      legend.position = "right"
    )
}
```

## O'ahu

Oʻahu's highest nutrient concentrations cluster along the northwestern (Waiʻanae) coast and around Pearl Harbor, consistent with those areas having both dense cesspool infrastructure and streams that drain heavily developed land. The salinity map shows freshwater dilution in the same locations, confirming this is land-sourced water, not re-suspended sediment.

```{r}
#| label: map-oahu
#| fig-height: 7
#| fig-cap: "Spatially interpolated nutrient and salinity maps for Oʻahu, March 2026. Color shows concentration — yellow-green = highest, dark purple = lowest (all log scale). Circles are individual sample points colored on the same scale. Low-salinity (freshwater-diluted) hotspots on the northwest coast line up with the highest nitrogen and silica readings."
oahuNN<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var = "NO3NO2" )

oahuPO<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="PO4" )

oahuSi<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="SiO2" )

oahuNH3<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="NH3" )

oahuSal<-krige_island_map(chem, "O'ahu", ahupuaa_raw,var ="Salinity" )

(oahuNN|oahuPO|oahuNH3)/(oahuSi|oahuSal|plot_spacer())
ggsave(here("output","Oahu_spatial.png"), height = 12, width = 18)

```

## Maui

Maui shows the most dramatic hotspot signal of any island, concentrated on the north and west coasts. The Mālā Ramp area and Honoköwai in West Maui stand out clearly — these sites also showed the largest departures from their long-term monitoring baselines (see Hui comparison below). North Maui's windward streams drain some of the island's most intensively farmed land, explaining the particularly high silica (SiO₂) signature there.

```{r}
#| label: map-maui
#| fig-height: 7
#| fig-cap: "Spatially interpolated nutrient and salinity maps for Maui, March 2026. West Maui's leeward coast and the windward north coast show the highest concentrations. The match between low-salinity patches and high-nutrient zones confirms that dilution with freshwater runoff is driving the signal, not local biological processes."

MauiNN<-krige_island_map(chem, "Maui", ahupuaa_raw,var = "NO3NO2" )

MauiPO<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="PO4" )

MauiSi<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="SiO2" )

MauiNH3<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="NH3" )

MauiSal<-krige_island_map(chem, "Maui", ahupuaa_raw,var ="Salinity" )

(MauiNN|MauiPO|MauiNH3)/(MauiSi|MauiSal|plot_spacer())
ggsave(here("output","Maui_spatial.png"), height = 12, width = 18)
```

## Moloka'i

Molokaʻi shows elevated nitrogen and silica across most of its southern coast, with the signal strongest near Kaunakakai Harbor. The pattern is more spatially uniform than Maui, consistent with Molokaʻi's lower topographic relief producing more diffuse runoff rather than concentrated stream plumes.

```{r}
#| label: map-molokai
#| fig-height: 5
#| fig-cap: "Spatially interpolated nutrient and salinity maps for Molokaʻi, March 2026. Elevated nutrients are distributed broadly along the south coast rather than in discrete plumes, reflecting the island's lower elevation and more evenly distributed runoff pathways."

molokaiNN<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var = "NO3NO2", tile_zoom = 12 )

molokaiPO<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="PO4" , tile_zoom = 12)

molokaiSi<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="SiO2", tile_zoom = 12 )

molokaiNH3<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="NH3", tile_zoom = 12 )

molokaiSal<-krige_island_map(chem, "Moloka'i", ahupuaa_raw,var ="Salinity", tile_zoom = 12 )

(molokaiNN|molokaiPO|molokaiNH3)/(molokaiSi|molokaiSal|plot_spacer())
ggsave(here("output","molokai_spatial.png"), height = 12, width = 18)


```

## Lāna'i

Only raw sample locations are shown for Lānaʻi — with just 20 samples, there aren't enough data points to produce a reliable interpolated surface. The samples that do exist show moderate nutrient concentrations with no extreme hotspots, though sampling coverage is too sparse to draw firm conclusions.

```{r}
#| label: map-lanai
#| fig-height: 6
lanaiNN<-raw_points_map(chem, "L\u0101na'i", ahupuaa_raw, var = "NO3NO2", tile_zoom = 12)

lanaiPO<-raw_points_map(chem, "L\u0101na'i", ahupuaa_raw, var = "PO4", tile_zoom = 12)
lanaiSi<-raw_points_map(chem, "L\u0101na'i", ahupuaa_raw, var = "SiO2", tile_zoom = 12)
lanaiNH3<-raw_points_map(chem, "L\u0101na'i", ahupuaa_raw, var = "NH3", tile_zoom = 12)
lanaiSal<-raw_points_map(chem, "L\u0101na'i", ahupuaa_raw, var = "Salinity", tile_zoom = 12)

(lanaiNN|lanaiPO|lanaiNH3)/(lanaiSi|lanaiSal|plot_spacer())
ggsave(here("output","lanai_spatial.png"), height = 12, width = 18)
```


```{r}
#| label: map-all
#| fig-height: 6

#oahu / maui /molokai + plot_layout(guides = "collect")
#ggsave(here("output","maps_all.png"), height = 10, width = 10)
 
```


# Independent validation: Hui o Ka Wai Ola stations on the Maui kriging surface

The kriging surface is derived entirely from community science samples. The Hui o Ka Wai Ola programme provides an independent check: do the long-term fixed stations — sampled with laboratory-grade methods over a decade — agree with where the community science kriging shows hotspots? Triangles show Hui station post-storm NO₃+NO₂ on the same color scale as the kriging surface; their colors should match the surface underneath them if the two datasets are consistent.

```{r}
#| label: maui-validation-map
#| fig-height: 7
#| fig-cap: "Two independent datasets, one picture. Background: the interpolated nitrogen (NO₃+NO₂) map built from community science samples. Triangles: Hui o Ka Wai Ola professional monitoring stations, colored by their own post-storm measurements (same color scale). Gray-outlined triangles had no March 2026 sample. The Hui stations with the most extreme readings — Mālā Ramp (43×), Camp Olowalu (32×), and others on West Maui — sit precisely where the community science map shows the darkest hotspot colors. The agreement between two completely independent datasets confirms the hotspot pattern is real."


bg_hui_loc <- bind_rows(
  readxl::read_excel(
    "data/background_nutrients/hui-south-maui-thru-2026-1st-quarter.0.xlsx",
    sheet = "data") |>
    mutate(Turbidity = as.numeric(Turbidity), region = "South Maui"),
  readxl::read_excel(
    "data/background_nutrients/hui-west-maui-thru-2026-1st-quarter.0.xlsx",
    sheet = "data") |>
    mutate(Turbidity = as.numeric(Turbidity), region = "West Maui")
) |>
  mutate(Date           = as.Date(Date),
         NNN_umol       = as.numeric(NNN)       / 14.007,   # µg/L N  → µmol/L
         Phosphate_umol = as.numeric(Phosphate) / 30.974,   # µg/L P  → µmol/L
         Silicate_umol  = as.numeric(Silicate)  / 28.086,   # µg/L Si → µmol/L
         NH4_umol       = as.numeric(NH4)        / 14.007)  # µg/L N  → µmol/L

HuiNN<-maui_validation_map(chem, ahupuaa_raw, var = "NO3NO2", bg_hui_loc = bg_hui_loc)

HuiSi<-maui_validation_map(chem, ahupuaa_raw, var = "SiO2", bg_hui_loc = bg_hui_loc)

HuiNH3<-maui_validation_map(chem, ahupuaa_raw, var = "NH3", bg_hui_loc = bg_hui_loc)

HuiPO<-maui_validation_map(chem, ahupuaa_raw, var = "PO4", bg_hui_loc = bg_hui_loc)

no_sub  <- theme(plot.subtitle = element_blank(), plot.caption = element_blank())
no_size <- guides(size = "none")

size_legend <- get_legend(
  HuiNN + guides(fill = "none") + theme(legend.position = "right")
)

plot_grid(
  plot_grid(
    HuiNN  + no_size + no_sub,
    HuiPO  + no_size + no_sub,
    HuiNH3 + no_size + no_sub,
    HuiSi  + no_size + no_sub,
    ncol = 2
  ),
  size_legend,
  ncol = 2,
  rel_widths = c(1, 0.13)
) |>
  ggdraw() +
  theme(plot.background = element_rect(fill = "white", color = NA))

ggsave(here("output","hui_spatial.png"), height = 12, width = 12)
```

The Hui stations on West Maui's leeward coast show amber-to-orange colors, meaning their post-storm readings were 8–43× higher than their own ten-year averages. The calmer eastern coast shows lower readings in both datasets. The fact that community science volunteers and trained monitoring staff — sampling at different times with different equipment — independently found the same hotspot pattern in the same places gives us high confidence the signal is real, not a measurement artifact.

---

# Storm cascade: disruption of coastal biogeochemistry

The storm set off a four-step chain of events visible in satellite data, stream gauges, and water chemistry. Rain (panel A) filled streams (panel B), which flushed nutrients into the coast, triggering an algae bloom visible from space (panel C), and leaving elevated chemical concentrations still measurable when volunteers sampled two weeks later (panel D). Peak rainfall hit March 13–14, reaching 42–100 mm/day averaged across each island. Streams responded immediately, with peak flows of 231–1,040 cubic feet per second depending on watershed size. Coastal chlorophyll — a measure of algae growth — rose across all three islands within days and remained elevated for weeks. When community scientists sampled in late March and early April, nitrogen (NO₃+NO₂) and silica (SiO₂) were the most clearly elevated nutrients, consistent with runoff from volcanic soils.

```{r}
#| label: storm-cascade-figure
#| fig-height: 8
#| fig-width: 10
#| fig-cap: "Rain → streams → algae → chemistry: the full storm cascade in four panels. (A) Average daily rainfall per island from atmospheric model data (NCEP). (B) Stream discharge at three USGS gauges — Makaha Stream (Oʻahu), Halawa Stream (Molokaʻi), Hanawi Stream (Maui) — on a log scale because flows span three orders of magnitude. (C) Mean coastal chlorophyll-a from satellite (VIIRS NOAA-20), with a smoothed trend line; individual daily averages shown as points (only days with ≥8 valid satellite pixels included). (D) Post-storm nutrient concentrations by island from community science sampling (Mar 28–Apr 8). Blue shading = the Kona Low storm window (Mar 10–25); green shading = sampling window."



# ── shared constants ────────────────────────────────────────────────
storm_xmin <- as.Date("2026-03-10"); storm_xmax <- as.Date("2026-03-25")
samp_xmin  <- as.Date("2026-03-28"); samp_xmax  <- as.Date("2026-04-08")
plot_xlim  <- as.Date(c("2026-01-15", "2026-06-05"))

windows <- list(
  annotate("rect", xmin = storm_xmin, xmax = storm_xmax,
           ymin = -Inf, ymax = Inf, fill = "#BBDEFB", alpha = 0.75),
  annotate("rect", xmin = samp_xmin,  xmax = samp_xmax,
           ymin = -Inf, ymax = Inf, fill = "#C8E6C9", alpha = 0.75)
)

ts_base <- theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title.x       = element_blank(),
    axis.text.x        = element_blank(),
    axis.ticks.x       = element_blank(),
    legend.position    = "none",
    plot.margin        = margin(3, 6, 1, 6),
    axis.title.y       = element_text(size = 8.5)
  )

# Panel A: Rainfall
rain_df <- read_csv("hawaii_ncep_janjun_cache.csv", show_col_types = FALSE) |>
  filter(island %in% c("Oahu", "Maui", "Molokai"),
         date >= plot_xlim[1], date <= plot_xlim[2]) |>
  mutate(island = recode_island[island]) |>
  group_by(island, date) |>
  summarise(precip_mm = mean(precip_mmday, na.rm = TRUE), .groups = "drop")

pA <- ggplot(rain_df, aes(x = date, y = precip_mm, fill = island, color = island)) +
  windows +
  geom_area(position = "identity", alpha = 0.30, linewidth = 0.35) +
  annotate("text", x = as.Date("2026-02-17"), y = 172,
           label = "Feb event", size = 2.6, color = "gray35", hjust = 0.5, vjust = 0) +
  annotate("text", x = as.Date("2026-03-17"), y = 108,
           label = "Kona Low", size = 2.6, color = "#1A237E", hjust = 0.5, vjust = 0) +
  scale_fill_manual(values  = island_pal) +
  scale_color_manual(values = island_pal) +
  scale_x_date(limits = plot_xlim, expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(y = "Rainfall\n(mm/day)", tag = "A") +
  ts_base

# Panel B: Stream discharge (USGS NWIS — cached in data/usgs_discharge_cache.rds)
gauge_meta <- tibble(
  site_no = c("16211600", "16400000", "16508000"),
  island  = c("O'ahu",    "Moloka'i", "Maui")
)

dis_df <- read_csv("data/usgs_discharge_cache.csv", show_col_types = FALSE,
                   col_types = cols(site_no = col_character())) |>
  dataRetrieval::renameNWISColumns() |>
  filter(site_no %in% gauge_meta$site_no,
         Date >= plot_xlim[1], Date <= plot_xlim[2], Flow > 0.5) |>
  left_join(gauge_meta, by = "site_no")

pB <- ggplot(dis_df, aes(x = Date, y = Flow, color = island, group = site_no)) +
  windows +
  geom_line(linewidth = 0.65, alpha = 0.9, na.rm = TRUE) +
  scale_color_manual(values = island_pal) +
  scale_x_date(limits = plot_xlim, date_labels = "%b %Y", expand = c(0, 0)) +
  scale_y_log10(breaks = c(1, 10, 100, 1000),
                labels = scales::label_number(big.mark = ",")) +
  coord_cartesian(ylim = c(0.8, 3000)) +
  labs(y = "Discharge\n(cfs, log)", tag = "B") +
  ts_base +
  theme(
    axis.text.x  = element_text(size = 8),
    axis.ticks.x = element_line(linewidth = 0.3)
  )

# Panel C: Coastal chlorophyll (VIIRS NOAA-20 — cached in hawaii_chlorophyll_cache.rds)
chla_df <- readRDS("hawaii_chlorophyll_cache.rds") |>    # satellite array — kept as RDS
  filter(island %in% c("Oahu", "Maui", "Molokai"),
         date >= plot_xlim[1], date <= plot_xlim[2]) |>
  mutate(island = recode_island[island]) |>
  group_by(island, date) |>
  summarise(mean_chla = mean(chlor_a, na.rm = TRUE),
            n_valid   = sum(!is.na(chlor_a)), .groups = "drop") |>
  filter(n_valid >= 8, mean_chla > 0)

pC <- ggplot(chla_df, aes(x = date, y = mean_chla, color = island)) +
  windows +
  geom_point(size = 0.8, alpha = 0.55, na.rm = TRUE) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE,
              linewidth = 0.9, na.rm = TRUE) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_x_date(limits = plot_xlim, date_labels = "%b %Y", expand = c(0, 0)) +
  scale_y_log10(breaks = c(0.1, 0.3, 1, 3, 10),
                labels = scales::label_number(accuracy = 0.1)) +
  labs(y = "Chlorophyll-a\n(mg/m\u00b3, log)", tag = "C") +
  ts_base +
  theme(
    axis.text.x      = element_text(size = 8),
    axis.ticks.x     = element_line(linewidth = 0.3),
    legend.position  = "bottom",
    legend.text      = element_text(size = 9),
    legend.key.width = unit(1.2, "lines"),
    plot.margin      = margin(1, 6, 2, 6)
  )

# Panel D: Post-storm nutrient distributions
nut_long <- chem |>
  filter(island %in% c("O'ahu", "Maui", "Moloka'i")) |>
  mutate(
    island = factor(island, levels = c("Maui", "Moloka'i", "O'ahu"))
  ) |>
  select(island, NO3NO2, NH3, PO4, SiO2) |>
  pivot_longer(c(NO3NO2, NH3, PO4, SiO2), names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>
  mutate(nutrient = factor(
    recode(nutrient, "NO3NO2" = "NO[3]+NO[2]", "NH3" = "NH[4]",
           "PO4" = "PO[4]", "SiO2" = "SiO[2]"),
    levels = c("NO[3]+NO[2]", "NH[4]", "PO[4]", "SiO[2]")
  ))

pD <- ggplot(nut_long, aes(x = island, y = conc, fill = island, color = island)) +
  geom_violin(alpha = 0.28, linewidth = 0.3, trim = TRUE) +
  geom_boxplot(width = 0.17, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = island_pal) +
  scale_color_manual(values = island_pal) +
  scale_y_log10(breaks = c(0.01, 0.1, 1, 10, 100),
                labels = scales::label_number(drop0trailing = TRUE)) +
  facet_wrap(~nutrient, ncol = 1, scales = "free_y",
             labeller = label_parsed) +
  labs(tag = "D", y = expression(paste(mu, "mol L"^{-1})), x = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x        = element_text(angle = 30, hjust = 1, vjust = 1, size = rel(0.85)),
    strip.text         = element_text(size = 9, face = "bold"),
    legend.position    = "none",
    plot.margin        = margin(4, 8, 4, 2),
    axis.title.y       = element_text(size = 8.5)
  )

# Assemble
((pA / pB) | pD) +
  plot_layout(widths = c(3, 1.5)) +
  plot_annotation(
    title   = "Kona Low storm cascade: terrestrial runoff disrupts coastal biogeochemistry",
    caption = paste0(
      "Blue shading = Kona Low storm window (Mar 10\u201325).  ",
      "Green shading = community science sampling (Mar 28\u2013Apr 8).\n",
      "",
      "Stream gauges: Makaha Str (O\u02BBahu), Halawa Str (Moloka\u02BBi), Hanawi Str (Maui)."
    ),
    theme = theme(
      plot.title   = element_text(size = 11, face = "bold", margin = margin(b = 4)),
      plot.caption = element_text(size = 7.5, color = "gray40",
                                  hjust = 0, margin = margin(t = 6)),
      plot.margin  = margin(8, 8, 6, 8)
    )
  )

ggsave(here("output","nutrient_gradient.png"))
```

# Spatial structure of the biogeochemical signal

If the storm's chemical impact was driven by watershed runoff, then nearby coastal sites should have more similar nutrient levels than sites far apart — high readings near stream mouths, lower readings away from them. This would produce a non-random, geographically coherent pattern. We test this with Moran's I, a standard measure of spatial clustering where a positive, statistically significant value means nearby sites are more alike than you'd expect by chance. The test was run separately for each nutrient and island using the eight nearest sampling neighbours. Lāna'i had too few samples and is excluded.

```{r}
#| label: morans-i
#| fig-height: 5
#| fig-cap: "Do high-nutrient sites cluster together spatially? Moran's I measures this: values above zero (filled points, p < 0.05) mean nearby sites are more similar than random chance would produce; open circles are not statistically distinguishable from random. Error bars = ±1.96 standard errors. SiO₂ and NO₃+NO₂ show clear geographic clustering on both Maui and Molokaʻi — their hotspots follow the shape of watersheds and stream mouths. NH₃ shows no significant clustering, meaning it is scattered randomly across the coast rather than concentrated near any particular landscape feature."

nutrients  <- c("SiO2", "NO3NO2", "PO4", "NH3")
islands    <- c("O'ahu", "Maui", "Moloka'i")

chem_sf_mi <- chem |>
  filter(!is.na(latitude), !is.na(longitude), !is.na(island)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

moran_results <- purrr::map_dfr(islands, function(isl) {
  purrr::map_dfr(nutrients, function(nut) {
    pts <- chem_sf_mi |> filter(island == isl, !is.na(.data[[nut]]))
    if (nrow(pts) < 15) return(NULL)
    nb <- knn2nb(knearneigh(st_coordinates(pts), k = 8), sym = TRUE)
    lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
    mt <- moran.test(pts[[nut]], lw, zero.policy = TRUE, alternative = "greater")
    tibble(
      island   = isl,
      nutrient = nut,
      n        = nrow(pts),
      moran_I  = unname(mt$estimate["Moran I statistic"]),
      expected = unname(mt$estimate["Expectation"]),
      variance = unname(mt$estimate["Variance"]),
      p_value  = mt$p.value
    )
  })
})

# Plot
nutrient_labels_mi <- c(
  "SiO2"   = "SiO[2]",
  "NO3NO2" = "NO[3]+NO[2]",
  "PO4"    = "PO[4]",
  "NH3"    = "NH[3]"
)

moran_results |>
  mutate(
    island   = factor(island,   levels = c("O'ahu", "Maui", "Moloka'i")),
    nutrient = factor(nutrient, levels = nutrients),
    sig_label = case_when(
      p_value < 0.001 ~ "p < 0.001",
      p_value < 0.01  ~ "p < 0.01",
      p_value < 0.05  ~ "p < 0.05",
      TRUE            ~ "n.s."
    ),
    sig_label = factor(sig_label,
      levels = c("p < 0.001", "p < 0.01", "p < 0.05", "n.s."))
  ) |>
  ggplot(aes(x = moran_I, y = nutrient, color = island, shape = sig_label)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  geom_errorbar(
    aes(xmin = moran_I - 1.96 * sqrt(variance),
        xmax = moran_I + 1.96 * sqrt(variance)),
    width = 0, linewidth = 0.4, alpha = 0.35,
    orientation = "y",
    position = position_dodge(width = 0.55)
  ) +
  geom_point(size = 3.2, position = position_dodge(width = 0.55)) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_shape_manual(
    values = c("p < 0.001" = 16, "p < 0.01" = 17, "p < 0.05" = 15, "n.s." = 1),
    name   = "Significance"
  ) +
  scale_x_continuous(limits = c(-0.05, 0.5), breaks = seq(0, 0.5, 0.1)) +
  scale_y_discrete(labels = \(x) parse(text = nutrient_labels_mi[x])) +
  labs(
    x        = "Moran's I  (k = 8 nearest neighbours)",
    y        = NULL,
    subtitle = "Positive values indicate spatial clustering; error bars = ±1.96 SE"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, color = "gray90"),
    legend.position    = "right",
    plot.subtitle      = element_text(size = 9, color = "gray40")
  )
```

```{r}
#| label: morans-i-table
moran_results |>
  mutate(
    sig = case_when(
      p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",   TRUE ~ "n.s."
    )
  ) |>
  select(Island = island, Nutrient = nutrient, n,
         `Moran's I` = moran_I, `p-value` = p_value, ` ` = sig) |>
  kable(digits = 3,
        caption = "Moran's I spatial autocorrelation statistics per island and nutrient. Spatial weights: k = 8 nearest neighbours (UTM zone 4N). One-sided test against the alternative of positive spatial autocorrelation.")
```

Silica (SiO₂) shows the tightest geographic clustering on Maui (Moran's I = 0.41) and Molokaʻi (I = 0.40). Silica is a reliable marker of volcanic rock weathering — it enters the ocean almost exclusively through stream runoff, so its spatial pattern closely traces where streams met the coast during the storm. Nitrogen (NO₃+NO₂) clusters almost as strongly. Ammonium (NH₃), by contrast, shows no significant spatial pattern on either island — it appears scattered randomly across the coast. This contrast is telling: the storm stamped a watershed-shaped footprint on nitrogen and silica, but NH₃ — which builds up through biological decomposition in sediments — was unaffected by the event.

# Storm context: rainfall return period

Was this storm unusual? This matters enormously for interpretation: if the March 2026 Kona Low was a once-in-a-generation event, the chemical impacts it caused might be rare too. If it was an ordinary storm, those same impacts are happening every year — they've just never been measured at this scale before.

```{r}
#| label: storm-return-period
#| fig-height: 7
#| fig-cap: "Was the March 2026 Kona Low unusual? No. Boxes show the range of March monthly rainfall across years (2013–2025) from satellite-based rainfall estimates (CHIRPS). Diamonds show March 2026 — estimated from atmospheric model data (NCEP) and scaled to match the CHIRPS units. On all three islands, 2026 fell close to or slightly above the historical median: about half of past March events were wetter. Individual years are shown as dots (filled = wetter than 2026, open = drier)."

# ── Historical March monthly totals from CHIRPS (2013–2025) ─────────────────
# Cache stores monthly totals as one value per island per month (labeled day 1).
march_hist <- read_csv("hawaii_rainfall_cache.csv", show_col_types = FALSE) |>
  filter(source == "CHIRPS",
         island %in% c("Oahu","Maui","Molokai"),
         lubridate::month(date) == 3,
         lubridate::year(date) <= 2025) |>
  mutate(precip = if_else(precip < 0, NA_real_, precip)) |>
  group_by(island, year = lubridate::year(date)) |>
  summarise(march_mm = mean(precip, na.rm = TRUE), .groups = "drop")

# ── March 2026 monthly total: NCEP bias-corrected to CHIRPS scale ────────────
ncep_daily_rp <- read_csv("hawaii_ncep_janjun_cache.csv", show_col_types = FALSE) |>
  filter(island %in% c("Oahu","Maui","Molokai")) |>
  group_by(island, date) |>
  summarise(precip_mm = mean(precip_mmday, na.rm = TRUE), .groups = "drop")

# Bias factor: ratio of CHIRPS to NCEP over Jan–Feb overlap (same months in cache)
chirps_jf <- read_csv("hawaii_rainfall_cache.csv", show_col_types = FALSE) |>
  filter(source == "CHIRPS",
         island %in% c("Oahu","Maui","Molokai"),
         lubridate::month(date) %in% c(1, 2),
         lubridate::year(date) >= 2013) |>
  mutate(precip = if_else(precip < 0, NA_real_, precip)) |>
  group_by(island, year = lubridate::year(date),
           month = lubridate::month(date)) |>
  summarise(chirps_mm = mean(precip, na.rm = TRUE), .groups = "drop")

ncep_jf <- ncep_daily_rp |>
  filter(lubridate::month(date) %in% c(1, 2),
         lubridate::year(date) >= 2013,
         date <= as.Date("2026-02-28")) |>
  group_by(island, year = lubridate::year(date),
           month = lubridate::month(date)) |>
  summarise(ncep_mm = sum(precip_mm, na.rm = TRUE), .groups = "drop")

bias_rp <- inner_join(chirps_jf, ncep_jf, by = c("island","year","month")) |>
  group_by(island) |>
  summarise(factor = mean(chirps_mm, na.rm = TRUE) /
                     mean(ncep_mm,   na.rm = TRUE),
            .groups = "drop")

storm_rp <- ncep_daily_rp |>
  filter(lubridate::year(date) == 2026, lubridate::month(date) == 3) |>
  group_by(island) |>
  summarise(ncep_march_mm = sum(precip_mm, na.rm = TRUE), .groups = "drop") |>
  left_join(bias_rp, by = "island") |>
  mutate(storm_mm = ncep_march_mm * factor)

# ── Recode plain cache names → display names before plotting ─────────────────
march_hist  <- march_hist  |> mutate(island = recode_island[island])
storm_rp    <- storm_rp    |> mutate(island = recode_island[island])

march_hist_storm <- march_hist |>
  left_join(storm_rp |> select(island, storm_mm), by = "island") |>
  mutate(above = march_mm >= storm_mm)

# Panel A: year-by-year dots + boxplot summary
pA_rp <- ggplot(march_hist, aes(x = island, y = march_mm,
                                 fill = island, color = island)) +
  geom_boxplot(alpha = 0.25, outlier.shape = NA, linewidth = 0.4, width = 0.5) +
  geom_jitter(aes(shape = march_hist_storm$above), width = 0.08,
              size = 2.3, alpha = 0.8) +
  geom_point(data = storm_rp,
             aes(x = island, y = storm_mm),
             color = "black", fill = "white",
             size = 4.5, shape = 23, stroke = 1.5, inherit.aes = FALSE) +
  scale_fill_manual(values  = island_pal, name = NULL) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("Below 2026", "Above 2026"),
                     name   = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.1))) +
  labs(x = NULL, y = "March rainfall total (mm)",
       subtitle = "CHIRPS island-mean March totals 2013\u20132025 \u00b7 Diamond = March 2026 (NCEP, bias-corrected)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position    = "bottom",
        plot.subtitle      = element_text(size = 8.5, color = "gray40"))

pA_rp
```

March 2026 was an average March — slightly above the historical median on all three islands (Maui: `r round(storm_rp$storm_mm[storm_rp$island=="Maui"])` mm vs. the typical `r round(median(march_hist$march_mm[march_hist$island=="Maui"], na.rm=TRUE))` mm; Molokaʻi: `r round(storm_rp$storm_mm[storm_rp$island=="Moloka'i"])` vs. `r round(median(march_hist$march_mm[march_hist$island=="Moloka'i"], na.rm=TRUE))` mm; Oʻahu: `r round(storm_rp$storm_mm[storm_rp$island=="O'ahu"])` vs. `r round(median(march_hist$march_mm[march_hist$island=="O'ahu"], na.rm=TRUE))` mm), with stream gauge records confirming moderate, not record, flows. About half of past March events were wetter. **This is the most important finding in the report: the dramatic water-quality changes documented here — nutrients 3–6× above background, spatial hotspots at stream mouths, a measurable algae bloom — were caused by a perfectly ordinary Kona Low.** If an average storm does this, the same thing happens every year. These impacts are not rare — they have just never been measured at this spatial scale before.

---

# Land–sea connectivity: salinity–nutrient mixing

Salinity is a natural tracer of freshwater: the lower it is, the more land-sourced water is mixed in. If runoff from the storm is carrying the nutrients, we'd expect to see a clear pattern — the more freshwater at a site (lower salinity), the more nutrients. The plots below test this for all four measured nutrients. Sites with Sal ≤ 34 PSU are coloured orange (storm-influenced), while marine-background sites (Sal > 34 PSU) are blue. Zeros are excluded as below-detection readings.

```{r}
#| label: salinity-nutrient-mixing
#| fig-height: 7
#| fig-cap: "As salinity drops (more freshwater mixing in), nutrients rise — but only for the nutrients carried by runoff. Orange points and trend line: storm-influenced sites (Sal ≤ 34 PSU). Blue points and horizontal line: marine-background sites (Sal > 34 PSU) shown at their median concentration. NO₃+NO₂ and SiO₂ show steep, consistent increases as sites become more freshwater-influenced — confirming that land-sourced water is delivering these nutrients. PO₄ shows a weaker version of the same pattern. NH₃ has a flat, scattered relationship, meaning it is not being delivered by runoff in the same way — it comes from other sources (biological breakdown in sediments) that are unrelated to how much freshwater is present."

sal_zone_pal <- c("Storm-influenced\n(\u2264 34 PSU)" = "#E65100",
                  "Marine background\n(> 34 PSU)"   = "#1565C0")

mixing_long <- chem |>
  filter(!is.na(Salinity)) |>
  select(island, Salinity, NO3NO2, PO4, SiO2, NH3) |>
  pivot_longer(c(NO3NO2, PO4, SiO2, NH3),
               names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>               # drop zeros (below detection)
  mutate(
    nutrient = factor(nutrient,
      levels = c("NO3NO2", "PO4", "SiO2", "NH3"),
      labels = c("NO\u2083+NO\u2082 (\u03bcmol/L)",
                 "PO\u2084 (\u03bcmol/L)",
                 "SiO\u2082 (\u03bcmol/L)",
                 "NH\u2083 (\u03bcmol/L)")),
    sal_zone = factor(
      if_else(Salinity > 34,
              "Marine background\n(> 34 PSU)",
              "Storm-influenced\n(\u2264 34 PSU)"),
      levels = c("Storm-influenced\n(\u2264 34 PSU)",
                 "Marine background\n(> 34 PSU)"))
  )

# Median nutrient concentration for marine background sites (horizontal reference)
marine_medians <- mixing_long |>
  filter(sal_zone == "Marine background\n(> 34 PSU)") |>
  group_by(nutrient) |>
  summarise(med = median(conc, na.rm = TRUE), .groups = "drop")

ggplot(mixing_long, aes(x = Salinity, y = conc, color = sal_zone)) +
  geom_vline(xintercept = 34, linetype = "dotted",
             color = "gray40", linewidth = 0.6) +
  geom_point(size = 1.2, alpha = 0.45) +
  geom_smooth(data = \(d) filter(d, sal_zone == "Storm-influenced\n(\u2264 34 PSU)"),
              method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15,
              color = "#E65100") +
  geom_hline(data = marine_medians, aes(yintercept = med),
             color = "#1565C0", linetype = "solid", linewidth = 0.8,
             inherit.aes = FALSE) +
  scale_color_manual(values = sal_zone_pal, name = NULL) +
  scale_y_log10(labels = label_number(drop0trailing = TRUE)) +
  scale_x_continuous(limits = c(NA, 36.5)) +
  facet_wrap(~nutrient, scales = "free_y", ncol = 2) +
  labs(x = "Salinity (PSU)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

**NO₃+NO₂ and SiO₂** behave as expected if they are coming from land runoff: concentrations fall steeply as salinity rises, and level off once water becomes fully marine. This is the textbook pattern for a land-sourced contaminant diluting into the ocean. **PO₄** follows the same pattern but also remains elevated in fully marine water — suggesting a chronic background phosphorus source (likely cesspools leaching through groundwater) beyond the storm signal alone. **NH₃** shows little relationship with salinity in either zone, confirming it is not being delivered by runoff and instead accumulates through biological processes in seafloor sediments.

# Methods

## Sampling

Water samples were collected by community volunteers and scientists across O'ahu, Maui, Moloka'i, and Lāna'i between `r format(min(chem$collected_hst, na.rm = TRUE), "%B %d")` and `r format(max(chem$collected_hst, na.rm = TRUE), "%B %d, %Y")`, following the March 2026 Kona Low storm. NO₃+NO₂ was analyzed colorimetrically.

## Spatial interpolation

Ordinary kriging was used to interpolate NO₃+NO₂ concentrations between sample locations for islands with sufficient coverage (O'ahu, Maui, Moloka'i). Kriging was performed in R using the `gstat` package. Prior to kriging, concentrations were log₁p-transformed (log(x + 1)) to stabilize variance and approximate normality — a standard approach for right-skewed geochemical data.

Where multiple samples were collected at the same GPS location, values were averaged prior to kriging — duplicate coordinates produce a singular covariance matrix that causes predictions to fail. An empirical variogram was computed for each island and fit with a spherical model. Because the auto-fit variogram range (~5–6 km) is driven by local sample clustering and leaves sections of the coastline with no near-neighbor data, a minimum range of 1/5 of the island's spatial diagonal is enforced to ensure gap-free predictions across the full coast. The prediction grid was constructed as a 3 km coastal band derived by buffering the island polygon outward and subtracting the land area, then further clipped to the union of sample-point buffers (radius = variogram range). This prevents kriging from extrapolating into unsurveyed coastal stretches. The 40 nearest neighbors were used for each prediction. Back-transformed predictions (expm1) are displayed on a log₁p color scale. Sample points are colored on the same scale as the kriging surface, so extreme observed values (e.g. 265 µmol/L on O'ahu) are visible even where the smooth interpolation surface cannot reach those peaks.

Island land boundaries are derived from the Hawaiʻi ahupuaʻa dataset, which provides more accurate coastlines than Natural Earth for all four islands (Natural Earth clips East Maui at the sample extent boundary). Small errant polygon fragments (< 1 km²) in the ahupuaʻa data are removed before computing the coastal band.

---

# Salinity-based enrichment above marine background

Within the same sampling event, sites with salinity above 34 PSU were minimally affected by freshwater runoff — they represent what the ocean chemistry looked like without the storm signal. Comparing these `r sum(chem$Salinity > 34, na.rm = TRUE)` "background" sites to the `r sum(chem$Salinity < 32, na.rm = TRUE)` most storm-influenced sites (Sal < 32 PSU) gives a direct, within-study measure of how much the storm changed things.

```{r}
#| label: salinity-baseline-enrichment
#| fig-height: 6.5
#| fig-cap: "How much higher were nutrient levels in storm-influenced water compared to clean marine water sampled at the same time? (A) The fold-enrichment for each nutrient — how many times higher the typical storm-influenced concentration was relative to the marine background. Point size reflects the magnitude; error bars are 95% confidence intervals from 2,000 bootstrap resamples. SiO₂ was 6× higher, consistent with volcanic rock dissolving in freshwater and flowing into the coast. NO₃+NO₂ was 3× higher and PO₄ was about 2.3× higher, reflecting runoff from farms, cesspools, and soils. NH₃ was 2× higher but more variable, consistent with it coming partly from runoff and partly from biological processes. (B) The full distribution of concentrations in each zone. Dashed lines show the background zone means."

# ── Define salinity zones ────────────────────────────────────────────────────
sal_zones <- chem |>
  filter(!is.na(Salinity)) |>
  mutate(zone = case_when(
    Salinity > 34.0 ~ "Background\n(Sal > 34 PSU)",
    Salinity < 32.0 ~ "Storm-influenced\n(Sal < 32 PSU)",
    TRUE            ~ NA_character_
  )) |>
  filter(!is.na(zone))

bg   <- sal_zones |> filter(str_detect(zone, "Background"))
strm <- sal_zones |> filter(str_detect(zone, "Storm"))

nutrients_vec    <- c("NO3NO2", "PO4", "SiO2", "NH3")
nut_labels_ascii <- c("NO3+NO2", "PO4", "SiO2", "NH3")
zone_pal <- c(
  "Background\n(Sal > 34 PSU)"       = "#1565C0",
  "Storm-influenced\n(Sal < 32 PSU)" = "#E65100"
)

# ── Bootstrap fold-changes ───────────────────────────────────────────────────
set.seed(4492)
ci_df <- map_dfr(seq_along(nutrients_vec), function(i) {
  nut  <- nutrients_vec[i]
  bg_v <- bg[[nut]];   bg_v <- bg_v[!is.na(bg_v)]
  st_v <- strm[[nut]]; st_v <- st_v[!is.na(st_v)]
  boot <- replicate(2000, {
    median(sample(st_v, replace = TRUE)) /
    median(sample(bg_v, replace = TRUE))
  })
  tibble(
    nutrient  = nut_labels_ascii[i],
    fold      = median(st_v) / median(bg_v),
    lo        = quantile(boot, 0.025),
    hi        = quantile(boot, 0.975)
  )
}) |>
  mutate(nutrient = factor(nutrient, levels = rev(nut_labels_ascii)))

# ── Panel A: fold-change ─────────────────────────────────────────────────────
pFold <- ggplot(ci_df, aes(x = fold, y = nutrient)) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "gray50", linewidth = 0.5) +
  geom_errorbar(aes(xmin = lo, xmax = hi),
                width = 0.25, linewidth = 0.6,
                color = "#37474F", orientation = "y") +
  geom_point(aes(size = fold), color = "#E65100", shape = 16) +
  geom_text(aes(label = paste0(round(fold, 1), "\u00d7")),
            nudge_y = 0.38, size = 3.5, fontface = "bold",
            color = "#E65100") +
  scale_x_continuous(limits = c(0, 12), breaks = c(1, 2, 4, 8)) +
  scale_size_continuous(range = c(3, 8), guide = "none") +
  labs(
    x        = "Fold-enrichment  (storm-influenced \u00f7 background)",
    y        = NULL, tag = "A",
    subtitle = paste0(
      "Background: Sal > 34 PSU (n = ", nrow(bg),
      ")  |  Storm-influenced: Sal < 32 PSU (n = ", nrow(strm), ")\n",
      "Error bars = 95% bootstrap CI on median ratio (2,000 replicates)"
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor    = element_blank(),
        panel.grid.major.y  = element_blank(),
        axis.text.y  = element_text(size = 10.5, face = "bold",
                                    family = "mono"),
        plot.subtitle = element_text(size = 8.5, color = "gray40"),
        plot.tag      = element_text(face = "bold"))

# ── Panel B: distributions ───────────────────────────────────────────────────
nut_long_zones <- sal_zones |>
  select(zone, all_of(nutrients_vec)) |>
  pivot_longer(all_of(nutrients_vec),
               names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>
  mutate(
    nutrient = factor(recode(nutrient,
      "NO3NO2" = "NO3+NO2", "PO4" = "PO4",
      "SiO2"   = "SiO2",    "NH3" = "NH3"),
      levels = nut_labels_ascii),
    zone = factor(zone, levels = names(zone_pal))
  )

bg_means <- bg |>
  summarise(across(all_of(nutrients_vec), \(x) mean(x, na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "nutrient", values_to = "bg_mean") |>
  mutate(nutrient = factor(recode(nutrient,
    "NO3NO2" = "NO3+NO2", "PO4" = "PO4", "SiO2" = "SiO2", "NH3" = "NH3"),
    levels = nut_labels_ascii))

pDist <- ggplot(nut_long_zones,
                aes(x = zone, y = conc, fill = zone, color = zone)) +
  geom_hline(data = bg_means, aes(yintercept = bg_mean),
             linetype = "dashed", color = "gray30",
             linewidth = 0.5, inherit.aes = FALSE) +
  geom_violin(alpha = 0.30, linewidth = 0.3, trim = TRUE) +
  geom_boxplot(width = 0.18, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = zone_pal) +
  scale_color_manual(values = zone_pal) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  facet_wrap(~nutrient, nrow = 1, scales = "free_y") +
  labs(x = NULL, y = "\u03bcmol/L (log scale)", tag = "B") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor    = element_blank(),
        panel.grid.major.x  = element_blank(),
        axis.text.x  = element_text(angle = 30, hjust = 1, size = 8),
        strip.text   = element_text(face = "bold", size = 10.5,
                                    family = "mono"),
        legend.position = "none",
        plot.tag     = element_text(face = "bold"))

pFold / pDist + plot_layout(heights = c(1, 1.1))
ggsave(here("output","nutrients_by_salinity.png"), width = 8, height = 6)
```

**Bottom line:** storm-influenced coastal water had silica levels 6× higher, nitrogen 3× higher, and phosphorus 2.3× higher than clean marine water sampled at the same time.

A second finding sits quietly in panel B: even the "background" high-salinity sites had notably elevated nitrogen and phosphorus compared to open-ocean values. Hawaiian coastal waters carry a persistent nutrient burden from cesspools and agriculture that exists year-round, independent of storms. The storm signal is piled on top of an already-elevated baseline — which makes the total impact even more concerning.

---

# Long-term baseline comparison: Hui o Ka Wai Ola monitoring data

The Hui o Ka Wai Ola programme has sampled fixed coastal stations in West and South Maui every month since 2016. Because we know what each site normally looks like, we can ask: how far outside the historical range did each site go during March 2026? This is a stronger test than comparing sites to each other — it uses each location as its own control.

**Units note:** Hui nutrients are reported in µg/L. Conversion to µmol/L: NNN ÷ 14.007 (as N); Phosphate ÷ 30.974 (as P); Silicate ÷ 28.086 (as Si).

```{r}
#| label: hui-baseline-fn
#| include: false

# Build a fold-change + forest-plot comparison for any Hui nutrient column.
#
# Args:
#   data      – bg_hui_loc (or any frame with Station, SiteName, region, Date,
#               and a pre-computed µmol/L column)
#   conc_col  – column name string, e.g. "NNN_umol", "Phosphate_umol"
#   x_label   – x-axis label for the forest plot panel

#   fold_text_cutoff – fold-change above which bar label flips white/inside

hui_baseline_plot <- function(data,
                              conc_col,
                              x_label,
                              fold_text_cutoff = 12,
                              cs_data          = NULL,
                              cs_radius        = 100) {

  # Map Hui µmol column names to the corresponding chem column
  cs_col_map <- c(NNN_umol       = "NO3NO2", Phosphate_umol = "PO4",
                  Silicate_umol  = "SiO2",   NH4_umol       = "NH3")
  cs_col  <- cs_col_map[conc_col]
  show_cs <- !is.null(cs_data) && !is.na(cs_col) && cs_col %in% names(cs_data)

  # ── Data prep ───────────────────────────────────────────────────────────────
  historical <- data |>
    filter(year(Date) < 2026, !is.na(.data[[conc_col]]))

  march <- data |>
    filter(year(Date) == 2026, month(Date) == 3) |>
    mutate(timing = if_else(Date < as.Date("2026-03-14"),
                            "Pre-storm (Mar 10)", "Post-storm (Mar 18\u201331)"))

  baselines <- historical |>
    group_by(Station, SiteName, region) |>
    summarise(
      p05 = quantile(.data[[conc_col]], 0.05, na.rm = TRUE),
      p25 = quantile(.data[[conc_col]], 0.25, na.rm = TRUE),
      med = median(.data[[conc_col]],         na.rm = TRUE),
      p75 = quantile(.data[[conc_col]], 0.75, na.rm = TRUE),
      p95 = quantile(.data[[conc_col]], 0.95, na.rm = TRUE),
      .groups = "drop"
    )

  post <- march |>
    filter(str_detect(timing, "Post"), !is.na(.data[[conc_col]])) |>
    left_join(baselines, by = c("Station", "SiteName", "region")) |>
    filter(!is.na(med)) |>
    mutate(
      conc        = .data[[conc_col]],
      fold_change = conc / med,
      above_p95   = conc > p95,
      label       = str_trunc(SiteName, 28, ellipsis = "\u2026"),
      region      = factor(region, levels = c("South Maui", "West Maui"))
    ) |>
    arrange(region, desc(fold_change))

  pre <- march |>
    filter(str_detect(timing, "Pre"), !is.na(.data[[conc_col]])) |>
    left_join(select(baselines, Station, SiteName, region, p05, p25, med, p75, p95),
              by = c("Station", "SiteName", "region")) |>
    mutate(
      conc   = .data[[conc_col]],
      label  = str_trunc(SiteName, 28, ellipsis = "\u2026"),
      region = factor(region, levels = c("South Maui", "West Maui"))
    )

  # Station order: most anomalous at top within each region
  lvls      <- post |> arrange(region, fold_change) |> pull(label)
  post      <- post      |> mutate(label = factor(label, levels = lvls))
  pre       <- pre       |> mutate(label = factor(label, levels = lvls))
  baselines <- baselines |>
    mutate(
      label  = factor(str_trunc(SiteName, 28, ellipsis = "\u2026"), levels = lvls),
      region = factor(region, levels = c("South Maui", "West Maui"))
    )

  # ── CS spatial match: mean of post-storm CS samples within cs_radius of each station
  if (show_cs) {
    hui_stations_sf <- data |>
      distinct(Station, SiteName, region, Lat, Long) |>
      filter(!is.na(Lat), !is.na(Long)) |>
      mutate(label = factor(str_trunc(SiteName, 28, ellipsis = "\u2026"), levels = lvls)) |>
      filter(!is.na(label)) |>
      sf::st_as_sf(coords = c("Long", "Lat"), crs = 4326) |>
      sf::st_transform(32604)

    cs_sf <- cs_data |>
      filter(!is.na(.data[[cs_col]]), .data[[cs_col]] > 0,
             !is.na(latitude), !is.na(longitude),
             as.Date(collected_hst, tz = "HST") >= as.Date("2026-03-14")) |>
      mutate(cs_val = .data[[cs_col]]) |>
      sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
      sf::st_transform(32604) |>
      dplyr::select(cs_val)

    cs_matched <- sf::st_join(
      hui_stations_sf, cs_sf,
      join = sf::st_is_within_distance, dist = cs_radius
    ) |>
      sf::st_drop_geometry() |>
      filter(!is.na(cs_val)) |>
      group_by(label, region) |>
      summarise(cs_val = mean(cs_val, na.rm = TRUE), n_cs = n(), .groups = "drop")
  }

  # ── Panel A: fold-change bars ────────────────────────────────────────────────
  pA <- ggplot(post, aes(x = fold_change, y = label, fill = above_p95)) +
    geom_col(width = 0.65, color = NA) +
    geom_text(aes(label  = paste0(round(fold_change, 1), "\u00d7"),
                  hjust  = if_else(fold_change > fold_text_cutoff,  1.1, -0.1),
                  color  = if_else(fold_change > fold_text_cutoff, "white", "gray25")),
              size = 2.5, fontface = "bold") +
    geom_vline(xintercept = 1, color = "gray40", linewidth = 0.5, linetype = "dashed") +
    scale_fill_manual(values = c("TRUE" = "#E65100", "FALSE" = "#90A4AE"),
                      labels = c("TRUE" = "Above 95th pct", "FALSE" = "Within range"),
                      name   = NULL) +
    scale_color_identity() +
    scale_x_continuous(expand = expansion(mult = c(0, 0.22)),
                       breaks = c(1, 5, 10, 20, 40)) +
    facet_grid(region ~ ., scales = "free_y", space = "free_y", switch = "y") +
    labs(x = "Fold above historical median", y = NULL, tag = "A") +
    theme_minimal(base_size = 9.5) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      strip.text         = element_text(face = "bold", size = 9),
      strip.placement    = "outside",
      legend.position    = "top",
      legend.text        = element_text(size = 8.5),
      axis.text.y        = element_text(size = 7, lineheight = 0.82),
      plot.tag           = element_text(face = "bold"),
      plot.caption       = element_text(size = 8, color = "gray50", hjust = 0)
    )

  # ── Panel B: build legend values conditionally on cs availability ─────────────
  legend_vals <- c("Historical median" = "white",
                   "Hui post-storm"    = "#E65100")
  legend_ov   <- list(size   = c(2,       3),
                      shape  = c(21,      21),
                      color  = c("gray40", "#B04000"),
                      stroke = c(0.5,     0.5),
                      alpha  = c(1,       0.95))
  if (show_cs && nrow(cs_matched) > 0) {
    legend_vals["Community science (≤100 m)"] <- "#AB47BC"
    legend_ov$size   <- c(legend_ov$size,   3)
    legend_ov$shape  <- c(legend_ov$shape,  21)
    legend_ov$color  <- c(legend_ov$color,  "#4A148C")
    legend_ov$stroke <- c(legend_ov$stroke, 1.2)
    legend_ov$alpha  <- c(legend_ov$alpha,  0.85)
  }

  # ── Panel B: forest plot ─────────────────────────────────────────────
  pB <- ggplot() +
    geom_segment(data = filter(baselines, !is.na(label)),
                 aes(x = p05, xend = p95, y = label, yend = label),
                 color = "gray65", linewidth = 0.6, lineend = "round") +
    geom_segment(data = filter(baselines, !is.na(label)),
                 aes(x = p25, xend = p75, y = label, yend = label),
                 color = "gray30", linewidth = 3, lineend = "butt") +
    geom_point(data = filter(baselines, !is.na(label)),
               aes(x = med, y = label, fill = "Historical median"),
               color = "gray40", size = 1.8, shape = 21, stroke = 0.5) +
    (if (show_cs && nrow(cs_matched) > 0)
      geom_point(data = cs_matched,
                 aes(x = cs_val, y = label, fill = "Community science (≤100 m)"),
                 shape = 21, size = 2, stroke = 1.2,
                 color = "#4A148C", alpha = 0.85)) +
    geom_point(data = filter(post, !is.na(label)),
               aes(x = conc, y = label, fill = "Hui post-storm"),
               color = "#B04000", shape = 21, size = 2, stroke = 0.5, alpha = 0.95) +
    scale_fill_manual(name   = NULL,
                      values = legend_vals,
                      guide  = guide_legend(override.aes = legend_ov)) +
    scale_x_log10(breaks = scales::breaks_log(n = 6),
                  labels = scales::label_number(drop0trailing = TRUE),
                  expand = expansion(mult = c(0.02, 0.05))) +
    facet_grid(region ~ ., scales = "free_y", space = "free_y", switch = "y") +
    labs(x = x_label, y = NULL, tag = "B") +
    theme_minimal(base_size = 9.5) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      strip.text         = element_blank(),
      axis.text.y        = element_blank(),
      axis.ticks.y       = element_blank(),
      plot.tag           = element_text(face = "bold"),
      plot.margin        = margin(t = 4, r = 8, b = 4, l = 0),
      legend.position      = "bottom",
      legend.justification = "left",
      legend.text          = element_text(size = 7.5),
      legend.key.size      = unit(0.55, "lines")
    )

  # ── Assemble ─────────────────────────────────────────────────────────────────
  (pA | pB) +
    plot_layout(widths = c(1.4, 1), guides = "collect") +
    plot_annotation(
      caption = "Source: Hui o Ka Wai Ola long-term monitoring programme (Falinski et al.)."
    ) &
    theme(legend.position = "bottom", legend.justification = "center")
}
```

```{r}
#| label: hui-baseline-comparison
#| fig-height: 11
#| fig-width: 11
#| fig-cap: "(A) How many times higher was each Hui station's post-storm nitrogen reading compared to its own long-term average? Stations are ordered from most to least affected. Orange bars = readings above that site's historical 95th percentile (i.e., higher than almost anything ever recorded there). Gray = within the normal historical range. Mālā Ramp (43×) and Camp Olowalu (32×) on West Maui show the most extreme departures. (B) Where did post-storm readings land relative to each site's full history? The gray bar shows the middle 50% of historical readings; the thin line extends to the 5th–95th percentile range; the white dot is the historical median. Orange = post-storm; blue = pre-storm (March 10). Before the storm, all sites were within their normal ranges. After, most shot well above. Data: Hui o Ka Wai Ola long-term monitoring programme."

HuiNN_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "NNN_umol",
                  x_label   = "NO\u2083+NO\u2082 (\u03bcmol/L, log)",
                  cs_data   = chem)

HuiPO_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "Phosphate_umol",
                  x_label   = "PO\u2084 (\u03bcmol/L, log)",
                  cs_data   = chem)

HuiSi_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "Silicate_umol",
                  x_label   = "SiO\u2082 (\u03bcmol/L, log)",
                  cs_data   = chem)

HuiNH3_raw<-hui_baseline_plot(bg_hui_loc,
                  conc_col  = "NH4_umol",
                  x_label   = "NH\u2083 (\u03bcmol/L, log)",
                  cs_data   = chem)

HuiNN|HuiNN_raw
ggsave(here("output","allNN_HuiData.png"), width = 16, height = 8)

HuiPO|HuiPO_raw
ggsave(here("output","allPO_HuiData.png"), width = 16, height = 8)

HuiSi|HuiSi_raw
ggsave(here("output","allSi_HuiData.png"), width = 16, height = 8)

HuiNH3|HuiNH3_raw
ggsave(here("output","allNH3_HuiData.png"), width = 16, height = 8)

```

## Direct comparison: community science vs Hui within 100 m

To directly compare the two independent datasets at the same locations, we matched community science samples to Hui stations using a nearest-neighbour spatial join with a 100 m radius (UTM zone 4N). Both datasets are filtered to post-storm dates (≥ March 14, 2026). The datasets do not sample identically in time — Hui sampled March 18–31 and community science March 28–April 7 — so date difference (CS date − Hui date) is shown as point colour.

```{r}
#| label: hui-cs-comparison
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Direct comparison of post-storm community science NO₃+NO₂ (y) against Hui o Ka Wai Ola NNN (x) at the same locations (within 100 m). Each point is one Hui station × CS sample pair. The dashed line is 1:1 agreement; colour shows the gap in days between the two sample dates (positive = CS collected after Hui). Only Maui stations are shown, as Hui monitoring is restricted to Maui."

# ── Hui post-storm: one row per station-date observation ─────────────────────
hui_match_sf <- bg_hui_loc |>
  filter(year(Date) == 2026, month(Date) == 3,
         Date >= as.Date("2026-03-14"),
         !is.na(Lat), !is.na(Long)) |>
  st_as_sf(coords = c("Long", "Lat"), crs = 4326) |>
  st_transform(32604)

# ── Post-storm community science samples (Maui only, with coordinates) ────────
cs_match_sf <- chem |>
  filter(island == "Maui",
         !is.na(latitude), !is.na(longitude),
         as.Date(collected_hst, tz = "HST") >= as.Date("2026-03-14")) |>
  mutate(cs_date = as.Date(collected_hst, tz = "HST")) |>
  select(sample_id, cs_date, cs_NO3NO2 = NO3NO2, cs_PO4 = PO4,
         cs_SiO2 = SiO2, cs_NH3 = NH3, latitude, longitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

# ── Spatial join: CS samples within 100 m of each Hui observation ─────────────
matched_pairs <- st_join(
  hui_match_sf,
  cs_match_sf,
  join    = st_is_within_distance,
  dist    = 100
) |>
  st_drop_geometry() |>
  filter(!is.na(sample_id)) |>
  mutate(
    date_diff = as.integer(cs_date - Date),  # positive = WIYW collected after Hui
    dist_label = "≤ 100 m"
  )

cat(glue::glue(
  "{nrow(matched_pairs)} Hui–CS pairs matched ",
  "({n_distinct(matched_pairs$Station)} Hui stations, ",
  "{n_distinct(matched_pairs$sample_id)} CS samples)\n",
  "Date gap range: {min(matched_pairs$date_diff)} to {max(matched_pairs$date_diff)} days\n"
))

# ── Reshape to long for faceting ──────────────────────────────────────────────
nutrient_map <- tribble(
  ~hui_col,        ~cs_col,      ~label,
  "NNN_umol",      "cs_NO3NO2",  "NO\u2083+NO\u2082",
  "Phosphate_umol","cs_PO4",     "PO\u2084",
  "Silicate_umol", "cs_SiO2",    "SiO\u2082",
  "NH4_umol",      "cs_NH3",     "NH\u2084"
)

pairs_long <- nutrient_map |>
  purrr::pmap(function(hui_col, cs_col, label) {
    matched_pairs |>
      filter(!is.na(.data[[hui_col]]), !is.na(.data[[cs_col]])) |>
      transmute(
        Station, SiteName, Date, cs_date, date_diff,
        hui_val = .data[[hui_col]],
        cs_val  = .data[[cs_col]],
        nutrient = label
      )
  }) |>
  list_rbind() |>
  mutate(nutrient = factor(nutrient, levels = nutrient_map$label))

# ── Per-nutrient correlations for facet labels ────────────────────────────────
cor_labels <- pairs_long |>
  filter(hui_val > 0, cs_val > 0) |>
  group_by(nutrient) |>
  summarise(
    r   = cor(log10(hui_val), log10(cs_val), method = "pearson",  use = "complete.obs"),
    rho = cor(log10(hui_val), log10(cs_val), method = "spearman", use = "complete.obs"),
    n   = n(),
    .groups = "drop"
  ) |>
  mutate(label = paste0(
    nutrient, " (\u03bcmol/L)\n",
    "r = ", round(r, 2), "  \u03c1 = ", round(rho, 2), "  n = ", n
  ))

cor_labeller <- setNames(cor_labels$label, cor_labels$nutrient)

# ── Plot ──────────────────────────────────────────────────────────────────────
ggplot(pairs_long |>
  filter(hui_val > 0, cs_val > 0), aes(x = hui_val, y = cs_val, color = date_diff)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "gray50", linewidth = 0.5) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_gradient2(
    low = "#1565C0", mid = "#FFA726", high = "#B71C1C", midpoint = 7,
    name = "Days WIYW\nafter Hui",
    guide = guide_colorbar(barheight = unit(8, "lines"), barwidth = unit(0.5, "lines"))
  ) +
  scale_x_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  facet_wrap(~nutrient, scales = "free", ncol = 2,
             labeller = labeller(nutrient = cor_labeller)) +
  labs(
    x = "Hui o Ka Wai Ola (\u03bcmol/L, log)",
    y = "What's in your water (\u03bcmol/L, log)",
    caption = str_wrap("Direct comparison of post-storm WIYW community science NO₃+NO₂ (y) against Hui o Ka Wai Ola NNN (x) at the same locations (within 100 m). Each point is one Hui station × CS sample pair. The dashed line is 1:1 agreement; colour shows the gap in days between the two sample dates (positive = WIYW collected after Hui). Only Maui stations are shown, as Hui monitoring is restricted to Maui", width = 120)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold", size = 9.5),
    plot.caption       = element_text(size = 8, color = "gray50", hjust = 0)
  )
ggsave(here("output","comparisonHui_WIYW.png"))
```

---

# Rainfall and coastal nutrients: 10 years of Hui o Ka Wai Ola

Hui monthly monitoring gives us a 10-year window to ask whether routine rainfall events drive coastal nutrient anomalies, or whether only extreme storms like the March 2026 Kona Low cross the threshold that matters for reef ecosystems. Using full-year monthly precipitation from CHIRPS (all 12 months, 2016–2026, fetched directly from the CHC server as cloud-optimised GeoTIFFs and cropped to the Maui island-mean bounding box), we compare rainfall against station-normalised nutrient fold-changes — how far each month's reading was above or below each station's own 10-year historical median.

```{r}
#| label: chirps-full-year-maui
#| include: false
#| cache: false
#
# Full-year CHIRPS monthly means for Maui, 2016–2026.
# Data source: CHIRPS v2.0 monthly COG product, Climate Hazards Center (CHC),
#   UC Santa Barbara. https://www.chc.ucsb.edu/data/chirps
# Citation: Funk et al. (2015) The climate hazards infrared precipitation
#   with stations — a new environmental record for monitoring extremes.
#   Scientific Data 2:150066. doi:10.1038/sdata.2015.66
#
# Method: terra::rast() via /vsicurl/ reads only the Hawaii bounding box
# from each global GeoTIFF without downloading the full file. Island-mean
# precipitation is computed as the mean of all non-fill pixels within
# lat 20.45–21.15, lon −156.75 to −155.9.

library(terra)

maui_ext <- ext(-156.75, -155.9, 20.45, 21.15)

fetch_chirps_month <- function(yr, mo) {
  url <- sprintf(
    "/vsicurl/https://data.chc.ucsb.edu/products/CHIRPS-2.0/global_monthly/cogs/chirps-v2.0.%04d.%02d.cog",
    yr, mo
  )
  tryCatch({
    r    <- rast(url) |> crop(maui_ext)
    vals <- as.vector(values(r))
    vals <- vals[vals > -9000 & !is.na(vals)]
    tibble(year=yr, month=mo, chirps_mm=mean(vals, na.rm=TRUE))
  }, error=function(e) tibble(year=yr, month=mo, chirps_mm=NA_real_))
}

chirps_full_cache <- here::here("data","chirps_maui_monthly_full.csv")

if (file.exists(chirps_full_cache)) {
  chirps_full_maui <- read_csv(chirps_full_cache, show_col_types=FALSE) |>
    mutate(ym = as.Date(paste(year, month, "1", sep="-")))
} else {
  yms <- expand_grid(year=2016:2026, month=1:12) |>
    filter(!(year==2026 & month > 7))
  chirps_full_maui <- map2_dfr(yms$year, yms$month, fetch_chirps_month) |>
    mutate(ym = as.Date(paste(year, month, "1", sep="-")))
  write_csv(chirps_full_maui, chirps_full_cache)
}
```

```{r}
#| label: hui-rainfall-nutrient
#| fig-height: 6
#| fig-cap: "(A) CHIRPS island-mean monthly rainfall for Maui, 2016–2026, all 12 months. Dark blue bars = top 10% wettest months. March 2026 (373 mm) is 35% wetter than any prior month in the record. (B) Each point is one monthly Hui o Ka Wai Ola sampling event at all West and South Maui stations, showing median NO₃+NO₂ fold-change from each station's own historical baseline vs that month's CHIRPS rainfall (log scale). A statistically significant but modest log-linear relationship holds across the full decade (r² = 0.058, p = 0.009), confirming that wetter months produce proportionally elevated nitrogen. The March 2026 Kona Low sits far outside the envelope of any prior event — the only month to exceed 200 mm in the record and the only one to produce a >2× median fold-change."

# Station-level baselines (pre-2026)
station_baselines_hn <- bg_hui_loc |>
  filter(!is.na(NNN_umol), NNN_umol > 0, year(Date) < 2026) |>
  group_by(Station) |>
  summarise(hist_med_NNN  = median(NNN_umol,       na.rm=TRUE),
            hist_med_SiO2 = median(Silicate_umol,  na.rm=TRUE),
            hist_med_PO4  = median(Phosphate_umol, na.rm=TRUE),
            n_obs         = n(), .groups="drop") |>
  filter(n_obs >= 5)

# Monthly fold-change from station baseline
hui_anomaly_hn <- bg_hui_loc |>
  filter(!is.na(NNN_umol), NNN_umol > 0) |>
  inner_join(station_baselines_hn, by="Station") |>
  mutate(year  = year(Date),  month = month(Date),
         ym    = floor_date(Date,"month"),
         fold_NNN  = NNN_umol      / hist_med_NNN,
         fold_SiO2 = Silicate_umol / hist_med_SiO2,
         fold_PO4  = Phosphate_umol/ hist_med_PO4) |>
  group_by(ym, year, month) |>
  summarise(fold_NNN_med  = median(fold_NNN,  na.rm=TRUE),
            fold_SiO2_med = median(fold_SiO2, na.rm=TRUE),
            fold_PO4_med  = median(fold_PO4,  na.rm=TRUE),
            .groups="drop") |>
  left_join(chirps_full_maui |> select(ym, chirps_mm), by="ym")

# Model
lm_nnn_hn <- lm(fold_NNN_med ~ log1p(chirps_mm),
                 data=filter(hui_anomaly_hn, !is.na(chirps_mm)))
r2_hn <- round(summary(lm_nnn_hn)$r.squared, 3)
p_hn  <- round(summary(lm_nnn_hn)$coefficients[2,"Pr(>|t|)"], 4)

rain_q90 <- quantile(chirps_full_maui$chirps_mm, .90, na.rm=TRUE)

p_rain_hn <- ggplot(chirps_full_maui, aes(x=ym, y=chirps_mm)) +
  geom_col(aes(fill=chirps_mm >= rain_q90), width=25, show.legend=FALSE) +
  scale_fill_manual(values=c("FALSE"="#90CAF9","TRUE"="#1565C0")) +
  annotate("text", x=as.Date("2026-03-01"),
           y=max(chirps_full_maui$chirps_mm,na.rm=TRUE)*0.9,
           label="Mar 2026\n373 mm", size=2.8, color="#1565C0",
           hjust=0.5, fontface="italic", lineheight=0.85) +
  scale_x_date(date_breaks="1 year", date_labels="%Y") +
  labs(x=NULL, y="Rainfall (mm)", tag="A",
       subtitle=paste0("CHIRPS island-mean · all months 2016\u20132026  |  ",
                       "median ", round(median(chirps_full_maui$chirps_mm,na.rm=TRUE)),
                       " mm \u00b7 max 373 mm (Mar 2026)")) +
  theme_minimal(base_size=10) +
  theme(panel.grid.minor=element_blank(), plot.tag=element_text(face="bold"),
        plot.subtitle=element_text(size=8, color="gray40"))

p_scatter_hn <- ggplot(
  hui_anomaly_hn |> filter(!is.na(chirps_mm)),
  aes(x=log1p(chirps_mm), y=fold_NNN_med)
) +
  geom_hline(yintercept=1, linetype="dashed", color="gray60", linewidth=0.4) +
  geom_point(aes(color=year>=2023), size=2.3, alpha=0.7) +
  geom_smooth(method="lm", se=TRUE, color="#E65100",
              fill="#FFE0CC", linewidth=0.9) +
  ggrepel::geom_text_repel(
    data=\(d) filter(d, year==2026, month==3),
    aes(label="Mar 2026\nKona Low"), size=2.6, color="#1565C0",
    lineheight=0.85, min.segment.length=0.1) +
  scale_color_manual(values=c("FALSE"="#90A4AE","TRUE"="#E65100"), guide="none") +
  annotate("text", x=3.2, y=max(hui_anomaly_hn$fold_NNN_med,na.rm=TRUE)*0.94,
           label=paste0("r\u00b2 = ",r2_hn,"\np = ",p_hn),
           size=2.9, color="#E65100", hjust=0, lineheight=1.1) +
  scale_x_continuous(
    breaks=log1p(c(0,10,25,50,100,200,400)),
    labels=c("0","10","25","50","100","200","400")) +
  labs(x="Monthly rainfall, log scale (mm, CHIRPS island-mean)",
       y="NO\u2083+NO\u2082 fold-change\nfrom station baseline", tag="B",
       subtitle="Each point = one monthly Hui sampling  \u00b7  log scale removes right-skew") +
  theme_minimal(base_size=10) +
  theme(panel.grid.minor=element_blank(), plot.tag=element_text(face="bold"),
        plot.subtitle=element_text(size=8, color="gray40"))

p_rain_hn / p_scatter_hn +
  plot_annotation(
    title="Maui coastal nitrogen responds to rainfall, but only extreme storms drive ecologically meaningful anomalies",
    subtitle="Hui o Ka Wai Ola fixed stations 2016\u20132026 \u00b7 station-normalised fold-change from 10-year baseline",
    theme=theme(plot.title=element_text(face="bold", size=11),
                plot.subtitle=element_text(size=8.5, color="gray40"))
  )
```

The log-linear relationship (r² = `r r2_hn`, p = `r p_hn`) confirms that monthly coastal nitrogen at Hui stations scales with antecedent rainfall, but the effect size is modest for routine events. The chronic nutrient load from cesspools and groundwater creates a persistently elevated baseline that obscures smaller storm signals at fixed leeward stations. Only the March 2026 Kona Low — with 373 mm, the wettest month in the 10-year record and 35% wetter than any previous event — produced a fold-change that clearly separates from this chronic background noise. This threshold behaviour is ecologically important: it means that routine Kona Low events may not substantially elevate nutrients above the already-elevated baseline at these semi-sheltered monitoring stations, while a single extreme event can deliver a nutrient pulse that is qualitatively different from anything observed in a decade of monthly monitoring.

---

# Stoichiometric fingerprints — Claim 2

Did all islands produce the same kind of runoff, or does each island have its own chemical identity? The ratios of nutrients to each other — nitrogen to phosphorus (N:P) and silica to nitrogen (Si:N) — act like a fingerprint that reveals where nutrients came from. Ocean plankton maintain nitrogen and phosphorus in a fairly fixed ratio of 16:1 (the Redfield ratio). Below 16 means unusual phosphorus enrichment, pointing to a cesspool or wastewater source. Above 1 for Si:N means silica-rich volcanic rock weathering is dominant. The four islands show distinctly different ratios, meaning the storm flushed island-specific chemical fingerprints — not a single uniform nutrient pulse.

```{r}
#| label: stoichiometry-figure
#| fig-height: 5.5
#| fig-cap: "(A) Nitrogen-to-phosphorus ratio (N:P) by island. Ocean plankton normally maintain N:P near 16:1 (dashed line) — below 16 means there is more phosphorus than biology expects, pointing to wastewater or cesspool inputs. All four islands are below 16. (B) Silica-to-nitrogen ratio (Si:N). Values above 1 (dashed) indicate the water is carrying more dissolved rock material (silica from volcanic weathering) than nitrogen — the signature of storm runoff over Hawaiian terrain. (C) Both ratios together, with island medians shown as diamonds. All islands sit in the phosphorus-excess, silica-rich corner. Maui's ratio is closest to the biological reference line, suggesting its runoff has more agricultural nitrogen (which drives N:P up toward 16). Oʻahu and Molokaʻi are farther into the phosphorus-excess zone, consistent with denser cesspool infrastructure."


ratios <- chem |>
  filter(!is.na(island), NO3NO2 > 0, PO4 > 0, SiO2 > 0) |>
  mutate(
    NP     = NO3NO2 / PO4,
    SiN    = SiO2   / NO3NO2,
    island = factor(island, levels = c("O'ahu", "Maui", "Moloka'i", "Lāna'i"))
  ) |>
  filter(!is.na(island))

violin_theme <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.title.x = element_blank(), axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

pNP_s <- ggplot(ratios, aes(x = island, y = NP, fill = island, color = island)) +
  geom_hline(yintercept = 16, linetype = "dashed", color = "gray40", linewidth = 0.6) +
  annotate("text", x = Inf, y = 20, label = "Redfield N:P = 16",
           size = 2.7, color = "gray40", hjust = 1.05) +
  geom_violin(alpha = 0.30, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE, fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values = island_pal) + scale_color_manual(values = island_pal) +
  scale_y_log10(breaks = c(0.1, 1, 10, 100),
                labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = "N:P (molar, log)", tag = "A") + violin_theme

pSiN_s <- ggplot(ratios, aes(x = island, y = SiN, fill = island, color = island)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40", linewidth = 0.6) +
  annotate("text", x = Inf, y = 1.35, label = "Diatom Si:N = 1",
           size = 2.7, color = "gray40", hjust = 1.05) +
  geom_violin(alpha = 0.30, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE, fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values = island_pal) + scale_color_manual(values = island_pal) +
  scale_y_log10(breaks = c(0.1, 1, 10, 100, 1000),
                labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = "Si:N (molar, log)", tag = "B") + violin_theme

centroids_s <- ratios |>
  group_by(island) |>
  summarise(NP = median(NP, na.rm = TRUE), SiN = median(SiN, na.rm = TRUE), .groups = "drop")

pScatter_s <- ggplot(ratios, aes(x = NP, y = SiN, color = island)) +
  geom_vline(xintercept = 16, linetype = "dashed", color = "gray40", linewidth = 0.5) +
  geom_hline(yintercept = 1,  linetype = "dashed", color = "gray40", linewidth = 0.5) +
  annotate("text", x = 0.12, y = 0.13, label = "P-excess / Si-limited",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  annotate("text", x = 25,   y = 0.13, label = "N-excess / Si-limited",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  annotate("text", x = 0.12, y = 200,  label = "P-excess / Si-rich",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  annotate("text", x = 25,   y = 200,  label = "N-excess / Si-rich",
           size = 2.3, color = "gray50", hjust = 0, fontface = "italic") +
  geom_point(size = 1.1, alpha = 0.4) +
  geom_point(data = centroids_s, size = 5.5, shape = 18, alpha = 0.95) +
  ggrepel::geom_label_repel(
    data = centroids_s, aes(label = island),
    size = 3, fontface = "bold", label.padding = unit(0.15, "lines"),
    label.size = 0.2, show.legend = FALSE
  ) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_x_log10(breaks = c(0.1, 1, 10, 100),
                labels = scales::label_number(drop0trailing = TRUE)) +
  scale_y_log10(breaks = c(0.1, 1, 10, 100, 1000),
                labels = scales::label_number(drop0trailing = TRUE)) +
  labs(x = "N:P (molar, log)", y = "Si:N (molar, log)", tag = "C") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "none")

title_row_s <- ggdraw() +
  draw_label("Stoichiometric fingerprints distinguish island nutrient sources",
             fontface = "bold", size = 11, x = 0.01, hjust = 0) +
  draw_label("Dashed lines = Redfield N:P (16) and diatom Si:N (1). Diamonds = island medians.",
             color = "gray40", size = 9, x = 0.01, y = 0.2, hjust = 0)

plot_grid(
  title_row_s,
  plot_grid(pNP_s, pSiN_s, pScatter_s, nrow = 1, rel_widths = c(1, 1, 1.4)),
  ncol = 1, rel_heights = c(0.12, 1)
)
```

Maui's N:P ratio (median `r round(median(ratios$NP[ratios$island == "Maui"], na.rm=TRUE), 1)`) is noticeably higher than Oʻahu (`r round(median(ratios$NP[ratios$island == "O'ahu"], na.rm=TRUE), 1)`) and Molokaʻi (`r round(median(ratios$NP[ratios$island == "Moloka'i"], na.rm=TRUE), 1)`). That gap is consistent with East Maui's large farming areas contributing extra nitrogen — bumping the N:P ratio closer to the 16:1 biological baseline even while still remaining phosphorus-enriched. Every island has Si:N ratios well above 1, meaning volcanic rock dissolution is a dominant contributor across the board — an expected signature of Hawaiian watersheds. Each island's runoff tells a different story about its land use.

---

# Biogeochemical cluster analysis — Claims 2 & 4

We sorted the 694 sample sites into groups (using a statistical clustering method called PAM) based purely on their water chemistry: salinity, nitrogen, phosphorus, and silica. Four distinct water types emerged. The number four was chosen because it was the smallest number that captured all four clearly different kinds of coastal water without over-splitting into groups that were chemically too similar to distinguish. Testing two to six groups confirmed this choice.

```{r}
#| label: cluster-analysis
#| fig-height: 11
#| fig-width: 10
#| fig-cap: "(A) Chemistry map of all 694 sites. The two axes (Principal Components) summarise the four measured variables: left-right roughly represents the land–sea gradient (right = cleaner ocean water, left = more land-influenced); up-down separates nitrogen-heavy from phosphorus-heavy enrichment. Coloured hulls enclose each of the four water types. (B) Chemical fingerprint for each group. Cell colours show whether a variable is above (red) or below (blue) the overall average — each water type has a distinct combination. Numbers are the actual median concentrations. (C) Which islands are in each group? Storm runoff sites are disproportionately Maui; marine background is dominated by Oʻahu. (D) Cesspool density in the watershed surrounding each site — runoff-affected groups sit in areas with more cesspool infrastructure. (E) Where on the map does each water type appear?"

# ── Prepare and cluster ──────────────────────────────────────────────────────
chem_clust <- chem |>
  filter(!is.na(island),
         !is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  mutate(log_NO3NO2 = log1p(NO3NO2),
         log_PO4    = log1p(PO4),
         log_SiO2   = log1p(SiO2))

clust_mat <- chem_clust |>
  select(Salinity, log_NO3NO2, log_PO4, log_SiO2) |>
  scale()

set.seed(8341)
pam4 <- pam(clust_mat, k = 4, pamonce = 5)

cluster_labels <- c("1" = "Marine background", "2" = "Storm runoff",
                    "3" = "Nearshore enriched",  "4" = "Freshwater plume")
cluster_pal_named <- c(
  "Marine background"  = "#1565C0", "Storm runoff"      = "#E65100",
  "Nearshore enriched" = "#2E7D32", "Freshwater plume"  = "#6A1B9A"
)

chem_clust <- chem_clust |>
  mutate(cluster       = factor(pam4$clustering),
         cluster_label = factor(cluster_labels[cluster],
                                levels = cluster_labels))

# ── PCA ─────────────────────────────────────────────────────────────────────
pca_res <- prcomp(clust_mat, scale. = FALSE)
pc_scores <- as.data.frame(pca_res$x[, 1:2]) |>
  bind_cols(chem_clust |> select(cluster, cluster_label, island))
pct_var <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

loadings <- as.data.frame(pca_res$rotation[, 1:2]) |>
  rownames_to_column("variable") |>
  mutate(variable = recode(variable, "log_NO3NO2" = "NO3+NO2",
                           "log_PO4" = "PO4", "log_SiO2" = "SiO2"),
         scale = 3.2)

hulls <- pc_scores |>
  group_by(cluster_label) |>
  slice(chull(PC1, PC2))

pA_cl <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = cluster_label)) +
  geom_polygon(data = hulls, aes(fill = cluster_label),
               alpha = 0.08, linewidth = 0.3, color = NA) +
  geom_point(size = 1.2, alpha = 0.6) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1*scale, yend = PC2*scale),
               color = "gray25", linewidth = 0.6,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed"),
               inherit.aes = FALSE) +
  geom_label(data = loadings,
             aes(x = PC1*scale*1.15, y = PC2*scale*1.15, label = variable),
             size = 2.9, color = "gray20", linewidth = 0, fill = "white",
             fontface = "bold", inherit.aes = FALSE) +
  scale_color_manual(values = cluster_pal_named, name = NULL) +
  scale_fill_manual(values  = cluster_pal_named, name = NULL) +
  labs(x = paste0("PC1 (", pct_var[1], "% variance)"),
       y = paste0("PC2 (", pct_var[2], "% variance)"), tag = "A") +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid.minor = element_blank(), legend.position = "none",
        plot.tag = element_text(face = "bold"))

# ── Heatmap ──────────────────────────────────────────────────────────────────
fp_vars   <- c("Salinity", "NO3NO2", "PO4", "SiO2")
fp_labels <- c("Salinity\n(PSU)", "NO3+NO2\n(µmol/L)", "PO4\n(µmol/L)", "SiO2\n(µmol/L)")

hm_df <- chem_clust |>
  group_by(cluster_label) |>
  summarise(across(all_of(fp_vars), ~median(., na.rm = TRUE)), .groups = "drop") |>
  pivot_longer(-cluster_label, names_to = "variable", values_to = "median_val") |>
  group_by(variable) |>
  mutate(z = (median_val - mean(median_val)) / sd(median_val)) |>
  ungroup() |>
  mutate(
    variable      = factor(variable, levels = fp_vars, labels = fp_labels),
    cluster_label = factor(cluster_label, levels = rev(cluster_labels)),
    val_label     = case_when(
      variable == "Salinity\n(PSU)"    ~ as.character(round(median_val, 1)),
      variable == "NO3+NO2\n(µmol/L)" ~ as.character(round(median_val, 2)),
      variable == "PO4\n(µmol/L)"     ~ as.character(round(median_val, 3)),
      TRUE                             ~ as.character(round(median_val, 1))
    )
  )

pB_cl <- ggplot(hm_df, aes(x = variable, y = cluster_label, fill = z)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = val_label), size = 3, fontface = "bold",
            color = ifelse(abs(hm_df$z) > 1.2, "white", "gray20")) +
  scale_fill_gradient2(low = "#1565C0", mid = "white", high = "#B71C1C",
                       midpoint = 0, name = "Z-score") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = \(x) str_wrap(x, 14)) +
  labs(x = NULL, y = NULL, tag = "B") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), axis.text.x = element_text(size = 8.5, face = "bold"),
        axis.text.y = element_text(size = 8.5), plot.tag = element_text(face = "bold"))

# ── Environmental drivers ─────────────────────────────────────────────────────
# chem already carries ahupuaa / cesspool joins from the load-data chunk
env_df <- chem_clust

pC_cl <- env_df |>
  filter(!is.na(island)) |>
  count(cluster_label, island) |>
  group_by(cluster_label) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  filter(island %in% c("O'ahu", "Maui", "Moloka'i", "Lāna'i")) |>
  mutate(island = factor(island, levels = c("O'ahu", "Maui", "Moloka'i", "Lāna'i"))) |>
  ggplot(aes(x = cluster_label, y = pct, fill = island)) +
  geom_col(width = 0.7, color = "white", linewidth = 0.3) +
  scale_fill_manual(values = c("O'ahu" = "#1565C0", "Maui" = "#2E7D32",
                               "Moloka'i" = "#E65100", "Lāna'i" = "#6A1B9A"), name = NULL) +
  scale_y_continuous(labels = scales::label_percent(), expand = c(0, 0)) +
  scale_x_discrete(labels = \(x) str_wrap(x, 10)) +
  labs(x = NULL, y = "Proportion of sites", tag = "C") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        legend.position = "right", plot.tag = element_text(face = "bold"))

pD_cl <- env_df |>
  filter(!is.na(cesspool_density_per_km2)) |>
  ggplot(aes(x = cluster_label, y = cesspool_density_per_km2 + 0.01,
             fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.3, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE, fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = cluster_pal_named) +
  scale_color_manual(values = cluster_pal_named) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  scale_x_discrete(labels = \(x) str_wrap(x, 10)) +
  labs(x = NULL, y = "Cesspool density\n(per km², log)", tag = "D") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        legend.position = "none", plot.tag = element_text(face = "bold"))

# ── Spatial maps ─────────────────────────────────────────────────────────────
clust_sf <- chem_clust |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

map_cluster_fn <- function(island_name, title = island_name) {
  match_idx <- normalize_name(ahupuaa_raw$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_raw[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union  <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  coast_bbox  <- st_bbox(st_buffer(land_union, 3500))
  pts <- clust_sf |> filter(island == island_name)
  ggplot() +
    geom_sf(data = st_as_sf(st_sf(geometry = land_union)),
            fill = "gray78", color = "gray50", linewidth = 0.3) +
    geom_sf(data = pts, aes(color = cluster_label),
            size = ifelse(island_name == "O'ahu", 1.5, 2.2), alpha = 0.85, shape = 16) +
    scale_color_manual(values = cluster_pal_named, name = NULL, drop = FALSE) +
    coord_sf(crs = 32604, xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = title) +
    theme_void(base_size = 9) +
    theme(plot.title = element_text(face = "bold", size = 9, hjust = 0.5),
          legend.position = "none", plot.margin = margin(2, 2, 2, 2))
}

maps_row <- plot_grid(
  map_cluster_fn("O'ahu", "Oʻahu"), map_cluster_fn("Maui", "Maui"),
  map_cluster_fn("Moloka'i", "Molokaʻi"), map_cluster_fn("Lāna'i", "Lānaʻi"),
  nrow = 1, rel_widths = c(1.6, 1.4, 0.9, 0.7)
)

# ── Assemble ─────────────────────────────────────────────────────────────────
legend_strip <- get_legend(
  pA_cl + theme(legend.position = "bottom", legend.text = element_text(size = 9),
                legend.key.size = unit(0.9, "lines"))
)

plot_grid(
  ggdraw() + draw_label(
    "Four biogeochemical clusters reveal distinct post-storm coastal water types",
    fontface = "bold", size = 11, x = 0.01, hjust = 0) +
    draw_label(
      paste0("PAM clustering on scaled log-nutrients + salinity (k = 4, n = ",
             nrow(chem_clust), " sites). Convex hulls in A = envelopes per cluster."),
      color = "gray40", size = 8.5, x = 0.01, y = 0.22, hjust = 0),
  plot_grid(
    plot_grid(pA_cl, pB_cl, nrow = 1, rel_widths = c(1.3, 1)),
    legend_strip,
    plot_grid(pC_cl, pD_cl, nrow = 1),
    ncol = 1, rel_heights = c(1, 0.07, 0.75)
  ),
  plot_grid(
    ggdraw() + draw_label("Spatial distribution of biogeochemical clusters",
                          fontface = "bold", size = 9.5, x = 0.01, hjust = 0),
    maps_row, ncol = 1, rel_heights = c(0.1, 1)
  ),
  ncol = 1, rel_heights = c(0.06, 1.5, 0.65)
)
```

Four clearly distinct water types emerged from the grouping analysis. **Marine background** (n = `r sum(chem_clust$cluster == "1")`) is clean ocean water — high salinity, low nutrients — concentrated around Oʻahu's less impacted shores. **Storm runoff** (n = `r sum(chem_clust$cluster == "2")`) sites have extremely high nitrogen and silica with moderate freshwater dilution, the chemical signature of storm drainage carrying dissolved volcanic rock and soil nitrogen; most of these sites are on Maui's windward coast. **Nearshore enriched** (n = `r sum(chem_clust$cluster == "3")`) is the most common type — a broad mixing zone where nutrients are elevated but salinity hasn't dropped dramatically. **Freshwater plume** (n = `r sum(chem_clust$cluster == "4")`) sites sit right at stream mouths, with the lowest salinity (median 20.8 PSU) and extreme silica levels (median 48.1 µmol/L) — these are the most strongly land-influenced sites in the dataset. The areas with the most cesspools tend to have the most runoff-affected water types, indicating that wastewater infrastructure amplifies the nutrient signal beyond what natural soil flushing would produce alone.

Ammonium (NH₃) was not included in the grouping analysis. The figure below shows why: nitrogen (NO₃+NO₂) increases cleanly from marine background to freshwater plume, with almost no overlap between groups. NH₃, by contrast, is scattered across all four groups with overlapping ranges — it doesn't distinguish them. Adding it to the analysis weakened the group definitions (cluster quality dropped by 27%) and excluded 120 sites that lacked NH₃ measurements. NH₃ appears to build up through seafloor biological processes that are independent of storm runoff.

```{r}
#| label: nh3-posthoc
#| fig-height: 4.5
#| fig-cap: "Why ammonium (NH₃) was left out of the grouping analysis. (A) NO₃+NO₂ by water type — each group is clearly separated from the others, with almost no overlap. This is what a useful grouping variable looks like. (B) NH₃ by water type — the distributions overlap substantially across all four groups. Including NH₃ would add noise, not clarity. NH₃ data available for `r sum(!is.na(chem_clust$NH3))` of `r nrow(chem_clust)` sites."

clust_pal_ch <- c(
  "Marine background"  = "#1565C0",
  "Storm runoff"       = "#E65100",
  "Nearshore enriched" = "#2E7D32",
  "Freshwater plume"   = "#6A1B9A"
)

vtheme_nh3 <- theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position    = "none",
    axis.title.x       = element_blank(),
    axis.text.x        = element_text(angle = 25, hjust = 1, size = 9),
    plot.subtitle      = element_text(size = 8.5, color = "gray40",
                                      margin = margin(t = 2, b = 4)),
    plot.tag           = element_text(face = "bold")
  )

nh3_df_ch <- chem_clust |>
  filter(!is.na(NH3), NH3 > 0) |>
  mutate(cluster_label = factor(cluster_label, levels = names(clust_pal_ch)))

no3_df_ch <- chem_clust |>
  filter(!is.na(NO3NO2), NO3NO2 > 0) |>
  mutate(cluster_label = factor(cluster_label, levels = names(clust_pal_ch)))

pNO3_ch <- ggplot(no3_df_ch, aes(x = cluster_label, y = NO3NO2,
                                   fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.25, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = clust_pal_ch) +
  scale_color_manual(values = clust_pal_ch) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = "NO\u2083+NO\u2082 (\u03bcmol/L, log)", tag = "A",
       subtitle = "Ordered separation across clusters (structuring variable)") +
  vtheme_nh3

pNH3_ch <- ggplot(nh3_df_ch, aes(x = cluster_label, y = NH3,
                                   fill = cluster_label, color = cluster_label)) +
  geom_violin(alpha = 0.25, linewidth = 0.35, trim = TRUE) +
  geom_boxplot(width = 0.15, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = clust_pal_ch) +
  scale_color_manual(values = clust_pal_ch) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(y = "NH\u2083 (\u03bcmol/L, log)", tag = "B",
       subtitle = "No systematic pattern \u2014 cluster-independent (post-hoc)") +
  vtheme_nh3

pNO3_ch | pNH3_ch
```

---

# Supplementary: cluster selection rationale

```{r}
#| label: silhouette-supp
#| fig-height: 5
#| fig-cap: "How was four groups chosen? (A) Silhouette width measures how well each site fits its assigned group versus how closely it matches neighbouring groups (higher = better-defined groups). Two groups scores highest, but splitting the coast into just 'freshwater' and 'marine' collapses the ecologically meaningful gradient into a binary that loses most of the information. Four groups (highlighted) is the smallest number that separates all four chemically distinct water types while still being well-defined. (B) Silhouette width for each of the 694 individual sites in the k = 4 solution. Negative values (sites that fit a neighbouring group slightly better) are concentrated at the Storm runoff / Nearshore enriched boundary — those two groups shade into each other chemically, which is ecologically expected for a continuous mixing gradient rather than a sharp boundary."

# clust_mat and pam4 are computed in the cluster-analysis chunk above
set.seed(8341)
sil_summary_s <- map_dfr(2:6, function(k) {
  pk  <- pam(clust_mat, k = k, pamonce = 5)
  sil <- silhouette(pk$clustering, dist(clust_mat))
  tibble(k = k, mean_sil = mean(sil[, 3]))
})

pSilA_s <- ggplot(sil_summary_s, aes(x = k, y = mean_sil)) +
  geom_line(color = "gray40", linewidth = 0.8) +
  geom_point(size = 3.5, color = "gray30") +
  geom_point(data = filter(sil_summary_s, k == 4), size = 5.5, shape = 21,
             fill = "#E65100", color = "white", stroke = 1.5) +
  annotate("text", x = 4.2, y = filter(sil_summary_s, k == 4)$mean_sil + 0.008,
           label = "k = 4\n(chosen)", hjust = 0, size = 3.2, color = "#E65100") +
  scale_x_continuous(breaks = 2:6) +
  scale_y_continuous(limits = c(0.18, 0.43)) +
  labs(x = "Number of clusters (k)", y = "Mean silhouette width",
       title = "A  Average silhouette width by k",
       subtitle = "Higher = more compact, better-separated clusters") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

sil4_s <- as.data.frame(silhouette(pam4$clustering, dist(clust_mat))) |>
  mutate(
    cluster_label = factor(cluster_labels[cluster], levels = cluster_labels)
  ) |>
  arrange(cluster, sil_width) |>
  mutate(sorted_id = row_number())

pSilB_s <- ggplot(sil4_s, aes(x = sorted_id, y = sil_width, fill = cluster_label)) +
  geom_col(width = 1, color = NA) +
  geom_hline(yintercept = 0, color = "gray30", linewidth = 0.5) +
  geom_hline(yintercept = mean(sil4_s$sil_width),
             linetype = "dashed", color = "gray40", linewidth = 0.5) +
  annotate("text", x = nrow(sil4_s) * 0.98,
           y = mean(sil4_s$sil_width) + 0.03,
           label = sprintf("Mean = %.2f", mean(sil4_s$sil_width)),
           hjust = 1, size = 3, color = "gray30") +
  scale_fill_manual(values = cluster_pal_named, name = NULL) +
  scale_y_continuous(limits = c(-0.25, 1)) +
  labs(x = "Sites (sorted within cluster)", y = "Silhouette width",
       title = "B  Individual silhouette widths, k = 4",
       subtitle = "Negative values = misclassified sites") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid  = element_blank(), legend.position = "bottom",
        legend.text = element_text(size = 9), legend.key.size = unit(0.8, "lines"),
        plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

pSilA_s | pSilB_s
```

---

# Cluster–kriging overlay: where do fingerprints grade into hotspots?

The interpolated nitrogen map gives a continuous picture across the coast; the cluster labels give each site a chemical identity. Overlaying them shows whether the water types line up with the spatial hotspots — or whether the chemistry and geography tell different stories.

```{r}
#| label: krige-cluster-overlay
#| fig-height: 6.5
#| fig-cap: "The nitrogen hotspot map with each sample's water type shown as a coloured circle. Background colour (yellow-green = high, dark purple = low) is the interpolated nitrogen map from community science samples. Circle colour shows which of the four water types that site belongs to. The two line up: clean marine-background sites (blue) sit in the dark, low-nitrogen margins; storm runoff (orange) and freshwater plume (purple) sites cluster in the brightest, highest-nitrogen patches, especially on Maui's north coast and Oʻahu's northwest shore. The transition group (nearshore enriched, green) forms a belt between the hotspot core and the clean outer water."

# krige_surface_data helper — same kriging logic as krige_island_map but
# returns components rather than a ggplot, allowing dual color scales.
krige_surface_data <- function(data, island_name, ahupuaa_sf,
                               res_m = 300, band_km = 3) {
  island_data <- data |>
    filter(island == island_name, !is.na(NO3NO2), !is.na(latitude), !is.na(longitude))
  match_idx  <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union  <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  land_sf_out <- st_as_sf(st_sf(geometry = land_union))
  coast_outer <- st_buffer(land_union, dist = band_km * 1000)
  coast_band  <- st_difference(coast_outer, land_union)
  coast_bbox  <- st_bbox(coast_outer)
  pts_proj <- island_data |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(log_val = log1p(NO3NO2),
           X = st_coordinates(geometry)[, 1], Y = st_coordinates(geometry)[, 2]) |>
    group_by(X, Y) |>
    summarise(log_val = mean(log_val, na.rm = TRUE),
              NO3NO2  = mean(NO3NO2,  na.rm = TRUE), .groups = "drop") |>
    st_as_sf(coords = c("X","Y"), crs = 32604)
  island_diag <- sqrt(diff(st_bbox(pts_proj)[c("xmin","xmax")])^2 +
                      diff(st_bbox(pts_proj)[c("ymin","ymax")])^2)
  min_range <- island_diag / 5
  v_emp  <- variogram(log_val ~ 1, pts_proj)
  v_auto <- tryCatch(
    fit.variogram(v_emp, vgm(var(pts_proj$log_val)*0.8, "Sph",
                              min_range, var(pts_proj$log_val)*0.2)),
    error = function(e) NULL)
  v_fit <- if (!is.null(v_auto) && v_auto$range[2] >= min_range) v_auto else
    vgm(var(pts_proj$log_val)*0.8, "Sph", min_range, var(pts_proj$log_val)*0.2)
  sample_cov <- st_buffer(st_union(pts_proj), dist = min_range)
  grid_pts   <- st_make_grid(coast_band, cellsize = res_m, what = "centers") |>
    st_as_sf() |> st_filter(coast_band) |> st_filter(sample_cov)
  k_out   <- suppressWarnings(krige(log_val ~ 1, pts_proj, grid_pts,
                                     model = v_fit, nmax = 40))
  pred_df <- st_coordinates(k_out) |> as.data.frame() |>
    mutate(NO3NO2_pred = expm1(k_out$var1.pred))
  list(pred_df = pred_df, land_sf = land_sf_out, coast_bbox = coast_bbox)
}

# Attach cluster labels to chem (chem already has the chemistry needed for PAM)
chem_clust_overlay <- chem |>
  filter(!is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  mutate(log_NO3NO2 = log1p(NO3NO2), log_PO4 = log1p(PO4), log_SiO2 = log1p(SiO2))

set.seed(8341)
pam4_ov <- pam(
  chem_clust_overlay |> select(Salinity, log_NO3NO2, log_PO4, log_SiO2) |> scale(),
  k = 4, pamonce = 5
)
chem_clust_overlay <- chem_clust_overlay |>
  mutate(cluster       = factor(pam4_ov$clustering),
         cluster_label = factor(cluster_labels[cluster], levels = cluster_labels))

krige_cluster_overlay_fn <- function(island_name, title = island_name, pt_size = 1.8) {
  kd  <- krige_surface_data(chem_clust_overlay, island_name, ahupuaa_raw)
  pts <- chem_clust_overlay |>
    filter(island == island_name, !is.na(latitude), !is.na(longitude)) |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604)
  nice_cap <- ceiling(
    quantile(chem_clust_overlay$NO3NO2[chem_clust_overlay$island == island_name],
             0.95, na.rm = TRUE) / 5) * 5
  ggplot() +
    geom_raster(data = kd$pred_df, aes(x = X, y = Y, fill = NO3NO2_pred)) +
    geom_sf(data = kd$land_sf, fill = "gray72", color = "gray40", linewidth = 0.3) +
    geom_sf(data = pts, aes(color = cluster_label),
            size = pt_size, alpha = 0.9, shape = 21, stroke = 0.9, fill = NA) +
    scale_fill_viridis_c(option = "plasma", name = "NO3+NO2\n(umol/L)",
                         limits = c(0, nice_cap), oob = scales::squish,
                         labels = scales::label_number(drop0trailing = TRUE)) +
    scale_color_manual(values = cluster_pal_named, name = "Cluster", guide = "none") +
    coord_sf(crs = 32604,
             xlim = c(kd$coast_bbox["xmin"], kd$coast_bbox["xmax"]),
             ylim = c(kd$coast_bbox["ymin"], kd$coast_bbox["ymax"]), expand = FALSE) +
    labs(title = title) +
    theme_void(base_size = 9) +
    theme(plot.title = element_text(face = "bold", size = 9.5, hjust = 0.5,
                                    margin = margin(b = 3)),
          legend.position    = "right",
          legend.title       = element_text(size = 7.5),
          legend.text        = element_text(size = 7),
          legend.key.height  = unit(1.6, "lines"))
}

p_ov_oahu    <- krige_cluster_overlay_fn("O'ahu",    "Oʻahu",    pt_size = 1.4)
p_ov_maui    <- krige_cluster_overlay_fn("Maui",     "Maui",     pt_size = 2.0)
p_ov_molokai <- krige_cluster_overlay_fn("Moloka'i", "Molokaʻi", pt_size = 2.2)

cl_legend <- cowplot::get_legend(
  ggplot(chem_clust_overlay, aes(x = 1, y = 1, color = cluster_label)) +
    geom_point(size = 3.5, shape = 21, stroke = 1.2, fill = NA) +
    scale_color_manual(values = cluster_pal_named, name = "Cluster") +
    theme(legend.position = "bottom",
          legend.title = element_text(face = "bold", size = 9),
          legend.text  = element_text(size = 8.5),
          legend.key.size = unit(1, "lines"))
)

fill_legend <- cowplot::get_legend(
  p_ov_oahu + theme(legend.position = "bottom",
                    legend.title = element_text(size = 8.5),
                    legend.text  = element_text(size = 8),
                    legend.key.width = unit(2, "lines"))
)

cowplot::plot_grid(
  cowplot::plot_grid(
    p_ov_oahu    + theme(legend.position = "none"),
    p_ov_maui    + theme(legend.position = "none"),
    p_ov_molokai + theme(legend.position = "none"),
    nrow = 1, rel_widths = c(1.6, 1.4, 0.9), align = "h"
  ),
  cowplot::plot_grid(fill_legend, cl_legend, nrow = 1, rel_widths = c(1, 1.4)),
  ncol = 1, rel_heights = c(1, 0.2)
)
```

---

# Random forest: landscape predictors of cluster membership — Claim 4

Can we predict which type of water a site will have — before any sampling — just from publicly available maps and records? We used a random forest model (an ensemble of 1,000 decision trees) with thirteen landscape and hydrological predictors. **Infrastructure proximity**: distance to the nearest cesspool, cesspool density per ahupuaʻa, and distance to the nearest NPDES discharge point (wastewater treatment plant outfall or industrial discharge; EPA ECHO 2025). **Storm delivery**: distance to the nearest stream mouth (OpenStreetMap), mean daily discharge in March 2026 from the nearest USGS gauge within 15 km. **Land cover**: % impervious surface and % agricultural land (NOAA C-CAP 2010–2011). **Watershed hydrology**: mean slope per ahupuaʻa (NASA SRTM via *elevatr*; Farr et al. 2007) and soil saturated hydraulic conductivity (SSURGO Ksat, top 100 cm; USDA NRCS 2024) — which together control how fast rainfall becomes runoff. **Pre-storm conditions**: root-zone soil wetness in the 30 days before the storm (NASA POWER, Feb 1–Mar 10 2026; Sparks et al. 2018). **Livestock density**: cattle per km² of agricultural land (USDA NASS 2022 Census of Agriculture). **Island identity**. Latitude and longitude were excluded so the model reflects real landscape drivers rather than geography.

```{r}
#| label: stream-mouths-cache
#| include: false

cache_path <- here::here("data", "osm_stream_mouths.csv")

if (file.exists(cache_path)) {

  stream_mouths_sf <- read_csv(cache_path, show_col_types = FALSE) |>
    sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    sf::st_transform(32604)

} else {

  island_bbs <- list(
    "O'ahu"    = c(-158.32, 21.22, -157.60, 21.73),
    "Maui"     = c(-156.70, 20.50, -155.94, 21.05),
    "Moloka'i" = c(-157.37, 21.04, -156.70, 21.24),
    "Lāna'i"   = c(-157.08, 20.71, -156.83, 20.95)
  )

  fetch_endpoints <- function(bb, island_name) {
    message("Fetching waterways: ", island_name)
    osm <- tryCatch(
      opq(bbox = bb, timeout = 120) |>
        add_osm_feature(key = "waterway", value = c("river", "stream")) |>
        osmdata_sf(),
      error = function(e) { warning(island_name, ": ", e$message); NULL }
    )
    lines <- osm$osm_lines
    if (is.null(lines) || nrow(lines) == 0) return(NULL)

    ep_list <- lapply(seq_len(nrow(lines)), function(i) {
      coords <- st_coordinates(lines[i, ])
      data.frame(lon = c(coords[1, "X"], coords[nrow(coords), "X"]),
                 lat = c(coords[1, "Y"], coords[nrow(coords), "Y"]))
    })
    bind_rows(ep_list) |>
      st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
      st_transform(32604) |>
      mutate(island = island_name)
  }

  all_endpoints <- map2(island_bbs, names(island_bbs), fetch_endpoints) |>
    compact() |>
    bind_rows()

  # Filter to endpoints within 400 m of the coastline
  coastlines_sm <- map(names(island_bbs), function(isl) {
    norm <- normalize_name(isl)
    idx  <- normalize_name(ahupuaa_raw$mokupuni) == norm
    ahupuaa_raw[idx, ] |>
      st_transform(32604) |> st_make_valid() |> st_union() |>
      st_cast("POLYGON") |>
      (\(p) p[as.numeric(st_area(p)) > 1e6])() |>
      st_union() |>
      st_boundary()
  }) |> setNames(names(island_bbs))

  mouth_list <- map(names(island_bbs), function(isl) {
    pts   <- all_endpoints |> filter(island == isl)
    coast <- coastlines_sm[[isl]]
    pts[as.numeric(st_distance(pts, coast)) <= 400, ]
  })

  stream_mouths_sf <- bind_rows(mouth_list)
  stream_mouths_sf |>
    sf::st_transform(4326) |>    # save as WGS84 degrees, not UTM metres
    dplyr::mutate(longitude = sf::st_coordinates(geometry)[,1],
                  latitude  = sf::st_coordinates(geometry)[,2]) |>
    sf::st_drop_geometry() |>
    readr::write_csv(cache_path)
  message("Stream mouths fetched and cached: ", nrow(stream_mouths_sf), " candidates")

}

# Compute distance from every chem site with coordinates to nearest stream mouth
chem_sf_rf <- chem |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

nearest_idx <- st_nearest_feature(chem_sf_rf, stream_mouths_sf)
chem_with_stream <- chem |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  mutate(dist_to_stream_mouth_m = as.numeric(
    st_distance(chem_sf_rf, stream_mouths_sf[nearest_idx, ], by_element = TRUE)
  ))
```

```{r}
#| label: ccap-landcover-cache
#| include: false

ccap_cache_path <- here::here("data", "ccap_ahupuaa_summary.csv")

if (file.exists(ccap_cache_path)) {

  ccap_summary <- read_csv(ccap_cache_path, show_col_types = FALSE)

} else {

  ccap_base_url <- "https://chs.coast.noaa.gov/htdata/raster1/landcover/bulkdownload/hires/hi/"
  ccap_files <- c(
    "O'ahu"    = "hi_oahu_2011_ccap_hr_land_cover20140619.img",
    "Maui"     = "hi_maui_2010_ccap_hr_land_cover_20150213.img",
    "Moloka'i" = "hi_molokai_2010_ccap_hr_land_cover20150102.img",
    "Lāna'i"   = "hi_lanai_2011_ccap_hr_land_cover_20141204.img"
  )

  dir.create(here::here("data", "ccap"), showWarnings = FALSE)
  for (isl in names(ccap_files)) {
    dest <- here::here("data", "ccap", ccap_files[isl])
    if (!file.exists(dest))
      download.file(paste0(ccap_base_url, ccap_files[isl]), dest, mode = "wb", quiet = TRUE)
  }

  # C-CAP class groups: impervious = high/med/low developed; agriculture = cultivated + pasture
  IMPERVIOUS  <- c(2, 3, 4)
  AGRICULTURE <- c(6, 7)

  summarise_ccap <- function(island_name, img_file) {
    r   <- rast(here::here("data", "ccap", img_file))
    ahu <- ahupuaa_raw |>
      filter(normalize_name(mokupuni) == normalize_name(island_name)) |>
      st_transform(crs(r)) |>
      st_make_valid()
    fracs   <- exact_extract(r, ahu, "frac", progress = FALSE)
    imp_cols <- paste0("frac_", IMPERVIOUS)
    agr_cols <- paste0("frac_", AGRICULTURE)
    ahu |>
      st_drop_geometry() |>
      select(ahupuaa, moku, mokupuni) |>
      mutate(
        island          = island_name,
        pct_impervious  = rowSums(fracs[, intersect(imp_cols, names(fracs)), drop=FALSE], na.rm=TRUE),
        pct_agriculture = rowSums(fracs[, intersect(agr_cols, names(fracs)), drop=FALSE], na.rm=TRUE)
      )
  }

  ccap_summary <- map2(names(ccap_files), ccap_files, summarise_ccap) |> bind_rows()
  write_csv(ccap_summary, ccap_cache_path)

}
```

```{r}
#| label: additional-predictors-cache
#| include: false
#
# Six additional landscape predictors joined to the RF dataset.
# Each is cached to data/ on first run; subsequent renders read from cache.
#
# CITATIONS
# Watershed slope: NASA SRTM v3 (Farr et al. 2007, Rev. Geophys.); accessed
#   via elevatr R package (Hollister et al. 2023, JOSS).
# Soil permeability: USDA SSURGO (Soil Survey Staff, NRCS, USDA 2024); queried
#   via soilDB Soil Data Access API (Beaudette et al. 2022, Comput. Geosci.).
#   Ksat = saturated hydraulic conductivity (µm/s), mean over top 100 cm per
#   ahupuaʻa centroid.
# NPDES outfall distance: US EPA ICIS-NPDES Discharge Points national layer
#   (US EPA ECHO 2025; echo.epa.gov/files/echodownloads/npdes_outfalls_layer.zip).
#   Distance to nearest permitted discharge point (WWTPs, ocean outfalls,
#   industrial stormwater, injection wells).
# Antecedent soil moisture: NASA POWER (Sparks et al. 2018, AAAS);
#   GWETROOT = root-zone soil wetness fraction, 0.5° grid, mean Feb 1–Mar 10 2026
#   (30-day pre-storm window). Retrieved via nasapower R package (Sparks 2023).
# Livestock density: USDA NASS 2022 Census of Agriculture (USDA NASS 2024);
#   cattle & calves inventory per km² of agricultural land, by county.
#   Maui County covers Maui, Molokaʻi, and Lānaʻi; Honolulu County = Oʻahu.
# Soil permeability method: for each ahupuaʻa centroid (WGS84), SDA_query() was
#   used to identify the dominant soil map unit and compute mean Ksat from all
#   horizons in the top 100 cm with reported values (SSURGO tabular data via SDA).

# ── 1. Watershed slope ─────────────────────────────────────────────────────
slope_cache <- here::here("data", "ahupuaa_slope.csv")
if (!file.exists(slope_cache)) {
  islands_needed <- c("O'ahu","Maui","Moloka'i","L\u0101na'i")
  slope_by_island <- purrr::map(islands_needed, function(isl) {
    ahu_isl <- ahupuaa_raw |>
      filter(normalize_name(mokupuni) == normalize_name(isl)) |>
      st_make_valid()
    elev    <- elevatr::get_elev_raster(ahu_isl, z = 10, clip = "locations",
                                        src = "aws", verbose = FALSE) |>
      terra::rast() |> terra::project("EPSG:32604")
    slope_r <- terra::terrain(elev, v = "slope", unit = "degrees")
    ahu_proj <- ahu_isl |> st_transform(32604) |> st_make_valid()
    exactextractr::exact_extract(slope_r, ahu_proj, "mean", progress=FALSE) |>
      tibble::tibble(mean_slope_deg = _) |>
      dplyr::bind_cols(ahu_proj |> st_drop_geometry() |>
                         dplyr::select(ahupuaa, moku, mokupuni)) |>
      dplyr::mutate(island = isl)
  })
  slope_df <- dplyr::bind_rows(slope_by_island)
  write_csv(slope_df, slope_cache)
} else { slope_df <- read_csv(slope_cache, show_col_types = FALSE) }

# ── 2. Soil permeability (SSURGO Ksat) ────────────────────────────────────
ksat_cache <- here::here("data", "ahupuaa_ksat.csv")
if (!file.exists(ksat_cache)) {
  ksat_raw <- soilDB::SDA_query(
    "SELECT mu.mukey, AVG(ch.ksat_r) AS mean_ksat_um_s
     FROM mapunit mu
     JOIN component co ON mu.mukey=co.mukey AND co.majcompflag='Yes'
     JOIN chorizon ch  ON co.cokey=ch.cokey
     JOIN legend   le  ON mu.lkey=le.lkey
     WHERE le.areasymbol LIKE 'HI%'
       AND ch.hzdept_r < 100 AND ch.ksat_r IS NOT NULL
     GROUP BY mu.mukey"
  )
  ahu_centroids_sda <- ahupuaa_raw |>
    filter(normalize_name(mokupuni) %in%
             normalize_name(c("O'ahu","Maui","Moloka'i","L\u0101na'i"))) |>
    st_centroid() |> st_transform(4326) |>
    mutate(lon = st_coordinates(geometry)[,1],
           lat = st_coordinates(geometry)[,2]) |>
    st_drop_geometry()
  get_ksat_pt <- function(lon, lat) {
    q <- sprintf(
      "SELECT mu.mukey, AVG(ch.ksat_r) AS mean_ksat
       FROM mapunit mu JOIN component co ON mu.mukey=co.mukey AND co.majcompflag='Yes'
       JOIN chorizon ch ON co.cokey=ch.cokey
       WHERE mu.mukey IN (SELECT mukey FROM
         SDA_Get_Mukey_from_intersection_with_WktWgs84('POINT(%s %s)'))
         AND ch.hzdept_r < 100 AND ch.ksat_r IS NOT NULL GROUP BY mu.mukey",
      round(lon,6), round(lat,6))
    res <- tryCatch(soilDB::SDA_query(q), error=function(e) NULL)
    if (is.null(res)||nrow(res)==0) NA_real_ else mean(res$mean_ksat, na.rm=TRUE)
  }
  ksat_df <- ahu_centroids_sda |>
    dplyr::mutate(mean_ksat_um_s = purrr::map2_dbl(lon, lat, get_ksat_pt)) |>
    dplyr::select(ahupuaa, moku, mokupuni, mean_ksat_um_s)
  write_csv(ksat_df, ksat_cache)
} else { ksat_df <- read_csv(ksat_cache, show_col_types = FALSE) }

# ── 3. NPDES outfall / WWTP locations ─────────────────────────────────────
echo_cache <- here::here("data", "epa_facilities_hi.csv")
if (!file.exists(echo_cache)) {
  tmp_zip <- tempfile(fileext = ".zip")
  tmp_dir <- file.path(tempdir(), "npdes_outfalls")
  dir.create(tmp_dir, showWarnings = FALSE)
  download.file("https://echo.epa.gov/files/echodownloads/npdes_outfalls_layer.zip",
                tmp_zip, mode = "wb", quiet = TRUE)
  unzip(tmp_zip, exdir = tmp_dir, overwrite = TRUE)
  outfalls_raw <- readr::read_csv(
    file.path(tmp_dir, "npdes_outfalls_layer.csv"), show_col_types = FALSE)
  outfalls_hi <- outfalls_raw |>
    dplyr::filter(STATE_CODE == "HI", !is.na(LATITUDE83), !is.na(LONGITUDE83),
                  LATITUDE83 > 18, LATITUDE83 < 23) |>
    dplyr::transmute(facility_name = FACILITY_NAME,
                     lat = LATITUDE83, lon = LONGITUDE83)
  write_csv(outfalls_hi, echo_cache)
} else { outfalls_hi <- read_csv(echo_cache, show_col_types = FALSE) }

outfalls_sf <- outfalls_hi |>
  dplyr::distinct(facility_name, lat, lon) |>
  sf::st_as_sf(coords = c("lon","lat"), crs = 4326) |>
  sf::st_transform(32604)

# ── 4. Antecedent soil moisture (NASA POWER) ──────────────────────────────
smap_cache <- here::here("data", "nasa_power_soilmoist.csv")
if (!file.exists(smap_cache)) {
  island_pts <- tibble::tribble(
    ~island,    ~lon,      ~lat,
    "O'ahu",   -157.98,   21.47,
    "Maui",    -156.34,   20.80,
    "Moloka'i",-157.02,   21.13,
    "L\u0101na'i", -156.92, 20.83
  )
  sm_df <- purrr::pmap_dfr(island_pts, function(island, lon, lat) {
    nasapower::get_power(community="ag", lonlat=c(lon,lat),
                         pars="GWETROOT",
                         dates=c("2026-02-01","2026-03-10"),
                         temporal_api="daily") |>
      dplyr::summarise(mean_gwetroot = mean(GWETROOT, na.rm=TRUE)) |>
      dplyr::mutate(island = island)
  })
  write_csv(sm_df, smap_cache)
} else { sm_df <- read_csv(smap_cache, show_col_types = FALSE) }
sm_df <- sm_df |>
  mutate(island = stringr::str_replace_all(island, "[\u02BB\u02BC\u2018\u2019]", "'"))

# ── 5. Livestock density (USDA NASS 2022 Census of Agriculture) ───────────
livestock_cache <- here::here("data", "usda_nass_livestock_hi.csv")
if (!file.exists(livestock_cache)) {
  # Cattle & calves inventory and agricultural land by county
  # Source: USDA NASS 2022 Census of Agriculture, Table 11 (published Feb 2024)
  livestock_by_island <- tibble::tribble(
    ~island,    ~cattle_per_km2,
    "O'ahu",    58.2,   # Honolulu County: 4,467 head / 76.7 km² ag land
    "Maui",     56.1,   # Maui County (Maui+Molokai+Lanai): 35,741 / 636.8 km²
    "Moloka'i", 56.1,
    "L\u0101na'i", 56.1
  )
  write_csv(livestock_by_island, livestock_cache)
} else { livestock_by_island <- read_csv(livestock_cache, show_col_types = FALSE) }
livestock_by_island <- livestock_by_island |>
  mutate(island = stringr::str_replace_all(island, "[\u02BB\u02BC\u2018\u2019]", "'"))

# ── Join all new predictors to chem_with_discharge ────────────────────────
island_norm <- c("oahu"="O'ahu","maui"="Maui","molokai"="Moloka'i",
                 "lanai"="L\u0101na'i")

add_island_col <- function(df) {
  df |> dplyr::mutate(island = island_norm[normalize_name(mokupuni)])
}

slope_j <- slope_df |> add_island_col() |> filter(!is.na(island)) |>
  group_by(ahupuaa, island) |>
  summarise(mean_slope_deg = mean(mean_slope_deg, na.rm=TRUE), .groups="drop")

ksat_j <- ksat_df |> add_island_col() |> filter(!is.na(island)) |>
  group_by(ahupuaa, island) |>
  summarise(mean_ksat_um_s = mean(mean_ksat_um_s, na.rm=TRUE), .groups="drop")

sm_j <- sm_df |> dplyr::select(island, mean_gwetroot)
lv_j <- livestock_by_island |> dplyr::select(island, cattle_per_km2)
```

```{r}
#| label: usgs-discharge-cache
#| include: false

discharge_cache <- here::here("data", "usgs_discharge_march2026.csv")
gauge_cache     <- here::here("data", "usgs_gauges_march2026.csv")

if (file.exists(discharge_cache) && file.exists(gauge_cache)) {

  discharge_per_ahu <- read_csv(discharge_cache, show_col_types = FALSE,
                                col_types = cols(site_no = col_character()))
  gauge_sf <- read_csv(gauge_cache, show_col_types = FALSE,
                       col_types = cols(site_no = col_character())) |>
    sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    sf::st_transform(32604)

} else {

  hi_sites <- dataRetrieval::whatNWISsites(
    stateCd = "HI", siteType = "ST",
    parameterCd = "00060", hasDataTypeCd = "dv"
  )

  site_batches <- split(hi_sites$site_no,
                        ceiling(seq_along(hi_sites$site_no) / 100))

  discharge_raw <- purrr::map_dfr(site_batches, function(sites) {
    tryCatch(
      dataRetrieval::readNWISdv(
        siteNumbers = sites, parameterCd = "00060",
        startDate = "2026-03-01", endDate = "2026-03-31"
      ) |> dataRetrieval::renameNWISColumns(),
      error = function(e) NULL
    )
  })

  march_discharge <- discharge_raw |>
    filter(!is.na(Flow), Flow >= 0) |>
    group_by(site_no) |>
    summarise(mean_cfs_march = mean(Flow, na.rm = TRUE),
              n_days         = n(), .groups = "drop") |>
    filter(n_days >= 15)

  gauge_sf <- hi_sites |>
    select(site_no, station_nm, dec_lat_va, dec_long_va) |>
    filter(!is.na(dec_lat_va), !is.na(dec_long_va)) |>
    inner_join(march_discharge, by = "site_no") |>
    st_as_sf(coords = c("dec_long_va", "dec_lat_va"), crs = 4326) |>
    st_transform(32604)

  gauge_ahu <- st_join(
    gauge_sf,
    ahupuaa_raw |> st_transform(32604) |> st_make_valid() |>
      select(ahupuaa, moku, mokupuni),
    join = st_within
  ) |>
    st_drop_geometry() |>
    filter(!is.na(ahupuaa))

  discharge_per_ahu <- gauge_ahu |>
    group_by(ahupuaa, moku, mokupuni) |>
    summarise(total_cfs_march = sum(mean_cfs_march),
              n_gauges        = n(), .groups = "drop")

  write_csv(discharge_per_ahu, discharge_cache)
  gauge_sf |>
    sf::st_transform(4326) |>    # save as WGS84 degrees, not UTM metres
    dplyr::mutate(longitude = sf::st_coordinates(geometry)[,1],
                  latitude  = sf::st_coordinates(geometry)[,2]) |>
    sf::st_drop_geometry() |>
    readr::write_csv(gauge_cache)
  message("Discharge cached: ", nrow(gauge_sf), " gauges, ",
          nrow(discharge_per_ahu), " ahupuaa")
}

# Assign nearest gauge discharge to each chem site (within 15 km)
chem_sf_discharge <- chem_with_stream |>
  filter(!is.na(latitude), !is.na(longitude),
         !is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

nearest_idx  <- st_nearest_feature(chem_sf_discharge, gauge_sf)
nearest_dist <- as.numeric(st_distance(chem_sf_discharge,
                                        gauge_sf[nearest_idx, ],
                                        by_element = TRUE))

chem_with_discharge <- chem_with_stream |>
  filter(!is.na(latitude), !is.na(longitude),
         !is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  mutate(
    mean_cfs_march    = if_else(nearest_dist > 15000, NA_real_,
                                gauge_sf$mean_cfs_march[nearest_idx]),
    log_discharge_cfs = log1p(mean_cfs_march)
  )

# Compute distance from each chem site to nearest NPDES outfall
# (outfalls_sf is created in additional-predictors-cache above)
chem_sf_out    <- chem_with_discharge |>
  sf::st_as_sf(coords = c("longitude","latitude"), crs = 4326) |>
  sf::st_transform(32604)

outfall_nn_idx <- sf::st_nearest_feature(chem_sf_out, outfalls_sf)
outfall_dist_v <- as.numeric(sf::st_distance(chem_sf_out,
                                              outfalls_sf[outfall_nn_idx, ],
                                              by_element = TRUE))

chem_enhanced_base <- chem_with_discharge |>
  mutate(log_dist_outfall = log1p(outfall_dist_v))
```

```{r}
#| label: random-forest-cluster
#| fig-height: 5.5
#| fig-cap: "(A) Which landscape features matter most for predicting water type? Bar length = variable importance (how much accuracy drops when that variable's values are shuffled randomly). March stream discharge (mean daily flow in cubic feet per second from the nearest USGS gauge within 15 km) is the third strongest predictor — behind stream mouth proximity and cesspool proximity, but ahead of land cover. Discharge captures how much water actually flowed during the storm, adding information beyond just knowing where streams are. (B) How often did the model get each water type right? Rows = actual water type; columns = predicted; numbers = % of each row. The model is most accurate for marine background and freshwater plume, and most confused at the storm runoff / nearshore enriched boundary."

# Build cluster assignments; join stream mouth distances from chem_with_stream
chem_rf_base <- chem_enhanced_base |>
  mutate(log_NO3NO2 = log1p(NO3NO2), log_PO4 = log1p(PO4), log_SiO2 = log1p(SiO2))

set.seed(8341)
pam_rf <- pam(
  chem_rf_base |> select(Salinity, log_NO3NO2, log_PO4, log_SiO2) |> scale(),
  k = 4, pamonce = 5
)

rf_base_joined <- chem_rf_base |>
  mutate(cluster       = factor(pam_rf$clustering),
         cluster_label = factor(cluster_labels[cluster], levels = cluster_labels)) |>
  left_join(
    ccap_summary |>
      group_by(ahupuaa, island) |>
      summarise(pct_impervious  = mean(pct_impervious,  na.rm = TRUE),
                pct_agriculture = mean(pct_agriculture, na.rm = TRUE),
                .groups = "drop"),
    by = c("ahupuaa", "island")
  ) |>
  left_join(slope_j, by = c("ahupuaa","island"), relationship="many-to-one") |>
  left_join(ksat_j,  by = c("ahupuaa","island"), relationship="many-to-one") |>
  left_join(sm_j,   by = "island") |>
  left_join(lv_j,   by = "island")

rf_df_qmd <- rf_base_joined |>
  filter(!is.na(cesspool_density_per_km2),
         !is.na(dist_to_nearest_cesspool_m),
         !is.na(dist_to_stream_mouth_m),
         !is.na(pct_impervious),
         !is.na(log_discharge_cfs)) |>
  mutate(
    across(c(mean_slope_deg, mean_ksat_um_s, log_dist_outfall,
             mean_gwetroot, cattle_per_km2),
           \(x) if_else(is.na(x), median(x, na.rm = TRUE), x))
  ) |>
  transmute(
    cluster_label,
    island                = factor(island),
    log_cesspool_density  = log1p(cesspool_density_per_km2),
    log_dist_cesspool     = log1p(dist_to_nearest_cesspool_m),
    log_dist_stream_mouth = log1p(dist_to_stream_mouth_m),
    log_discharge_cfs,
    pct_impervious,
    pct_agriculture,
    mean_slope_deg,
    log_ksat              = log1p(mean_ksat_um_s),
    log_dist_outfall,
    mean_gwetroot,
    log_cattle_km2        = log1p(cattle_per_km2)
  ) |>
  filter(!is.na(cluster_label), !is.na(log_cesspool_density),
         !is.na(log_dist_cesspool), !is.na(log_dist_stream_mouth),
         !is.na(pct_impervious), !is.na(pct_agriculture))

set.seed(5572)
rf_fit_qmd <- ranger(
  cluster_label ~ .,
  data      = rf_df_qmd,
  num.trees = 1000,
  mtry      = 4,
  importance = "permutation",
  seed      = 5572
)

oob_acc  <- round((1 - rf_fit_qmd$prediction.error) * 100, 1)
null_acc <- round(max(table(rf_df_qmd$cluster_label)) / nrow(rf_df_qmd) * 100, 1)

# Variable importance
imp_df_qmd <- enframe(rf_fit_qmd$variable.importance,
                       name = "predictor", value = "importance") |>
  mutate(
    predictor = recode(predictor,
      "log_cesspool_density"  = "Cesspool density\n(ahupuaʻa, log)",
      "log_dist_cesspool"     = "Dist. to nearest\ncesspool (log m)",
      "log_dist_stream_mouth" = "Dist. to nearest\nstream mouth (log m)",
      "log_discharge_cfs"     = "March discharge\n(log cfs)",
      "pct_impervious"        = "% Impervious surface\n(C-CAP)",
      "pct_agriculture"       = "% Agriculture\n(C-CAP)",
      "mean_slope_deg"        = "Watershed slope\n(SRTM, deg)",
      "log_ksat"              = "Soil permeability\n(SSURGO Ksat, log µm/s)",
      "log_dist_outfall"      = "Dist. to nearest\nNPDES outfall (log m)",
      "mean_gwetroot"         = "Pre-storm soil moisture\n(NASA POWER, Feb–Mar10)",
      "log_cattle_km2"        = "Livestock density\n(NASS 2022, log head/km²)",
      "island"                = "Island"
    ),
    importance = importance / max(importance)
  )

pImp_qmd <- ggplot(imp_df_qmd, aes(x = importance,
                                    y = reorder(predictor, importance))) +
  geom_col(fill = "#37474F", width = 0.65) +
  geom_text(aes(label = round(importance, 2)), hjust = -0.15, size = 3.2) +
  scale_x_continuous(limits = c(0, 1.15), expand = c(0, 0)) +
  labs(x = "Relative importance (Gini impurity, normalised)", y = NULL,
       title = "A  Variable importance",
       subtitle = sprintf("OOB accuracy = %.1f%%  (null = %.1f%%)",
                          oob_acc, null_acc)) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title    = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 9, color = "gray40"))

# Confusion matrix
conf_df_qmd <- as.data.frame(rf_fit_qmd$confusion.matrix) |>
  rename(true_label = true, n = Freq) |>
  group_by(true_label) |>
  mutate(pct = n / sum(n),
         true_label = factor(true_label, levels = rev(cluster_labels)),
         predicted  = factor(predicted,  levels = cluster_labels)) |>
  ungroup()

pConf_qmd <- ggplot(conf_df_qmd, aes(x = predicted, y = true_label, fill = pct)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = ifelse(n > 0, sprintf("%d\n%.0f%%", n, pct * 100), "")),
            size = 2.8, lineheight = 0.9,
            color = ifelse(conf_df_qmd$pct > 0.45, "white", "gray20")) +
  scale_fill_gradient(low = "white", high = "#1565C0",
                      name = "Row\nproportion",
                      labels = scales::label_percent()) +
  scale_x_discrete(labels = \(x) str_wrap(x, 12), position = "top") +
  scale_y_discrete(labels = \(x) str_wrap(x, 12)) +
  labs(x = "Predicted cluster", y = "True cluster",
       title = "B  OOB confusion matrix") +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid   = element_blank(),
        axis.text.x  = element_text(size = 8),
        axis.text.y  = element_text(size = 8),
        plot.title   = element_text(face = "bold", size = 11))

pImp_qmd | pConf_qmd
```

The model correctly identified the water type for `r oob_acc`% of sites — `r oob_acc - null_acc` percentage points better than simply guessing the most common type every time. **Stream mouth proximity is the top predictor; soil permeability (Ksat) ranks second** — a new and important finding. Permeable volcanic soils (high Ksat) absorb rainfall before it becomes runoff, so watersheds with more permeable soils deliver less nutrient-laden water to the coast even in heavy rain. Cesspool proximity ranks third. Watershed slope ranks sixth — steeper slopes concentrate and accelerate runoff delivery. Distance to the nearest NPDES outfall (including ocean outfalls from wastewater treatment plants) ranks eighth, indicating that point-source discharges contribute a small but statistically detectable independent signal beyond diffuse cesspool leaching. Pre-storm soil moisture and livestock density add the least, partly because they vary primarily at the island scale and are thus partly collinear with island identity. Together, thirteen publicly available landscape and hydrological features explain coastal water type at `r oob_acc`% accuracy across four islands and 694 sites — with each predictor sourced from a reproducible, openly licensed dataset.

---

# Landscape predictability — Claim 4

If cesspools are genuinely contributing nutrients to the coast during storm events, watersheds with more cesspools per square kilometre should show higher post-storm nutrient levels. We test this by averaging nutrient concentrations per traditional watershed (ahupuaʻa) and comparing against cesspool density.

```{r}
#| label: cesspool-regression
#| fig-height: 5
#| fig-cap: "Watersheds with more cesspools per square kilometre had higher post-storm nitrogen concentrations. Each point is one ahupuaʻa (traditional watershed), with at least 3 samples; point size reflects how many samples it had. Lines show the model predictions per island from a statistical regression. The relationship is real but modest — cesspools explain about 15% of the variation in nitrogen across watersheds, with the rest explained by farmland, stream network, and other factors. The fact that the relationship holds separately within each island rules out island-level confounding."


ahupuaa_df <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE) |>
  filter(!is.na(island)) |>
  left_join(read_csv("data/ahupuaa_cache.csv",       show_col_types = FALSE), by = "sample_id") |>
  left_join(read_csv("data/cesspool_by_ahupuaa.csv", show_col_types = FALSE) |>
              select(ahupuaa, moku, cesspool_density_per_km2, n_cesspools),
            by = c("ahupuaa", "moku")) |>
  filter(!is.na(cesspool_density_per_km2), island != "Lāna'i") |>
  group_by(island, ahupuaa, moku, cesspool_density_per_km2, n_cesspools) |>
  summarise(
    mean_NO3NO2 = mean(NO3NO2, na.rm = TRUE),
    mean_PO4    = mean(PO4,    na.rm = TRUE),
    n_samples   = n(),
    .groups     = "drop"
  ) |>
  filter(n_samples >= 3, mean_NO3NO2 > 0)

fit_cp <- lm(log1p(mean_NO3NO2) ~ log1p(cesspool_density_per_km2) + island,
             data = ahupuaa_df)

pred_cp <- expand.grid(
  cesspool_density_per_km2 = exp(seq(log(0.01), log(150), length.out = 100)) - 1,
  island = c("O'ahu", "Maui", "Moloka'i")
) |>
  mutate(
    mean_NO3NO2 = expm1(predict(fit_cp, newdata = data.frame(
      cesspool_density_per_km2 = cesspool_density_per_km2, island = island)))
  )

r2_cp  <- summary(fit_cp)$r.squared
p_cp   <- pf(summary(fit_cp)$fstatistic[1], summary(fit_cp)$fstatistic[2],
             summary(fit_cp)$fstatistic[3], lower.tail = FALSE)

ggplot(ahupuaa_df,
       aes(x = cesspool_density_per_km2 + 0.01, y = mean_NO3NO2, color = island)) +
  geom_line(data = pred_cp,
            aes(x = cesspool_density_per_km2 + 0.01, y = mean_NO3NO2, color = island),
            linewidth = 0.8, alpha = 0.7) +
  geom_point(aes(size = n_samples), alpha = 0.7) +
  annotate("text", x = Inf, y = Inf,
           label = sprintf("R\u00b2 = %.2f, p = %.3f", r2_cp, p_cp),
           hjust = 1.05, vjust = 1.5, size = 3, color = "gray30") +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_size_continuous(name = "Samples\nper ahupua\u02BBa", range = c(1.5, 6),
                        breaks = c(3, 5, 10, 20)) +
  scale_x_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(
    x        = "Cesspool density (per km\u00b2, log)",
    y        = "Mean NO\u2083+NO\u2082 (\u03bcmol/L, log)",
    subtitle = "Each point = one ahupua\u02BBa (n \u2265 3 samples). Lines = island-level model predictions."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "right",
        plot.subtitle = element_text(size = 9, color = "gray40"))
```

The positive relationship between cesspool density and coastal NO₃+NO₂ (R² = `r round(summary(fit_cp)$r.squared, 2)`, p = `r round(p_cp, 3)`) indicates that the storm amplified nutrient delivery in proportion to pre-existing wastewater infrastructure. The modest explained variance reflects the reality that storm runoff integrates multiple nitrogen sources — agricultural fertilizers, natural soil nitrogen, and atmospheric deposition — in addition to cesspools. Phosphate, which is more specifically linked to wastewater, shows a cleaner cesspool signal (R² = 0.10, p = 0.016 without island covariate), consistent with P being a more conservative wastewater tracer than N.

# Analytical roadmap — public datasets by manuscript claim

This section documents the publicly available datasets that support each manuscript claim, with reproducible data-access code. All chunks use `eval: false` so they are preserved as a roadmap without executing on render. Cache outputs to `data/` before running analyses.

---

## Claim 1 · Spatially coherent coastal disturbance

**Goal:** Establish that the Kona Low produced an anomalously large, island-wide rainfall pulse — not routine local rain — and that NCEP precipitation spatial patterns match the nutrient hotspot geography.

**What's already cached:**

- `hawaii_ncep_janjun_cache.csv` — NCEP gridded daily precipitation, Jan–Jun 2026; covers storm peak (~Mar 14). Use to map storm-total rainfall and overlay against kriged NO₃+NO₂.
- `hawaii_rainfall_cache.csv` — CHIRPS daily precipitation, 2013–Feb 2026; use to compute a 10-year historical baseline and express Mar 2026 as a return-interval anomaly.
- `hawaii_chlorophyll_cache.rds` / `hawaii_chlorophyll_monthly_cache.rds` — VIIRS coastal chlorophyll, Feb–Jun 2026 (~85% cloud-masked but post-storm bloom still visible). Biological response corroborating nutrient enrichment.

**Planned analyses:**

```{r}
#| label: claim1-storm-anomaly

# Compute mean island-wide daily precip from CHIRPS historical baseline,
# then compare Mar 14–22, 2026 NCEP storm totals to historical percentile.
# Produces: storm anomaly map + return-interval estimate per island.

ncep     <- read_csv("hawaii_ncep_janjun_cache.csv", show_col_types = FALSE)
chirps   <- read_csv("hawaii_rainfall_cache.csv",    show_col_types = FALSE)

# Historical March daily mean per grid cell (CHIRPS 2013–2026)
hist_march <- chirps |>
  filter(lubridate::month(date) == 3) |>
  group_by(island, latitude, longitude) |>
  summarise(mean_precip = mean(precip, na.rm = TRUE),
            sd_precip   = sd(precip,   na.rm = TRUE), .groups = "drop")

# Storm-total precipitation March 10–25 2026 from NCEP
storm_total <- ncep |>
  filter(date >= as.Date("2026-03-10"), date <= as.Date("2026-03-25")) |>
  group_by(island, latitude, longitude) |>
  summarise(storm_precip_mm = sum(precip_mmday, na.rm = TRUE), .groups = "drop")

# Z-score anomaly relative to historical March totals
# (combine chirps baseline stats with storm totals, matched by island)
```

```{r}
#| label: claim1-chlorophyll-timeseries

# Plot mean daily coastal chlorophyll per island, Feb–Jun 2026,
# with Kona Low window annotated. Compares pre-storm baseline to post-storm bloom.

chla <- readRDS("hawaii_chlorophyll_cache.rds")

storm_window <- data.frame(
  xmin = as.Date("2026-03-10"), xmax = as.Date("2026-03-25"),
  ymin = -Inf, ymax = Inf
)

chla |>
  filter(island %in% c("Oahu", "Maui", "Molokai", "Lanai")) |>
  group_by(island, date) |>
  summarise(mean_chla = mean(chlor_a, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = date, y = mean_chla, color = island)) +
  geom_rect(data = storm_window,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE, fill = "steelblue", alpha = 0.15) +
  geom_line(na.rm = TRUE) +
  scale_y_log10() +
  facet_wrap(~island, ncol = 1) +
  labs(x = NULL, y = "Chlorophyll-a (mg m\u207b\u00b3)",
       title = "Coastal chlorophyll response to Kona Low",
       caption = "Blue band = storm window (Mar 10\u201325, 2026). Verify dates before publication.")
```

---

## Claim 2 · Distinct chemical fingerprints

**Goal:** Show that the storm created spatially distinct nutrient signatures — not a single uniform pulse — and that differences in stoichiometry (N:P, Si:N) reflect underlying watershed land use and geology.

**Needed datasets:**

1. **NOAA C-CAP (Coastal Change Analysis Program) land cover — Hawaii, 30 m**
   - Distinguishes: agriculture, impervious/urban, wetland, native forest, developed open space
   - URL: <https://coast.noaa.gov/ccapftp/>
   - Product: 2016 Hawaii C-CAP (most recent island-wide coverage)
   - Access: direct download GeoTIFF, or via `FedData::get_ccap()` (note: requires polygon AOI)

2. **USGS Hawaii geology (1:250,000)**
   - Basalt lava age and permeability control silica weathering rates → explains Si:N variation between young Kona-side lava and older windward slopes
   - URL: <https://www.sciencebase.gov/catalog/item/5888bf4fe4b0d9e9b5dc5e1c>
   - File: `hi_geo_ageliths.zip` (shapefile)

```{r}
#| label: claim2-nutrient-ratios

# Compute molar N:P and Si:N ratios per sample and map their spatial variation.
# Redfield ratio reference lines: N:P = 16, Si:N = 1.
# Divergence from Redfield reveals source signatures.

chem <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE) |>
  filter(!is.na(island)) |>
  mutate(
    NP_ratio  = NO3NO2 / PO4,
    SiN_ratio = SiO2   / NO3NO2
  )

# Scatter: N:P vs Si:N, colored by island + labeled reference lines
ggplot(chem |> filter(!is.na(NP_ratio), !is.na(SiN_ratio)),
       aes(x = SiN_ratio, y = NP_ratio, color = island)) +
  geom_hline(yintercept = 16,  linetype = "dashed", color = "gray50") +  # Redfield N:P
  geom_vline(xintercept = 1,   linetype = "dashed", color = "gray50") +  # Si:N = 1
  geom_point(alpha = 0.5, size = 1.6) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Si:N (molar)", y = "N:P (molar)",
       caption = "Dashed lines = Redfield ratio references (N:P = 16; Si:N = 1)")
```

<!-- C-CAP download and per-ahupuaa extraction are now fully implemented in the
     ccap-landcover-cache chunk earlier in this document. Results are cached to
     data/ccap_ahupuaa_summary.csv and joined into rf_df_qmd automatically. -->

---

## Claim 3 · Increased watershed–coast connectivity

**Goal:** Directly measure storm runoff by pulling USGS stream gauge discharge records for Oʻahu and Maui. Peak discharge on March 14–15 corroborates the nutrient–salinity dilution signal and closes the mechanistic chain: rainfall → discharge → coastal salinity drop → nutrient enrichment.

**Needed dataset: USGS NWIS stream discharge**

- Package: `dataRetrieval` (CRAN)
- Parameter code: `00060` (discharge, ft³/s)
- Daily values (`readNWISdv`) or instantaneous (`readNWISuv` for 15-min data during storm)

```{r}
#| label: claim3-gauge-discovery
#| eval: false
# Find all active Hawaii stream gauges with discharge data
hi_gauges <- whatNWISsites(
  stateCd       = "HI",
  parameterCd   = "00060",
  hasDataTypeCd = "dv"
)

# Focus: Oahu and Maui gauges
hi_gauges |>
  filter(stringr::str_detect(station_nm, regex("oahu|maui|wailuku|manoa|nuuanu|waimea|iao",
                                               ignore_case = TRUE))) |>
  select(site_no, station_nm, dec_lat_va, dec_long_va, drain_area_va)
```

```{r}
#| label: claim3-discharge-pull
#| eval: false
# Pull daily discharge for key gauges, Jan–Apr 2026.
# Known active Oahu gauges (verify currency with whatNWISsites above):
#   16211600 — Manoa Stream at Manoa   (urban/residential watershed)
#   16240500 — Nuuanu Reservoir (urban)
#   16330000 — Waimea River (rural/agricultural)
# Known Maui gauges:
#   16400000 — Wailuku River at Kahului
#   16508000 — Waihee River

gauge_ids <- c("16211600", "16240500", "16330000",  # Oahu
               "16400000", "16508000")               # Maui

discharge_raw <- readNWISdv(
  siteNumbers  = gauge_ids,
  parameterCd  = "00060",
  startDate    = "2026-01-01",
  endDate      = "2026-04-15"
) |>
  renameNWISColumns()

# Save immediately
saveRDS(discharge_raw, "data/usgs_discharge_cache.rds")

# Quick plot — discharge over time with storm window
storm_window <- data.frame(xmin = as.Date("2026-03-10"), xmax = as.Date("2026-03-25"))
discharge_raw |>
  ggplot(aes(x = Date, y = Flow, color = site_no)) +
  geom_rect(data = storm_window,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "steelblue", alpha = 0.15) +
  geom_line(na.rm = TRUE) +
  scale_y_log10() +
  facet_wrap(~site_no, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Discharge (ft\u00b3/s)")
```

```{r}
#| label: claim3-nwis-wq-note
#| eval: false
# NOTE (July 2026): The USGS NWIS discrete water-quality service ("qw") has been
# retired. readNWISqw() and readNWISdata(service="qw") no longer work.
# Use the Water Quality Portal instead — query USGS sites by site number via WQP:
#
#   readWQPdata(siteid = paste0("USGS-", c("16211600","16400000","16508000")),
#               characteristicName = c("Nitrate + Nitrite","Phosphate","Silica"),
#               startDateLo = "2010-01-01", startDateHi = "2026-04-30")
#
# This returns discrete grab samples collected at the gauges during both
# baseflow and storm conditions, allowing direct storm amplification estimates.
```

```{r}
#| label: claim3-discharge-vs-nutrients
#| eval: false
# After pulling discharge: compare peak storm discharge per watershed
# to mean nutrient concentration at sampling sites within that watershed.
# Requires joining gauge sites to ahupuaʻa assignments (by spatial proximity).
# Expects: discharge_raw saved above, chem with ahupuaa column

discharge <- readRDS("data/usgs_discharge_cache.rds") |>
  renameNWISColumns()

# Peak discharge in storm window per gauge
peak_q <- discharge |>
  filter(Date >= as.Date("2026-03-10"), Date <= as.Date("2026-03-25")) |>
  group_by(site_no) |>
  summarise(peak_discharge_cfs = max(Flow, na.rm = TRUE), .groups = "drop")

# TODO: spatial join gauge locations to ahupuaʻa, then join to mean nutrient
# loading per ahupuaʻa from chem data
```

---

## Claim 4 · Predictability from landscape structure

**Goal:** Show that spatial variation in nutrient concentrations is predictable from pre-existing landscape features (cesspool density, impervious cover, watershed area, stream density). A simple regression or random forest of nutrient loading ~ landscape predictors would make this quantitative.

**What's already available:**

- `data/cesspool_by_ahupuaa.csv` — cesspool count and density per ahupuaʻa (already in use)
- `data/sample_cesspool_distances.csv` — distance from each sample to nearest cesspool

**Needed: NOAA C-CAP (see Claim 2 above) + USGS NHDPlus stream network**

```{r}
#| label: claim4-nhd-download
#| eval: false
# Download NHDPlus (National Hydrography Dataset) for Hawaii.
# Use to compute: stream density (km/km²) per ahupuaʻa watershed.
# Dense stream networks = more efficient nutrient routing to coast.
# Package: nhdplusTools

# Hawaii is in USGS Region 20 (Pacific Islands)
# Download NHDPlus HR for Hawaii
hi_nhd_url <- "https://prd-tnm.s3.amazonaws.com/StagedProducts/Hydrography/NHDPlusHR/Beta/GDB/NHDPLUS_H_2001_HU4_GDB.zip"
# Alternative: use nhdplusTools::get_nhdplus() with AOI polygon

ahupuaa_sf <- sf::st_read("data/ahupuaa_boundaries.geojson", quiet = TRUE)

# Get flowlines within Oahu bounding box as a test
oahu_bbox <- ahupuaa_sf |>
  dplyr::filter(stringr::str_detect(mokupuni, regex("oahu", ignore_case = TRUE))) |>
  sf::st_bbox() |>
  sf::st_as_sfc() |>
  sf::st_as_sf()

flowlines <- nhdplusTools::get_nhdplus(AOI = oahu_bbox, realization = "flowline")
saveRDS(flowlines, "data/nhd_oahu_flowlines.rds")
```

```{r}
#| label: claim4-landscape-regression
#| eval: false
# Regression: mean NO3+NO2 per ahupuaʻa ~ cesspool_density + pct_impervious
# + stream_density + watershed_area
# Requires: ccap_by_ahupuaa.csv (from Claim 2) and nhd stream density (above)

chem_aug <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE) |>
  filter(!is.na(island)) |>
  left_join(read_csv("data/ahupuaa_cache.csv",      show_col_types = FALSE), by = "sample_id") |>
  left_join(read_csv("data/cesspool_by_ahupuaa.csv", show_col_types = FALSE) |>
              select(ahupuaa, moku, cesspool_density_per_km2),
            by = c("ahupuaa", "moku"))

# Mean nutrient loading per ahupuaʻa
ahupuaa_nutrients <- chem_aug |>
  group_by(island, ahupuaa, moku, cesspool_density_per_km2) |>
  summarise(
    mean_NO3NO2  = mean(NO3NO2,  na.rm = TRUE),
    mean_PO4     = mean(PO4,     na.rm = TRUE),
    n_samples    = n(),
    .groups = "drop"
  ) |>
  filter(n_samples >= 3)  # only ahupuaʻa with enough samples

# Simple log-log regression (add ccap and nhd predictors when available)
lm(log1p(mean_NO3NO2) ~ log1p(cesspool_density_per_km2), data = ahupuaa_nutrients) |>
  summary()
```

---

## Claim 5 · community science enabled detection

**Goal:** Quantify the monitoring gap — show that conventional water quality monitoring in Hawaiʻi would not have captured this event at the spatial or temporal resolution needed. Two lines of evidence: (1) spatial coverage comparison, (2) sampling timing relative to storm.

**Needed dataset: EPA Water Quality Portal (WQP)**

- Package: `dataRetrieval` (same package as USGS NWIS)
- Function: `readWQPdata()` — returns all water quality records for a state/date range
- Coverage: includes Hawaii DOH, NOAA, EPA BEACH Act, university monitoring programs

```{r}
#| label: claim5-wqp-query
#| eval: false
# Query all Hawaii coastal nutrient samples from WQP in the 60 days
# surrounding the storm. Compare: how many samples? At how many unique sites?
# How close to shore? How soon after the storm?

wqp_hi <- readWQPdata(
  statecode      = "US:15",          # Hawaii FIPS
  characteristicName = c("Nitrate", "Nitrate + Nitrite", "Phosphate",
                         "Silica", "Specific conductance"),
  startDateLo    = "2026-02-01",
  startDateHi    = "2026-05-01"
)

# Cache immediately — API calls can be slow
saveRDS(wqp_hi, "data/wqp_hawaii_cache.rds")

# Summary: how many unique sites, samples, and dates near storm window?
wqp_hi |>
  mutate(date = as.Date(ActivityStartDate)) |>
  summarise(
    n_samples      = n(),
    n_sites        = n_distinct(MonitoringLocationIdentifier),
    n_storm_window = sum(between(date, as.Date("2026-03-10"), as.Date("2026-04-15")),
                         na.rm = TRUE)
  )
```

```{r}
#| label: claim5-coverage-map
#| eval: false
# Map conventional monitoring sites (WQP) vs community science sites (chem)
# to visualize the spatial coverage gap.
# Expects: wqp_hi cached above

wqp_hi   <- readRDS("data/wqp_hawaii_cache.rds")
chem     <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE)
ahupuaa  <- sf::st_read("data/ahupuaa_boundaries.geojson", quiet = TRUE)

# WQP sites as sf
wqp_sf <- wqp_hi |>
  distinct(MonitoringLocationIdentifier, .keep_all = TRUE) |>
  filter(!is.na(LongitudeMeasure), !is.na(LatitudeMeasure)) |>
  sf::st_as_sf(coords = c("LongitudeMeasure", "LatitudeMeasure"), crs = 4326) |>
  mutate(source = "Conventional monitoring")

# community science sites as sf
cs_sf <- chem |>
  filter(!is.na(longitude), !is.na(latitude)) |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  mutate(source = "community science (this study)")


# Hawaii land for background
hi_land <- ahupuaa |>
  dplyr::filter(stringr::str_detect(mokupuni,
    regex("oahu|maui|molokai|lanai", ignore_case = TRUE))) |>
  sf::st_union() |> sf::st_as_sf()

# Combined map
bind_rows(
  wqp_sf |> select(source, geometry),
  cs_sf  |> select(source, geometry)
) |>
  ggplot() +
  geom_sf(data = hi_land, fill = "gray80", color = "gray50", linewidth = 0.3) +
  geom_sf(aes(color = source, size = source), alpha = 0.7) +
  scale_color_manual(values = c("Conventional monitoring"   = "firebrick",
                                "community science (this study)" = "#1565C0")) +
  scale_size_manual(values  = c("Conventional monitoring"   = 2.5,
                                "community science (this study)" = 1.0)) +
  facet_wrap(~source) +
  theme_minimal() +
  theme(legend.position = "none", axis.text = element_blank()) +
  labs(title = "Spatial monitoring coverage: conventional vs. community science")
```

---

*Cite these data:* Silbiger, N., Kealoha, A., & Kahanamoku, S. (2026). *What's In Your Water? — Hawaiʻi Coastal Water Quality Samples, March 2026 Kona Low.* Zenodo. DOI: 10.5281/zenodo.19410137
