---
title: "What's In Your Water — Paper Figures"
subtitle: "4 main figures for manuscript submission"
author: "Silbiger, N., Kealoha, A., & Kahanamoku, S."
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: flatly
    code-fold: true
    fig-width: 10
    fig-height: 7
    lang: en
execute:
  warning: false
  message: false
---

```{r}
#| label: packages

library(tidyverse)
library(scales)
library(patchwork)
library(here)
library(cowplot)
library(ggrepel)

# Spatial
library(sf)
library(gstat)
library(stars)
library(terra)
library(maptiles)
library(tidyterra)
library(rnaturalearth)
library(rnaturalearthdata)

# Clustering & ML
library(cluster)
library(ranger)

# Misc
library(readxl)
library(lubridate)
library(dataRetrieval)
```

```{r}
#| label: helpers

normalize_name <- function(x) {
  x |>
    stringr::str_replace_all("[ʻʼ'']", "") |>
    stringr::str_replace_all("ā", "a") |>
    stringr::str_to_lower() |>
    stringr::str_trim()
}

# Island colour palette (consistent across all figures)
island_pal <- c(
  "O'ahu"    = "#1565C0",
  "Maui"     = "#2E7D32",
  "Moloka'i" = "#E65100",
  "Lāna'i"   = "#6A1B9A"
)

recode_island <- c(
  "Oahu"    = "O'ahu",
  "Maui"    = "Maui",
  "Molokai" = "Moloka'i",
  "Lanai"   = "Lāna'i"
)

# Water-type cluster palette
cluster_pal <- c(
  "Marine background"  = "#1565C0",
  "Nearshore enriched" = "#2E7D32",
  "Runoff-dominated"   = "#E65100"
)

# Reef health thresholds (µmol/L)
reef_thresholds <- c(NO3NO2 = 1, PO4 = 0.1, SiO2 = 5, NH3 = 1)

# Shared time-window constants (used in Fig 1 and across the paper)
storm_xmin <- as.Date("2026-03-10"); storm_xmax <- as.Date("2026-03-25")
samp_xmin  <- as.Date("2026-03-28"); samp_xmax  <- as.Date("2026-04-08")
plot_xlim  <- as.Date(c("2026-01-15", "2026-06-05"))

window_rects <- list(
  annotate("rect", xmin = storm_xmin, xmax = storm_xmax,
           ymin = -Inf, ymax = Inf, fill = "#BBDEFB", alpha = 0.75),
  annotate("rect", xmin = samp_xmin,  xmax = samp_xmax,
           ymin = -Inf, ymax = Inf, fill = "#C8E6C9", alpha = 0.75)
)

# Shared time-series theme (stacked panels — no x axis on upper panels)
ts_theme <- theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title.x       = element_blank(),
    axis.text.x        = element_blank(),
    axis.ticks.x       = element_blank(),
    legend.position    = "none",
    plot.margin        = margin(3, 6, 1, 6),
    axis.title.y       = element_text(size = 8.5)
  )

# kriging scale metadata
var_meta <- list(
  NO3NO2 = list(
    label     = expression(atop(NO[3]+NO[2], "("*mu*"mol/L)")),
    limits    = c(0, 10),
    breaks    = c(0.5, 1, 3, 5, 10),
    labels    = c("0.5", "1", "3", "5", "≥10"),
    trans     = "log1p",
    direction = 1
  ),
  PO4 = list(
    label     = expression(atop(PO[4], "("*mu*"mol/L)")),
    limits    = c(0, 1),
    breaks    = c(0.05, 0.1, 0.3, 0.5, 1),
    labels    = c("0.05", "0.1", "0.3", "0.5", "≥1"),
    trans     = "log1p",
    direction = 1
  ),
  SiO2 = list(
    label     = expression(atop(SiO[2], "("*mu*"mol/L)")),
    limits    = c(0, 50),
    breaks    = c(1, 5, 10, 20, 50),
    labels    = c("1", "5", "10", "20", "≥50"),
    trans     = "log1p",
    direction = 1
  ),
  NH3 = list(
    label     = expression(atop(NH[3], "("*mu*"mol/L)")),
    limits    = c(0, 3),
    breaks    = c(0.1, 0.5, 1, 2, 3),
    labels    = c("0.1", "0.5", "1", "2", "≥3"),
    trans     = "log1p",
    direction = 1
  ),
  Salinity = list(
    label     = expression(atop("Salinity", "(PSU)")),
    limits    = c(28, 36),
    breaks    = c(28, 30, 32, 34, 36),
    labels    = c("≤28", "30", "32", "34", "36"),
    trans     = "identity",
    direction = -1
  )
)
```

```{r}
#| label: load-data

chem <- read_csv("data/clean_chemdata.csv", show_col_types = FALSE) |>
  filter(!is.na(island)) |>
  mutate(island = stringr::str_replace_all(island, "[ʻʼ'']", "'")) |>
  left_join(read_csv("data/ahupuaa_cache.csv",             show_col_types = FALSE), by = "sample_id") |>
  left_join(read_csv("data/sample_cesspool_distances.csv", show_col_types = FALSE), by = "sample_id") |>
  left_join(
    read_csv("data/cesspool_by_ahupuaa.csv", show_col_types = FALSE) |>
      select(ahupuaa, moku, cesspool_density_per_km2, n_cesspools),
    by = c("ahupuaa", "moku")
  ) |>
  left_join(read_csv("data/cesspool_buffer_counts.csv", show_col_types = FALSE), by = "sample_id") |>
  left_join(
    read_csv("data/wave_fetch.csv", show_col_types = FALSE) |>
      select(sample_id, fetch_mean_km, fetch_min_km),
    by = "sample_id"
  ) |>
  left_join(read_csv("data/sample_waveheight.csv", show_col_types = FALSE), by = "sample_id") |>
  left_join(
    read_csv("data/pop_density_ahupuaa.csv", show_col_types = FALSE) |>
      select(ahupuaa, island, pop_density_km2),
    by = c("ahupuaa", "island")
  )

ahupuaa_raw <- st_read("data/ahupuaa_boundaries.geojson", quiet = TRUE)
```

```{r}
#| label: kriging-functions

krige_island_map <- function(data, island_name, ahupuaa_sf,
                              var = "NO3NO2", res_m = 250,
                              band_km = 3, tile_zoom = 11) {
  meta    <- var_meta[[var]]
  use_log <- !is.null(meta) && meta$trans == "log1p"
  if (is.null(meta)) meta <- list(
    label = var, limits = NULL, breaks = waiver(), labels = waiver(),
    trans = "identity", direction = 1
  )

  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]),
           !is.na(latitude), !is.na(longitude))

  match_idx  <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  land_sf    <- st_as_sf(st_sf(geometry = land_union))

  coast_outer <- st_buffer(land_union, dist = band_km * 1000)
  coast_band  <- st_difference(coast_outer, land_union)
  coast_bbox  <- st_bbox(coast_outer)

  pts_proj <- island_data |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(raw_val = .data[[var]],
           log_val = if (use_log) log1p(raw_val) else raw_val,
           X = st_coordinates(geometry)[, 1],
           Y = st_coordinates(geometry)[, 2]) |>
    group_by(X, Y) |>
    summarise(log_val = mean(log_val, na.rm = TRUE),
              raw_val = mean(raw_val, na.rm = TRUE), .groups = "drop") |>
    st_as_sf(coords = c("X", "Y"), crs = 32604)

  pts_proj_all <- island_data |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(val = .data[[var]],
           X   = st_coordinates(geometry)[, 1],
           Y   = st_coordinates(geometry)[, 2]) |>
    filter(X >= coast_bbox["xmin"], X <= coast_bbox["xmax"],
           Y >= coast_bbox["ymin"], Y <= coast_bbox["ymax"])

  island_diag <- sqrt(diff(st_bbox(pts_proj)[c("xmin", "xmax")])^2 +
                      diff(st_bbox(pts_proj)[c("ymin", "ymax")])^2)
  min_range <- island_diag / 5
  v_emp     <- variogram(log_val ~ 1, pts_proj)
  init_ps   <- var(pts_proj$log_val, na.rm = TRUE)
  v_auto <- tryCatch(
    fit.variogram(v_emp, vgm(init_ps * 0.8, "Sph", min_range, init_ps * 0.2)),
    error = function(e) NULL
  )
  v_fit <- if (!is.null(v_auto) && v_auto$range[2] >= min_range) v_auto else
    vgm(init_ps * 0.8, "Sph", min_range, init_ps * 0.2)

  sample_cov <- st_buffer(st_union(pts_proj), dist = min_range)
  grid_pts <- st_make_grid(coast_band, cellsize = res_m, what = "centers") |>
    st_as_sf() |> st_filter(coast_band) |> st_filter(sample_cov)

  k_out  <- suppressWarnings(krige(log_val ~ 1, pts_proj, grid_pts, model = v_fit, nmax = 40))
  pred_df <- st_coordinates(k_out) |>
    as.data.frame() |>
    mutate(pred = if (use_log) expm1(k_out$var1.pred) else k_out$var1.pred)

  ggplot() +
    geom_spatraster_rgb(
      data = get_tiles(st_transform(coast_outer, 4326),
                       provider = "Esri.WorldImagery", zoom = tile_zoom)
    ) +
    geom_raster(data = pred_df, aes(x = X, y = Y, fill = pred), alpha = 0.8) +
    geom_sf(data = land_sf, fill = NA, color = "white", linewidth = 0.4) +
    geom_sf(data = pts_proj_all, aes(fill = val),
            size = 1.0, shape = 21, color = "white", stroke = 0.3, alpha = 0.5) +
    scale_fill_viridis_c(
      name      = meta$label,
      option    = "plasma", direction = meta$direction,
      trans     = meta$trans, limits = meta$limits,
      breaks    = meta$breaks, labels = meta$labels,
      oob       = scales::squish, na.value = "transparent",
      guide     = guide_colorbar(barheight = unit(10, "lines"),
                                 barwidth  = unit(0.6, "lines"),
                                 ticks.linewidth = 0.5)
    ) +
    coord_sf(crs = 32604,
             xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = island_name) +
    theme_minimal(base_size = 11) +
    theme(axis.title = element_blank(), axis.text = element_text(size = 7),
          plot.title = element_text(face = "bold", size = 12),
          legend.position = "right")
}

# Returns pred_df + land polygon + coast_bbox for external compositing
krige_surface_data <- function(data, island_name, ahupuaa_sf,
                               var = "NO3NO2", res_m = 300, band_km = 3) {
  meta    <- var_meta[[var]]
  use_log <- !is.null(meta) && meta$trans == "log1p"
  island_data <- data |>
    filter(island == island_name, !is.na(.data[[var]]),
           !is.na(latitude), !is.na(longitude))
  match_idx  <- normalize_name(ahupuaa_sf$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_sf[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  coast_outer <- st_buffer(land_union, dist = band_km * 1000)
  coast_band  <- st_difference(coast_outer, land_union)
  coast_bbox  <- st_bbox(coast_outer)
  pts_proj <- island_data |>
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
    st_transform(32604) |>
    mutate(raw_val = .data[[var]],
           log_val = if (use_log) log1p(raw_val) else raw_val,
           X = st_coordinates(geometry)[, 1],
           Y = st_coordinates(geometry)[, 2]) |>
    group_by(X, Y) |>
    summarise(log_val = mean(log_val, na.rm = TRUE),
              raw_val = mean(raw_val, na.rm = TRUE), .groups = "drop") |>
    st_as_sf(coords = c("X", "Y"), crs = 32604)
  island_diag <- sqrt(diff(st_bbox(pts_proj)[c("xmin", "xmax")])^2 +
                      diff(st_bbox(pts_proj)[c("ymin", "ymax")])^2)
  min_range <- island_diag / 5
  v_emp <- variogram(log_val ~ 1, pts_proj)
  v_auto <- tryCatch(
    fit.variogram(v_emp, vgm(var(pts_proj$log_val) * 0.8, "Sph",
                              min_range, var(pts_proj$log_val) * 0.2)),
    error = function(e) NULL
  )
  v_fit <- if (!is.null(v_auto) && v_auto$range[2] >= min_range) v_auto else
    vgm(var(pts_proj$log_val) * 0.8, "Sph", min_range, var(pts_proj$log_val) * 0.2)
  sample_cov <- st_buffer(st_union(pts_proj), dist = min_range)
  grid_pts <- st_make_grid(coast_band, cellsize = res_m, what = "centers") |>
    st_as_sf() |> st_filter(coast_band) |> st_filter(sample_cov)
  k_out  <- suppressWarnings(krige(log_val ~ 1, pts_proj, grid_pts, model = v_fit, nmax = 40))
  pred_df <- st_coordinates(k_out) |>
    as.data.frame() |>
    mutate(pred = if (use_log) expm1(k_out$var1.pred) else k_out$var1.pred)
  list(pred_df    = pred_df,
       land_sf    = st_as_sf(st_sf(geometry = land_union)),
       coast_bbox = coast_bbox,
       coast_outer = coast_outer)
}
```

---

# Figure 1 — Storm Cascade & Monitoring Context

*Rain → streams → coastal nutrients → bloom; conventional monitoring absent throughout.*

```{r}
#| label: fig1-data

rain_df <- read_csv("hawaii_ncep_janjun_cache.csv", show_col_types = FALSE) |>
  filter(island %in% c("Oahu", "Maui", "Molokai"),
         date >= plot_xlim[1], date <= plot_xlim[2]) |>
  mutate(island = recode_island[island]) |>
  group_by(island, date) |>
  summarise(precip_mm = mean(precip_mmday, na.rm = TRUE), .groups = "drop")

dis_df <- read_csv("data/usgs_discharge_cache.csv", show_col_types = FALSE,
                   col_types = cols(site_no = col_character())) |>
  dataRetrieval::renameNWISColumns() |>
  filter(site_no %in% c("16211600", "16400000", "16508000"),
         Date >= plot_xlim[1], Date <= plot_xlim[2], Flow > 0.5) |>
  left_join(
    tibble(site_no = c("16211600", "16400000", "16508000"),
           island  = c("O'ahu",    "Moloka'i", "Maui")),
    by = "site_no"
  )

chla_df <- readRDS("hawaii_chlorophyll_cache.rds") |>
  filter(island %in% c("Oahu", "Maui", "Molokai"),
         date >= plot_xlim[1], date <= plot_xlim[2]) |>
  mutate(island = recode_island[island]) |>
  group_by(island, date) |>
  summarise(mean_chla = mean(chlor_a, na.rm = TRUE),
            n_valid   = sum(!is.na(chlor_a)), .groups = "drop") |>
  filter(n_valid >= 8, mean_chla > 0)

nut_long_fig1 <- chem |>
  filter(island %in% c("O'ahu", "Maui", "Moloka'i")) |>
  mutate(island = factor(island, levels = c("Maui", "Moloka'i", "O'ahu"))) |>
  select(island, NO3NO2, NH3, PO4, SiO2) |>
  pivot_longer(c(NO3NO2, NH3, PO4, SiO2), names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>
  mutate(nutrient = factor(
    recode(nutrient, "NO3NO2" = "NO3+NO2", "NH3" = "NH3",
           "PO4" = "PO4",     "SiO2" = "SiO2"),
    levels = c("NO3+NO2", "NH3", "PO4", "SiO2")
  ))

thresh_long_fig1 <- tibble(
  nutrient  = c("NO3+NO2", "NH3", "PO4", "SiO2"),
  threshold = c(1, 1, 0.1, 5)
)
```

```{r}
#| label: fig1-panels

# Panel A: Rainfall
pA_f1 <- ggplot(rain_df, aes(x = date, y = precip_mm, fill = island, color = island)) +
  window_rects +
  geom_area(position = "identity", alpha = 0.30, linewidth = 0.35) +
  annotate("text", x = as.Date("2026-03-17"), y = 108,
           label = "Kona Low", size = 2.6, color = "#1A237E", hjust = 0.5, vjust = 0) +
  scale_fill_manual(values  = island_pal) +
  scale_color_manual(values = island_pal) +
  scale_x_date(limits = plot_xlim, expand = c(0, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(y = "Rainfall\n(mm/day)", tag = "A") +
  ts_theme

# Panel B: Stream discharge (log scale)
pB_f1 <- ggplot(dis_df, aes(x = Date, y = Flow, color = island, group = site_no)) +
  window_rects +
  geom_line(linewidth = 0.65, alpha = 0.9, na.rm = TRUE) +
  scale_color_manual(values = island_pal) +
  scale_x_date(limits = plot_xlim, date_labels = "%b %Y", expand = c(0, 0)) +
  scale_y_log10(breaks = c(1, 10, 100, 1000),
                labels = scales::label_number(big.mark = ",")) +
  coord_cartesian(ylim = c(0.8, 3000)) +
  labs(y = "Discharge\n(cfs, log)", tag = "B") +
  ts_theme

# Panel C: Coastal chlorophyll-a (VIIRS 4 km)
pC_f1 <- ggplot(chla_df, aes(x = date, y = mean_chla, color = island)) +
  window_rects +
  geom_point(size = 0.7, alpha = 0.50, na.rm = TRUE) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE, linewidth = 0.9, na.rm = TRUE) +
  scale_color_manual(values = island_pal, name = NULL) +
  scale_x_date(limits = plot_xlim, date_labels = "%b %Y", expand = c(0, 0)) +
  scale_y_log10(breaks = c(0.1, 0.3, 1, 3, 10),
                labels = scales::label_number(accuracy = 0.1)) +
  labs(y = expression("Chl-"*italic(a)*"\n(mg m"^{-3}*", log)"), tag = "C") +
  ts_theme +
  theme(axis.text.x  = element_text(size = 8),
        axis.ticks.x = element_line(linewidth = 0.3),
        legend.position = "right",
        legend.text     = element_text(size = 8.5),
        plot.margin     = margin(1, 6, 2, 6))

# Panel D: Post-storm nutrient distributions by island (violin + box + threshold)
pD_f1 <- ggplot(nut_long_fig1, aes(x = island, y = conc, fill = island, color = island)) +
  geom_hline(data = thresh_long_fig1, aes(yintercept = threshold),
             linetype = "dashed", color = "#B71C1C", linewidth = 0.5,
             inherit.aes = FALSE) +
  geom_violin(alpha = 0.28, linewidth = 0.3, trim = TRUE) +
  geom_boxplot(width = 0.17, outliers = FALSE,
               fill = "white", linewidth = 0.45, alpha = 0.9) +
  scale_fill_manual(values  = island_pal) +
  scale_color_manual(values = island_pal) +
  scale_y_log10(breaks = c(0.01, 0.1, 1, 10, 100),
                labels = scales::label_number(drop0trailing = TRUE)) +
  facet_wrap(~nutrient, ncol = 1, scales = "free_y") +
  labs(y = expression(mu*"mol L"^{-1}), x = NULL, tag = "D") +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x  = element_text(angle = 30, hjust = 1, vjust = 1, size = rel(0.85)),
    strip.text   = element_text(size = 8.5, face = "bold"),
    legend.position = "none",
    plot.margin  = margin(4, 8, 4, 2),
    axis.title.y = element_text(size = 8.5),
    plot.tag     = element_text(face = "bold")
  )
```

```{r}
#| label: fig1-assemble
#| fig-height: 9
#| fig-width: 12
#| fig-cap: "**Figure 1.** Storm cascade and citizen-science response. **(A)** Island-mean daily rainfall (NCEP GDAS, mm/day). **(B)** Stream discharge at three USGS gauges (log scale; Makaha Stream, Oʻahu; Halawa Stream, Molokaʻi; Hanawi Stream, Maui). **(C)** VIIRS NOAA-20 4 km island-mean coastal chlorophyll-a (log scale; loess smoother). **(D)** Post-storm nutrient concentrations by island from citizen-science sampling (Mar 28–Apr 8; log scale). Dashed red lines = reef health thresholds. Blue shading = Kona Low storm window (Mar 10–25); green shading = citizen-science sampling window. Conventional monitoring records in WQP = 0 during this period (last record: Dec 18, 2025)."

fig1 <- ((pA_f1 / pB_f1 / pC_f1) | pD_f1) +
  plot_layout(widths = c(3, 1.4)) +
  plot_annotation(
    theme = theme(plot.margin = margin(6, 6, 4, 6))
  )

fig1
ggsave(here("output", "fig1_storm_cascade.pdf"), fig1,
       width = 12, height = 9, device = cairo_pdf)
ggsave(here("output", "fig1_storm_cascade.png"), fig1,
       width = 12, height = 9, dpi = 300)
```

---

# Figure 2 — Spatial Fingerprints of Nutrient Enrichment

*Island-scale hotspot pattern; historically unprecedented; independently validated.*

```{r}
#| label: fig2-kriging
#| fig-height: 6
#| fig-width: 15
#| fig-cap: "**Figure 2A.** Ordinary kriged NO₃+NO₂ surfaces for Oʻahu, Maui, and Molokaʻi (250 m grid, 3 km coastal band, log₁p colour scale). Circles = citizen-science sample locations on same scale. Colour scale is uniform across islands. Lānaʻi shown in Extended Data Fig. 1."

oahu_nn  <- krige_island_map(chem, "O'ahu",    ahupuaa_raw, var = "NO3NO2") +
  theme(legend.position = "none")
maui_nn  <- krige_island_map(chem, "Maui",     ahupuaa_raw, var = "NO3NO2") +
  theme(legend.position = "none")
molo_nn  <- krige_island_map(chem, "Moloka'i", ahupuaa_raw, var = "NO3NO2", tile_zoom = 12)

# All three panels share identical colour scales — one legend on the right is sufficient
fig2a <- oahu_nn | maui_nn | molo_nn

fig2a
ggsave(here("output", "fig2a_kriging_NO3NO2.pdf"), fig2a,
       width = 15, height = 6, device = cairo_pdf)
ggsave(here("output", "fig2a_kriging_NO3NO2.png"), fig2a,
       width = 15, height = 6, dpi = 300)
```

```{r}
#| label: fig2-hui-data

bg_hui_loc <- bind_rows(
  read_excel("data/background_nutrients/hui-south-maui-thru-2026-1st-quarter.0.xlsx",
             sheet = "data") |>
    mutate(Turbidity = as.numeric(Turbidity), region = "South Maui"),
  read_excel("data/background_nutrients/hui-west-maui-thru-2026-1st-quarter.0.xlsx",
             sheet = "data") |>
    mutate(Turbidity = as.numeric(Turbidity), region = "West Maui")
) |>
  mutate(
    Date           = as.Date(Date),
    NNN_umol       = as.numeric(NNN)       / 14.007,
    Phosphate_umol = as.numeric(Phosphate) / 30.974,
    Silicate_umol  = as.numeric(Silicate)  / 28.086,
    NH4_umol       = as.numeric(NH4)        / 14.007
  )
```

```{r}
#| label: fig2-hui-validation
#| fig-height: 7
#| fig-width: 8
#| fig-cap: "**Figure 2B.** Maui NO₃+NO₂ kriging surface with independent Hui o Ka Wai Ola monitoring stations overlaid. Triangle size proportional to fold-change above station-specific 10-year historical median. Labels = stations with fold-change > 15×. Agreement between two fully independent datasets confirms the hotspot pattern is real."

# Build Hui station sf with fold-change encoding
hui_col <- "NNN_umol"
hui_post <- bg_hui_loc |>
  filter(year(Date) == 2026, month(Date) == 3,
         Date >= as.Date("2026-03-14"), !is.na(.data[[hui_col]])) |>
  group_by(Station, SiteName, region, Lat, Long) |>
  summarise(val_umol = mean(.data[[hui_col]], na.rm = TRUE), .groups = "drop")

hui_hist_med <- bg_hui_loc |>
  filter(!is.na(.data[[hui_col]])) |>
  group_by(Station) |>
  summarise(med_hist = median(.data[[hui_col]], na.rm = TRUE), .groups = "drop")

log_lo   <- log1p(1); log_hi <- log1p(20)
grey_pal <- colorRampPalette(c("gray20", "white"))(1000)

hui_sf <- bg_hui_loc |>
  distinct(Station, SiteName, region, Lat, Long) |>
  filter(!is.na(Lat), !is.na(Long)) |>
  left_join(hui_post    |> select(Station, val_umol), by = "Station") |>
  left_join(hui_hist_med,                              by = "Station") |>
  mutate(
    fold_change = val_umol / med_hist,
    fc_safe     = pmin(pmax(replace_na(fold_change, 1), 1), 20),
    tri_fill    = grey_pal[round((log1p(fc_safe) - log_lo) /
                                    (log_hi - log_lo) * 999) + 1],
    tri_fill    = case_when(
      is.na(fold_change) & !is.na(med_hist) ~ "gray50",
      is.na(med_hist)                        ~ "gray70",
      TRUE                                   ~ tri_fill
    ),
    fold_change = replace_na(fold_change, 1)
  ) |>
  st_as_sf(coords = c("Long", "Lat"), crs = 4326) |>
  st_transform(32604)

# Build kriging surface for Maui
kd_maui <- krige_surface_data(chem, "Maui", ahupuaa_raw, var = "NO3NO2")
cs_maui  <- chem |>
  filter(island == "Maui", !is.na(NO3NO2), !is.na(latitude), !is.na(longitude)) |>
  mutate(val = NO3NO2) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)
sat_maui <- get_tiles(
  st_as_sfc(kd_maui$coast_bbox) |> st_set_crs(32604) |> st_transform(4326),
  provider = "Esri.WorldImagery", zoom = 11
)

meta_nn <- var_meta[["NO3NO2"]]

fig2b <- ggplot() +
  geom_spatraster_rgb(data = sat_maui) +
  geom_raster(data = kd_maui$pred_df, aes(x = X, y = Y, fill = pred), alpha = 0.8) +
  geom_sf(data = kd_maui$land_sf, fill = NA, color = "white", linewidth = 0.35) +
  geom_sf(data = cs_maui, aes(fill = val),
          shape = 21, size = 1.0, color = "white", stroke = 0.3, alpha = 0.5) +
  scale_fill_viridis_c(
    name = meta_nn$label, option = "plasma", direction = meta_nn$direction,
    trans = meta_nn$trans, limits = meta_nn$limits,
    breaks = meta_nn$breaks, labels = meta_nn$labels,
    oob = scales::squish, na.value = "transparent",
    guide = guide_colorbar(barheight = unit(10, "lines"), barwidth = unit(0.6, "lines"),
                           ticks.linewidth = 0.5)
  ) +
  geom_sf(data = hui_sf,
          aes(size = fold_change, fill = I(tri_fill)),
          color = "white", shape = 24, stroke = 0.8, alpha = 0.85) +
  geom_label_repel(
    data = hui_sf |> filter(fold_change > 15),
    aes(label    = paste0(SiteName, "\n", round(fold_change, 0), "×"),
        geometry = geometry),
    stat = "sf_coordinates", size = 2.5, fontface = "bold", color = "#B71C1C",
    fill = alpha("white", 0.85), label.size = 0.2,
    nudge_y = 4500, min.segment.length = 0
  ) +
  scale_size_continuous(
    name   = "Fold-change\n(Mar 2026 /\nhistorical median)",
    range  = c(1.0, 8), limits = c(1, 20),
    breaks = c(1, 3, 5, 10, 20), labels = c("1", "3", "5", "10", "≥20"),
    guide  = guide_legend(
      theme = theme(legend.key = element_rect(fill = "gray30", color = NA)),
      override.aes = list(shape = 24, color = "white", stroke = 0.8, alpha = 0.85,
                          fill = c("#333333", "#6F6F6F", "#929292", "#C6C6C6", "#FFFFFF"))
    )
  ) +
  coord_sf(crs = 32604,
           xlim = c(kd_maui$coast_bbox["xmin"], kd_maui$coast_bbox["xmax"]),
           ylim = c(kd_maui$coast_bbox["ymin"], kd_maui$coast_bbox["ymax"]),
           expand = FALSE) +
  labs(caption = "△ Hui o Ka Wai Ola long-term monitoring stations (Falinski et al.)") +
  theme_minimal(base_size = 11) +
  theme(axis.title = element_blank(), axis.text = element_text(size = 7),
        plot.caption = element_text(size = 8, color = "gray50", hjust = 0),
        legend.position = "right")

fig2b
ggsave(here("output", "fig2b_maui_hui_validation.pdf"), fig2b,
       width = 8, height = 7, device = cairo_pdf)
ggsave(here("output", "fig2b_maui_hui_validation.png"), fig2b,
       width = 8, height = 7, dpi = 300)
```

```{r}
#| label: fig2-salinity-mixing
#| fig-height: 5.5
#| fig-width: 9
#| fig-cap: "**Figure 2C.** Salinity–nutrient mixing relationships. Orange = storm-influenced (Sal ≤ 34 PSU) with OLS trend ± 95% CI. Blue horizontal line = median marine background (Sal > 34 PSU). Dotted vertical line marks Sal = 34 PSU. Zeros excluded as below-detection."

sal_zone_pal <- c(
  "Storm-influenced\n(≤ 34 PSU)" = "#E65100",
  "Marine background\n(> 34 PSU)" = "#1565C0"
)

mixing_long <- chem |>
  filter(!is.na(Salinity)) |>
  select(island, Salinity, NO3NO2, PO4, SiO2, NH3) |>
  pivot_longer(c(NO3NO2, PO4, SiO2, NH3), names_to = "nutrient", values_to = "conc") |>
  filter(!is.na(conc), conc > 0) |>
  mutate(
    nutrient = factor(nutrient,
      levels = c("NO3NO2", "PO4", "SiO2", "NH3"),
      labels = c("NO3+NO2 (umol/L)", "PO4 (umol/L)",
                 "SiO2 (umol/L)",    "NH3 (umol/L)")),
    sal_zone = factor(
      if_else(Salinity > 34,
              "Marine background\n(> 34 PSU)",
              "Storm-influenced\n(≤ 34 PSU)"),
      levels = c("Storm-influenced\n(≤ 34 PSU)", "Marine background\n(> 34 PSU)"))
  )

marine_medians <- mixing_long |>
  filter(sal_zone == "Marine background\n(> 34 PSU)") |>
  group_by(nutrient) |>
  summarise(med = median(conc, na.rm = TRUE), .groups = "drop")

fig2c <- ggplot(mixing_long, aes(x = Salinity, y = conc, color = sal_zone)) +
  geom_vline(xintercept = 34, linetype = "dotted", color = "gray40", linewidth = 0.6) +
  geom_point(size = 1.2, alpha = 0.40) +
  geom_smooth(data = \(d) filter(d, sal_zone == "Storm-influenced\n(≤ 34 PSU)"),
              method = "lm", se = TRUE, linewidth = 0.85, alpha = 0.15,
              color = "#E65100") +
  geom_hline(data = marine_medians, aes(yintercept = med),
             color = "#1565C0", linetype = "solid", linewidth = 0.8,
             inherit.aes = FALSE) +
  scale_color_manual(values = sal_zone_pal, name = NULL) +
  scale_y_log10(labels = label_number(drop0trailing = TRUE)) +
  scale_x_continuous(limits = c(NA, 36.5)) +
  facet_wrap(~nutrient, scales = "free_y", ncol = 2) +
  labs(x = "Salinity (PSU)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "bold"))

fig2c
ggsave(here("output", "fig2c_salinity_mixing.pdf"), fig2c,
       width = 9, height = 5.5, device = cairo_pdf)
ggsave(here("output", "fig2c_salinity_mixing.png"), fig2c,
       width = 9, height = 5.5, dpi = 300)
```

---

# Figure 3 — Causes: Landscape Predictability

*Pre-existing landscape infrastructure predicts post-storm water type.*

```{r}
#| label: fig3-pam-clustering

# PAM clustering on log-nutrients + salinity (k = 3)
chem_clust <- chem |>
  filter(!is.na(island), !is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  mutate(log_NO3NO2 = log1p(NO3NO2), log_PO4 = log1p(PO4), log_SiO2 = log1p(SiO2))

clust_mat <- chem_clust |>
  select(Salinity, log_NO3NO2, log_PO4, log_SiO2) |>
  scale()

set.seed(8341)
pam3 <- pam(clust_mat, k = 3, pamonce = 5)

# Reorder clusters 1 = lowest NO3NO2 (marine bg), 3 = highest (runoff-dominated)
chem_clust <- chem_clust |>
  mutate(temp_cluster = pam3$clustering)
cluster_medians <- chem_clust |>
  group_by(temp_cluster) |>
  summarise(med_NO3 = median(NO3NO2, na.rm = TRUE), .groups = "drop") |>
  arrange(med_NO3) |>
  mutate(new_cluster = row_number())
cluster_remap <- setNames(cluster_medians$new_cluster, cluster_medians$temp_cluster)

chem_clust <- chem_clust |>
  mutate(cluster = as.numeric(cluster_remap[temp_cluster])) |>
  select(-temp_cluster)

cluster_labels <- c("1" = "Marine background",
                    "2" = "Nearshore enriched",
                    "3" = "Runoff-dominated")

chem_clust <- chem_clust |>
  mutate(
    cluster       = factor(cluster),
    cluster_label = factor(cluster_labels[cluster],
                           levels = c("Marine background",
                                      "Nearshore enriched",
                                      "Runoff-dominated"))
  )

pam3$clustering <- as.numeric(cluster_remap[pam3$clustering])
```

```{r}
#| label: fig3-pam-heatmap
#| fig-height: 3.5
#| fig-width: 5.5

fp_vars   <- c("Salinity", "NO3NO2", "PO4", "SiO2")
fp_labels <- c("Salinity\n(PSU)", "NO₃+NO₂\n(µmol/L)", "PO₄\n(µmol/L)", "SiO₂\n(µmol/L)")

hm_df <- chem_clust |>
  group_by(cluster_label) |>
  summarise(across(all_of(fp_vars), ~median(., na.rm = TRUE)), .groups = "drop") |>
  pivot_longer(-cluster_label, names_to = "variable", values_to = "median_val") |>
  group_by(variable) |>
  mutate(z = (median_val - mean(median_val)) / sd(median_val)) |>
  ungroup() |>
  mutate(
    variable      = factor(variable, levels = fp_vars, labels = fp_labels),
    cluster_label = factor(cluster_label,
                           levels = c("Runoff-dominated", "Nearshore enriched", "Marine background")),
    val_label = case_when(
      variable == "Salinity\n(PSU)"    ~ as.character(round(median_val, 1)),
      variable == "NO₃+NO₂\n(µmol/L)" ~ as.character(round(median_val, 2)),
      variable == "PO₄\n(µmol/L)"     ~ as.character(round(median_val, 3)),
      TRUE                             ~ as.character(round(median_val, 1))
    )
  )

pA_f3 <- ggplot(hm_df, aes(x = variable, y = cluster_label, fill = z)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = val_label), size = 3, fontface = "bold",
            color = ifelse(abs(hm_df$z) > 1.2, "white", "gray20")) +
  scale_fill_gradient2(low = "#1565C0", mid = "white", high = "#B71C1C",
                       midpoint = 0, name = "Z-score") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = \(x) str_wrap(x, 14)) +
  labs(x = NULL, y = NULL, tag = "A") +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(size = 8.5, face = "bold"),
        axis.text.y = element_text(size = 9),
        plot.tag = element_text(face = "bold"))

pA_f3
```

```{r}
#| label: fig3-cluster-maps
#| fig-height: 3
#| fig-width: 12

clust_sf <- chem_clust |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

map_cluster_fn <- function(island_name, dot_size = 2.2) {
  match_idx  <- normalize_name(ahupuaa_raw$mokupuni) == normalize_name(island_name)
  land_union <- ahupuaa_raw[match_idx, ] |>
    st_transform(32604) |> st_make_valid() |> st_union()
  land_parts <- st_cast(land_union, "POLYGON")
  land_union <- land_parts[as.numeric(st_area(land_parts)) > 1e6] |> st_union()
  coast_bbox <- st_bbox(st_buffer(land_union, 3500))
  pts <- clust_sf |> filter(island == island_name)
  ggplot() +
    geom_sf(data = st_as_sf(st_sf(geometry = land_union)),
            fill = "gray78", color = "gray55", linewidth = 0.3) +
    geom_sf(data = pts, aes(color = cluster_label),
            size = dot_size, alpha = 0.85, shape = 16) +
    scale_color_manual(values = cluster_pal, name = NULL, drop = FALSE) +
    coord_sf(crs = 32604,
             xlim = c(coast_bbox["xmin"], coast_bbox["xmax"]),
             ylim = c(coast_bbox["ymin"], coast_bbox["ymax"]), expand = FALSE) +
    labs(title = island_name) +
    theme_void(base_size = 9) +
    theme(plot.title = element_text(face = "bold", size = 9, hjust = 0.5),
          legend.position = "none", plot.margin = margin(2, 2, 2, 2))
}

map_oahu  <- map_cluster_fn("O'ahu",    dot_size = 1.5)
map_maui  <- map_cluster_fn("Maui")
map_molo  <- map_cluster_fn("Moloka'i", dot_size = 2.5)
map_lanai <- map_cluster_fn("Lāna'i",   dot_size = 3.0)

legend_strip <- get_legend(
  ggplot(chem_clust, aes(x = NO3NO2, y = PO4, color = cluster_label)) +
    geom_point() +
    scale_color_manual(values = cluster_pal, name = "Water type (PAM k = 3)") +
    theme_minimal() +
    theme(legend.position = "bottom", legend.text = element_text(size = 9),
          legend.key.size = unit(0.9, "lines"))
)

pB_f3 <- plot_grid(
  plot_grid(map_oahu, map_maui, map_molo, map_lanai,
            nrow = 1, rel_widths = c(1.6, 1.4, 0.9, 0.7)),
  legend_strip,
  ncol = 1, rel_heights = c(1, 0.2)
)

pB_f3
```

```{r}
#| label: fig3-rf-data

# Load cached predictor files (all pre-computed)
stream_mouths_sf <- read_csv(here("data", "osm_stream_mouths.csv"), show_col_types = FALSE) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

ccap_summary <- read_csv(here("data", "ccap_ahupuaa_summary.csv"), show_col_types = FALSE)

slope_df <- read_csv(here("data", "ahupuaa_slope.csv"), show_col_types = FALSE)
ksat_df  <- read_csv(here("data", "ahupuaa_ksat.csv"),  show_col_types = FALSE)

outfalls_hi <- read_csv(here("data", "epa_facilities_hi.csv"), show_col_types = FALSE)
outfalls_sf <- outfalls_hi |>
  distinct(facility_name, lat, lon) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(32604)

discharge_per_ahu <- read_csv(here("data", "usgs_discharge_march2026.csv"),
                              show_col_types = FALSE,
                              col_types = cols(site_no = col_character()))
gauge_sf <- read_csv(here("data", "usgs_gauges_march2026.csv"),
                     show_col_types = FALSE,
                     col_types = cols(site_no = col_character())) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

sm_df <- read_csv(here("data", "nasa_power_soilmoist.csv"), show_col_types = FALSE) |>
  mutate(island = stringr::str_replace_all(island, "[ʻʼ'']", "'"))

livestock_by_island <- read_csv(here("data", "usda_nass_livestock_hi.csv"),
                                show_col_types = FALSE) |>
  mutate(island = stringr::str_replace_all(island, "[ʻʼ'']", "'"))

island_norm <- c("oahu" = "O'ahu", "maui" = "Maui",
                 "molokai" = "Moloka'i", "lanai" = "Lāna'i")
add_island_col <- function(df) {
  df |> mutate(island = island_norm[normalize_name(mokupuni)])
}

slope_j <- slope_df |> add_island_col() |> filter(!is.na(island)) |>
  group_by(ahupuaa, island) |>
  summarise(mean_slope_deg = mean(mean_slope_deg, na.rm = TRUE), .groups = "drop")
ksat_j <- ksat_df  |> add_island_col() |> filter(!is.na(island)) |>
  group_by(ahupuaa, island) |>
  summarise(mean_ksat_um_s = mean(mean_ksat_um_s, na.rm = TRUE), .groups = "drop")

# Compute stream mouth distances for all chem sites with coordinates
chem_sf_rf <- chem |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

nearest_sm  <- st_nearest_feature(chem_sf_rf, stream_mouths_sf)
chem_w_stream <- chem |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  mutate(dist_to_stream_mouth_m = as.numeric(
    st_distance(chem_sf_rf, stream_mouths_sf[nearest_sm, ], by_element = TRUE)
  ))

# Assign nearest gauge discharge (within 15 km)
chem_sf_d <- chem_w_stream |>
  filter(!is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)
nearest_g  <- st_nearest_feature(chem_sf_d, gauge_sf)
nearest_gd <- as.numeric(st_distance(chem_sf_d, gauge_sf[nearest_g, ], by_element = TRUE))

chem_w_dis <- chem_w_stream |>
  filter(!is.na(Salinity), !is.na(NO3NO2), !is.na(PO4), !is.na(SiO2)) |>
  mutate(
    mean_cfs_march    = if_else(nearest_gd > 15000, NA_real_, gauge_sf$mean_cfs_march[nearest_g]),
    log_discharge_cfs = log1p(mean_cfs_march)
  )

# Compute distance to nearest NPDES outfall
chem_sf_o  <- chem_w_dis |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)
outfall_nn  <- st_nearest_feature(chem_sf_o, outfalls_sf)
outfall_d   <- as.numeric(st_distance(chem_sf_o, outfalls_sf[outfall_nn, ], by_element = TRUE))

chem_enhanced <- chem_w_dis |>
  mutate(log_dist_outfall = log1p(outfall_d))
```

```{r}
#| label: fig3-rf-fit

# Build RF dataset (placeholder predictor set — to be finalised)
rf_base <- chem_enhanced |>
  mutate(log_NO3NO2 = log1p(NO3NO2), log_PO4 = log1p(PO4), log_SiO2 = log1p(SiO2))

# Cluster assignments on RF-eligible rows (k = 3, consistent with Fig 3A)
set.seed(8341)
pam_rf <- pam(
  rf_base |> select(Salinity, log_NO3NO2, log_PO4, log_SiO2) |> scale(),
  k = 3, pamonce = 5
)
# Reorder by NO3NO2 median (marine bg → nearshore → runoff)
rf_med <- tibble(temp = pam_rf$clustering, NO3NO2 = rf_base$NO3NO2) |>
  group_by(temp) |>
  summarise(med = median(NO3NO2, na.rm = TRUE), .groups = "drop") |>
  arrange(med) |>
  mutate(new = row_number())
rf_remap <- setNames(rf_med$new, rf_med$temp)

rf_base <- rf_base |>
  mutate(
    cluster       = factor(rf_remap[pam_rf$clustering]),
    cluster_label = factor(cluster_labels[cluster],
                           levels = c("Marine background", "Nearshore enriched", "Runoff-dominated"))
  )

rf_joined <- rf_base |>
  left_join(
    ccap_summary |>
      group_by(ahupuaa, island) |>
      summarise(pct_impervious  = mean(pct_impervious,  na.rm = TRUE),
                pct_agriculture = mean(pct_agriculture, na.rm = TRUE),
                .groups = "drop"),
    by = c("ahupuaa", "island")
  ) |>
  left_join(slope_j,           by = c("ahupuaa", "island"), relationship = "many-to-one") |>
  left_join(ksat_j,            by = c("ahupuaa", "island"), relationship = "many-to-one") |>
  left_join(sm_df     |> select(island, mean_gwetroot), by = "island") |>
  left_join(livestock_by_island |> select(island, cattle_per_km2), by = "island") |>
  mutate(across(
    c(mean_slope_deg, mean_ksat_um_s, log_dist_outfall,
      cattle_per_km2, n_cess_1km, fetch_mean_km,
      hs_mean_m, hs_at_collection_m, pop_density_km2),
    \(x) if_else(is.na(x), median(x, na.rm = TRUE), x)
  ))

rf_df <- rf_joined |>
  filter(!is.na(cesspool_density_per_km2),
         !is.na(dist_to_nearest_cesspool_m),
         !is.na(dist_to_stream_mouth_m),
         !is.na(pct_impervious),
         !is.na(log_discharge_cfs),
         !is.na(cluster_label)) |>
  transmute(
    cluster_label,
    log_n_cess_1km        = log1p(n_cess_1km),
    log_dist_cesspool     = log1p(dist_to_nearest_cesspool_m),
    log_fetch_mean        = log1p(fetch_mean_km),
    hs_mean_m,
    hs_at_collection_m,
    log_pop_density       = log1p(pop_density_km2),
    log_dist_stream_mouth = log1p(dist_to_stream_mouth_m),
    log_discharge_cfs,
    pct_impervious,
    sqrt_ag               = sqrt(pct_agriculture),
    mean_slope_deg,
    log_ksat              = log1p(mean_ksat_um_s),
    log_dist_outfall,
    log_cattle_km2        = log1p(cattle_per_km2)
  ) |>
  mutate(
    cess_x_ksat  = log_n_cess_1km * log_ksat,
    imp_x_ag     = pct_impervious * sqrt_ag,
    dist_x_slope = log_dist_stream_mouth * mean_slope_deg,
    disc_x_ag    = log_discharge_cfs * sqrt_ag,
    disc_x_cess  = log_discharge_cfs * log_n_cess_1km,
    imp_x_slope  = pct_impervious * mean_slope_deg,
    cluster_label = factor(cluster_label,
                           levels = c("Marine background", "Nearshore enriched", "Runoff-dominated"))
  )

set.seed(5572)
rf_fit <- ranger(
  cluster_label ~ .,
  data       = rf_df,
  num.trees  = 1000,
  mtry       = 4,
  importance = "permutation",
  seed       = 5572
)

oob_acc  <- round((1 - rf_fit$prediction.error) * 100, 1)
null_acc <- round(max(table(rf_df$cluster_label)) / nrow(rf_df) * 100, 1)
```

```{r}
#| label: fig3-rf-importance
#| fig-height: 6
#| fig-width: 7

label_map <- c(
  "log_n_cess_1km"        = "Cesspool density (1 km radius)",
  "log_dist_cesspool"     = "Distance to nearest cesspool",
  "log_dist_stream_mouth" = "Stream mouth proximity",
  "log_fetch_mean"        = "Coastal wave exposure (fetch)",
  "hs_mean_m"             = "Significant wave height (storm mean)",
  "hs_at_collection_m"    = "Significant wave height (at sampling)",
  "log_pop_density"       = "Population density",
  "log_discharge_cfs"     = "Storm discharge (March USGS)",
  "pct_impervious"        = "Impervious surface cover",
  "sqrt_ag"               = "Agricultural land cover",
  "mean_slope_deg"        = "Watershed slope",
  "log_ksat"              = "Soil permeability (Ksat)",
  "log_dist_outfall"      = "Distance to wastewater outfall",
  "log_cattle_km2"        = "Livestock density",
  "cess_x_ksat"           = "Cesspool density \u00d7 soil permeability",
  "imp_x_ag"              = "Impervious \u00d7 agricultural cover",
  "dist_x_slope"          = "Stream proximity \u00d7 watershed slope",
  "disc_x_ag"             = "Storm discharge \u00d7 agricultural cover",
  "disc_x_cess"           = "Storm discharge \u00d7 cesspool density",
  "imp_x_slope"           = "Impervious cover \u00d7 watershed slope"
)

anthropogenic_labels <- c(
  "Cesspool density (1 km radius)", "Distance to nearest cesspool",
  "Impervious surface cover", "Agricultural land cover",
  "Distance to wastewater outfall", "Livestock density", "Population density",
  "Cesspool density \u00d7 soil permeability",
  "Impervious \u00d7 agricultural cover",
  "Storm discharge \u00d7 agricultural cover",
  "Storm discharge \u00d7 cesspool density",
  "Impervious cover \u00d7 watershed slope"
)

imp_df <- enframe(rf_fit$variable.importance, name = "predictor", value = "importance") |>
  mutate(
    predictor  = recode(predictor, !!!label_map),
    importance = importance / max(importance),
    driver     = if_else(predictor %in% anthropogenic_labels,
                         "Anthropogenic", "Natural / physical")
  ) |>
  arrange(importance)

pC_f3 <- ggplot(imp_df, aes(x = importance, y = reorder(predictor, importance),
                             fill = driver)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = round(importance, 2)), hjust = -0.15, size = 3.0) +
  scale_x_continuous(limits = c(0, 1.2), expand = c(0, 0)) +
  scale_fill_manual(
    values = c("Anthropogenic" = "#BF360C", "Natural / physical" = "#37474F"),
    name   = NULL
  ) +
  labs(x = "Relative permutation importance (normalised)", y = NULL, tag = "C",
       subtitle = sprintf("OOB accuracy = %.1f%%  (null = %.1f%%)", oob_acc, null_acc)) +
  theme_minimal(base_size = 10.5) +
  theme(panel.grid.minor   = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.subtitle      = element_text(size = 8.5, color = "gray40"),
        legend.position    = "bottom",
        legend.text        = element_text(size = 9),
        plot.tag           = element_text(face = "bold"))

pC_f3
```

```{r}
#| label: fig3-pdp
#| fig-height: 3.5
#| fig-width: 10

# Probability RF (same predictors, for PDP)
set.seed(5572)
rf_prob <- ranger(
  cluster_label ~ .,
  data        = rf_df,
  num.trees   = 1000,
  mtry        = 4,
  probability = TRUE,
  seed        = 5572
)

# Partial dependence: cesspool density × Ksat (top interaction)
ksat_breaks <- quantile(rf_df$log_ksat, c(1/3, 2/3))
ksat_rep <- c(
  "Low"    = median(rf_df$log_ksat[rf_df$log_ksat <= ksat_breaks[1]]),
  "Medium" = median(rf_df$log_ksat[rf_df$log_ksat > ksat_breaks[1] &
                                    rf_df$log_ksat <= ksat_breaks[2]]),
  "High"   = median(rf_df$log_ksat[rf_df$log_ksat > ksat_breaks[2]])
)

cess_seq <- seq(min(rf_df$log_n_cess_1km), max(rf_df$log_n_cess_1km), length.out = 60)

pdp_df <- map_dfr(names(ksat_rep), function(grp) {
  kv <- ksat_rep[[grp]]
  map_dfr(cess_seq, function(cv) {
    nd    <- rf_df |>
      mutate(log_n_cess_1km = cv, log_ksat = kv, cess_x_ksat = cv * kv)
    probs <- colMeans(predict(rf_prob, data = nd)$predictions)
    tibble(ksat_group = grp, log_n_cess_1km = cv, n_cess_1km = expm1(cv), as_tibble_row(probs))
  })
})

ksat_pal <- c("Low"    = "#1565C0",
              "Medium" = "#FB8C00",
              "High"   = "#E53935")

pD_f3 <- pdp_df |>
  pivot_longer(
    cols = c(`Marine background`, `Nearshore enriched`, `Runoff-dominated`),
    names_to = "cluster", values_to = "probability"
  ) |>
  mutate(
    cluster    = factor(cluster, levels = c("Marine background",
                                             "Nearshore enriched",
                                             "Runoff-dominated")),
    ksat_group = factor(ksat_group, levels = c("Low", "Medium", "High"),
                        labels = c("Low Ksat", "Medium Ksat", "High Ksat"))
  ) |>
  ggplot(aes(x = log_n_cess_1km, y = probability, color = ksat_group)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.5, linewidth = 1.1) +
  scale_color_manual(values = ksat_pal, name = "Soil permeability\n(Ksat tertile)") +
  scale_x_continuous(
    name   = "Cesspools within 1 km (count, log scale)",
    breaks = log1p(c(0, 1, 5, 20, 50, 100, 200)),
    labels = c("0", "1", "5", "20", "50", "100", "200")
  ) +
  scale_y_continuous(name = "Predicted probability",
                     labels = label_percent(accuracy = 1)) +
  facet_wrap(~cluster, nrow = 1) +
  labs(tag = "D") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "bold"),
        legend.position  = "right",
        legend.title     = element_text(size = 9),
        legend.text      = element_text(size = 8.5),
        plot.tag         = element_text(face = "bold"))

pD_f3
```

```{r}
#| label: fig3-assemble
#| fig-height: 13
#| fig-width: 12
#| fig-cap: "**Figure 3.** Landscape predictability of post-storm coastal water type. **(A)** Biogeochemical fingerprint heatmap for each PAM cluster (k = 3; Z-scores of cluster-median concentrations; values = median µmol/L or PSU). **(B)** Spatial distribution of PAM water-type clusters across the four study islands. **(C)** Random forest variable importance (permutation, normalised to maximum). Red = anthropogenic drivers; gray = natural/physical drivers. OOB accuracy vs null shown in subtitle. **(D)** Partial dependence of predicted cluster probability on cesspool density within 1 km, stratified by soil permeability (Ksat tertile). High-Ksat soils amplify the cesspool signal: runoff-dominated probability climbs steeply with cesspool density only when permeability is high."

fig3 <- (pA_f3 / pB_f3 / pC_f3 / pD_f3) +
  plot_layout(heights = c(1, 0.9, 2.5, 1)) +
  plot_annotation(theme = theme(plot.margin = margin(6, 6, 4, 6)))

fig3
ggsave(here("output", "fig3_landscape_causes.pdf"), fig3,
       width = 12, height = 13, device = cairo_pdf)
ggsave(here("output", "fig3_landscape_causes.png"), fig3,
       width = 12, height = 13, dpi = 300)
```

---

# Figure 4 — Consequences: Reef Risk & Biological Response

*Post-storm concentrations exceeded reef health thresholds across all islands; phytoplankton responded.*

```{r}
#| label: fig4-aca-data
#| include: false
#| cache: false

# ── Download Allen Coral Atlas Coral/Algae polygons for Hawaii (WFS; cached) ─
aca_cache <- here("data", "aca_hawaii_coral.rds")

if (!file.exists(aca_cache)) {
  message("Fetching Allen Coral Atlas benthic data from WFS…")
  islands_aca <- tribble(
    ~island,    ~lon_min, ~lat_min, ~lon_max, ~lat_max,
    "O'ahu",   -158.40,   21.20,  -157.60,   21.75,
    "Maui",    -156.75,   20.50,  -155.90,   21.05,
    "Moloka'i",-157.40,   21.00,  -156.65,   21.30
  )

  fetch_aca <- function(island, lon_min, lat_min, lon_max, lat_max) {
    url <- paste0(
      "https://allencoralatlas.org/geoserver/ows?",
      "service=WFS&version=2.0.0&request=GetFeature",
      "&typeNames=coral-atlas:benthic_data_verbose",
      "&bbox=", paste(lon_min, lat_min, lon_max, lat_max, sep = ","), ",CRS:84",
      "&outputFormat=application/json"
    )
    out <- tryCatch(sf::st_read(url, quiet = TRUE),
                   error = function(e) { warning(island, ": ", e$message); NULL })
    if (!is.null(out) && nrow(out) > 0) out |> mutate(island = island) else NULL
  }

  reef_raw   <- pmap_dfr(islands_aca, fetch_aca)
  reef_coral <- reef_raw |> filter(class_name == "Coral/Algae")
  saveRDS(reef_coral, aca_cache)
  message("Cached ", nrow(reef_coral), " Coral/Algae polygons.")
}
reef_coral <- readRDS(aca_cache)

# ── Compute kriging rasters and extract per reef polygon (cached) ─────────────
reef_krige_cache <- here("data", "aca_reef_kriged_NO3.rds")

if (!file.exists(reef_krige_cache)) {
  message("Computing kriging surfaces and extracting per reef polygon…")

  krige_to_raster <- function(island_name) {
    kd     <- krige_surface_data(chem, island_name, ahupuaa_raw,
                                 var = "NO3NO2", res_m = 250)
    # Rasterize the irregular grid of point predictions
    pred_v <- terra::vect(
      as.matrix(kd$pred_df[, c("X", "Y")]),
      type = "points",
      atts = data.frame(pred = kd$pred_df$pred),
      crs  = "EPSG:32604"
    )
    r_tmpl <- terra::rast(terra::ext(pred_v) + 300,
                          resolution = 250, crs = "EPSG:32604")
    terra::rasterize(pred_v, r_tmpl, field = "pred", fun = "mean")
  }

  reef_utm <- reef_coral |> st_transform(32604)

  reef_with_krige <- map_dfr(
    c("O'ahu", "Maui", "Moloka'i"),
    function(isl) {
      message("  ", isl)
      r   <- krige_to_raster(isl)
      sub <- reef_utm |> filter(island == isl)
      if (nrow(sub) == 0) return(NULL)
      sub |> mutate(
        mean_NO3NO2 = exactextractr::exact_extract(r, sub, "mean",
                                                   progress = FALSE)
      )
    }
  )
  saveRDS(reef_with_krige, reef_krige_cache)
}
reef_with_krige <- readRDS(reef_krige_cache)

# ── Assign nearest PAM cluster (≤ 3 km) ───────────────────────────────────────
chem_clust_sf_f4 <- chem_clust |>
  filter(!is.na(latitude), !is.na(longitude), !is.na(cluster_label)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32604)

reef_cents <- reef_with_krige |> st_centroid()
nn_idx     <- st_nearest_feature(reef_cents, chem_clust_sf_f4)
nn_dist    <- as.numeric(
  st_distance(reef_cents, chem_clust_sf_f4[nn_idx, ], by_element = TRUE)
)

reef_classified <- reef_with_krige |>
  st_drop_geometry() |>
  mutate(
    cluster_label    = as.character(chem_clust_sf_f4$cluster_label[nn_idx]),
    dist_to_sample_m = nn_dist,
    # Only assign cluster if nearest sample is within 3 km coastal kriging range
    cluster_label    = if_else(dist_to_sample_m > 3000, NA_character_, cluster_label),
    island           = factor(island, levels = c("O'ahu", "Maui", "Moloka'i")),
    cluster_label    = factor(cluster_label,
                              levels = c("Marine background",
                                         "Nearshore enriched",
                                         "Runoff-dominated"))
  ) |>
  filter(!is.na(cluster_label), !is.na(mean_NO3NO2))
```

```{r}
#| label: fig4-reef-vulnerability
#| fig-height: 5.5
#| fig-width: 8.5
#| fig-cap: "**Figure 4A.** Coral reef area (km²) co-occurring with each post-storm PAM water-type cluster, by island. Benthic polygons from Allen Coral Atlas v2.0 (Coral/Algae class; <10 m depth). PAM cluster assigned by nearest citizen-science sampling site within 3 km. Mean kriged NO₃+NO₂ (µmol/L) annotated for each cluster. Total reef area label shown above each bar group. Runoff-dominated water masses overlap substantially more mapped coral area on Maui and Molokaʻi, where post-storm nutrient concentrations were highest."

# ── Per-cluster summary stats ──────────────────────────────────────────────────
reef_summary <- reef_classified |>
  group_by(island, cluster_label) |>
  summarise(
    total_km2 = sum(area_sqkm, na.rm = TRUE),
    mean_NO3  = mean(mean_NO3NO2, na.rm = TRUE),
    n_patches = n(),
    .groups   = "drop"
  )

cluster_totals <- reef_classified |>
  group_by(cluster_label) |>
  summarise(
    total_km2 = sum(area_sqkm, na.rm = TRUE),
    mean_NO3  = median(mean_NO3NO2, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  mutate(
    km2_lab = paste0(round(total_km2, 1), " km²"),
    no3_lab = paste0("median\n", round(mean_NO3, 2), " µmol/L")
  )

y_max <- max(reef_summary |> group_by(cluster_label) |>
               summarise(s = sum(total_km2)) |> pull(s))

pA_f4 <- ggplot(reef_summary,
                aes(x = cluster_label, y = total_km2, fill = island)) +
  geom_col(width = 0.68, color = "white", linewidth = 0.3) +
  # Total km² label above each group
  geom_text(data = cluster_totals,
            aes(x = cluster_label, y = total_km2, label = km2_lab),
            inherit.aes = FALSE, vjust = -0.55,
            size = 3.2, fontface = "bold") +
  # Median NO3NO2 label below x-axis (requires clip = "off")
  geom_text(data = cluster_totals,
            aes(x = cluster_label, y = 0, label = no3_lab),
            inherit.aes = FALSE, vjust = 1.35,
            size = 2.6, color = "#5D4037", lineheight = 0.9) +
  scale_fill_manual(values = island_pal, name = NULL) +
  scale_x_discrete(labels = \(x) str_wrap(x, 11)) +
  scale_y_continuous(
    expand  = expansion(mult = c(0, 0.18)),
    labels  = \(x) paste0(x, " km²")
  ) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = "Coral reef area (km²)", tag = "A",
       subtitle = "Allen Coral Atlas Coral/Algae class  \u00b7  nearest sample \u22643 km") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.margin        = margin(6, 10, 28, 10),   # bottom space for NO3 labels
    plot.subtitle      = element_text(size = 8.5, color = "gray45"),
    legend.position    = "right",
    plot.tag           = element_text(face = "bold")
  )

pA_f4
```

```{r}
#| label: fig4-viirs-data

library(rerddap)
library(marmap)

# Load cached VIIRS 750 m data (downloaded in nutrient_hotspot_maps.qmd)
all_750  <- readRDS("data/viirs_750m_bloom_cache.rds")
all_ext  <- readRDS("data/viirs_750m_prestorm_extended.rds")
bathy_raw <- readRDS("data/hawaii_bathy_1min.rds")

bathy_df_qc <- fortify(bathy_raw) |>
  as_tibble() |>
  rename(longitude = x, latitude = y, depth_m = z)

bathy_lon_step <- diff(sort(unique(bathy_df_qc$longitude)))[1]
bathy_lat_step <- diff(sort(unique(bathy_df_qc$latitude)))[1]

add_depth <- function(df) {
  df |>
    mutate(lon_key = round(longitude / bathy_lon_step) * bathy_lon_step,
           lat_key = round(latitude  / bathy_lat_step) * bathy_lat_step) |>
    left_join(
      bathy_df_qc |>
        mutate(lon_key = round(longitude / bathy_lon_step) * bathy_lon_step,
               lat_key = round(latitude  / bathy_lat_step) * bathy_lat_step) |>
        select(lon_key, lat_key, depth_m),
      by = c("lon_key", "lat_key")
    ) |>
    select(-lon_key, -lat_key)
}

bloom_pixels_qc <- all_750 |>
  filter(period == "peak_bloom", !is.na(chla), chla >= 0.001, chla <= 100.1) |>
  add_depth() |>
  filter(!is.na(depth_m), depth_m <= -10) |>
  group_by(island, latitude, longitude) |>
  summarise(bloom_chl = mean(chla, na.rm = TRUE), n_days = n(), .groups = "drop")

pre_pixels_ext <- all_ext |>
  filter(!is.na(chla), chla >= 0.001, chla <= 100.1) |>
  add_depth() |>
  filter(!is.na(depth_m), depth_m <= -10) |>
  group_by(island, latitude, longitude) |>
  summarise(pre_chl_ext = mean(chla, na.rm = TRUE), n_days_ext = n(), .groups = "drop")

pixel_matched_ext <- bloom_pixels_qc |>
  inner_join(pre_pixels_ext |> select(island, latitude, longitude, pre_chl_ext, n_days_ext),
             by = c("island", "latitude", "longitude")) |>
  mutate(fold_change_ext = bloom_chl / pre_chl_ext,
         chl_anomaly_ext = bloom_chl - pre_chl_ext)

# Match sample sites to nearest 750 m pixel (≤ 5 km ≈ 0.045 decimal degrees)
find_nearest_chl <- function(site_lat, site_lon, site_island, bloom_df) {
  map_dbl(seq_along(site_lat), function(i) {
    px <- bloom_df[bloom_df$island == site_island[i], ]
    if (nrow(px) == 0) return(NA_real_)
    d   <- sqrt((px$latitude - site_lat[i])^2 + (px$longitude - site_lon[i])^2)
    idx <- which.min(d)
    if (d[idx] > 0.045) NA_real_ else px$bloom_chl[idx]
  })
}

chem_sites <- chem_clust |>
  filter(island %in% c("Maui", "Moloka'i", "O'ahu"),
         !is.na(latitude), !is.na(longitude), !is.na(NO3NO2))
island_750 <- recode(chem_sites$island, "Moloka'i" = "Molokai", "O'ahu" = "Oahu")

chem_750m_match <- chem_sites |>
  mutate(
    island_750     = island_750,
    bloom_chl_750m = find_nearest_chl(latitude, longitude, island_750, bloom_pixels_qc),
    fc_ext         = find_nearest_chl(latitude, longitude, island_750,
                                       pixel_matched_ext |>
                                         select(island, latitude, longitude,
                                                bloom_chl = fold_change_ext)),
    chl_anom_ext   = find_nearest_chl(latitude, longitude, island_750,
                                       pixel_matched_ext |>
                                         select(island, latitude, longitude,
                                                bloom_chl = chl_anomaly_ext))
  ) |>
  filter(!is.na(bloom_chl_750m))
```

```{r}
#| label: fig4-bloom-map
#| fig-height: 5.5
#| fig-width: 10
#| fig-cap: "**Figure 4B.** VIIRS NOAA-20 750 m peak-bloom Chl-a (Mar 20–Apr 8, 2026; pixels with ≥ 2 cloud-free days; depth ≥ 10 m). Overlaid points = citizen-science sites coloured by PAM water-type cluster."

hawaii_sf <- ne_states(country = "United States of America", returnclass = "sf") |>
  filter(name == "Hawaii")

df_for_map <- chem_clust |>
  filter(island %in% c("Maui", "Moloka'i", "O'ahu"),
         !is.na(cluster_label), !is.na(latitude), !is.na(longitude))

pB_f4 <- ggplot() +
  geom_tile(
    data = bloom_pixels_qc |>
      mutate(island = recode(island, "Molokai" = "Moloka'i", "Oahu" = "O'ahu")),
    aes(x = longitude, y = latitude, fill = log10(bloom_chl)), na.rm = TRUE
  ) +
  geom_sf(data = hawaii_sf, fill = "gray45", color = "white", linewidth = 0.3) +
  geom_point(data = df_for_map,
             aes(x = longitude, y = latitude, color = cluster_label),
             size = 1.2, alpha = 0.75, shape = 16) +
  scale_fill_gradientn(
    colors   = c("#08306B", "#2171B5", "#6BAED6", "#C6DBEF",
                 "#FDD0A2", "#FDAE6B", "#F16913", "#D94801", "#7F2704"),
    na.value = "transparent",
    name     = expression("Chl-"*italic(a)*"\n(mg m"^{-3}*")"),
    breaks   = log10(c(0.05, 0.1, 0.3, 1, 3, 10)),
    labels   = c("0.05", "0.1", "0.3", "1", "3", "10")
  ) +
  scale_color_manual(values = cluster_pal, name = "Water type") +
  coord_sf(xlim = c(-158.6, -155.85), ylim = c(20.45, 21.80), expand = FALSE) +
  guides(
    fill  = guide_colorbar(order = 1, barheight = unit(3.5, "cm"),
                           title.position = "top"),
    color = guide_legend(order = 2, override.aes = list(size = 3, alpha = 1))
  ) +
  labs(x = NULL, y = NULL, tag = "B",
       subtitle = "VIIRS 750 m  ·  peak bloom Mar 20–Apr 8, 2026") +
  theme_minimal(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "#c8dff0", color = NA),
    legend.position  = "right",
    axis.text        = element_text(size = 8.5),
    plot.subtitle    = element_text(size = 9, color = "gray45"),
    plot.tag         = element_text(face = "bold")
  )

pB_f4
```

```{r}
#| label: fig4-hui-summary
#| fig-height: 5.5
#| fig-width: 7
#| fig-cap: "**Figure 4C.** Post-storm Hui o Ka Wai Ola NO₃+NO₂ fold-change above station-specific historical median (all records 2016–Jan 2026). Orange = readings above the station's historical 95th percentile. Stations ordered within West/South Maui regions by magnitude."

hui_hist <- bg_hui_loc |>
  filter(year(Date) < 2026, !is.na(NNN_umol)) |>
  group_by(Station, SiteName, region) |>
  summarise(med_hist = median(NNN_umol, na.rm = TRUE),
            p95_hist = quantile(NNN_umol, 0.95, na.rm = TRUE), .groups = "drop")

hui_post_nn <- bg_hui_loc |>
  filter(year(Date) == 2026, month(Date) == 3,
         Date >= as.Date("2026-03-14"), !is.na(NNN_umol)) |>
  group_by(Station, SiteName, region) |>
  summarise(val_umol = mean(NNN_umol, na.rm = TRUE), .groups = "drop") |>
  left_join(hui_hist, by = c("Station", "SiteName", "region")) |>
  filter(!is.na(med_hist)) |>
  mutate(
    fold_change = val_umol / med_hist,
    above_p95   = val_umol > p95_hist,
    label       = str_trunc(SiteName, 26, ellipsis = "…"),
    region      = factor(region, levels = c("South Maui", "West Maui"))
  ) |>
  arrange(region, fold_change) |>
  mutate(label = factor(label, levels = unique(label)))

pC_f4 <- ggplot(hui_post_nn, aes(x = fold_change, y = label, fill = above_p95)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = paste0(round(fold_change, 1), "×"),
        hjust = if_else(fold_change > 12, 1.1, -0.1),
        color = if_else(fold_change > 12, "white", "gray25")),
    size = 2.5, fontface = "bold"
  ) +
  geom_vline(xintercept = 1, color = "gray40", linewidth = 0.5, linetype = "dashed") +
  scale_fill_manual(
    values = c("TRUE" = "#E65100", "FALSE" = "#90A4AE"),
    labels = c("TRUE" = "Above 95th percentile", "FALSE" = "Within historical range"),
    name   = NULL
  ) +
  scale_color_identity() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.25)),
                     breaks = c(1, 5, 10, 20, 40)) +
  facet_grid(region ~ ., scales = "free_y", space = "free_y", switch = "y") +
  labs(x = expression("Fold above historical median — NO"[3]*"+NO"[2]*" (µmol/L)"),
       y = NULL, tag = "C") +
  theme_minimal(base_size = 9.5) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank(),
    strip.text   = element_text(face = "bold", size = 9),
    strip.placement = "outside",
    legend.position = "bottom",
    legend.text  = element_text(size = 8.5),
    axis.text.y  = element_text(size = 7, lineheight = 0.85),
    plot.tag     = element_text(face = "bold")
  )

pC_f4
```

```{r}
#| label: fig4-assemble
#| fig-height: 12
#| fig-width: 12
#| fig-cap: "**Figure 4.** Reef risk and biological response. **(A)** Proportion of samples exceeding published coral reef health thresholds. **(B)** VIIRS 750 m peak-bloom chlorophyll-a; PAM water-type clusters overlaid. **(C)** Independent validation: Hui o Ka Wai Ola NO₃+NO₂ fold-change above each station's 10-year historical median for March 2026 post-storm readings. Orange = above 95th percentile."

fig4 <- (pA_f4 / pB_f4 / pC_f4) +
  plot_layout(heights = c(1, 1.4, 1.5)) +
  plot_annotation(theme = theme(plot.margin = margin(6, 6, 4, 6)))

fig4
ggsave(here("output", "fig4_consequences.pdf"), fig4,
       width = 12, height = 12, device = cairo_pdf)
ggsave(here("output", "fig4_consequences.png"), fig4,
       width = 12, height = 12, dpi = 300)
```
