
---
title: "Water Quality Results — Hawaiʻi Kona Low, March 2026"
subtitle: "Summary of chemistry results across all sampled sites"
author: "Silbiger, N., Kealoha, A., & Kahanamoku, S."
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: flatly
    code-fold: true
---



# Introduction

This report summarizes water chemistry results from the *What's In Your Water?* community science project, which collected ocean water samples across Oʻahu and Maui Nui following the March 2026 Kona Low storm. Samples were gathered by volunteer community members and scientists, and are being analyzed for salinity, nutrients (nitrate+nitrite, phosphate, silicate, and ammonium), and dissolved organic matter — key indicators of coral reef health.

Reef health thresholds for Hawaiian nearshore waters used throughout this report: NO₃+NO₂ > 1 µmol/L, PO₄ > 0.1 µmol/L, SiO₂ > 5 µmol/L, NH₃ > 1 µmol/L.

```{r}
#| label: load-packages
#| message: false
#| warning: false
library(tidyverse)
library(lubridate)
library(ggforce)
library(ggridges)
library(patchwork)
library(ggrepel)
library(ggh4x)
library(scales)
```

```{r}
#| label: load-data
#| message: false
#| warning: false
source("01_load_clean_data.R")

cesspool_dist       <- read_csv("data/sample_cesspool_distances.csv", show_col_types = FALSE)
cesspool_by_ahupuaa <- read_csv("data/cesspool_by_ahupuaa.csv",       show_col_types = FALSE)

chem_data <- read_csv("data/ChemData.csv", show_col_types = FALSE) |>
  rename(
    sample_id = sample_ID,
    NO3NO2    = `NO3+NO2 (umol/L)`,
    PO4       = `PO4 (umol/L)`,
    SiO2      = `SiO2 (umol/L)`,
    NH3       = `NH3 (umol/L)`
  ) |>
  mutate(
    sample_id = toupper(sample_id),
    across(c(Salinity, NO3NO2, PO4, SiO2, NH3),
           ~if (is.character(.)) readr::parse_number(.) else .)
  )

df_report <- df_clean |>
  mutate(
    hour        = hour(collected_hst),
    minute      = minute(collected_hst),
    time_of_day = if_else(
      (hour * 60 + minute) >= (6 * 60) & (hour * 60 + minute) < (18 * 60 + 45),
      "Daytime", "Nighttime"
    ),
    date = as.Date(collected_hst, tz = "Pacific/Honolulu")
  ) |>
  full_join(chem_data, by = "sample_id") |>
  left_join(cesspool_dist, by = "sample_id")
```

```{r}
#| label: setup-constants
thresholds    <- c(NO3NO2 = 1, PO4 = 0.1, SiO2 = 5, NH3 = 1)
island_palette <- c("#1565C0", "#2E7D32", "#E65100", "#6A1B9A")

normalize_island <- function(x) {
  x |>
    stringr::str_replace_all("[\u02BB\u2018\u2019']", "") |>
    stringr::str_replace_all("\u0101", "a") |>
    stringr::str_to_lower() |>
    stringr::str_trim()
}
```

```{r}
#| label: prepare-chem-data
# Exclude open-ocean samples and a known extreme outlier (M226, Kaluakoʻi)
df_chem <- df_report |>
  filter(
    ahupuaa != "Open Ocean" | is.na(ahupuaa),
    sample_id != "M226"
  )

islands       <- sort(unique(na.omit(df_chem$island)))
island_colors <- setNames(island_palette[seq_along(islands)], islands)
```

# Reef health report card

Percentage of water samples exceeding reef health thresholds for Hawaiian nearshore waters.

```{r}
#| label: gauge-helper
make_gauge <- function(pct, nutrient_name, threshold_label, units = "\u00b5mol/L") {
  grade_color <- case_when(
    pct < 25 ~ "#43A047",
    pct < 50 ~ "#FB8C00",
    TRUE     ~ "#E53935"
  )
  fill_end <- (pct / 100) * pi

  ggplot() +
    geom_arc_bar(
      data = tibble(x0 = 0, y0 = 0, r0 = 0.58, r = 1, start = 0, end = pi),
      aes(x0 = x0, y0 = y0, r0 = r0, r = r, start = start, end = end),
      fill = "grey88", color = NA
    ) +
    geom_arc_bar(
      data = tibble(x0 = 0, y0 = 0, r0 = 0.58, r = 1, start = 0, end = fill_end),
      aes(x0 = x0, y0 = y0, r0 = r0, r = r, start = start, end = end),
      fill = grade_color, color = NA
    ) +
    annotate("segment",
             x = 0, xend = 0, y = 0.58, yend = 1.08,
             color = "white", linewidth = 0.8) +
    annotate("text", x = 0, y = 0.22,
             label = paste0(round(pct), "%"),
             size = 9, fontface = "bold", color = grade_color, hjust = 0.5) +
    annotate("text", x =  1.2, y = -0.02, label = "0%",
             size = 3, color = "grey55", hjust = 0.5) +
    annotate("text", x = -1.2, y = -0.02, label = "100%",
             size = 3, color = "grey55", hjust = 0.5) +
    coord_cartesian(xlim = c(-1.6, 1.6), ylim = c(-0.45, 1.25)) +
    ggtitle(
      nutrient_name,
      subtitle = paste0("threshold: >", threshold_label, " ", units)
    ) +
    theme_void() +
    theme(
      aspect.ratio  = 1.7 / 3.2,
      plot.title    = element_text(hjust = 0.5, face = "bold", size = 13,
                                   color = "grey15", margin = margin(b = 1)),
      plot.subtitle = element_text(hjust = 0.5, size = 9,
                                   color = "grey55", margin = margin(b = 2)),
      plot.margin   = margin(8, 8, 8, 8)
    )
}
```

```{r}
#| label: plot-report-card
#| fig-width: 12
#| fig-height: 4
exceed_pcts <- df_chem |>
  select(NO3NO2, PO4, SiO2, NH3) |>
  summarise(across(
    everything(),
    ~round(mean(. > thresholds[cur_column()], na.rm = TRUE) * 100)
  )) |>
  unlist()

(make_gauge(exceed_pcts["NO3NO2"], "Nitrate + Nitrite", 1)  +
 make_gauge(exceed_pcts["PO4"],    "Phosphate",         0.1) +
 make_gauge(exceed_pcts["SiO2"],   "Silicate",          5)   +
 make_gauge(exceed_pcts["NH3"],    "Ammonium",          1)) +
  plot_layout(ncol = 4, nrow = 1) +
  plot_annotation(
    title    = "Hawaiian Nearshore Water Quality Report",
    subtitle = "Percentage of water samples exceeding reef health thresholds",
    caption  = "Thresholds based on normal ranges for Hawaiian nearshore reefs",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 17, hjust = 0.5,
                                   margin = margin(b = 4)),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40",
                                   margin = margin(b = 8)),
      plot.caption  = element_text(size = 9, color = "grey55", hjust = 0.5)
    )
  )
```

::: {.callout-tip title="Interpretation"}
Phosphate and silicate had the highest exceedance rates among analyzed samples, followed closely by nitrate+nitrite and ammonium. Exceedances across all four nutrients — well above baseline conditions — are consistent with significant terrestrial runoff reaching nearshore reefs during and after the Kona Low storm. As more samples are processed, these percentages may shift.
:::

# Nutrient concentrations — all samples

Each dot is one water sample. The dashed red line marks the reef health threshold. Dots to the right of the line exceeded the threshold.

```{r}
#| label: strip-helper
make_one_strip <- function(df_col, display_label, thresh) {
  if (nrow(df_col) == 0) return(ggplot() + theme_void())
  ggplot(df_col, aes(x = value, y = island, color = island)) +
    geom_jitter(height = 0.2, width = 0, alpha = 0.55, size = 1.8) +
    geom_vline(xintercept = thresh, color = "firebrick",
               linetype = "dashed", linewidth = 0.8) +
    scale_color_manual(values = island_colors, guide = "none") +
    scale_x_continuous(trans  = "log1p",
                       labels = label_number(drop0trailing = TRUE)) +
    labs(title = display_label, x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.title         = element_text(face = "bold", size = 11)
    )
}
```

```{r}
#| label: plot-strip
#| fig-width: 10
#| fig-height: 8
strip_df <- df_chem |>
  filter(!is.na(island)) |>
  select(sample_id, island, NO3NO2, PO4, SiO2, NH3) |>
  mutate(island = factor(island, levels = sort(unique(island))))

p_no3  <- make_one_strip(strip_df |> filter(!is.na(NO3NO2)) |> mutate(value = NO3NO2),
                         "Nitrate+Nitrite (\u00b5mol/L)", 1)
p_po4  <- make_one_strip(strip_df |> filter(!is.na(PO4))   |> mutate(value = PO4),
                         "Phosphate (\u00b5mol/L)", 0.1)
p_sio2 <- make_one_strip(strip_df |> filter(!is.na(SiO2))  |> mutate(value = SiO2),
                         "Silicate (\u00b5mol/L)", 5)
p_nh3  <- make_one_strip(strip_df |> filter(!is.na(NH3))   |> mutate(value = NH3),
                         "Ammonium (\u00b5mol/L)", 1)

(p_no3 + p_po4) / (p_sio2 + p_nh3)
```

::: {.callout-tip title="Interpretation"}
Most samples cluster near or below each threshold, but a meaningful tail of high-concentration outliers drives the exceedance rates — pointing to localized hotspots rather than uniform contamination across all sites. High outliers appear on all four islands, suggesting the storm's runoff impact was widespread but uneven.
:::

# Salinity distribution by island

Each curve shows the full spread of salinity readings across samples from that island. The dotted green lines mark normal seawater (34–36 psu) — values to the left indicate freshwater mixing from storm runoff.

```{r}
#| label: plot-ridgeline
#| fig-width: 9
#| fig-height: 5
sal_df <- df_chem |> filter(!is.na(Salinity), !is.na(island))

ggplot(sal_df, aes(x = Salinity, y = island, fill = island)) +
  geom_density_ridges(
    alpha          = 0.75,
    scale          = 0.9,
    quantile_lines = TRUE,
    quantiles      = 2,
    color          = "white"
  ) +
  geom_vline(xintercept = 34, color = "#2E7D32", linetype = "dotted", linewidth = 1.1) +
  geom_vline(xintercept = 36, color = "#2E7D32", linetype = "dotted", linewidth = 1.1) +
  scale_fill_manual(values = island_colors, guide = "none") +
  labs(
    title    = "Salinity distribution by island",
    subtitle = "Median shown as white line. Dotted green lines = normal seawater (34\u202636\u00a0psu). Lower values = more freshwater.",
    x        = "Salinity (psu)",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold")
  )
```

::: {.callout-tip title="Interpretation"}
Maui shows the most pronounced freshwater mixing, with a large portion of samples falling well below 34 psu — indicating strong land-ocean connectivity during the storm. Oʻahu and Molokaʻi also show clear freshwater pulses, while Lānaʻi samples remained close to normal salinity. Low salinity is often the first warning signal that storm runoff has reached a reef.
:::

# Pollution burden by island

Share of analyzed samples by how many nutrient thresholds were simultaneously exceeded. Darker colors indicate more nutrients above threshold at the same site.

```{r}
#| label: plot-burden
#| fig-width: 8
#| fig-height: 4
burden_df <- df_chem |>
  filter(!is.na(island)) |>
  filter(!is.na(NO3NO2) | !is.na(PO4) | !is.na(SiO2) | !is.na(NH3)) |>
  mutate(
    n_exceeded =
      (!is.na(NO3NO2) & NO3NO2 > thresholds["NO3NO2"]) +
      (!is.na(PO4)    & PO4    > thresholds["PO4"])    +
      (!is.na(SiO2)   & SiO2   > thresholds["SiO2"])   +
      (!is.na(NH3)    & NH3    > thresholds["NH3"]),
    n_exceeded = factor(
      n_exceeded,
      levels = 0:4,
      labels = c("0 \u2014 None", "1", "2", "3", "4 \u2014 All")
    )
  ) |>
  count(island, n_exceeded) |>
  group_by(island) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

ggplot(burden_df, aes(x = prop, y = island, fill = n_exceeded)) +
  geom_col(position = "stack", width = 0.55) +
  scale_fill_manual(
    values = c(
      "0 \u2014 None" = "#43A047",
      "1"            = "#A5D6A7",
      "2"            = "#FB8C00",
      "3"            = "#E53935",
      "4 \u2014 All"  = "#7B1FA2"
    ),
    name = "Nutrients above\nthreshold"
  ) +
  scale_x_continuous(
    labels = percent,
    expand = expansion(mult = c(0, 0.03))
  ) +
  labs(
    title    = "Pollution burden by island",
    subtitle = "Share of analyzed samples by number of nutrient thresholds simultaneously exceeded",
    x        = NULL,
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold"),
    legend.position    = "right"
  )
```

::: {.callout-tip title="Interpretation"}
At every island, the large majority of analyzed samples exceeded at least two nutrient thresholds simultaneously, indicating multi-nutrient contamination rather than a single-source event. Oʻahu shows the most bimodal pattern — a higher share of both "clean" samples and highly impacted ones — suggesting greater spatial variability in runoff exposure across the island.
:::

# Water quality stoplight by moku

Percentage of samples exceeding reef health thresholds for each nutrient, grouped by traditional Hawaiian land division (moku). Moku with fewer than 3 analyzed samples are excluded.

```{r}
#| label: plot-stoplight
#| fig-width: 8
#| fig-height: 10
heatmap_df <- df_chem |>
  filter(!is.na(moku)) |>
  group_by(island, moku) |>
  summarise(
    n                 = n(),
    `Nitrate+Nitrite` = mean(NO3NO2 > thresholds["NO3NO2"], na.rm = TRUE) * 100,
    `Phosphate`       = mean(PO4    > thresholds["PO4"],    na.rm = TRUE) * 100,
    `Silicate`        = mean(SiO2   > thresholds["SiO2"],   na.rm = TRUE) * 100,
    `Ammonium`        = mean(NH3    > thresholds["NH3"],     na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  filter(n >= 3) |>
  pivot_longer(
    c(`Nitrate+Nitrite`, Phosphate, Silicate, Ammonium),
    names_to = "nutrient", values_to = "pct"
  ) |>
  filter(!is.nan(pct)) |>
  mutate(
    label = if_else(
      moku %in% c("L\u0101hain\u0101", "Kona"),
      paste0(moku, " (", island, ")"),
      moku
    ),
    label    = fct_reorder(label, pct, mean, .na_rm = TRUE),
    nutrient = factor(nutrient,
                      levels = c("Nitrate+Nitrite", "Phosphate", "Silicate", "Ammonium")),
    grade = case_when(
      pct < 25 ~ "Low",
      pct < 50 ~ "Moderate",
      TRUE     ~ "High"
    ),
    grade = factor(grade, levels = c("Low", "Moderate", "High"))
  )

ggplot(heatmap_df, aes(x = nutrient, y = label, fill = grade)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(
    aes(label = paste0(round(pct), "%"),
        color = grade == "Moderate"),
    size = 3.5, fontface = "bold"
  ) +
  scale_fill_manual(
    values = c("Low" = "#43A047", "Moderate" = "#FB8C00", "High" = "#E53935"),
    name   = NULL,
    labels = c("Low (<25%)", "Moderate (25\u201350%)", "High (>50%)")
  ) +
  scale_color_manual(
    values = c("TRUE" = "grey15", "FALSE" = "white"),
    guide  = "none"
  ) +
  labs(
    title    = "Water quality stoplight by moku",
    subtitle = "% of samples exceeding reef health threshold. Moku with \u22653 analyzed samples shown.",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid      = element_blank(),
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(face = "bold", size = 11),
    legend.position = "bottom",
    legend.key.size = unit(1, "lines")
  )
```

::: {.callout-tip title="Interpretation"}
Several moku show high (≥50%) exceedance rates across multiple nutrients simultaneously, pointing to areas with compounding pollution pressure. Phosphate and silicate tend to show the broadest exceedances, consistent with their mixed sources in runoff (fertilizers, eroding soils). Moku with lower exceedance rates may reflect either cleaner conditions or limited sampling — more samples will sharpen this picture.
:::

# Nutrient concentration vs. distance to nearest cesspool

Each dot is one water sample. X-axis shows distance (m, log scale) to the nearest cesspool; y-axis shows nutrient concentration (log₁p scale). Lānaʻi is excluded — no cesspool location data are publicly available for that island. Open-ocean samples are excluded.

```{r}
#| label: plot-cesspool-proximity
#| fig-width: 12
#| fig-height: 9
nutrient_labels <- c(
  NO3NO2 = "Nitrate+Nitrite (\u00b5mol/L)",
  PO4    = "Phosphate (\u00b5mol/L)",
  SiO2   = "Silicate (\u00b5mol/L)",
  NH3    = "Ammonium (\u00b5mol/L)"
)

proximity_df <- df_report |>
  filter(
    !is.na(dist_to_nearest_cesspool_m),
    !is.na(island),
    !grepl("L\u0101na", island)
  ) |>
  pivot_longer(c(NO3NO2, PO4, SiO2, NH3),
               names_to  = "nutrient",
               values_to = "concentration") |>
  filter(!is.na(concentration)) |>
  mutate(nutrient = factor(nutrient,
           levels = names(nutrient_labels),
           labels = unname(nutrient_labels)))

ggplot(proximity_df, aes(x = dist_to_nearest_cesspool_m, y = concentration)) +
  geom_point(alpha = 0.45, size = 1.4) +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(
    trans  = "log1p",
    labels = label_number(drop0trailing = TRUE)
  ) +
  facet_wrap(island ~ nutrient, scales = "free", ncol = 4) +
  labs(
    x = "Distance to nearest cesspool (m, log\u2081\u2080 scale)",
    y = "Concentration (\u00b5mol/L, log\u2081p scale)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text    = element_text(face = "bold"),
    panel.spacing = unit(0.6, "lines")
  )
```

::: {.callout-tip title="Interpretation"}
No consistent gradient is visible between straight-line distance to the nearest cesspool and nutrient concentrations. This likely reflects the complexity of groundwater and surface flow paths: contamination from cesspools does not travel in straight lines, and storm-driven surface runoff can carry nutrients far from their source quickly. Distance to the nearest cesspool alone is not a reliable predictor of water quality at these sites.
:::

**Data source:** Cesspool locations from the Hawaiʻi Statewide GIS Program, *Hawaiʻi Cesspool Prioritization Tool (HCPT) Cesspools* (layer 32) and *Onsite Sewage Disposal Systems — Molokaʻi* (layer 23). [geodata.hawaii.gov](https://geodata.hawaii.gov/arcgis/rest/services/Infrastructure/MapServer). Molokaʻi data filtered to Class I (cesspool) systems. Accessed April 28, 2026.

# Mean nutrient concentration by cesspool density

Each dot is one ahupuaʻa, sized by number of water samples. X-axis shows cesspool density (cesspools per km², log scale); y-axis shows mean nutrient concentration (log scale, labeled on normal scale). Error bars show ±1 SE. Only ahupuaʻa with ≥3 analyzed samples shown. Lānaʻi excluded — no cesspool location data are publicly available.

```{r}
#| label: plot-cesspool-density
#| fig-width: 12
#| fig-height: 9
cp_join <- cesspool_by_ahupuaa |>
  mutate(island_key = normalize_island(island)) |>
  select(ahupuaa, island_key, cesspool_density_per_km2)

density_df <- df_chem |>
  filter(!grepl("L\u0101na", island)) |>
  mutate(island_key = normalize_island(island)) |>
  inner_join(cp_join, by = c("ahupuaa", "island_key")) |>
  filter(!is.na(cesspool_density_per_km2)) |>
  pivot_longer(c(NO3NO2, PO4, SiO2, NH3),
               names_to  = "nutrient",
               values_to = "concentration") |>
  filter(!is.na(concentration)) |>
  group_by(ahupuaa, island, cesspool_density_per_km2, nutrient) |>
  summarise(
    mean_conc = mean(concentration),
    se_conc   = sd(concentration) / sqrt(n()),
    n         = n(),
    .groups   = "drop"
  ) |>
  filter(n >= 3) |>
  mutate(nutrient = factor(nutrient,
           levels = names(nutrient_labels),
           labels = unname(nutrient_labels)))

label_df <- density_df |>
  group_by(island, nutrient) |>
  slice_max(mean_conc, n = 2) |>
  ungroup()

ggplot(density_df, aes(x = cesspool_density_per_km2, color = island)) +
  geom_errorbar(
    aes(y = mean_conc, ymin = mean_conc - se_conc, ymax = mean_conc + se_conc),
    width = 0, linewidth = 0.8
  ) +
  geom_point(aes(y = mean_conc, size = n)) +
  geom_text_repel(
    data = label_df,
    aes(y = mean_conc, label = ahupuaa),
    size = 3, show.legend = FALSE,
    min.segment.length = 0.2, box.padding = 0.4, max.overlaps = 20
  ) +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(
    trans  = "log1p",
    labels = label_number(drop0trailing = TRUE)
  ) +
  scale_color_manual(values = island_colors, name = "Island") +
  scale_size_continuous(
    name   = "Samples\nin ahupua\u02BBa",
    range  = c(2, 6),
    breaks = c(3, 5, 10, 20)
  ) +
  facet_grid2(island ~ nutrient, scales = "free", independent = "y") +
  labs(
    x = "Cesspool density (per km\u00b2, log10 scale)",
    y = "Mean concentration (\u00b5mol/L, log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom"
  )
```

::: {.callout-tip title="Interpretation"}
No strong or consistent relationship is apparent between ahupuaʻa-level cesspool density and mean nutrient concentrations. This likely reflects the influence of other nutrient sources (agricultural runoff, stormwater, stream discharge) that the storm mobilized alongside cesspool leachate, making it difficult to isolate the cesspool signal at this spatial scale with the current sample sizes.
:::

**Data source:** Cesspool locations from the Hawaiʻi Statewide GIS Program, *Hawaiʻi Cesspool Prioritization Tool (HCPT) Cesspools* (layer 32) and *Onsite Sewage Disposal Systems — Molokaʻi* (layer 23). [geodata.hawaii.gov](https://geodata.hawaii.gov/arcgis/rest/services/Infrastructure/MapServer). Molokaʻi data filtered to Class I (cesspool) systems. Accessed April 28, 2026.

---

*Cite these data:* Silbiger, N., Kealoha, A., & Kahanamoku, S. (2026). *What's In Your Water? — Hawaiʻi Coastal Water Quality Samples, March 2026 Kona Low.* Zenodo. DOI: 10.5281/zenodo.19410137
