---
title: "Science Paper — Analytical & Figure Roadmap"
subtitle: "Extreme storms reorganize coastal biogeochemistry across tropical island seascapes"
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    code-fold: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
```

---

# Paper architecture

## Core argument

Extreme storms are intensifying under climate change, yet their coastal
biogeochemical effects remain poorly characterized because conventional
monitoring is too slow and too sparse to capture episodic events at island
scale. We show — using a citizen science rapid-response network deployed
within 15 days of the March 2026 Kona Low across 746 sites on four Hawaiian
islands — that extreme storms do not simply elevate coastal nutrients: they
*temporarily reorganize land–sea biogeochemical connectivity*, creating
spatially coherent, chemically distinct nutrient fingerprints whose structure
is predictable from pre-existing landscape features. This reframes storms from
hydrological disturbances to biogeochemical restructuring events, with
implications for reef exposure forecasting and monitoring design globally.

## Claim hierarchy

| # | Claim | Key evidence | Figure |
|---|---|---|---|
| 1 | Kona Low produced an island-scale, spatially coherent nutrient anomaly | Kriging maps; Moran's I significant for NO₃+NO₂ & SiO₂ across all islands | Fig 1 |
| 2 | Storm generated chemically distinct fingerprints, not one uniform pulse | PAM k=4 clusters; stoichiometric divergence from Redfield | Fig 3 |
| 3 | Storm temporarily elevated land–sea biogeochemical connectivity | Salinity–nutrient mixing; post-storm VIIRS chlorophyll bloom | Fig 1D + Fig 2B |
| 4 | Reorganization was predictable from landscape structure | RF OOB 61 % (null 36 %); top predictors = coastal position + cesspool distance | Fig 4 |
| 5 | Citizen science enabled detection; conventional monitoring was absent | 0 WQP records in storm window vs 746 CS samples | Fig 2 |

## Numbers that must appear in the paper

- **746 sites**, 4 islands, collected within **15 ± 3 days** of storm peak (Mar 14)
- **0 conventional WQP records** during storm window (last record: Dec 18, 2025)
- Maui **73%** / Oʻahu **40%** / Molokaʻi **65%** of sites exceeded NO₃+NO₂ reef threshold (1 µmol/L)
- Storm peak: **68–124 mm/day** rainfall (NCEP); discharge spike **231–1,040 cfs** (USGS)
- RF OOB accuracy **61.3%** vs null **36.3%**; top predictor = coastal geographic position

---

# Figure plan

## Figure 1 — The biogeochemical footprint (impact figure)

**One-sentence message:** The Kona Low imprinted a spatially coherent,
island-specific pattern of coastal nutrient enrichment across four Hawaiian
islands within days of landfall.

**Layout:** 2-column Science figure (~18.3 cm wide × ~14 cm tall)

| Panel | Content | Data | Status |
|---|---|---|---|
| **A** Oʻahu kriging map | NO₃+NO₂ kriged surface, plasma scale capped 20 µmol/L, hotspots obvious | `chem`, `ahupuaa_raw` | ✅ done |
| **B** Maui kriging map | Same scale | `chem`, `ahupuaa_raw` | ✅ done |
| **C** Molokaʻi kriging map | Same scale; Lānaʻi raw points inset | `chem`, `ahupuaa_raw` | ✅ done |
| **D** Salinity–nutrient mixing | Scatter: salinity (x) vs NO₃+NO₂ (y, log), all islands, reef threshold line | `chem` | ✅ done |

**Key design notes:**
- All three kriging panels use the **same fixed color scale** (0–20 µmol/L, plasma)
  so readers can compare islands directly.
- Panel D anchor: add two reference points — (1) Hawaiian open ocean endmember
  from HOT (Station ALOHA surface NO₃+NO₂ ≈ 0.1 µmol/L) and (2) median
  salinity >34.5 PSU sub-group from our own data as "background" marker.
  This makes the "reorganization" claim concrete.

**Missing piece — HIGH PRIORITY:**
> Download HOT (Hawaii Ocean Time-series) surface water NO₃+NO₂ data from
> <https://hahana.soest.hawaii.edu/hot/hot-dogs/> (monthly since 1988).
> Plot the mean ± SD as a horizontal reference band on panel D.
> This baseline contrast is essential for Science — without it, we cannot
> claim the values are anomalous, only that they vary.

```{r fig1-baseline-stub, eval=FALSE, echo=TRUE}
# Download HOT nutrient data (do once)
# File: https://hahana.soest.hawaii.edu/hot/hot-dogs/bextracted.jgofs
# Relevant columns: date, NO3NO2 (surface, <25 m), salinity

hot <- read_table("data/hot_nutrients.txt", comment = "#") |>
  filter(depth < 25, !is.na(NO3NO2)) |>
  summarise(
    mean_NO3NO2 = mean(NO3NO2, na.rm = TRUE),
    sd_NO3NO2   = sd(NO3NO2,   na.rm = TRUE),
    mean_sal    = mean(salinity, na.rm = TRUE)
  )

# Add to mixing plot as horizontal band
# annotate("rect", xmin=34, xmax=36, ymin=hot$mean_NO3NO2 - hot$sd_NO3NO2,
#          ymax=hot$mean_NO3NO2 + hot$sd_NO3NO2, fill="gray70", alpha=0.4)
```

---

## Figure 2 — The storm cascade and the detection gap (methods / context)

**One-sentence message:** The Kona Low generated a measurable hydrological
discharge pulse and a post-storm chlorophyll bloom; citizen science was the
only monitoring source present during the event.

**Layout:** 2-column figure (~18.3 cm × ~12 cm tall), two rows

| Panel | Content | Data | Status |
|---|---|---|---|
| **A** Storm cascade time series | NCEP rainfall (top) + USGS discharge (middle) + VIIRS chlorophyll (bottom); storm & sampling windows highlighted | `ncep`, `discharge_raw`, `chla` | ✅ done |
| **B** Detection gap | WQP monthly counts 2015–2026 (top panel) vs citizen science burst (bottom panel); Kona Low window marked | `wqp_hist`, `chem` | ✅ done |

**Key design notes:**
- Panel A: three stacked time-series panels with shared x-axis. Collapse to
  island means for clarity; show individual islands as translucent ribbons if
  space allows.
- Panel B: visually stark contrast is the point. Make the conventional
  monitoring panel's y-axis start at 0 and end at ~250/month, so the
  sudden drop to 0 in 2026 is obvious.

**Missing piece — MODERATE PRIORITY:**
> Add CHIRPS historical rainfall anomaly (2013–Feb 2026 baseline) to show
> how unusual the March storm was in terms of return period. A single
> annotation on Panel A — "e.g., 1-in-X year event" — greatly strengthens
> the motivation.

```{r fig2-storm-anomaly-stub, eval=FALSE, echo=TRUE}
# CHIRPS historical March daily precip per island (already cached)
# Compare Mar 14-16 2026 NCEP values to CHIRPS historical distribution

chirps_march <- chirps |>
  filter(island %in% c("Oahu","Maui","Molokai"),
         lubridate::month(date) == 3) |>
  group_by(island, date) |>
  summarise(precip = mean(precip, na.rm = TRUE), .groups = "drop")

storm_peak <- ncep |>
  filter(island %in% c("Oahu","Maui","Molokai"),
         date == as.Date("2026-03-14")) |>
  group_by(island) |>
  summarise(peak_mm = mean(precip_mmday, na.rm = TRUE), .groups = "drop")

# Percentile rank of storm peak within 11-year March distribution
chirps_march |>
  group_by(island) |>
  summarise(pct_rank = mean(precip < storm_peak$peak_mm[match(first(island),
                                                               storm_peak$island)]),
            .groups = "drop")
```

---

## Figure 3 — Chemically distinct fingerprints (novelty figure)

**One-sentence message:** The storm generated four chemically distinct coastal
water types whose stoichiometric signatures reveal source-specific contributions
from rock weathering, agricultural nitrogen, and wastewater phosphorus.

**Layout:** 2-column figure (~18.3 cm × ~12 cm), two rows

| Panel | Content | Data | Status |
|---|---|---|---|
| **A** PCA biplot | PC1 (64%) vs PC2 (17%); points colored by cluster; loading arrows; convex hulls | `chem`, PAM k=4 | ✅ done |
| **B** Fingerprint heatmap | Rows = clusters; columns = Salinity, NO₃+NO₂, PO₄, SiO₂; Z-score fill; median values annotated | `chem`, PAM k=4 | ✅ done |
| **C** Stoichiometric fingerprint | N:P (y) vs Si:N (x) scatter, log-log; island medians as diamonds; Redfield (N:P=16) and diatom (Si:N=1) reference lines; cluster background shading | `chem` | ✅ done |

**Key design notes:**
- Panels A and B together tell the clustering story; C extends it to
  the source-attribution story (what *generates* the fingerprints).
- In panel C, the key Science-level message: all islands fall in
  P-excess / Si-rich quadrant (volcanic geology + wastewater), but
  **Maui is closest to Redfield** (more agricultural N), while
  **Oʻahu and Molokaʻi show greater P-enrichment** (cesspool density).
  This is the distinction between islands that makes Claim 2 strong.

**Missing piece — MODERATE PRIORITY:**
> Add literature-derived reference points to panel C for:
> (1) Open ocean (HOT surface water N:P and Si:N as a single point)
> (2) Published Hawaiian reef water quality baselines (Smith et al. 2005;
>     Stimson & Larned 2000) — these are tabulated in literature, no
>     download needed.
> These reference points anchor the stoichiometric fingerprint in a global
> context and make the departure from background explicit.

---

## Figure 4 — Landscape predictability (forecasting figure)

**One-sentence message:** The spatial structure of the biogeochemical anomaly
was predictable from pre-existing landscape features — coastal geographic
position and proximity to wastewater infrastructure — enabling a framework
for reef exposure forecasting before future events.

**Layout:** 2-column figure (~18.3 cm × ~12 cm), two rows

| Panel | Content | Data | Status |
|---|---|---|---|
| **A** RF variable importance | Horizontal bar chart; normalized Gini impurity; OOB vs null accuracy in subtitle | `chem_full`, `ranger` | ✅ done |
| **B** OOB confusion matrix | 4×4 heat map; row-normalized; % correct on diagonal | `ranger` | ✅ done |
| **C** Cluster × geography | Maps of Oʻahu + Maui side by side; points colored by cluster; kriging surface background | `chem`, PAM, kriging | ✅ done |

**Key design notes:**
- Panel A's most important message: coastal position (lat/lon) dominates,
  reflecting windward/leeward rainfall gradient. This is not just "where
  you are" — it is a proxy for watershed exposure to orographic rainfall,
  which is a fundamental landscape feature.
- A short annotation on Panel A: "Windward/leeward position captures
  storm exposure better than any single infrastructure metric" — this is
  the conceptual bridge from empirical RF to landscape theory.

**Missing pieces — LOWER PRIORITY (but would strengthen for Science):**
> 1. **NOAA C-CAP impervious surface** — % impervious cover per ahupuaʻa
>    watershed would replace lat/lon as a more mechanistically interpretable
>    predictor. Access: <https://coast.noaa.gov/ccapftp/>
>    Expected improvement: RF OOB probably rises from 61% to ~70-75%.
>
> 2. **NHDPlus stream density** — stream network density per watershed
>    captures drainage efficiency. Access: `nhdplusTools` package.
>
> 3. **Watershed area** — from ahupuaʻa boundaries (already have polygons;
>    just compute `st_area()`). Larger watersheds → more runoff volume.

```{r fig4-watershed-area-stub, eval=FALSE, echo=TRUE}
# Compute ahupuaa watershed area from existing polygons
ahupuaa_area <- ahupuaa_raw |>
  sf::st_transform(32604) |>
  mutate(area_km2 = as.numeric(sf::st_area(geometry)) / 1e6) |>
  sf::st_drop_geometry() |>
  select(ahupuaa, moku, mokupuni, area_km2)

# Join to chem_full and add to RF predictor set
# Then re-run ranger and compare OOB accuracy
```

---

# Extended data (supplementary figures)

Science papers use "Extended Data" — numbered ED figures that appear after
the main text in the online supplement but are peer-reviewed as part of the
manuscript. Aim for 6–8 ED figures.

| ED# | Content | Status |
|---|---|---|
| **ED1** | Sampling design map: all 4 islands with n per island, collection date distribution, and WQP conventional station locations for comparison | ✅ buildable from existing data |
| **ED2** | Kriging diagnostics: empirical variograms + fitted models for each island | ✅ buildable from krige function internals |
| **ED3** | Full salinity–nutrient mixing (4-panel: NO₃+NO₂, PO₄, SiO₂, NH₃ vs salinity) with reef threshold lines | ✅ done (`salinity-nutrient-mixing` chunk) |
| **ED4** | Silhouette analysis (k=2–6 + individual widths for k=4) | ✅ done (`silhouette-supp` chunk) |
| **ED5** | Moran's I spatial autocorrelation by island × nutrient | ✅ done (`morans-i` chunk) |
| **ED6** | CHIRPS/NCEP rainfall return-period analysis — how anomalous was the Kona Low vs 11-year record? | ⚠️ needs storm anomaly code (stub in Fig 2 above) |
| **ED7** | VIIRS chlorophyll anomaly — post-storm bloom relative to 2015–2025 March climatology | ⚠️ needs VIIRS climatology baseline (see stub below) |
| **ED8** | Cesspool density → coastal NO₃+NO₂ regression (ahupuaʻa level, by island, log-log) | ✅ done (`cesspool-regression` chunk) |

```{r ed7-chlorophyll-anomaly-stub, eval=FALSE, echo=TRUE}
# ED7: VIIRS chlorophyll anomaly
# Need: March climatology from NOAA CoastWatch ERDDAP (2015-2025 March mean per pixel)
# Already have: hawaii_chlorophyll_cache.rds (Feb-Jun 2026)

# Step 1: Pull monthly VIIRS from ERDDAP for March 2015-2025
# library(rerddap)
# viirs_march <- info("nesdisVHNSQchlaMonthly", url="coastwatch.pfeg.noaa.gov/erddap")
# ... griddap() call for Hawaii bbox, March only, 2015-2025

# Step 2: Compute pixel-level March mean (climatology)
# Step 3: Subtract 2026 post-storm (late March) from climatology
# Step 4: Map anomaly — positive = bloom, negative = no change
```

---

# Data gaps summary

## Priority 1 — Required before Science submission

These gaps directly affect the paper's core claims:

| Gap | What it provides | Source | R package | Est. effort |
|---|---|---|---|---|
| **HOT surface nutrients** (1988–2026) | Open-ocean baseline for Fig 1D mixing plot; anchors "reorganization" claim | <https://hahana.soest.hawaii.edu/hot/hot-dogs/> | `read_table()` | 1 hour |
| **Internal salinity baseline** | High-salinity (>34.5 PSU) sites as within-study control; clean enrichment factor | Already in `chem_full` | none | 30 min |
| **CHIRPS storm return period** | How unusual was March 14? "1-in-X year event" annotation on Fig 2A | Already in `chirps` cache | `dplyr` | 1 hour |

## Priority 2 — Strengthens manuscript; reviewer-expected

| Gap | What it provides | Source | R package | Est. effort |
|---|---|---|---|---|
| **NOAA C-CAP impervious surface** | Replaces lat/lon as RF predictor; more mechanistic than position alone | NOAA coast.noaa.gov/ccapftp | `terra`, `FedData` | 4 hours |
| **Watershed area** (from ahupuaʻa polygons) | Runoff volume proxy; compute from existing data | `ahupuaa_raw` polygons | `sf::st_area()` | 30 min |
| **WQP site locations** (mapped) | Spatial comparison for detection gap figure | `wqp_hist` + `whatWQPsites()` | `dataRetrieval` | 2 hours |
| **VIIRS chlorophyll climatology** | Biological response anomaly (ED7) | NOAA CoastWatch ERDDAP | `rerddap` | 3 hours |

## Priority 3 — Nice to have; supplements narrative

| Gap | What it provides | Source | R package | Est. effort |
|---|---|---|---|---|
| **NOAA reef extent polygons** | % Hawaiian reef area above NO₃+NO₂ threshold | NOAA CRCP / CoRIS | `sf` | 3 hours |
| **USGS WQP stream nutrients** (baseflow) | Storm amplification factor: storm-flow vs baseflow at gauged streams | WQP via `readWQPdata(siteid="USGS-XXXXXX")` | `dataRetrieval` | 2 hours |
| **NHDPlus stream density** | Drainage efficiency per watershed | `nhdplusTools` | `nhdplusTools` | 4 hours |
| **GPM IMERG rainfall** (0.1° res) | Higher-resolution storm footprint than NCEP | NASA Earthdata / ERDDAP | `httr2` | 3 hours |

---

# Analysis checklist

## Done ✅

- [x] Kriging maps (Oʻahu, Maui, Molokaʻi) — fixed plasma scale
- [x] Salinity–nutrient mixing (all 4 nutrients)
- [x] PAM k=4 cluster analysis with PCA biplot
- [x] Chemical fingerprint heatmap (cluster profiles)
- [x] Stoichiometric fingerprints (N:P, Si:N by island)
- [x] Moran's I spatial autocorrelation
- [x] Random forest: landscape → cluster membership (OOB 61.3%)
- [x] Confusion matrix (OOB)
- [x] Ahupuaʻa cesspool regression (NO₃+NO₂ and PO₄)
- [x] WQP detection gap timeline
- [x] Storm cascade figure (NCEP + USGS discharge + VIIRS chlorophyll)
- [x] Silhouette analysis (k=2–6 supplementary)
- [x] Kriging + cluster overlay maps

## Needed ⚠️

- [ ] **HOT baseline** — download and add to Fig 1D mixing plot
- [ ] **Internal salinity baseline** — median nutrients at Sal >34.5 PSU vs <32 PSU
- [ ] **CHIRPS storm anomaly** — percentile rank of March 14 peak vs 11-year record
- [ ] **Watershed area** — `st_area()` on ahupuaʻa polygons → add to RF
- [ ] **C-CAP impervious surface** — download, zonal stats per ahupuaʻa → add to RF
- [ ] **VIIRS chlorophyll climatology** — March 2015–2025 mean → 2026 anomaly map
- [ ] **WQP site map** — spatial visualization of conventional monitoring coverage
- [ ] **Reef exposure** — % reef area above threshold (needs reef polygons)

---

# Quick-start code for highest-priority gaps

## 1. Internal salinity baseline (30 minutes)

```{r salinity-baseline, eval=FALSE, echo=TRUE}
library(tidyverse)

# Within-study control: high-salinity sites = near-background marine conditions
baseline_vs_storm <- chem_full |>
  filter(!is.na(Salinity), !is.na(NO3NO2)) |>
  mutate(
    condition = case_when(
      Salinity > 34.5 ~ "Background\n(Sal > 34.5)",
      Salinity < 32.0 ~ "Storm-influenced\n(Sal < 32)",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(condition)) |>
  group_by(condition) |>
  summarise(
    n            = n(),
    median_NO3   = median(NO3NO2, na.rm = TRUE),
    median_PO4   = median(PO4,    na.rm = TRUE),
    median_SiO2  = median(SiO2,   na.rm = TRUE),
    pct_above_N  = mean(NO3NO2 > 1, na.rm = TRUE),
    .groups = "drop"
  )

# Enrichment factor
baseline_vs_storm |>
  summarise(
    NO3_enrichment = median_NO3[condition == "Storm-influenced\n(Sal < 32)"] /
                     median_NO3[condition == "Background\n(Sal > 34.5)"]
  )
```

## 2. Watershed area from existing polygons (30 minutes)

```{r watershed-area, eval=FALSE, echo=TRUE}
library(sf); library(tidyverse)

watershed_area <- ahupuaa_raw |>
  sf::st_transform(32604) |>
  mutate(area_km2 = as.numeric(sf::st_area(geometry)) / 1e6) |>
  sf::st_drop_geometry() |>
  group_by(ahupuaa, moku, mokupuni) |>
  summarise(area_km2 = sum(area_km2), .groups = "drop")

# Join to chem_full and add to RF training set
# Then re-run ranger() and check if OOB improves
```

## 3. HOT open-ocean baseline (1 hour)

```{r hot-baseline, eval=FALSE, echo=TRUE}
# Download from: https://hahana.soest.hawaii.edu/hot/hot-dogs/
# Navigate to: HOT-DOGS > Nutrients > NO3+NO2 at surface

# After download to data/hot_nutrients.csv:
hot <- read_csv("data/hot_nutrients.csv", show_col_types = FALSE) |>
  filter(depth_m < 25) |>
  summarise(
    mean_NO3 = mean(NO3NO2_umol_L, na.rm = TRUE),
    sd_NO3   = sd(NO3NO2_umol_L,   na.rm = TRUE),
    n        = n()
  )

# HOT surface NO3+NO2 typically: mean ≈ 0.03 umol/L, range 0.01–0.30
# Use as reference line/band in Fig 1D mixing diagram
```

## 4. CHIRPS storm return period (1 hour)

```{r chirps-return-period, eval=FALSE, echo=TRUE}
library(tidyverse)

# Already have chirps (2013-2026 Feb) and ncep for storm period

# Step 1: Island-mean March precipitation from CHIRPS
chirps_march_daily <- chirps |>
  filter(island %in% c("Oahu","Maui","Molokai"),
         lubridate::month(date) == 3) |>
  group_by(island, date) |>
  summarise(precip_mm = mean(precip, na.rm = TRUE), .groups = "drop")

# Step 2: 3-day rolling max (storm is multi-day)
chirps_3day <- chirps_march_daily |>
  arrange(island, date) |>
  group_by(island) |>
  mutate(roll3 = zoo::rollmax(precip_mm, 3, fill = NA, align = "right")) |>
  ungroup()

# Step 3: Storm peak 3-day total from NCEP (Mar 13-15)
ncep_storm <- ncep |>
  filter(island %in% c("Oahu","Maui","Molokai"),
         date >= as.Date("2026-03-13"), date <= as.Date("2026-03-15")) |>
  group_by(island) |>
  summarise(storm_3day = sum(precip_mmday, na.rm = TRUE), .groups = "drop")

# Step 4: Exceedance probability
chirps_3day |>
  left_join(ncep_storm, by = "island") |>
  group_by(island) |>
  summarise(
    pct_exceeded = mean(roll3 >= storm_3day, na.rm = TRUE),
    return_period_yrs = 1 / pct_exceeded,
    .groups = "drop"
  )
```

---

# Recommended next steps (prioritized)

1. **Run the internal salinity baseline** (30 min) — produces the enrichment
   factor number for the abstract ("X-fold above background").

2. **Compute watershed area** (30 min) — immediate improvement to RF predictor
   set with no data download needed.

3. **Download HOT nutrients** (1 hr) — adds the open-ocean endmember to the
   mixing diagram. This single addition makes the Fig 1D panel publication-ready
   for Science's standards.

4. **Run CHIRPS storm anomaly** (1 hr) — produces the return period estimate
   for Fig 2 context ("1-in-N year event").

5. **Download C-CAP impervious surface** (4 hrs, most involved) — transforms
   the RF analysis from empirical (position-based) to mechanistic (land cover),
   which is the difference between "interesting" and "impactful" for reviewers.

---

*Roadmap last updated: `r Sys.Date()`*
*Data directory: `data/`*
*Main analysis QMD: `nutrient_hotspot_maps.qmd`*
